---
title: "nb-bench-v3"
output:
  pdf_document: default
  html_document: default
date: "2026-06-15"
---


results from more extensive simulations and maybe real data


```{r, echo=FALSE, cache=TRUE}
rm(list=ls())
library(tidyverse)


source("~/.Rprofile")
source("~/katsevich-lab/code/sceptre3-project-v2/benchmarking/guide-assignment/assignment-analysis/script_runs_analysis_functions.R")
source("~/katsevich-lab/code/sceptre3-project-v2/benchmarking/guide-assignment/sceptre-nb/nb-bench_v3-helpers.R")
load_assns_crispat <- function(path_to_assns, all_guides, cell_name_to_idx) {
  crisp_assign_raw <- read.csv(path_to_assns) |>
     left_join(cell_name_to_idx, by=join_by(cell_id == cell_name)) |>
     dplyr::select(-cell_id)
      # mutate(cell_idx = gsub("cell_idx_", "", cell_id) |> as.integer())
  unordered_cell_list <- split(crisp_assign_raw$cell_idx, crisp_assign_raw$grna_id)
  
  guide_to_assns = vector("list", length(all_guides)) |>
    setNames(all_guides)
  
  guide_to_assns[names(unordered_cell_list)] <- unordered_cell_list
  guide_to_assns
}

load_run <- function(run_name, dataset_name) {
  base_path = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/outputs"
  )
  input_data_path = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data",
    dataset_name
  )
  if(file.exists(true_perts_fp <- file.path(input_data_path, "true_pert_matrix.rds"))) {
    true_perts = read_rds(true_perts_fp)
  }
  grna_matrix = file.path(input_data_path, "sceptre/grna_matrix.rds") |>
    read_rds()
  
  
  data_path = file.path(base_path, run_name)
  
  cov_df <- read_csv(file.path(input_data_path, "sceptre/cell_covariates.csv"), show_col_types = FALSE) |>
    mutate(
      grna_n_nonzero_subset = Matrix::colSums(grna_matrix > 0),
      grna_n_umis_subset = Matrix::colSums(grna_matrix)
    )
  
  
  
  
  fnames = dir(data_path)
  
  # 1. load assignment matrices
  assn_mat_fnames = fnames[grepl("^assignment_matrix", fnames) & grepl(dataset_name, fnames, fixed = TRUE)]
  all_assns = list()
  guide_names = rownames(grna_matrix)
  for(i in seq_along(assn_mat_fnames)) {
    curr_fname = assn_mat_fnames[i]
    script_name = gsub("^assignment_matrix_", "", curr_fname) |>
      gsub(pattern = "^script_", replacement="", fixed=TRUE) |>
      gsub(pattern =paste0("_", dataset_name, ".rds"), replacement="", fixed=TRUE)
    
    mat = read_rds(file.path(data_path, curr_fname))
    
    curr_assns = lapply(guide_names, function(guide) which(mat[guide,]) |> as.integer()) |>
      setNames(guide_names) |>
      list() |>
      setNames(script_name)
    
    all_assns = c(all_assns, curr_assns)
    
  }
  
  # 2. load extras from script runs
  extras_fnames = fnames[grepl("^assignment_extras", fnames) & grepl(dataset_name, fnames, fixed = TRUE)]
  extras_all <- lapply(extras_fnames, function(fname) read_rds(file.path(data_path, fname)))
  names(extras_all) <- gsub("assignment_extras_script_", "", extras_fnames) |>
    gsub(pattern = paste0("_", dataset_name, ".rds"), replacement="", fixed=TRUE)
  
  
  # 3. load other methods if there
  
  # i need this to change cell name assignments to cell index assignments
  
  cell_name_to_idx <- data.frame(
    cell_name = colnames(grna_matrix),
    cell_idx = 1:ncol(grna_matrix)
  )
  
  if(length(fname_crispat <- fnames[grepl("crispat", fnames) & grepl(dataset_name, fnames)]) > 0) {
    all_assns$crispat <- load_assns_crispat(file.path(data_path, fname_crispat), guide_names, cell_name_to_idx)
  }
  
  out = list(
    run_name = run_name, dataset_name = dataset_name,
    input_data_path = input_data_path,
    grna_matrix = grna_matrix,
    cov_df = cov_df,
    all_assns = all_assns,
    all_extras = extras_all
  )
  if(exists("true_perts")) {
    out$true_perts = true_perts
    
    out$true_assns = lapply(guide_names, function(guide) {
      which(true_perts[guide,] == 1 & grna_matrix[guide,] > 0) |> as.integer()
      # which(sim_results$true_perts[guide,] == 1 ) |> as.integer()
    }) |>
      setNames(guide_names)
    
    
  }
  out
}


sim_pois0nb_2np_3p <- load_run(
  run_name =  "pois0nb_sims_sum_2np_3p",
  dataset_name = "sims_sum_2np_3p"
)


sim_pois0nb_1np_3p_disp <-  load_run(
  run_name =  "pois0nb_sims_sum_1np_3p_disp",
  dataset_name = "sims_sum_1np_3p_disp"
)



sim_pois0nb_repeat_old <-  load_run(
  run_name =  "pois0nb_sims_repeat_old",
  dataset_name = "sims_sum_repeat_old"
)


repl_results <-  load_run(
  run_name =  "script_real_data",
  dataset_name = "replogle-rd7_assign_nguides=800_ncells=100k"
)

gasp_results <-  load_run(
  run_name =  "script_real_data",
  dataset_name = "gasperini_assign_nguides=800_ncells=100k"
)
# guide_names = rownames(sim_results$grna_matrix)

# sim_pois0nb_2np_3p$all_assns$thresh_10 <- lapply(rownames(sim_pois0nb_2np_3p$grna_matrix), function(guide) {
#   which(sim_pois0nb_2np_3p$grna_matrix[guide,] > 10)
# }) |>
#   setNames(rownames(sim_pois0nb_2np_3p$grna_matrix))
# sim_pois0nb_1np_3p_disp$all_assns$thresh_10 <- lapply(rownames(sim_pois0nb_1np_3p_disp$grna_matrix), function(guide) {
#   which(sim_pois0nb_1np_3p_disp$grna_matrix[guide,] > 10)
# }) |>
#   setNames(rownames(sim_pois0nb_1np_3p_disp$grna_matrix))

```



# 0. introduce my new simulations


main problem: no single nice distribution looks anything like the NPs

The point of this plot: to show how we need the complexity of the summed process.

```{r, echo=FALSE, fig.cap="Standard distributions do not decay so rapidly but then become nearly uniform. A single Pois or NB will not provide a realistic non-pertrubed model."}

grna_odm_repl = ondisc::initialize_odm_from_backing_file(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7/sceptre-pipeline/grna.odm"
)

repl_guide_stats = read_csv(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7/sceptre-pipeline/grna_summary_stats.csv",
  show_col_types = FALSE
)

grna_odm_gasp = ondisc::initialize_odm_from_backing_file(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/gasperini/sceptre-pipeline/grna.odm"
)

gasp_guide_stats = read_csv(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/gasperini/sceptre-pipeline/grna_summary_stats.csv",
  show_col_types = FALSE
)

get_guides <- function(guide_stats, n, min_idx = 200) {
   guide_stats |>
  arrange(grna_total) |>
  pull(grna) %>%
  `[`(seq(min_idx, nrow(guide_stats) - 50, length = n) |> round())
}

# let's pick a sequence of guides in terms of expression
repl_guides <- get_guides(repl_guide_stats, 4)

gasp_guides <-  get_guides(gasp_guide_stats, 4) 


plotlist <- c(
  lapply(seq_along(repl_guides), function(i) {
    plot_umi_histogram(grna_odm_repl[repl_guides[i],],
                       title = paste0("Replogle: ", i, " / ", length(repl_guides)),
                       x_breaks = c(1,10,100,1000,10000,100000))
  }),
  lapply(seq_along(gasp_guides), function(i) {
    plot_umi_histogram(grna_odm_gasp[gasp_guides[i],],
                       title = paste0("Gasperini: ", i, " / ", length(gasp_guides)),
                       x_breaks = c(1,10,100,1000,10000,100000))
  })
)

cowplot::plot_grid(plotlist = plotlist)
```



so what am i doing now? inspired by Cellbender model:

$$
Y_i = \text{Pois}(L_i \mu_{\text{exo}}) + \text{Ber}(\pi_{\text{endo}}) \text{Pois}(L_i \mu_{\text{endo}}) + \text{Ber}(\pi_{\text{pert}}) \text{NB}(L_i \mu_{\text{pert}}, \theta_{\text{pert}})
$$


The idea: 

- the perturbed distribution is the same: NegBin with cell-specific library size rescaling factors 

- but now the NP is a sum of two processes:
  
  - exogeneous noise, representing random stray bits getting in there. $\mu_{\text{exo}}$ is small
  
  - endogenous noise, from droplet contamination, chimeras, etc, giving rare but larger values. $\pi_{\text{endo}}$ is very small but $\mu_{\text{endo}}$ is larger.



I am able to get satisfying looking distributions from this, and in particular I can now simulate distributions where precision is the hard part. Before, with such light tails, precision was often trivial because the NPs were so concentrated near zero.

```{r, echo=FALSE, fig.cap="Example of my new simulations. Note that now precision will not be trivial.", include=FALSE, eval=FALSE}

# sim_dataset_name <- "replogle-rd7_sims_from_clns_50k_4_mu"
# 
# grna_matrix_sims <- read_rds(
#   file.path( "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data",
#              sim_dataset_name, "sceptre/grna_matrix.rds")
# )
# 
# true_perts_sim <- read_rds(
#    file.path( "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data",
#              sim_dataset_name, "true_pert_matrix.rds")
# )
# 
# i=1
# plot_umi_histogram_real_vs_sim(
#   umis_real = grna_odm_repl[tail(repl_guides, 1),],
#   umis_sim = grna_matrix_sims[i,],
#   is_pert = true_perts_sim[i,],
#   title = paste0("Repl. vs ", rownames(grna_matrix_sims)[i])
# )

```


```{r, echo=FALSE, fig.cap="example simulation datasets, both new and old"}
plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    NP_lowvar_P_low = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPlowvar_P1_1",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPlowvar_P1_1",]
    ),
      NP_lowvar_P_high = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPlowvar_P3_2",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPlowvar_P3_2",]
    ),
    NP_highvar_P_low = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPhighvar_P1_1",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPhighvar_P1_1",]
    ),
      NP_highvar_P_high = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPhighvar_P3_1",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPhighvar_P3_1",]
    )
  ),
   title = "Real vs simulated: sims_sum_2np_3p"
)

plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    NP_highvar_P_lowdisp = list(
      umis_sim = sim_pois0nb_1np_3p_disp$grna_matrix["sim_NPhighvar_Plowdisp_1",],
  is_pert = sim_pois0nb_1np_3p_disp$true_perts["sim_NPhighvar_Plowdisp_1",]
    ),
      NP_highvar_P_meddisp = list(
      umis_sim = sim_pois0nb_1np_3p_disp$grna_matrix["sim_NPhighvar_Pmeddisp_1",],
  is_pert = sim_pois0nb_1np_3p_disp$true_perts["sim_NPhighvar_Pmeddisp_1",]
    ),
    NP_highvar_P_highdisp = list(
      umis_sim = sim_pois0nb_1np_3p_disp$grna_matrix["sim_NPhighvar_Phighdisp_1",],
  is_pert = sim_pois0nb_1np_3p_disp$true_perts["sim_NPhighvar_Phighdisp_1",]
    )
  ),
   title = "Real vs simulated: sims_sum_1np_3p_disp"
)



plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    NP_pois_P_small = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Psmall_1",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Psmall_1",]
    ),
      NP_pois_P_med = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Pmed_74",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Pmed_74",]
    ),
    NP_pois_P_large = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Plarge_26",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Plarge_26",]
    )
  ),
   title = "Real vs simulated: sims_repeat_old"
)
```




# 1. summary metrics


```{r, echo=FALSE, fig.cap="Comparing sceptre, crispat, and the bset new method I've found, on the new simulations sims_sum_2np_3p"}
# for "pois0nb_sims_sum_2np_3p" ~~~~~~~~~~~~~~~~~~~~~~

P_to_mean <- tibble(
  p = c("P1", "P2", "P3"),
  mu_pert = c(500, 750, 1000)
)

guide_meta <- tibble(
  grna_id = rownames(sim_pois0nb_2np_3p$grna_matrix),
  np = extract_part(grna_id, 2),
  p = extract_part(grna_id, 3)
) %>%
  left_join(P_to_mean, by = "p")


## get metrics and avg ~~~~~~~~~~~~~~~~~~~~~~~~~~~~

guide_metric_tbl <- make_guide_metric_tbl(sim_pois0nb_2np_3p) |>
  left_join(guide_meta, by = "grna_id")


avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "np", "p", "mu_pert")
)


plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )


## 1. comparing the ones we're most interested in


plot_df |>
  filter(
    method_name %in% c("crispat", "sceptre", "script_threshglmpois1000_pois0nb", "script_threshglmpois1000_poisbug", "script_threshglmpois1000_poissumlog", 
                       "script_glmpois_poissumlog")
  ) |>
make_avg_metrics_plot() +
  ggtitle("Metrics for key methods on sims_sum_2np_3p")



## 2. different thresholds

# plot_df |>
#   filter(
#     grepl("threshglmpois", method_name),
#     grepl("pois0nb", method_name)
#   ) |>
# make_avg_metrics_plot() +
#   ggtitle("Metrics for key methods")



```

```{r, echo=FALSE, fig.cap = "simulations with much bigger and more overdispersed values in sims_sum_1np_3p_disp"}
# for "sims_sum_1np_3p_disp" ~~~~~~~~~~~~~~~~~~~~~~

P_to_mean <- tibble(
  p = c("Phighdisp", "Pmeddisp", "Plowdisp"),
  mu_pert = c(10001000, 1001000, 101000)
)

guide_meta <- tibble(
  grna_id = rownames(sim_pois0nb_1np_3p_disp$grna_matrix),
  np = extract_part(grna_id, 2),
  p = extract_part(grna_id, 3)
) %>%
  left_join(P_to_mean, by = "p")


## get metrics and avg ~~~~~~~~~~~~~~~~~~~~~~~~~~~~

guide_metric_tbl <- make_guide_metric_tbl(sim_pois0nb_1np_3p_disp) |>
  left_join(guide_meta, by = "grna_id")


avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "np", "p", "mu_pert")
)


plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )


## 1. comparing the ones we're most interested in


plot_df |>
  filter(
    method_name %in% c(
      "crispat", "script_threshglmpois1000_pois0nb", "script_glmpois_poisbug", "thresh_10", "script_threshglmpois1000_poissumlog", "script_threshglmpois1000_poisbug", "script_glmpois_poissumlog" # "thresh_10" #  "script_threshglmpois_pois0nb",
    )
  ) |>
  mutate(method_name = ifelse(method_name == "script_glmpois_poisbug", "sceptre", method_name)) |>
make_avg_metrics_plot(x_breaks = c(1e3,1e4,1e5,1e6,1e7)) +
  ggtitle("Metrics for more overdispersed data")




```



```{r, echo=FALSE, fig.cap="repeating old simulations"}
# for "sims_repeat_old" ~~~~~~~~~~~~~~~~~~~~~~

P_to_mean <- tibble(
  p = c("Plarge", "Pmed", "Psmall"),
  mu_pert = c(970, 970/4, 970/8)
)

guide_meta <- tibble(
  grna_id = rownames(sim_pois0nb_repeat_old$grna_matrix),
  np = extract_part(grna_id, 2),
  p = extract_part(grna_id, 3)
) %>%
  left_join(P_to_mean, by = "p")


## get metrics and avg ~~~~~~~~~~~~~~~~~~~~~~~~~~~~

guide_metric_tbl <- make_guide_metric_tbl(sim_pois0nb_repeat_old) |>
  left_join(guide_meta, by = "grna_id")


avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "np", "p", "mu_pert")
)


plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )


## 1. comparing the ones we're most interested in


plot_df |>
  # filter(! method_name %in% c("script_trimglmpois_poisfix", "script_glmpois_poisbug")) |>
  filter(
    method_name %in% c(
      "crispat", "sceptre", "script_threshglmpois1000_pois0nb", "script_threshglmpois1000_poissumlog",
      "script_threshglmpois1000_poisbug", "script_glmpois_poissumlog"
    )
  ) |>
make_avg_metrics_plot(x_breaks = c(1e3,1e4,1e5,1e6,1e7)) +
  ggtitle("Metrics for old simulations")




```








# 1.5 why does bugged work so well?


```{r, echo=FALSE, fig.cap="these plots show the poor precision on some guides for the current sceptre"}
plot_y_vs_o_one_run(
  extras = sim_pois0nb_repeat_old$all_extras$trimglmpois_poisfix,
  grna_matrix = sim_pois0nb_repeat_old$grna_matrix,
  true_perts           = sim_pois0nb_repeat_old$true_perts,
  input_dir            = file.path(sim_pois0nb_repeat_old$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_repeat_old$grna_matrix)[c(1:6)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
  color_is_pert = TRUE,
  # exclude_0 = TRUE,
  seed                 = 1L
) 


plot_y_vs_o_one_run(
  extras = sim_pois0nb_repeat_old$all_extras$glmpois_poisbug,
  grna_matrix = sim_pois0nb_repeat_old$grna_matrix,
  true_perts           = sim_pois0nb_repeat_old$true_perts,
  input_dir            = file.path(sim_pois0nb_repeat_old$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_repeat_old$grna_matrix)[c(1:6)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
  color_is_pert = TRUE,
  # exclude_0 = TRUE,
  seed                 = 1L
) 

plot_y_vs_o_one_run(
  extras = sim_pois0nb_repeat_old$all_extras$glmpois_poissumlog,
  grna_matrix = sim_pois0nb_repeat_old$grna_matrix,
  true_perts           = sim_pois0nb_repeat_old$true_perts,
  input_dir            = file.path(sim_pois0nb_repeat_old$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_repeat_old$grna_matrix)[c(1:6)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
  color_is_pert = TRUE,
  # exclude_0 = TRUE,
  seed                 = 1L
) 


plot_y_vs_o_one_run(
  extras = sim_pois0nb_repeat_old$all_extras$threshglmpois1000_pois0nb,
  grna_matrix = sim_pois0nb_repeat_old$grna_matrix,
  true_perts           = sim_pois0nb_repeat_old$true_perts,
  input_dir            = file.path(sim_pois0nb_repeat_old$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_repeat_old$grna_matrix)[c(1:6)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
    color_is_pert = TRUE,
  seed                 = 1L
) 







```





experiment: what if we just do 
$$
\mathbf 1\{y_i \geq 2 \max\{\mu_i, \gamma\}\}
$$
or something?

```{r}
thresh_est_func <- function(sim_results, run_name, glm_thresh = 1000) {
  
  grna_matrix = sim_results$grna_matrix
  cov_df <- build_covariate_matrix(file.path(sim_results$input_data_path, "sceptre"))
  
  assns <- vector("list", nrow(grna_matrix)) |>
    setNames(rownames(grna_matrix))
  extras <- list(
    per_guide = assns
  )
  for(i in 1:nrow(grna_matrix)) {
    
    y = grna_matrix[i,]
    keep <- y <= glm_thresh
    mod <- glm.fit(x = cov_df[keep,], y = y[keep])
    
    # are these lined up? i think so...
    stopifnot(colnames(cov_df) == names(mod$coefficients))
    offsets <- as.numeric(cov_df %*% mod$coefficients)
    
    # classifying
    g_est <- log(mean(y[y >= 20]))
    pi_est <- mean(y >= 20)
    
    # simplified version of decision boundary
    ## this is a moer accurate one
    # assns[[i]] <- which(y > (g_est - log(pi_est)) / (log(1 + g_est / exp(offsets))))
    ## this is even simpler
    assns[[i]] <- which(y > (1 - log(pi_est) / g_est) * max(exp(offsets), g_est))
    
    extras$per_guide[[i]] <- list(
      assignments = assns[[i]],
      offset_model_summary = list(
        coefficients = mod$coefficients
      )
    )
  }
  sim_results$all_assns[[run_name]] <- assns
  sim_results$all_extras[[run_name]] <- extras
  sim_results
}


sim_pois0nb_repeat_old <- thresh_est_func(sim_pois0nb_repeat_old, run_name = "test_thresh", 1000)


```

i suspect that the model that current sceptre does badly at isn't 
```{r}

```



ok so i see why current sceptre has poor precision on the largest mean from old sims: we have tiny offsets, so we're in a regime where we call it pert if approx $y_i > 2 \mu_{0i}$. These $o_i$ are super negative, so we get way too small of a thresh.

is this the same thing that probably happens on replogle? heavy tailed leads to tiny $o_i$, then we get too small of a threshold. 

are we relying on the coincidence that $\log \gamma$ is around 5, the typical UMI threshold...?










what about just adding a thing like "we will never call a cell with <= 5 UMI counts perturbed..." ... ??
--> use pois-pois to approx gamma and pi, but add a guard like this? 

oh well the problem also is that we don't want to use the actual multiplicative model... 

IDEA: so let's try using a robustified offset model. that's a simple change. does that fix that poor precision model?

if so, then the bugged model reduces to ?

```{r}
plot_y_vs_o_one_run(
  extras = sim_pois0nb_repeat_old$all_extras$glmpois_poisbug,
  grna_matrix = sim_pois0nb_repeat_old$grna_matrix,
  true_perts           = sim_pois0nb_repeat_old$true_perts,
  input_dir            = file.path(sim_pois0nb_repeat_old$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_repeat_old$grna_matrix)[c(1:6)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
  color_is_pert = TRUE,
  # exclude_0 = TRUE,
  seed                 = 1L
) 

plot_y_vs_o_one_run(
  extras = sim_pois0nb_repeat_old$all_extras$glmpois_poisbug,
  grna_matrix = sim_pois0nb_repeat_old$grna_matrix,
  true_perts           = sim_pois0nb_repeat_old$true_perts,
  input_dir            = file.path(sim_pois0nb_repeat_old$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_repeat_old$grna_matrix)[c(101:106)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
  color_is_pert = TRUE,
  # exclude_0 = TRUE,
  seed                 = 1L
) 
```


ok so here's my current hail mary: we still fit the pois-pois, but use a different way to get the decisions out of it... like 
$$
\mathbf 1\{y_i \geq 2 \max\{\mu_i, \gamma\}\}
$$
or something...



ok here it is:

Suppose we model $Y_i \sim \text{Pois}(e^{o_i}) + Z_i \text{Pois}(\Delta)$ for $Z_i \stackrel{\text{iid}}\sim \text{Pois}(\pi)$ indicating a successful perturbation. We then have
$$
Y_i \sim \begin{cases}
\text{Pois}(e^{o_i}) & Z_i = 0 \\ \text{Pois}(e^{o_i} + \Delta) & Z_i = 1
\end{cases}
$$
I think I made a convincing argument on Monday for using the sum of multiple count models for these data. This is a very simplified version of Cellbender with $P(\text{endogeneous error}) = 0$, $\theta_{\text{pert}} = \infty$, and no library size scaling in the perturbed distribution. 

I think that's pretty solid justification for having a model of this form.

The other problem is that we want $\Delta$ to be on the log scale. If we directly fit this model, $\Delta$ will be like $e^\gamma$ in our current model, and we have most of the same problems that we get from the bug-fixed version. 

Here is how we can justify the use of $\gamma \approx \log \Delta$: 

we want to reduce the ability of poor offsets to mess with us. So consider the log likelihood ratio of the perturbed to non-perturbed that we get, for a test observation $y$:
$$
\ell_+ = \frac{\text{Pois}(y| e^o+\Delta)}{\text{Pois}(y|e^o)} = 
$$




some guides sceptre failed on:

```{r, echo=FALSE}
plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    high_prec = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Plarge_2",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Plarge_2",]
    ),
      low_prec = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Plarge_6",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Plarge_6",]
    )
  ),
   title = "old simulated guides with good and bad scep prec"
)
```



<!-- Does bugged reduce the meaning of the offsets? -->

<!-- For bugged, we use the ratio -->
<!-- $$ -->
<!-- \frac{\text{Pois}(y|e^{o_i} + \gamma)}{\text{Pois}(y|e^{o_i})} = \frac{}{} -->
<!-- $$ -->



```{r, echo=FALSE, fig.cap="these plots show the performance on more realistic simulations."}
plot_y_vs_o_one_run(
  extras = sim_pois0nb_2np_3p$all_extras$glmpois_poisbug,
  grna_matrix = sim_pois0nb_2np_3p$grna_matrix,
  true_perts           = sim_pois0nb_2np_3p$true_perts,
  input_dir            = file.path(sim_pois0nb_2np_3p$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_2np_3p$grna_matrix)[c(1,2,101,102,201,202)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
  color_is_pert = TRUE,
  # exclude_0 = TRUE,
  seed                 = 1L
) 
plot_y_vs_o_one_run(
  extras = sim_pois0nb_2np_3p$all_extras$threshglmpois1000_pois0nb,
  grna_matrix = sim_pois0nb_2np_3p$grna_matrix,
  true_perts           = sim_pois0nb_2np_3p$true_perts,
  input_dir            = file.path(sim_pois0nb_2np_3p$input_data_path, "sceptre"),
  # covariate_matrix     = NULL,
  grna_ids             = rownames(sim_pois0nb_2np_3p$grna_matrix)[c(1,2,101,102,201,202)],
  # n_guides             = 6L,
  max_points_per_panel = 10000L,
  name_                = NULL,
  # log_y                = TRUE,
  cap_y                = 1000,
  facet_scales         = "free",
  show_est="p",
    color_is_pert = TRUE,
  seed                 = 1L
) 
```








# 2. offset exploration


how do my different thresholdings work?

<!-- do i see evidence that a small ridge penalty would be worth it? what about a way to check that the coefs are well-behaved? -->

```{r, echo=FALSE}
plot_coef_vs_coef(
  extras_x = sim_pois0nb_2np_3p$all_extras$threshglmpois1000_pois0nb,
  extras_y = sim_pois0nb_2np_3p$all_extras$glmpois_poisbug,
  "2np_3_p sims: new method", "current sceptre",
                  coef_names = colnames(build_covariate_matrix(file.path(sim_pois0nb_2np_3p$input_data_path, "sceptre"))))

plot_coef_vs_coef(
  extras_x = sim_pois0nb_1np_3p_disp$all_extras$threshglmpois1000_pois0nb,
  extras_y = sim_pois0nb_1np_3p_disp$all_extras$glmpois_poisbug,
  label_x = "Overdisp sims: new method", label_y = "current sceptre",
                  coef_names = colnames(build_covariate_matrix(file.path(sim_pois0nb_1np_3p_disp$input_data_path, "sceptre"))))

```


The betas vary a lot, but that doesn't seem to be a huge performance effect. Maybe because bugged is robust to them?



### real data



big question: how similar is my additive model to sceptre?

- i want jaccard, prec, and recall, using sceptre as the ground truth

```{r}

# agreement on replogle

methods_to_use <- c("script_glmpois_poisbug", "crispat", "script_glmpois_poissumlog", "script_threshglmpois1000_poissumlog", "script_threshglmpois1000_poisbug")
stopifnot(all(methods_to_use %in% names(repl_results$all_assns)))

guide_metric_tbl = make_guide_metric_tbl_vs_ref(
  all_assns = repl_results$all_assns[methods_to_use],
  grna_matrix = repl_results$grna_matrix,
    ref_method = "script_glmpois_poisbug"
)

avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "tail_mean_group")
)

make_avg_metrics_plot_ref <- function(df, x_breaks=NULL) {
  if(is.null(x_breaks)) x_breaks <- c(200,400,800,1600)
  ggplot(
    df,
    aes(
      x = tail_mean_group,
      y = mean,
      group = method_name,
      color = method_name,
      shape = method_name
      # linetype = method_name
    )
  ) +
    geom_line() +
    geom_errorbar(
      aes(ymin = ymin, ymax = ymax),
      width = 0.04,
      alpha = 0.7
    ) +
    geom_point(size = 2) +
    facet_grid( ~ metric) +
    # scale_x_continuous(
    #   trans = "log2",
    #   breaks = x_breaks
    # ) +
    scale_shape_manual(
      values = c(16, 17, 15, 3, 7, 8, 4, 5, 6, 9,10,11)
    ) +
    labs(
      x = expression(mu[pert]),
      y = "Mean",
      color = "Method",
      shape = "Method"
      # linetype = "Method"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom"
    )
}

plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )

plot_df |>
  make_avg_metrics_plot_ref()

```

```{r}
plot_avg_n_assignments_by_group(
  all_assns = repl_results$all_assns[methods_to_use],
  grna_matrix = repl_results$grna_matrix
)
```


```{r}

# agreement on gasperini

methods_to_use <- c("script_glmpois_poisbug", "crispat", "script_glmpois_poissumlog", "script_threshglmpois1000_poissumlog", "script_threshglmpois1000_poisbug")
stopifnot(all(methods_to_use %in% names(gasp_results$all_assns)))

guide_metric_tbl = make_guide_metric_tbl_vs_ref(
  all_assns = gasp_results$all_assns[methods_to_use],
  grna_matrix = gasp_results$grna_matrix,
    ref_method = "script_glmpois_poisbug"
)

avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "tail_mean_group")
)

make_avg_metrics_plot_ref <- function(df, x_breaks=NULL) {
  if(is.null(x_breaks)) x_breaks <- c(200,400,800,1600)
  ggplot(
    df,
    aes(
      x = tail_mean_group,
      y = mean,
      group = method_name,
      color = method_name,
      shape = method_name
      # linetype = method_name
    )
  ) +
    geom_line() +
    geom_errorbar(
      aes(ymin = ymin, ymax = ymax),
      width = 0.04,
      alpha = 0.7
    ) +
    geom_point(size = 2) +
    facet_grid( ~ metric) +
    # scale_x_continuous(
    #   trans = "log2",
    #   breaks = x_breaks
    # ) +
    scale_shape_manual(
      values = c(16, 17, 15, 3, 7, 8, 4, 5, 6, 9,10,11)
    ) +
    labs(
      x = expression(mu[pert]),
      y = "Mean",
      color = "Method",
      shape = "Method"
      # linetype = "Method"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom"
    )
}

plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )

plot_df |>
  make_avg_metrics_plot_ref()

```

```{r}
plot_avg_n_assignments_by_group(
  all_assns = gasp_results$all_assns[methods_to_use],
  grna_matrix = gasp_results$grna_matrix
)
```

