---
title: "assign-for-assoc"
output: html_document
date: "2026-06-27"
---


the purpose of this notebook is to do EDA to verify the assignment choices I'm making for my association benchmarking.


Gasperini: they use threshold of 5, although in the paper they also say something about "read counts" seaprate from UMI counts, and i don't know what those are.

# 1. does thresh=5 seem ok on Gasperini?


```{r}
rm(list=ls())
library(tidyverse)


source("~/.Rprofile")
source("~/katsevich-lab/code/sceptre3-project-v2/benchmarking/guide-assignment/assignment-analysis/script_runs_analysis_functions.R")
source("~/katsevich-lab/code/sceptre3-project-v2/benchmarking/guide-assignment/sceptre-nb/nb-bench_v3-helpers.R")

get_guides <- function(guide_stats, n, min_idx = 200) {
   guide_stats |>
  arrange(grna_total) |>
  pull(grna) %>%
  `[`(seq(min_idx, nrow(guide_stats) - 50, length = n) |> round())
}
```



```{r}



grna_odm_gasp = ondisc::initialize_odm_from_backing_file(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/gasperini/sceptre-pipeline/grna.odm"
)

gasp_guide_stats = read_csv(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/gasperini/sceptre-pipeline/grna_summary_stats.csv",
  show_col_types = FALSE
)



# let's pick a sequence of guides in terms of expression
# repl_guides <- get_guides(repl_guide_stats, 4)

gasp_guides <-  get_guides(gasp_guide_stats, 16) 


plotlist <- lapply(seq_along(gasp_guides), function(i) {
    plot_umi_histogram(grna_odm_gasp[gasp_guides[i],],
                       title = paste0("Gasperini: ", i, " / ", length(gasp_guides)),
                       x_breaks = c(1,10,100,1000,10000,100000)) 
  })


cowplot::plot_grid(plotlist = plotlist)
```


ok yeah this set of 16 at least all looks reasonable, and we have the justification of it being what the scientists did.




# 2. Replogle

From the paper:

Reads from the sgRNA libraries were mapped with Cell Ranger. To account for differences in sequencing depths across GEM
groups from the same experiment, reads were downsampled to produce a more even distribution of the number of reads per cell
across gemgroups, with a threshold of 1000 reads per cell in the K562 day 8 experiment, 800 reads per cell in the K562 day 6 experiment, and 3000 reads per cell in the RPE1 experiment. Guide calling was performed with a Poisson-Gaussian mixture model as previously described. For each guide, the mixture model was fit 100 times, selecting the maximum likelihood model from among the fits.
After guide calling, each cell was categorized according to its guide identities as representing a single genetic perturbation or a multiplet (which may arise from lentiviral recombination or multiple cell encapsulation during droplet generation). Only cells bearing two
guides targeting the same gene or a single guide were used for downstream analysis


So: they used a Poisson-Gaussian mixture on the log2 counts. exactly crispat.

so should i do max method, or pois-gauss?

let's look at max. how can i verify reasonableness? i should look at the difference between the max and the next for each cell. if it's big, then this is easy.

```{r}
grna_odm_repl = ondisc::initialize_odm_from_backing_file(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7/sceptre-pipeline/grna.odm"
)

repl_guide_stats = read_csv(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7/sceptre-pipeline/grna_summary_stats.csv",
  show_col_types = FALSE
)

thresh <- 1  # with this thresh, we actually get the real answer
# each entry is a cell, 
large_umis <- vector("list", ncol(grna_odm_repl))
for(i in 1:nrow(grna_odm_repl)) {
  y <- grna_odm_repl[i,]
  
  big_locs <- which(y >= thresh)
  # large_umis[big_locs] 
  for(loc in big_locs) {
    large_umis[[loc]] <- c(large_umis[[loc]], 
                           y[loc] |> setNames(i)
                           )
  }
}

sapply(large_umis, length) |> table()


gaps = sapply(large_umis, function(l) {
  if(length(l) == 0) {
    0
  } else if(length(l) == 1) {
    l[1]
  } else {
    sort(l) |> tail(5) |> diff() |> rev()
  }
})


cell_df <- data.frame(
    grna_n_nonzero = numeric(length(large_umis)),
  grna_n_umis = 0,
  grna_max = 0,
  gap_1 = 0,
  gap_2 = 0
)

cell_df$grna_n_nonzero <- sapply(large_umis, length)
cell_df$grna_n_umis <- sapply(large_umis, sum)
cell_df$grna_max <- sapply(large_umis, max)
cell_df$gap_1 = sapply(gaps, `[`, 1)
cell_df$gap_2 = sapply(gaps, `[`, 2)
```

TEST: from sceptre max assign, is the lot of "0 or 2+" in agreement with what i compute here?
```{r}
zero_or_twoplus <- which(cell_df$grna_n_umis < 5 | (cell_df$grna_max / cell_df$grna_n_umis <= .8))

jacc <- function(a,b) length(intersect(a,b)) / length(union(a,b))

jacc(scep@cells_w_zero_or_twoplus_grnas, zero_or_twoplus)

setdiff(scep@cells_w_zero_or_twoplus_grnas, zero_or_twoplus)
setdiff(zero_or_twoplus, scep@cells_w_zero_or_twoplus_grnas)

diff_idx = setdiff(scep@cells_w_zero_or_twoplus_grnas, zero_or_twoplus)
cell_df[diff_idx,]

```



hmm! so there are 120k cells that don't have an umis 50 or bigger!

how big are gaps?
```{r}
cell_df %>%
  filter(complete.cases(.)) %>%
  `+`(1) |> log10() |>
  sample_n(20000) |>
  GGally::ggpairs()

# cell_df %>%
#   filter(complete.cases(.)) %>%
#   # `+`(1) |> log10() |>
# ggplot(aes(x = grna_max, y = gap_1)) +
#   geom_hex(bins = 40) +
#     scale_fill_viridis_c(
#     # trans = "log10",
#     option = "plasma"
#   ) +
#   # scale_fill_continuous(trans = "log10") +
#   theme_minimal() +
#   labs(
#     x = "max guide UMI for this cell",
#     y = "Difference between largest and 2nd largest guide UMI for this cell",
#     title = "Max vs top difference in guide UMI for each cell"
#   )
```

```{r}
log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = log1p,
  inverse = expm1
)

nice_log1p_breaks <- function(x) {
  rng <- range(x, finite = TRUE)
  breaks <- c(
    0, 1, 2, 5,
    10, 20, 50,
    100, 200, 500,
    1000, 2000, 5000,
    10000, 20000, 50000,
    1e5, 2e5, 5e5,
    1e6
  )
  breaks[breaks >= rng[1] & breaks <= rng[2]]
}

cell_df %>%
  filter(complete.cases(.)) %>%
  # `+`(1) |> log10() |>
ggplot(aes(x = grna_max, y = gap_1)) +
  geom_hex(bins = 40) +
  scale_x_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = scales::label_number(accuracy = 1)
  ) +
  scale_y_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = scales::label_number(accuracy = 1)
  ) +
  scale_fill_viridis_c(
    # trans = "log10",
    option = "plasma",
    name = "Count"
  ) +
  theme_minimal() +
    labs(
    x = "max guide UMI for this cell",
    y = "Difference between largest and 2nd largest guide UMI for this cell",
    title = "Max vs top difference in guide UMI for non-empty cells"
  )
```

ok i don't think max alone will be good


what threshold should i do?







ok let's look at some cell dists?

- sample 

```{r}

path_to_data <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data",
  "replogle-rd7",
  "sceptre-pipeline"
)
scep <- sceptre::read_ondisc_backed_sceptre_object(
  sceptre_object_fp = file.path(path_to_data, "sceptre_object.rds"),
  response_odm_file_fp = file.path(path_to_data, "response.odm"),
  grna_odm_file_fp = file.path(path_to_data, "grna.odm")
)

# good check!!
cor(scep@covariate_data_frame$grna_n_nonzero, sapply(large_umis, length))
cor(scep@covariate_data_frame$grna_n_umis, sapply(large_umis, sum))


scep@covariate_data_frame |>
  ggplot(aes(x = grna_n_umis)) +
  geom_histogram() +
  scale_x_log10()
```




do replogle cells have barcodes that i can use for confident removal? nope. just int idx





ok, so how to get these assignments? do it as a sceptre run, or just directly ... ? well, why not. but i don't need the pipeline. in-mem will be fine probably.


GASPERINI

```{r}
rm(list=ls())
library(sceptre)
source("~/.Rprofile")

dataset <- "gasperini"
THRESHOLD = 5

path_to_data <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data",
  dataset,
  "sceptre-pipeline"
)

path_to_outputs <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  dataset,
  paste0("thresholding-", THRESHOLD)
)
dir.create(path_to_outputs, showWarnings = FALSE)

scep <- sceptre::read_ondisc_backed_sceptre_object(
  sceptre_object_fp = file.path(path_to_data, "sceptre_object.rds"),
  response_odm_file_fp = file.path(path_to_data, "response.odm"),
  grna_odm_file_fp = file.path(path_to_data, "grna.odm")
)

scep = scep |>
  set_analysis_parameters(resampling_mechanism = "permutations", formula_object = ~1) |>
  assign_grnas(method = "thresholding", threshold = THRESHOLD)

assn_mat = get_grna_assignments(scep)

## some checks before writing ~~~~~~~~~~~~~~~~

grna_odm <- file.path(path_to_data, "grna.odm") |> ondisc::initialize_odm_from_backing_file()

# check that guides haven't been shuffled
if(!identical(rownames(assn_mat), rownames(grna_odm))) {
  stop("assignment matrix rownames and odm rownames do not match.")
}

# checking that cells haven't been shuffled
guide_to_check <- seq(1, nrow(grna_odm), by = 100)
guides_agree <- sapply(guide_to_check, function(idx) all.equal(grna_odm[idx,] >= 5, assn_mat[idx,])) |>
  all()
if(!guides_agree) {
  stop("directly computing the thresholding on the ODM leads to different cell-wise results than in the assignment matrix.")
}

## write ~~~~~~~~~~~~~~~~~~~
saveRDS(assn_mat, file.path(path_to_outputs, "grna_assignment_matrix.rds"))

```


now repeating for replogle, except i also need to decide what cells to remove

```{r}
rm(list=ls())
library(sceptre)
source("~/.Rprofile")

dataset <- "replogle-rd7"

path_to_data <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data",
  dataset,
  "sceptre-pipeline"
)

# path_to_outputs <- file.path(
#   .get_config_path("LOCAL_BENCHMARKING_DIR"),
#   "guide_assignment/outputs",
#   dataset,
#   paste0("thresholding-", THRESHOLD)
# )
# dir.create(path_to_outputs, showWarnings = FALSE)

scep <- sceptre::read_ondisc_backed_sceptre_object(
  sceptre_object_fp = file.path(path_to_data, "sceptre_object.rds"),
  response_odm_file_fp = file.path(path_to_data, "response.odm"),
  grna_odm_file_fp = file.path(path_to_data, "grna.odm")
)

scep = scep |>
  set_analysis_parameters(resampling_mechanism = "permutations", formula_object = ~1) |>
  assign_grnas(method = "maximum")
```

```{r}
assn_mat = get_grna_assignments(scep)

## some checks before writing ~~~~~~~~~~~~~~~~

grna_odm <- file.path(path_to_data, "grna.odm") |> ondisc::initialize_odm_from_backing_file()

# check that guides haven't been shuffled
if(!identical(rownames(assn_mat), rownames(grna_odm))) {
  stop("assignment matrix rownames and odm rownames do not match.")
}

# checking that cells haven't been shuffled
guide_to_check <- seq(1, nrow(grna_odm), by = 100)
guides_agree <- sapply(guide_to_check, function(idx) all.equal(grna_odm[idx,] >= 5, assn_mat[idx,])) |>
  all()
if(!guides_agree) {
  stop("directly computing the thresholding on the ODM leads to different cell-wise results than in the assignment matrix.")
}

## write ~~~~~~~~~~~~~~~~~~~
saveRDS(assn_mat, file.path(path_to_outputs, "grna_assignment_matrix.rds"))
```







CRISPAT ANALYSIS

```{r}
path_to_assns = "~/data/projects/sceptre3/benchmarking/guide_assignment/outputs/run_all_3guides_100k/"
curr_grna_mat = read_rds("~/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7_assign_nguides=400_ncells=100k/sceptre/grna_matrix.rds")
crisp_assign_raw <- file.path(path_to_assns, "assignments_crispat_replogle-rd7_assign_nguides=400_ncells=100k.csv") |>
  read_csv(show_col_types = FALSE)
all(crisp_assign_raw$grna_id %in% rownames(curr_grna_mat))
guide2cells =  split(crisp_assign_raw$cell_id, crisp_assign_raw$grna_id)
setequal(names(guide2cells), rownames(curr_grna_mat))



# now i can get umi histograms for the pert cells

min_pert = sapply(names(guide2cells), function(guide) {
  assns = guide2cells[[guide]]
  y = curr_grna_mat[guide,]
  min(y[assns])
}) 

hist(min_pert |> log10(), 50)
sort(min_pert)[1:20]

lowest_guide = "10821_non-targeting_non-targeting_non-targeting"
low_guide = "10073_ZFC3H1_P1P2_ENSG00000133858"
high_guide = "1399_CDC40_P1P2_ENSG00000168438"
highest_guide = "1586_CHMP4B_P1P2_ENSG00000101421"
# ok... very variable. we are going as low as 5 for a bunch of these... 
# so am i comfortable using a umi count of 5? sure, i am for gasp
# so let's do thresh of 5 combined with max ... ?
plot_umi_histogram(curr_grna_mat[lowest_guide,])
mean(curr_grna_mat[lowest_guide,] >= 5) * 100  # veery few considered pert here

plot_umi_histogram(curr_grna_mat[low_guide,])
mean(curr_grna_mat[low_guide,] >= 5) * 100  # veery few considered pert here

plot_umi_histogram(curr_grna_mat[high_guide,])
mean(curr_grna_mat[high_guide,] >= 5) * 100  # veery few considered pert here

# ohhh... the relly high thresholds are for BAD ones ... ??
plot_umi_histogram(curr_grna_mat[highest_guide,])
mean(curr_grna_mat[highest_guide,] >= 5) * 100  # veery few considered pert here


```


ok. i need to justify my choice here. how can i do that?

i pick a threshold that is:

1. scientifically reasonable

2. generally matches the data: i agree that cells with UMI counts below this should not be called pert


is 5 fine? how many cells are likely below this? 



```{r}
mean(min_pert < 5)


cell_has_large_enough = Matrix::colSums(curr_grna_mat >= 5)
mean(cell_has_large_enough >= 1)
# 20% of cells have at least one guide with a UMI >= 5



```

ok.. now how will i go about removing these? loop thru, compute max, ? or just loop thru and any cells with guide >= 5 are safe


```{r}
min_umi_to_keep = 5

cells_too_small <- rep(TRUE, ncol(grna_odm_repl))

for(i in 1:nrow(grna_odm_repl)) {
  y = grna_odm_repl[i,]
  
  cells_too_small[y >= min_umi_to_keep] <- FALSE
}

table(cells_too_small)

# ok yep these agree with the full results!
cell_df$grna_max[cells_too_small] |> table()
cell_df$grna_max[!cells_too_small] |> min()

```

















checking moi of my old assignment methods

```{r}
rm(list=ls())
source("~/.Rprofile")
source_data <- "replogle-rd7"
gene_qc_thresh = 7

# Set up paths
path_to_data <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data", source_data, "sceptre-pipeline"
)

path_to_assigns <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs", source_data, "sceptre-pipeline"
)

scep <- sceptre::read_ondisc_backed_sceptre_object(
  sceptre_object_fp = file.path(path_to_data, "sceptre_object.rds"),
  response_odm_file_fp = file.path(path_to_data, "response.odm"),
  grna_odm_file_fp = file.path(path_to_data, "grna.odm")
)

# scep_assn_mat = readRDS(file.path(path_to_assigns, "grna_assignment_matrix.rds"))
# scep_assn_mat = readRDS(file.path("~/data/projects/sceptre3/benchmarking/guide_assignment/outputs", source_data, "thresholding-5", "grna_assignment_matrix.rds"))
scep_assn_mat = readRDS(file.path("~/data/projects/sceptre3/benchmarking/guide_assignment/outputs", source_data, "maximum", "grna_assignment_matrix.rds"))


grna_target_df = scep@grna_target_data_frame

moi_observed <- Matrix::colSums(scep_assn_mat) |> mean()

total_num_guides <- grna_target_df |>
  filter(! grna_target %in% c("nt_off_target", "unknown")) |>
  pull(grna_id) |>
  unique() |>
  length()

avg_num_guides_per_target <- grna_target_df |>
  filter(! grna_target %in% c("nt_off_target", "unknown")) |>
  group_by(grna_target) |>
  summarize(n=n()) |>
  pull(n) |>
  mean()

gene_summary_stats = read_csv(file.path(path_to_data, "gene_summary_stats.csv"), show_col_types = FALSE)

genes_passing_qc <- gene_summary_stats |>
  filter(gene_n_nonzero * moi_observed / total_num_guides * avg_num_guides_per_target >= gene_qc_thresh) |>
  pull(gene)

cat(length(genes_passing_qc), " genes (out of ", nrow(gene_summary_stats), ") pass gene QC.\n", sep="")
```




tim question: what about using the filtering `remove cells for which the top gRNA contributes <80% of the UMIs`

--> look at max / sum, see where we end up. i suspect we'll have bad behavior from small ones...

i need to see the symmetric diffs between this and my QC




```{r}

cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis
  ) |>
  filter(grna_n_umis > 0) |>
  sample_n(10000) |>
  ggplot(aes(x = grna_n_umis, y = max_over_tot)) +
  # geom_point() +
  geom_jitter(width = .05, height = .025, size = .5, alpha = .4) +
  scale_x_log10() +
  geom_hline(yintercept = .8, color = "red4", linetype="dashed")

cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis
  ) |>
  filter(grna_n_umis > 0) |>
  # sample_n(10000) |>
  ggplot(aes(x = grna_n_umis, y = max_over_tot)) +
  geom_hex(bins = 40) +
  scale_x_log10() +
  # scale_y_continuous(
  #   trans = log1p_trans,
  #   breaks = nice_log1p_breaks,
  #   labels = scales::label_number(accuracy = 1)
  # ) +
  scale_fill_viridis_c(
    trans = "log10",
    option = "plasma",
    name = "Count"
  ) +
  theme_minimal() +
  geom_hline(yintercept = .8, linetype="dashed", color="green") +
  labs(
    x = "cell UMI total (log10 scale)",
    y = "cell max / cell total",
    title = "Max/Tot vs Tot for Replogle cells",
    subtitle = "aaa"
  )

```

what is my QC? max < 5 are removed

```{r}
cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis,
    scep_qc = max_over_tot >= .8,
    my_qc = grna_max >= 5
  ) |>
  group_by(scep_qc, my_qc) |>
  summarize(n=n())
```

```{r}
log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = log1p,
  inverse = expm1
)

nice_log1p_breaks <- function(x) {
  rng <- range(x, finite = TRUE)
  breaks <- c(
    0, 1, 2, 5,
    10, 20, 50,
    100, 500,
    1000,  5000,
    10000,  50000,
    1e5, 5e5,
    1e6
  )
  breaks[breaks >= rng[1] & breaks <= rng[2]]
}


cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis,
    scep_qc = max_over_tot >= .8,
    thresh_qc = grna_max >= 5
  ) |>
  ggplot(aes(x = grna_n_umis)) +
  geom_histogram(bins=50) +
  facet_wrap(scep_qc ~ thresh_qc) +
   scale_x_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = scales::label_number(accuracy = 1)
  ) +
  theme_bw()


```


```{r}
log1p_trans <- scales::trans_new(
  name = "log1p",
  transform = log1p,
  inverse = expm1
)

nice_log1p_breaks <- function(x) {
  rng <- range(x, finite = TRUE)
  breaks <- c(
    0, 1, 2, 5,
    10, 
    100, 
    1000,
    10000, 
    1e5,
    1e6
  )
  breaks[breaks >= rng[1] & breaks <= rng[2]]
}

label_scientific_10 <- function(x) {
  sapply(x, function(z) {
    if (is.na(z)) return(NA_character_)
    if (z <= 10) return(as.character(z))

    exponent <- floor(log10(z))
    coefficient <- z / 10^exponent

    if (abs(coefficient - 1) < 1e-10) {
      paste0("10^", exponent)
    } else {
      paste0(coefficient, " %*% 10^", exponent)
    }
  }) |>
    parse(text = _)
}

cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis,
    pass_sceptre_qc = max_over_tot >= .8,
    pass_thresh_5_qc = grna_max >= 5
  ) |>
  ggplot(aes(x = grna_n_umis)) +
  geom_histogram(bins = 50) +
  facet_wrap(
    pass_sceptre_qc ~ pass_thresh_5_qc,
    labeller = label_both
  ) +
  scale_x_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = label_scientific_10
  ) +
  theme_bw()

cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis,
    pass_sceptre_qc = max_over_tot >= .8,
    pass_thresh_5_qc = grna_max >= 5
  ) |>
  ggplot(aes(x = grna_n_umis, y = grna_max)) +
  # geom_point(size=.5, alpha = .4) +
  geom_hex(bins = 40) +
  facet_wrap(
    pass_sceptre_qc ~ pass_thresh_5_qc,
    labeller = label_both
  ) +
  scale_x_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = label_scientific_10
  ) +
  scale_y_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = label_scientific_10
  ) +
  theme_bw() +
    scale_fill_viridis_c(
    trans = "log10",
    option = "plasma",
    name = "Count"
  ) +
  labs(title = "Replogle: QC differences", x = "cell UMI total", y = "cell max UMI")
```


```{r}
cell_df |>
  mutate(
    max_over_tot = grna_max / grna_n_umis,
    pass_sceptre_qc = max_over_tot >= .8,
    pass_thresh_5_qc = grna_max >= 5
  ) |>
  ggplot(aes(x = grna_n_umis, y = grna_max)) +
  # geom_point(size=.5, alpha = .4) +
  geom_hex(bins = 40) +
  facet_wrap(
    pass_sceptre_qc ~ pass_thresh_5_qc,
    labeller = label_both
  ) +
  scale_x_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = label_scientific_10
  ) +
  scale_y_continuous(
    trans = log1p_trans,
    breaks = nice_log1p_breaks,
    labels = label_scientific_10
  ) +
  theme_bw() +
    scale_fill_viridis_c(
    trans = "log10",
    option = "plasma",
    name = "Count"
  ) +
  labs(title = "Replogle: QC differences", x = "cell UMI total", y = "cell max UMI")
```



ok so scep_qc let's very few in that i don't. those are all junk cells, but only a couple thousand isn't a lot.

my_qc lets MANY cells in that sceptre doesn't. i keep 500k cells. scep_qc keeps 300k. 

let's look at some dists for this
