---
title: "gRNA Assignment Results"
output:
  pdf_document: default
  html_document: default
date: "2026-01-10"
---

```{r, echo=FALSE, include=FALSE}
rm(list=ls())
library(sceptre)
library(tidyverse)
library(purrr)
library(knitr)
library(kableExtra)
library(scales)
library(stringr)

source("~/.Rprofile")

logp1_trans <- function(base = 10) {
  trans <- function(x) log(x + 1, base)
  inv   <- function(x) base^x - 1  # correct inverse
  
  # Major breaks: at base^k - 1
  brks <- function(limits) {
    rng <- limits + 1
    scales::log_breaks(base = base)(rng) - 0
  }
  
  scales::trans_new(
    name = paste0("logp1-", base),
    transform = trans,
    inverse   = inv,
    breaks    = brks,
    domain = c(0, Inf)
  )
}

read_trace_multipart <- function(path, parts) {
  parse_nextflow_time <- function(time_str) {
    hours <- stringr::str_extract(time_str, "\\d+(?=h)") |> as.numeric()
    minutes <- stringr::str_extract(time_str, "\\d+(?=m)") |> as.numeric()
    seconds <- stringr::str_extract(time_str, "[\\d.]+(?=s)") |> as.numeric()
    total_minutes <- 0
    if (!is.na(hours)) total_minutes <- total_minutes + hours * 60
    if (!is.na(minutes)) total_minutes <- total_minutes + minutes
    if (!is.na(seconds)) total_minutes <- total_minutes + seconds / 60
    return(total_minutes)
  }
  
  parse_memory <- function(mem_str) {
    if (is.na(mem_str) || mem_str == "-") return(NA_real_)
    num <- stringr::str_extract(mem_str, "[\\d.]+") |> as.numeric()
    unit <- stringr::str_extract(mem_str, "[A-Z]+")
    if (is.na(unit)) return(NA_real_)
    if (unit == "MB") return(num / 1024)
    else if (unit == "GB") return(num)
    else if (unit == "KB") return(num / (1024^2))
    else return(NA_real_)
  }
  
  traces <- lapply(parts, function(part_name) {
    if (part_name == "") {
      trace_file <- file.path(path, "trace.tsv")
    } else {
      trace_file <- file.path(path, part_name, "trace.tsv")
    }
    read.table(trace_file, header = TRUE, sep = "\t",
               stringsAsFactors = FALSE, check.names = FALSE) |>
      dplyr::mutate(part = ifelse(part_name == "", "single", part_name))
  })
  
  dplyr::bind_rows(traces) |>
    dplyr::mutate(
      method = tolower(gsub("_ASSIGN", "", process)),
      dataset = tag,
      runtime_minutes = sapply(realtime, parse_nextflow_time),
      peak_memory_gb = sapply(peak_rss, parse_memory)
    ) |>
    dplyr::filter(!is.na(runtime_minutes) & !is.na(peak_memory_gb)) |>
    dplyr::arrange(method, dataset, desc(runtime_minutes)) |>
    dplyr::group_by(method, dataset) |>
    dplyr::slice(1) |>
    dplyr::ungroup()
}

plot_workers_per_time_scep_pipeline <- function(trace_file, dataset_name) {
  assign <- trace_file %>%
    filter(process == "assign_grnas") %>%
    mutate(
      start    = as.POSIXct(start,    format = "%Y-%m-%d %H:%M:%OS", tz = "America/New_York"),
      complete = as.POSIXct(complete, format = "%Y-%m-%d %H:%M:%OS", tz = "America/New_York")
    ) %>%
    filter(!is.na(start), !is.na(complete))
  
  events <- bind_rows(
    transmute(assign, time = start,    delta =  1L),
    transmute(assign, time = complete, delta = -1L)
  ) %>%
    arrange(time, delta) %>%   # important: process -1 before +1 if ties (complete is exclusive)
    mutate(active_workers = cumsum(delta))
  
  # Add an initial point so the plot starts at 0
  events <- bind_rows(
    tibble(time = min(events$time), delta = 0L, active_workers = 0L),
    events
  )
  
  p <- ggplot(events, aes(time, active_workers)) +
    geom_step(direction = "hv") +
    labs(x = "Time", y = "Active assign_grnas workers",
         title = paste0("`sceptre-pipeline` parallelization for ", dataset_name)) +
    theme_bw()
  p
}

jaccard <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  if (union == 0) return(NA_real_)
  return(intersection / union)
}

precision <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  denom <- length(set1)
  if (denom == 0) return(NA_real_)
  return(intersection / denom)
}

recall <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  denom <- length(set2)
  if (denom == 0) return(NA_real_)
  return(intersection / denom)
}
F1 <- function(set1, set2) {
  inter <- length(intersect(set1, set2))
  denom <- length(set1) + length(set2)
  if (denom == 0) return(NA_real_)  # or return(1) if you want "both empty => perfect match"
  2 * inter / denom
}
metrics <- list(
  Jaccard = jaccard,
  Precision = precision,
  Recall = recall,
  F1 = F1
)

objs_to_not_rm <- c("objs_to_not_rm", "read_trace_multipart", "plot_workers_per_time_scep_pipeline",
                    ".get_config_path", "logp1_trans", "metrics")
```



Outline

1. Overview of what we are benchmarking

  a. summary of methods
  
  b. summary of datasets

2. Results

  a. Simulated data: ground truth analysis, and computational results
  
  b. Real data: comparison between methods, and computational results


    
    
# 1. Overview of what we are benchmarking

## 1a. Methods

For this part of the project, we are assessing five methods (two of which are versions of sceptre) on two real datasets and one simulated dataset. 


a. `pertpy`'s `assign_mixture_model()`

- They fit a 2-component mixture to the non-zero log2-transformed counts, with the (log2) counts for the non-perturbed cells being modeled as $\text{Pois}(\lambda)$ and the (log2) perturbed counts are modeled as $\mathcal N(\mu, \sigma)$. The mixing proportions are $\pi = (\pi_0, \pi_1)$, for 4 parameters in total

- They do a "separation penalty" where they add 10 to the log posterior when $\lambda < \mu$, to encourage the Poisson component to stay left of the Gaussian and to avoid component flipping

- Cells with 0 UMI counts for a given guide are always assigned to not express that guide. Guides with 1 non-zero cell are skipped.

- There is no normalizing of any sort, and no information is shared between the assignment steps for each guide. No cell-level statistics are used for normalization, or anything like that

- Low and high MOI are handled the exact same

- The model is fit using NUTS MCMC via NumPyro

- Priors: $\mu \sim \mathcal N(\mu_0, \sigma_0)$ with a default of $(\mu_0, \sigma_0) = (3, 2)$; $\sigma \sim \text{HalfNormal}(\tau_0)$ with a default of $\tau_0 = 1$; $\lambda \sim \text{Exp}(r)$ with a default of $r = 0.2$; and $\pi \sim \text{Dirichlet}(\alpha_0, \alpha_1)$ with a default of $\alpha_1 = 0.15$. 

- By default, 50 warmup samples are done and then 100 samples are kept

- Hard-coded is that each guide starts with the same seed. This may have weird consequences


b. `CLEANSER`

- There are two models here: CROP-seq (`--cs`) and Direct Capture (`--dc`)

- CROP-seq: ambient UMI counts modeled as $\text{Pois}(\lambda)$, perturbed counts as $\text{NegBin}(\mu_1,\theta_1)$ (mean and overdispersion, respectively). Parameter bounds: $\lambda \in [10^{-6}, 5]$; $\mu_1 \geq \lambda$; and $\theta_1 \geq 10^{-6}$ 

- Direct Capture: ambient model is $\text{NegBin}(\mu_0,\theta_0)$, and perturbed model is $\text{NegBin}(\mu_1,\theta_1)$. Parameter bounds: $\mu_0 \geq 10^{-6}$; $\theta_0 \geq 0.1$; $\mu_1 \geq 5$; $\theta_1 \geq 0.1$.

- In both cases, the mixing proportion is $r \in [10^{-6}, 10^{-1}]$

- Each guide only uses cells with non-zero counts. A cell with all zero counts will be completely excluded. They do a "zero truncation correction" to remove the bias introduced in the parameter estimates due to only fitting on the positive counts.

- Normalization: each cell $c$ gets a factor $L_c$ defined as (cell library size) / (average library size).  $L_c$ is used to scale each cell's perturbed component mean parameter, so e.g. the mean parameter for cell $c$ in CROP-seq would be $L_c \mu_1$, not just $\mu_1$. For DC, the ambient mean parameter is also scaled by $L_c$. This is the only information shared between guides. One complexity: only cell UMI counts at or below the `--lpf` value are counted towards the cell library size. By default `lpf = 2`, so really this is like the ambient library size. The idea is to exclude the perturbed UMI counts from the library size. Setting `lpf = 0` means no filtering is done. [Note: in the fishash paper, they stick to the default `lpf` everywhere as we do too]. If a cell's library size is computed as $0$ due to the `lpf` filtering, they set it to $1$. 

- They directly model the raw counts, aside from the filtering and normalization described above

- No MOI settings

- The model is fit via Stan, with the following parameters: in both cases $r \sim \text{Beta}(1, 10)$.

- For CS, $\lambda \sim \text{LogNormal}(\log(1.1),\log(1.1))$; $\mu_1 \sim \text{LogNormal}(\log(100),\log(100))$; and $\theta_1 \sim \text{Beta}(1, 10)$

- For DC, $\mu_0 \sim \text{LogNormal}(\log(1.075),\log(1.1))$ and $\theta_0 \sim \text{LogNormal}(\log(1.1),\log(1.1))$; and $\mu_1 \sim \text{LogNormal}(\log(100),\log(3))$ and $\theta_1 \sim \text{LogNormal}(\log(3),\log(2))$

- Other MCMC parameters: 4 chains by default; 300 warmup iterations and then 1000 samples kept per chain; HMC/NUTS via Stan 

- Each guide runs on a unique seed and is a totally independent fitting, aside from the $L_c$

- The outputted posterior probabilities for each guide-cell pair is the median over all MCMC samples (so over all 4000 samples by default)

c. `crispat`'s `ga_poisson_gauss()`

- They model the log2 ambient counts as $\text{Pois}(\lambda)$ (technically a continuous Poisson) with a prior of $\lambda \sim \text{LogNormal}(0, 1)$, and the log2 perturbed counts as $\mathcal N (\mu, \sigma)$ with priors of $\mu \sim \mathcal N(3,2)$ and $\sigma \sim \text{LogNormal}(2, 1)$. The mixing weights are $\pi = (\pi_0, \pi_1) \sim \text{Dirichlet}(0.9, 0.1)$. 

- Only the non-zero raw counts are considered for each guide. UMIs of 0 do not get used. Guides with only 1 non-zero cell or where the max UMI count is at most 1 are skipped. 

- Each guide is fit totally independently. No information is shared

- There are no MOI settings

- The model is fit using Stochastic Variational Inference via Pyro. Then they do MAP estimation, and assign a cell as perturbed if P(Gaussian component | data) > 0.5. 

d. `SCEPTRE`

For `SCEPTRE`, we have two different implementations of the same underlying algorithm:

1. The `sceptre` R package

2. The `sceptre-pipeline` nextflow pipeline

TODO: fill this in

## 1b. Datasets

Currently we are benchmarking on the following datasets:

1. Simulated

2. Gasperini (high MOI)

3. Replogle-RD7 (low MOI)

TODO: add details on real datasets
```{r, eval=FALSE, echo=FALSE}
real_dataset_names <- c("gasperini", "replogle-rd7")
simulated_dataset_names <- c()
```



### 1b(i): The simulated datasets 

- **Assignment step:** Run `sceptre-pipeline` on Replogle-rd7 with all default settings to get guide assignments for this dataset using the mixture method.

- **Guide choice:** We only consider the guide `8645_TACC3_P1_ENSG00000013810`. This guide was chosen by first restricting to the guides between the 95\% and 99\% quantiles for total number of expressed cells, as determined by `sceptre-pipeline`, and then picking the guide from among those with the greatest number of non-zero cells. The idea behind this was to find a highly expressed guide that we could downsample and still have realistic results for.

- **Model fitting:** Fit two models on the real data using cell covariates $\log(\text{grna\_n\_nonzero}+1)$ and
$\log(\text{grna\_n\_umis}+1)$:
- Negative binomial GLM for UMI counts: $\text{umi\_counts} \sim \text{covariates}$
- Logistic regression for perturbation probability: $\text{sceptre\_assignments} \sim \text{covariates}$

- **Cell subsampling:** Subsample to 100,000 cells, keeping all originally perturbed cells plus randomly sampled
non-perturbed cells

- **Simulation procedure:** For each simulated guide:
- Generate perturbation indicators via $\text{Bernoulli}(\text{expit}(\hat{\eta}_{\text{pert}} + \beta_0))$,
where $\hat{\eta}_{\text{pert}}$ is the fitted logit probability and $\beta_0$ is an intercept adjustment. $\beta_0$ will be varied to change the overall perturbation rate.
- Generate UMI counts from a two-component negative binomial mixture:
$$Y_i \mid Z_i \sim \text{NegBin}(\mu_i, \theta_i)$$
$$\log(\mu_i) = \hat{\eta}_{\text{umi},i} + \beta_{\text{pert}} \cdot Z_i + \beta_{\text{non-pert}} \cdot (1 -
Z_i)$$
where $Z_i$ is the perturbation indicator, $\hat{\eta}_{\text{umi},i}$ is the fitted log-mean, $\theta_i =
\theta_{\text{pert}}$ if perturbed or $\theta_{\text{non-pert}}$ otherwise
- Clip simulated UMI counts to the maximum observed value in real data

- **Run 1 (replogle-rd7_simulated_100k_0.015-pert):**
- 25 simulated guides
- Parameters: $\beta_0 = 2$, $\beta_{\text{non-pert}} = -5$, $\beta_{\text{pert}} = 5.25$,
$\theta_{\text{pert}} = 10$, $\theta_{\text{non-pert}} = 0.1$
- Results in approximately 1.5% perturbation rate

- **Run 2 (replogle-rd7_simulated_100k_0.005-pert):**
- 10 simulated guides
- Parameters: $\beta_0 = 1$, $\beta_{\text{non-pert}} = -5$, $\beta_{\text{pert}} = 5.25$,
$\theta_{\text{pert}} = 10$, $\theta_{\text{non-pert}} = 0.1$
- Results in approximately 0.5% perturbation rate


<!-- i want to make a better plot of my simulated UMI counts. It will be a plot with two rows. The first row will be -->
<!--   histograms of the real and simulated gRNAs, done via faceting, so right now it'll be 3 histograms. Do these on -->
<!--   the log1p-log1p scale that i currently have. Then the bottom row will be 4 plots, where the first two are the -->
<!--   non-perturbed and perturbed UMI counts for simulation1, respectively, and then the same for the second -->
<!--   simulation. I want to do this as a 1x4 faceted histogram so i can have one facet be "dataset", and then the -->
<!--   other "perturbation status". -->

<!--, fig.width=10, fig.height=7, out.width="\\linewidth", fig.align="center"}-->
```{r, echo=FALSE} 
# y_row_idx <- 58
guide <-  "8645_TACC3_P1_ENSG00000013810"

sim_names <- c("replogle-rd7_simulated_100k_0.015-pert", "replogle-rd7_simulated_100k_0.005-pert")
short_names <- c("Sim: 1.5% Pert.", "Sim: 0.5% Pert.")

# getting simulated data
sim_data <- list()
for(sim_name in sim_names) {
  sim_data_fp <- file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data", sim_name
  )
  
  true_pert_status = readRDS(file.path(sim_data_fp, "true_perturbation_status.rds"))[1,]
  simulated_umis = readRDS(file.path(sim_data_fp, "sceptre/grna_matrix.rds"))[1,]
  
  sim_data[[sim_name]] <- data.frame(
    true_pert_status,
    simulated_umis,
    cell_idx =  gsub("cell_idx_", "", names(true_pert_status)) |> as.integer()
  )
}

# quick check: do all of these simulations use the same real cells?
lapply(sim_data, `[[`, "cell_idx") |>
  do.call(what = cbind) |>
  apply(1, function(v) all(v == v[1])) |>
  stopifnot()

# since they do, we can just take the first cell_idx
cell_idx <- sim_data[[1]]$cell_idx

# get actual UMI counts
actual_data_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data/replogle-rd7/sceptre-pipeline/"
)
grna_odm <- ondisc::initialize_odm_from_backing_file(paste0(actual_data_fp, "grna.odm"))
real_umis <- grna_odm[guide, ][cell_idx]

p1_df <- data.frame(
  umis = c(real_umis, lapply(sim_data, `[[`, "simulated_umis") |> unlist()),
  type = rep(c("Real", short_names), each = length(cell_idx)) |>
    factor(levels = c("Real", short_names))
)
p1 <- p1_df |>
  ggplot(aes(x = umis, fill = type)) +
  geom_histogram(bins=30) +
  facet_wrap(~type) +
  theme_bw()  +
  scale_y_continuous(trans = logp1_trans(10)) +
  scale_x_continuous(trans = logp1_trans(10)) +
  ggtitle(bquote("Real vs simulated guides (guide=" * .(guide) * ")")) +
  labs(x = "UMI", fill = "Type", y = "Count")


p2_df <- data.frame(
  umis = lapply(sim_data, `[[`, "simulated_umis") |> unlist(),
  pert_status = lapply(sim_data, `[[`, "true_pert_status") |> unlist(),
  type = rep(short_names, each = length(cell_idx)) |> factor(levels = short_names)
)

p2 <- p2_df |>
  mutate(pert_status = ifelse(pert_status == 1, "Pert.", "Non pert.")) |>
  ggplot(aes(x = umis, fill = pert_status)) +
  geom_histogram(bins=30) +
  facet_wrap(type~pert_status, nrow = 1) +
  theme_bw()  +
  scale_y_continuous(trans = logp1_trans(10)) +
  scale_x_continuous(trans = logp1_trans(10)) +
  ggtitle("Simulated UMI counts by perturbation status") +
  labs(x = "UMI", fill = "Pert. status", y = "Count")


cowplot::plot_grid(p1, p2, nrow = 2)
```


<!-- Method notes: -->

<!-- - CLEANSER had to be run separately on each piece, as it ran out of memory when all 25 simulated gRNAs were provided at once -->

<!-- - sceptre was allowed to control for `grna_n_nonzero` and `grna_n_umis`  -->



# 2. Results

## 2a. Results on simulated data

```{r, echo=FALSE, include=FALSE}
rm(list=setdiff(ls(), objs_to_not_rm))

process_sceptre <- function(scep_output_fp, assn_list) {
  scep_assign_raw <- readRDS(scep_output_fp)
  
  scep_assign <- assn_list
  for(grna_name in names(scep_assign)) {
    scep_assign[[grna_name]] <- which(scep_assign_raw[grna_name,])
  }
  scep_assign
}

process_crispat <- function(crisp_output_fp, assn_list) {
  crisp_assign_raw <- read.csv(crisp_output_fp)
  unordered_cell_list <- split(crisp_assign_raw$cell_id, crisp_assign_raw$grna_id)
  crisp_assign <- assn_list
  for(grna_name in names(unordered_cell_list)) {
    crisp_assign[[grna_name]] <- unordered_cell_list[[grna_name]]
  }
  crisp_assign
}

process_pert <- function(pert_output_fp, assn_list) {
  pert_assign_raw <- read.csv(pert_output_fp)
  
  pert_assign <- assn_list

  for(grna_name in names(pert_assign)) {
    pert_assign[[grna_name]] <- pert_assign_raw$cell_id[grepl(pattern = grna_name, x = pert_assign_raw$grna_id)]
  }
  pert_assign
}

process_clns <- function(clns_output_fp, assn_list, ground_truth) {
  clns_assign_raw <- read.csv(clns_output_fp)
  unordered_cell_list <- split(clns_assign_raw$cell_id, clns_assign_raw$grna_id)
  clns_assign <- assn_list
  for(i in names(unordered_cell_list)) {
    grna_name <- rownames(ground_truth)[[as.integer(i)]]
    clns_assign[[grna_name]] <- unordered_cell_list[[i]]
  }
  clns_assign
}

# process_clns_parts <- function(clns_parts_fp, assn_list) {
#   clns_assign <- assn_list
#   fnames <- dir(clns_parts_fp)
#   for(i in 1:length(fnames)) {
#     clns_assign_guide <- read.table(
#       paste0(clns_parts_fp, "/posteriors_row_", i, ".csv"),
#       header = TRUE,
#       sep = "",
#       col.names = c("grna_id", "cell_id", "posterior")
#     )
#     # assn_list matches with the order of these files
#     clns_assign[[i]] <- clns_assign_guide$cell_id[clns_assign_guide$posterior > .8]
#   }
#   clns_assign
# }



sim_results_objs_to_keep <- c("process_sceptre", "process_pert", "process_crispat", "process_clns", "sim_results_objs_to_keep")
```


```{r, echo=FALSE, include=FALSE}
rm(list=setdiff(ls(), c(objs_to_not_rm, sim_results_objs_to_keep)))

use_cell_names <- TRUE # use cell names vs integer cell idx
verbose <- TRUE

sim_names <- list(
  list(dataset_name = "replogle-rd7_simulated_100k_0.015-pert",
       run_name = "run_all_sims_100k",
       short_name = "Sim: 1.5% Pert."),
  list(dataset_name = "replogle-rd7_simulated_100k_0.005-pert",
       run_name = "run_all_sims_100k_0.005-pert",
       short_name = "Sim: 0.5% Pert.")
)

all_results <- vector("list", length(sim_names)) |>
  setNames(sapply(sim_names, `[[`, "short_name"))
ground_truths <- all_results

for(sim_name in sim_names) {
  
  dataset_name <- sim_name$dataset_name
  run_name <- sim_name$run_name
  short_name <- sim_name$short_name
  
  input_base_fp <- file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data",
    dataset_name
  )
  
  output_base_fp <- file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/outputs",
    run_name
  )
  
  # 1. loading the ground truth matrix, and turning into a list indexed by guide of
  # the cells actually perturbed for that guide 
  ground_truth_mat <- readRDS(file.path(input_base_fp, "true_perturbation_status.rds"))
  
  # doing this to ensure the same structure for each method's assignments on this dataset 
  assn_list_template <- vector("list", length = nrow(ground_truth_mat)) |>
    setNames(rownames(ground_truth_mat))
  
  true_assn_list <- assn_list_template
  for(grna_name in names(true_assn_list)) {
    true_assn_list[[grna_name]] <- which(ground_truth_mat[grna_name, ] == 1)
    if(use_cell_names) {
      true_assn_list[[grna_name]] <- names(true_assn_list[[grna_name]])
    }
  }
  
  ground_truths[[short_name]] <- true_assn_list
  
  # 2. loading the results for each method and processing to make sure we're using
  # either cell names or cell idx consistently
  scep_assign <- process_sceptre(
    paste0(output_base_fp, "/assignment_matrix_sceptre_", dataset_name, ".rds"),
    assn_list = assn_list_template
  )
  crisp_assign <- process_crispat(
    paste0(output_base_fp, "/assignments_crispat_", dataset_name, ".csv"),
    assn_list = assn_list_template
  )
  pert_assign <- process_pert(
    paste0(output_base_fp, "/assignments_pertpy_", dataset_name, ".csv"),
    assn_list = assn_list_template
  )
  clns_assign <- process_clns(
    paste0(output_base_fp, "/assignments_cleanser_", dataset_name, ".csv"),
    assn_list = assn_list_template,
    ground_truth = ground_truth_mat
  )
  
  if(use_cell_names) {
    # sceptre and cleanser have integer idx not cell names, so they need to be transformed
    cell_names <- colnames(ground_truth_mat)
    for(guide in names(scep_assign)) {
      scep_assign[[guide]] <- cell_names[scep_assign[[guide]]]
      clns_assign[[guide]] <- cell_names[clns_assign[[guide]]]
    }
  }
  
  # 3. collecting into a single data structure
  all_results[[short_name]] <- list(
    scep = scep_assign,
    crisp = crisp_assign,
    pert = pert_assign,
    clns = clns_assign
  )
  
  if(verbose) {
    cat("All methods loaded for dataset_name =", dataset_name, "\n")
    cat("\t", nrow(ground_truth_mat), "guides and", ncol(ground_truth_mat), "cells.\n")
    
  }
}

```



```{r, echo = FALSE}
dataset_names <- names(all_results)
method_names  <- names(all_results[[1]])
method_renamer <- c(scep = "SCEPTRE", crisp = "crispat", clns = "CLEANSER", pert = "pertpy")

# metrics is your named list of functions: jaccard, precision, recall
# ground_truths[[i]][[k]] is the truth set for dataset i, guide k

make_metric_df <- function(metric_fun, metric_name) {
  rows <- list()
  idx <- 1L

  for (d in seq_along(all_results)) {
    ds_name <- names(all_results)[d]
    truth_d <- ground_truths[[d]]

    for (m in seq_along(all_results[[d]])) {
      meth_name <- names(all_results[[d]])[m]
      preds <- all_results[[d]][[m]]  # list over guides

      stopifnot(length(preds) == length(truth_d))

      vals <- vapply(seq_along(preds), function(k) {
        metric_fun(preds[[k]], truth_d[[k]])
      }, numeric(1))

      rows[[idx]] <- data.frame(
        dataset = ds_name,
        method  = meth_name,
        value   = vals,
        metric  = metric_name
      )
      idx <- idx + 1L
    }
  }

  bind_rows(rows) %>%
    mutate(
      dataset = factor(dataset, levels = dataset_names),
      method  = factor(method, levels = method_names),
      metric  = factor(metric, levels = names(metrics))
    )
}

# Build one combined DF with all metrics
metric_df <- bind_rows(
  lapply(names(metrics), function(nm) make_metric_df(metrics[[nm]], nm))
) |>
  mutate(
    value = ifelse(is.na(value), 0, value),
    method = method_renamer[as.character(method)] |>
      factor(levels = method_renamer)
  )

# Plot: one 2x4 histogram grid per metric
plots <- lapply(levels(metric_df$metric), function(metric_name) {
  df <- metric_df %>% filter(metric == metric_name)

  ggplot(df, aes(x = value)) +
    geom_histogram(bins = 40, boundary = 0) +
    facet_grid(dataset ~ method) +
    theme_bw() +
    labs(
      title = paste("Guide-level", metric_name, "vs True Perturbation Status"),
      x = metric_name,
      y = "Number of guides"
    ) +
    coord_cartesian(xlim = c(0, 1)) +
    theme(axis.text.x = element_text(size = 8))
})

names(plots) <- levels(metric_df$metric)

plots$Jaccard
plots$Precision
plots$Recall

```


```{r, echo=FALSE}
rm(make_metric_df, metric_df, plots)

summ_mean_q <- function(x, probs = c(0.025, 0.975)) {
  x <- x[!is.na(x)]
  if (!length(x)) return(c(avg = NA_real_, q025 = NA_real_, q975 = NA_real_))
  qs <- as.numeric(quantile(x, probs = probs, names = FALSE, type = 7))
  c(avg = mean(x), q025 = qs[1], q975 = qs[2])
}

# ---- guide-level table ----
guide_df <- bind_rows(lapply(seq_along(all_results), function(di) {
  ds_name <- names(all_results)[di]
  truth   <- ground_truths[[di]]

  bind_rows(lapply(seq_along(all_results[[di]]), function(mj) {
    m_name <- names(all_results[[di]])[mj]
    preds  <- all_results[[di]][[mj]]

    stopifnot(length(preds) == length(truth))

    tibble(
      dataset = ds_name,
      method  = m_name,
      guide   = seq_along(preds),
      jaccard = vapply(seq_along(preds), \(k) metrics$Jaccard(preds[[k]], truth[[k]]), numeric(1)),
      precision = vapply(seq_along(preds), \(k) metrics$Precision(preds[[k]], truth[[k]]), numeric(1)),
      recall    = vapply(seq_along(preds), \(k) metrics$Recall(preds[[k]], truth[[k]]), numeric(1)),
      num_assn_cells = vapply(seq_along(preds), \(k) length(preds[[k]]), numeric(1))
    )
  }))
}))

# ---- dataset × method summary table (what you asked for) ----
summary_df <- guide_df %>%
  group_by(dataset, method) %>%
  summarise(
    Jaccard_avg        = summ_mean_q(jaccard)["avg"],
    Jaccard_lower_0.025 = summ_mean_q(jaccard)["q025"],
    Jaccard_upper_0.975 = summ_mean_q(jaccard)["q975"],

    Precision_avg        = summ_mean_q(precision)["avg"],
    Precision_lower_0.025 = summ_mean_q(precision)["q025"],
    Precision_upper_0.975 = summ_mean_q(precision)["q975"],

    Recall_avg        = summ_mean_q(recall)["avg"],
    Recall_lower_0.025 = summ_mean_q(recall)["q025"],
    Recall_upper_0.975 = summ_mean_q(recall)["q975"],

    num_assn_cells_avg        = summ_mean_q(num_assn_cells)["avg"],
    num_assn_cells_lower_0.025 = summ_mean_q(num_assn_cells)["q025"],
    num_assn_cells_upper_0.975 = summ_mean_q(num_assn_cells)["q975"],
    .groups = "drop"
  ) %>%
  mutate(
    dataset = factor(dataset, levels = dataset_names),
    method  = factor(method, levels = method_names)
  ) %>%
  arrange(dataset, method)

# summary_df

```

<!-- Dataset-level table -->
```{r, include=FALSE}
dataset_attributes <- lapply(ground_truths, function(ground_truth) data.frame(
  avg_num_pert = sapply(ground_truth, length) |> mean() |> round(),
  num_guides = length(ground_truth)
)) |>
  do.call(what = rbind)
# print(dataset_attributes)
```


```{r, echo=FALSE, results='asis', include=FALSE}
dataset_renamer <- c(`Sim: 1.5% Pert.`="1.5% Pert", `Sim: 0.5% Pert.`="0.5% Pert")

# Rounding controls
digits_avg <- 3    # digits for averages (J/P/R)
digits_q   <- 3    # digits for quantiles (J/P/R)

fmt_band <- function(avg, lo, hi, d_avg, d_q) {
  sprintf(paste0("%.", d_avg, "f (%.", d_q, "f, %.", d_q, "f)"), avg, lo, hi)
}

disp_df <- summary_df %>%
  mutate(
    dataset = dplyr::recode(as.character(dataset), !!!dataset_renamer),
    method  = dplyr::recode(as.character(method),  !!!method_renamer),

    Jaccard   = fmt_band(Jaccard_avg,   Jaccard_lower_0.025,   Jaccard_upper_0.975,   digits_avg, digits_q),
    Precision = fmt_band(Precision_avg, Precision_lower_0.025, Precision_upper_0.975, digits_avg, digits_q),
    Recall    = fmt_band(Recall_avg,    Recall_lower_0.025,    Recall_upper_0.975,    digits_avg, digits_q),

    # Num assigned cells: round avg + quantiles to nearest int
    `Num. Assn. Cells` = sprintf(
      "%d (%d, %d)",
      as.integer(round(num_assn_cells_avg)),
      as.integer(round(num_assn_cells_lower_0.025)),
      as.integer(round(num_assn_cells_upper_0.975))
    )
  ) %>%
  select(dataset, method, Jaccard, Precision, Recall, `Num. Assn. Cells`)

disp_df %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Mean (2.5\\%, 97.5\\%) over guides"
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("hold_position", "striped", "scale_down"),
    font_size = 8
  ) %>%
  kableExtra::column_spec(1, width = "1.0in") %>%  # dataset
  kableExtra::column_spec(2, width = "0.9in") %>%  # method
  kableExtra::column_spec(3, width = "1.25in") %>% # Jaccard
  kableExtra::column_spec(4, width = "1.25in") %>% # Precision
  kableExtra::column_spec(5, width = "1.25in") %>% # Recall
  kableExtra::column_spec(6, width = "1.15in")     # Num cells

```

<!-- NOTE: i manually move the \addlinespace up one, and bold the best ones per dataset -->

\begin{table}[!h]
\centering
\caption{Mean (2.5\%, 97.5\%) over guides}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{8}{10}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{1.0in}>{\raggedright\arraybackslash}p{0.9in}>{\raggedright\arraybackslash}p{1.25in}>{\raggedright\arraybackslash}p{1.25in}>{\raggedright\arraybackslash}p{1.25in}>{\raggedright\arraybackslash}p{1.15in}}
\toprule
dataset & method & Jaccard & Precision & Recall & Num. Assn. Cells\\
\midrule

\cellcolor{gray!10}{1.5\% Pert} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{0.958 (0.940, 0.969)} & \cellcolor{gray!10}{0.975 (0.959, 0.989)} & \cellcolor{gray!10}{0.982 (0.975, 0.990)} & \cellcolor{gray!10}{1467 (1400, 1544)}\\

1.5\% Pert & crispat & \textbf{0.959 (0.950, 0.971)} & \textbf{0.998 (0.994, 0.999)} & 0.961 (0.952, 0.974) & 1403 (1357, 1469)\\

\cellcolor{gray!10}{1.5\% Pert} & \cellcolor{gray!10}{pertpy} & \cellcolor{gray!10}{0.075 (0.000, 0.809)} & \cellcolor{gray!10}{0.260 (0.002, 1.000)} & \cellcolor{gray!10}{0.081 (0.000, 0.809)} & \cellcolor{gray!10}{1013 (0, 5544)}\\

1.5\% Pert & CLEANSER & 0.431 (0.416, 0.448) & 0.432 (0.417, 0.449) & \textbf{0.995 (0.991, 0.998)} & 3354 (3230, 3447)\\

\addlinespace

\cellcolor{gray!10}{0.5\% Pert} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{0.928 (0.889, 0.964)} & \cellcolor{gray!10}{0.960 (0.904, 0.990)} & \cellcolor{gray!10}{0.966 (0.949, 0.984)} & \cellcolor{gray!10}{538 (488, 596)}\\

0.5\% Pert & crispat & \textbf{0.950 (0.934, 0.964)} & \textbf{0.980 (0.972, 0.988)} & 0.969 (0.949, 0.981) & 528 (497, 563)\\

\cellcolor{gray!10}{0.5\% Pert} & \cellcolor{gray!10}{pertpy} & \cellcolor{gray!10}{0.108 (0.000, 0.750)} & \cellcolor{gray!10}{0.510 (0.014, 0.989)} & \cellcolor{gray!10}{0.110 (0.000, 0.751)} & \cellcolor{gray!10}{258 (0, 1537)}\\

0.5\% Pert & CLEANSER & 0.217 (0.205, 0.229) & 0.217 (0.205, 0.230) & \textbf{0.997 (0.991, 1.000)} & 2454 (2401, 2531)\\
\bottomrule
\end{tabular}}
\end{table}


<!-- another experiment: should sceptre be run on high MOI for these experiments...? that does seem right since each row is independent, so a cell's assignment in one place shouldn't affect it in another!  (TODO sceptre was high low MOI for these) -->

<!-- ```{r} -->
<!-- rm(list=setdiff(ls(), c(objs_to_not_rm, -->
<!--       "process_sceptre", "process_pert", "process_crispat", "process_clns", "process_clns_parts", "jaccard"))) -->

<!-- dataset_names <- c("replogle-rd7_simulated_100k_0.005-pert", "replogle-rd7_simulated_100k_0.015-pert") -->
<!-- run_name <- "run_sceptre_all_sims" -->

<!-- output_base_fp <- file.path( -->
<!--   .get_config_path("LOCAL_BENCHMARKING_DIR"), -->
<!--   "guide_assignment/outputs", -->
<!--   run_name -->
<!-- ) -->

<!-- sims <- list() -->
<!-- for(dataset_name in dataset_names) { -->
<!--   input_base_fp <- file.path( -->
<!--     .get_config_path("LOCAL_BENCHMARKING_DIR"), -->
<!--     "guide_assignment/input_data", -->
<!--     dataset_name -->
<!--   ) -->

<!--   ground_truth <- readRDS(file.path(input_base_fp, "true_perturbation_status.rds")) -->
<!--   assn_list <- vector("list", length = nrow(ground_truth)) |> -->
<!--     setNames(rownames(ground_truth)) -->

<!--   true_assign <- assn_list -->
<!--   for(grna_name in names(true_assign)) { -->
<!--     true_assign[[grna_name]] <- which(ground_truth[grna_name, ] == 1)  -->
<!--   } -->

<!--   scep_assns <- process_sceptre(paste0(output_base_fp, "/assignment_matrix_sceptre_", dataset_name, ".rds"), assn_list = assn_list) -->

<!--   sims[[dataset_name]] <- Map(jaccard, scep_assns, true_assign) |> as.numeric() -->

<!-- } -->

<!-- lapply(sims, sort) -->
<!-- ``` -->





```{r, echo=FALSE}
# rm(list=setdiff(ls(), objs_to_not_rm))
library(dplyr)
library(ggplot2)
library(tidyr)

dataset_renamer <- c(`replogle-rd7_simulated_100k_0.015-pert`="1.5% Pert (25 guides)",
                     `replogle-rd7_simulated_100k_0.005-pert`="0.5% Pert (10 guides)")
method_renamer <- c(sceptre = "SCEPTRE", crispat = "crispat", cleanser = "CLEANSER", pertpy = "pertpy")


trace_data_list <- list()

for(sim_name in sim_names) {
  dataset_name <- sim_name$dataset_name
  run_name <- sim_name$run_name
  short_name <- sim_name$short_name
  
  output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  run_name
)
  
  trace_data_list[[short_name]] <- read_trace_multipart(output_base_fp, parts = "")
}

trace_data <- do.call(rbind, trace_data_list) |>
  mutate(
    dataset = dataset_renamer[as.character(dataset)],
    method = method_renamer[as.character(method)]
  )

# # Create combined faceted plot
trace_data_long <- trace_data |>
  filter(!is.na(runtime_minutes) | !is.na(peak_memory_gb)) |>
  pivot_longer(
    cols = c(runtime_minutes, peak_memory_gb),
    names_to = "metric",
    values_to = "value"
  ) |>
  filter(!is.na(value)) |>
  mutate(
    metric = recode(metric,
                    "runtime_minutes" = "Runtime (minutes)",
                    "peak_memory_gb" = "Peak Memory (GB)")
  )


combined_plot <- ggplot(trace_data_long,
                        aes(x = method, y = value, color = dataset)) +
  geom_point(
    position = position_dodge(width = 0.6),
    size = 3
  ) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_y_log10() +
  theme_bw() +
  labs(
    title = "Simulations: Computational Performance",
    x = "Method",
    y = "Value (log10 scale)",
    color = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(combined_plot)

# Summary table
trace_summary <- trace_data |>
  dplyr::select(method, dataset, runtime_minutes, peak_memory_gb) |>
  arrange(method, dataset)
# print(trace_summary)
```

```{r, echo=FALSE, results='asis', include=FALSE}
tab <- trace_summary |>
  mutate(method = factor(method, levels = method_renamer)) |>
  arrange(dataset, method) |>
  mutate(
    runtime_minutes = round(runtime_minutes, 2),
    peak_memory_gb  = round(peak_memory_gb, 2)
  )

# row ranges per dataset (so group_rows() knows what to span)
grp <- tab |>
  count(dataset, name = "n") |>
  mutate(end = cumsum(n),
         start = end - n + 1)

tab |>
  select(-dataset) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Simulations: Computational Performance",
    col.names = c("Method", "Runtime (min)", "Peak memory (GB)"),
    align = c("l", "r", "r")
  ) |>
  kable_styling(latex_options = "hold_position") %>%
  { 
    out <- .
    for (i in seq_len(nrow(grp))) {
      out <- group_rows(out, grp$dataset[i], grp$start[i], grp$end[i])
    }
    out
  }
```

<!-- NOTE: i manually bolded the best values for each  -->

\begin{table}[!h]
\centering
\caption{Simulations: Computational Performance}
\centering
\begin{tabular}[t]{lrr}
\toprule
Method & Runtime (min) & Peak memory (GB)\\
\midrule
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{0.5\% Pert (10 guides)}}\\
\hspace{1em}SCEPTRE & \textbf{0.21} & \textbf{0.31}\\
\hspace{1em}crispat & 8.70 & 1.10\\
\hspace{1em}CLEANSER & 13.80 & 2.90\\
\hspace{1em}pertpy & 7.77 & 3.90\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{1.5\% Pert (25 guides)}}\\
\hspace{1em}SCEPTRE & \textbf{0.37} & \textbf{0.31}\\
\hspace{1em}crispat & 10.53 & 1.10\\
\hspace{1em}CLEANSER & 50.18 & 8.60\\
\hspace{1em}pertpy & 12.98 & 5.40\\
\bottomrule
\end{tabular}
\end{table}

# 2b. Results on real data

For this section, all methods were run on downsampled versions of the data. This is due to computation time for `pertpy` and memory for `CLEANSER`. These datasets, denoted `replogle-rd7_medium` and `gasperini_medium`, each contain the first 2,500 guides and the first 50,000 cells as they appear in the data. 
  
```{r, echo=FALSE, include=FALSE}
rm(list=setdiff(ls(), objs_to_not_rm))
find_file_in_parts <- function(base_dir, parts, filename) {
  for (part in parts) {
    filepath <- file.path(base_dir, part, filename)
    if (file.exists(filepath)) {
      return(filepath)
    }
  }
  return(NULL)  # Not found in any part
}

# Modified version that works with multi-part runs
make_assn_lists_multipart <- function(dataset_name, output_base_fp, parts = NULL) {
  
  # If parts is NULL, assume single directory (backward compatible)
  if (is.null(parts)) {
    parts <- c("")  # Empty string means output_base_fp itself
  }
  
  process_crispat <- function(assn_list_template) {
    crisp_output_fp <- find_file_in_parts(
      output_base_fp, parts,
      paste0("assignments_crispat_", dataset_name, ".csv")
    )
    
    if (is.null(crisp_output_fp)) {
      warning("No crispat assignments detected. Returning all-empty list.")
      return(assn_list_template)
    }
    
    cat("   crispat found in:", dirname(crisp_output_fp), "\n")
    crisp_assign_raw <- read.csv(crisp_output_fp) |>
      mutate(cell_id = gsub("CELL_", "", cell_id) |> as.integer())
    unordered_cell_list <- split(crisp_assign_raw$cell_id, crisp_assign_raw$grna_id)
    crisp_assign <- assn_list_template
    crisp_assign[names(unordered_cell_list)] <- unordered_cell_list
    crisp_assign
  }
  
  process_pert <- function(assn_list_template) {
    pert_output_fp <- find_file_in_parts(
      output_base_fp, parts,
      paste0("assignments_pertpy_", dataset_name, ".csv")
    )
    
    if (is.null(pert_output_fp)) {
      warning("No pertpy assignments detected. Returning all-empty list.")
      return(assn_list_template)
    }
    
    cat("   pertpy found in:", dirname(pert_output_fp), "\n")
    pert_assign_raw <- read.csv(pert_output_fp) |>
      filter(grna_id != "negative") |>
      mutate(cell_id = gsub("CELL_", "", cell_id) |> as.integer())
    pert_assign <- assn_list_template
    
    dt <- as.data.table(pert_assign_raw)
    
    # 1) Split each cell's "g1+g2+g3" into multiple columns
    spl <- dt[, tstrsplit(grna_id, "\\+", fixed = FALSE)]
    
    # 2) Go long: one row per (cell_id, grna)
    long <- melt(
      cbind(dt[, .(cell_id)], spl),
      id.vars = "cell_id",
      value.name = "grna",
      na.rm = TRUE
    )[, .(cell_id, grna)]
    
    # (optional) drop duplicates if the same guide appears twice for a cell
    setkey(long, grna, cell_id)
    long <- unique(long)
    
    # 3) Aggregate: list of cells per gRNA
    grna_cells_list <- long[, .(cells = list(cell_id)), by = grna]
    grna_cells_list <- setNames(grna_cells_list$cells, grna_cells_list$grna)
    pert_assign[names(grna_cells_list)] <- grna_cells_list
    
    pert_assign
  }
  
  process_clns <- function(assn_list_template, posterior_thresh = 0.8) {
    clns_output_fp <- find_file_in_parts(
      output_base_fp, parts,
      paste0("assignments_cleanser_", dataset_name, ".csv")
    )
    
    if (is.null(clns_output_fp)) {
      warning("No cleanser assignments detected. Returning all-empty list.")
      return(assn_list_template)
    }
    
    cat("   cleanser found in:", dirname(clns_output_fp), "\n")
    clns_assign_raw <- read.csv(clns_output_fp) |>
      filter(posterior >= posterior_thresh) |>
      mutate(grna_id = names(assn_list_template)[grna_id])
    unordered_cell_list <- split(clns_assign_raw$cell_id, clns_assign_raw$grna_id)
    clns_assign <- assn_list_template
    clns_assign[names(unordered_cell_list)] <- unordered_cell_list
    clns_assign
  }
  
  # 0. load sceptre assignment matrix
  scep_output_fp <- find_file_in_parts(
    output_base_fp, parts,
    paste0("assignment_matrix_sceptre_", dataset_name, ".rds")
  )
  
  if (is.null(scep_output_fp)) {
    stop("Sceptre assignment matrix not found in any part!")
  }
  
  cat("   sceptre found in:", dirname(scep_output_fp), "\n")
  scep_assn_mat <- readRDS(scep_output_fp)
  
  # this list is used to keep all methods in the same order
  grna_names <- rownames(scep_assn_mat)
  assn_list_template <- vector("list", length = length(grna_names)) |>
    setNames(grna_names)
  
  # 1. process scep
  scep_assn <- lapply(grna_names, function(grna) which(scep_assn_mat[grna,])) |>
    setNames(grna_names)
  cat("   sceptre processed.\n")
  
  # 2. process other methods
  crisp_assn <- process_crispat(assn_list_template)
  cat("   crispat processed.\n")
  
  pertpy_assn <- process_pert(assn_list_template)
  cat("   pertpy processed.\n")
  
  cleanser_assn <- process_clns(assn_list_template)
  cat("   cleanser processed.\n")
  
  all_assns <- list(
    scep = scep_assn,
    crisp = crisp_assn,
    pert = pertpy_assn,
    cleanser = cleanser_assn
  )
  
  return(all_assns)
}



source("~/.Rprofile")
library(data.table)

output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  "run_all_medium_parts"
)
parts <- paste0("part", 1:3)
repl_assns <- make_assn_lists_multipart(
  "replogle-rd7_medium",
  output_base_fp,
  parts = parts
)
gasp_assns <- make_assn_lists_multipart(
  "gasperini_medium",
  output_base_fp,
  parts = parts
)

all_assns <- list(
  gasperini_medium = gasp_assns,
  replogle_medium = repl_assns
)


```


```{r, echo=FALSE}
method_levels <- c("SCEPTRE", "crispat", "pertpy", "CLEANSER")

compute_pairwise_metrics <- function(ds_assns, ds_name, metrics) {
  stopifnot(length(ds_assns) >= 2)

  method_names <- names(ds_assns)
  if (is.null(method_names) || any(method_names == "")) {
    method_names <- paste0("method_", seq_along(ds_assns))
  }

  # all ordered method pairs by *index*
  pairs <- expand.grid(m1 = seq_along(ds_assns),
                       m2 = seq_along(ds_assns),
                       KEEP.OUT.ATTRS = FALSE,
                       stringsAsFactors = FALSE) |>
    dplyr::filter(m1 != m2)

  purrr::imap_dfr(metrics, function(f, metric_name) {
    purrr::pmap_dfr(pairs, function(m1, m2) {
      a1 <- ds_assns[[m1]]
      a2 <- ds_assns[[m2]]

      # compute per-guide metric
      vals <- purrr::map2_dbl(a1, a2, f)

      tibble(
        dataset  = ds_name,
        metric   = metric_name,
        method_1 = method_names[m1],
        method_2 = method_names[m2],
        guide    = seq_along(vals),
        value    = vals
      )
    })
  })
}
pairwise_metrics <- imap_dfr(all_assns, ~ compute_pairwise_metrics(.x, .y, metrics)) |>
  mutate(
    # nice label that also makes direction obvious for Precision/Recall
    method_pair = paste(method_1, "\u2192", method_2),
    metric = factor(metric, levels = c("Jaccard", "Precision", "Recall"))
  )
pairwise_metrics <- pairwise_metrics |>
  mutate(
    method_1 = recode(method_1, scep = "SCEPTRE", crisp = "crispat", pert = "pertpy", cleanser = "CLEANSER") |>
      factor(levels  = method_levels),
    method_2 = recode(method_2, scep = "SCEPTRE", crisp = "crispat", pert = "pertpy", cleanser = "CLEANSER") |>
      factor(levels  = method_levels),
    method_pair = paste(method_1, "\u2192", method_2),
    value = ifelse(is.na(value), 0, value)
  )

```



```{r, echo=FALSE}

method_levels <- c("SCEPTRE","crispat","pertpy","CLEANSER")

# desired x-axis order for the 6 unique pairs
pair_levels <- expand.grid(m1 = method_levels, m2 = method_levels) |>
  dplyr::filter(match(m1, method_levels) < match(m2, method_levels)) |>
  transmute(pair = paste(m1, "vs", m2)) |>
  pull(pair)

jaccard_unique <- pairwise_metrics |>
  filter(metric == "Jaccard", !is.na(value)) |>
  mutate(
    method_1 = factor(method_1, levels = method_levels),
    method_2 = factor(method_2, levels = method_levels)
  ) |>
  filter(as.integer(method_1) < as.integer(method_2)) |>
  mutate(
    pair = paste(as.character(method_1), "vs", as.character(method_2)),
    pair = factor(pair, levels = pair_levels)
  )

p_jaccard <- ggplot(jaccard_unique, aes(x = pair, y = value)) +
  geom_violin(trim = FALSE, scale = "width", width = 0.9) +
  # geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.4) +
  facet_wrap(~dataset, ncol = 1) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_x_discrete(labels = \(x) str_wrap(x, width = 12)) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text  = element_text(size = 10, face = "bold")
  ) +
  labs(
    title = "Jaccard similarity by method pair",
    x = NULL,
    y = "Jaccard"
  )

p_jaccard
```

```{r, echo=FALSE}
method_levels <- c("SCEPTRE","crispat","pertpy","CLEANSER")

# Ordered list of the 6 unordered pairs (for consistent x-axis ordering)
pair_levels <- as.vector(combn(method_levels, 2, FUN = \(z) paste(z[1], "vs", z[2])))

prep_pair_dir <- function(metric_name) {
  pairwise_metrics |>
    filter(metric == metric_name, !is.na(value)) |>
    mutate(
      method_1 = factor(method_1, levels = method_levels),
      method_2 = factor(method_2, levels = method_levels),

      # Define unordered pair (A,B) using the factor level order
      A = if_else(as.integer(method_1) < as.integer(method_2),
                  as.character(method_1), as.character(method_2)),
      B = if_else(as.integer(method_1) < as.integer(method_2),
                  as.character(method_2), as.character(method_1)),

      pair = factor(paste(A, "vs", B), levels = pair_levels),

      # Binary direction within each unordered pair:
      # left half = A→B, right half = B→A
      side = if_else(as.character(method_1) == A, "A->B", "B->A")
    )
}

plot_split_violin <- function(df, title, ylab) {
  if (requireNamespace("gghalves", quietly = TRUE)) {
    ggplot(df, aes(x = pair, y = value)) +
      gghalves::geom_half_violin(
        data = df |> filter(side == "A->B"),
        aes(fill = side),
        side = "l", trim = FALSE, scale = "width", width = 0.9
      ) +
      gghalves::geom_half_violin(
        data = df |> filter(side == "B->A"),
        aes(fill = side),
        side = "r", trim = FALSE, scale = "width", width = 0.9
      ) +
      facet_wrap(~dataset, ncol = 1) +
      scale_fill_manual(
        values = c("grey75", "grey35"),
        breaks = c("A->B", "B->A"),
        labels = c("A -> B (left half)", "B -> A (right half)")
      ) +
      coord_cartesian(ylim = c(0, 1)) +
      scale_x_discrete(labels = \(x) str_wrap(x, width = 14)) +
      theme_bw(base_size = 9) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top"
      ) +
      labs(title = title, x = NULL, y = ylab, fill = NULL)

  } else {
    # fallback: still readable, but not split halves
    ggplot(df, aes(x = pair, y = value, fill = side)) +
      geom_violin(
        position = position_dodge(width = 0.8),
        trim = FALSE, scale = "width", width = 0.9
      ) +
      facet_wrap(~dataset, ncol = 1) +
      scale_fill_manual(values = c("grey75", "grey35")) +
      coord_cartesian(ylim = c(0, 1)) +
      scale_x_discrete(labels = \(x) str_wrap(x, width = 14)) +
      theme_bw(base_size = 9) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top"
      ) +
      labs(title = paste0(title, " (install gghalves for split violins)"),
           x = NULL, y = ylab, fill = NULL)
  }
}

# Precision
prec_df <- prep_pair_dir("Precision")
p_precision <- plot_split_violin(
  prec_df,
  "Precision by method pair (split by direction)",
  "Precision"
)

# Recall
rec_df <- prep_pair_dir("Recall")
p_recall <- plot_split_violin(
  rec_df,
  "Recall by method pair (split by direction)",
  "Recall"
)

p_precision
p_recall


```



```{r, eval=FALSE, echo=FALSE}

pair_levels <- as.vector(combn(method_levels, 2, FUN = \(z) paste(z[1], "vs", z[2])))

metric_for_table <- pairwise_metrics |>
  filter(metric %in% c("Jaccard","Precision","Recall"), !is.na(value)) |>
  mutate(
    method_1 = factor(method_1, levels = method_levels),
    method_2 = factor(method_2, levels = method_levels),

    A = if_else(as.integer(method_1) < as.integer(method_2),
                as.character(method_1), as.character(method_2)),
    B = if_else(as.integer(method_1) < as.integer(method_2),
                as.character(method_2), as.character(method_1)),
    pair = factor(paste(A, "vs", B), levels = pair_levels),

    direction = paste(as.character(method_1), "->", as.character(method_2))

  ) |>
  # remove Jaccard double counting
  filter(metric != "Jaccard" | as.integer(method_1) < as.integer(method_2))

summary_tbl <- metric_for_table |>
  group_by(dataset, metric, pair, direction) |>
  summarise(
    n = n(),
    mean = mean(value),
    sd = sd(value),
    se = sd / sqrt(n),
    tcrit = qt(0.975, df = n - 1),
    halfwidth = tcrit * se,
    .groups = "drop"
  ) |>
  # Jaccard doesn't have direction (cleaner display)
  mutate(direction = if_else(metric == "Jaccard", "", direction)) |>
  # clamp to [0,1] for display
  mutate(
    lo = pmax(0, mean - halfwidth),
    hi = pmin(1, mean + halfwidth),
    halfwidth = (hi - lo) / 2,
    mean_pm = sprintf("%.3f $\\pm$ %.3f", mean, halfwidth)
  ) |>
  arrange(dataset, metric, pair, direction)

```

```{r, eval=FALSE, echo=FALSE}


method_levels <- c("SCEPTRE","crispat","pertpy","CLEANSER")
pair_levels <- as.vector(combn(method_levels, 2, FUN = \(z) paste(z[1], "vs", z[2])))

jaccard_summary <- pairwise_metrics |>
  filter(metric == "Jaccard", !is.na(value)) |>
  mutate(
    method_1 = factor(method_1, levels = method_levels),
    method_2 = factor(method_2, levels = method_levels)
  ) |>
  # keep only one direction (avoid double count)
  filter(as.integer(method_1) < as.integer(method_2)) |>
  mutate(
    pair = factor(
      paste(as.character(method_1), "vs", as.character(method_2)),
      levels = pair_levels
    )
  ) |>
  group_by(dataset, pair) |>
  summarise(
    n = n(),
    mean = mean(value),
    sd = sd(value),
    se = sd / sqrt(n),
    tcrit = qt(0.975, df = n - 1),
    halfwidth = tcrit * se,
    .groups = "drop"
  ) |>
  # clamp to [0,1] for display
  mutate(
    lo = pmax(0, mean - halfwidth),
    hi = pmin(1, mean + halfwidth),
    halfwidth = (hi - lo) / 2,
    `Mean $\\pm$ 95\\% CI` = sprintf("%.3f $\\pm$ %.3f", mean, halfwidth)
  ) |>
  select(
    Dataset = dataset,
    `Method pair` = pair,
    `Mean $\\pm$ 95\\% CI`
  ) |>
  arrange(Dataset, `Method pair`)

jaccard_summary_print <- jaccard_summary |>
  mutate(
    Dataset = gsub("_", "\\\\_", Dataset),
    `Method pair` = gsub("_", "\\\\_", `Method pair`)  # probably unnecessary, but safe
  )

jaccard_summary_print |>
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,  # keep $\pm$ working
    caption = "Pairwise Jaccard similarity across guides (mean $\\pm$ 95\\% CI).",
    align = c("l", "l", "r")
  ) |>
  kable_styling(latex_options = "hold_position", font_size = 9) |>
  collapse_rows(columns = 1, valign = "top")

```

\begin{table}[!h]
\centering
\caption{Pairwise Jaccard similarity across guides (mean $\pm$ 95\% CI).}
\centering
\fontsize{9}{11}\selectfont
\begin{tabular}[t]{llr}
\toprule
Dataset & Method pair & Mean $\pm$ 95\% CI\\
\midrule
 & SCEPTRE vs crispat & 0.960 $\pm$ 0.006\\
\cmidrule{2-3}
 & SCEPTRE vs pertpy & 0.264 $\pm$ 0.016\\
\cmidrule{2-3}
 & SCEPTRE vs CLEANSER & 0.919 $\pm$ 0.002\\
\cmidrule{2-3}
 & crispat vs pertpy & 0.253 $\pm$ 0.016\\
\cmidrule{2-3}
 & crispat vs CLEANSER & 0.899 $\pm$ 0.006\\
\cmidrule{2-3}
\multirow[t]{-6}{*}{\raggedright\arraybackslash gasperini\_medium} & pertpy vs CLEANSER & 0.256 $\pm$ 0.016\\
\cmidrule{1-3}
 & SCEPTRE vs crispat & 0.953 $\pm$ 0.005\\
\cmidrule{2-3}
 & SCEPTRE vs pertpy & 0.756 $\pm$ 0.015\\
\cmidrule{2-3}
 & SCEPTRE vs CLEANSER & 0.940 $\pm$ 0.006\\
\cmidrule{2-3}
 & crispat vs pertpy & 0.785 $\pm$ 0.015\\
\cmidrule{2-3}
 & crispat vs CLEANSER & 0.972 $\pm$ 0.003\\
\cmidrule{2-3}
\multirow[t]{-6}{*}{\raggedright\arraybackslash replogle\_medium} & pertpy vs CLEANSER & 0.775 $\pm$ 0.015\\
\bottomrule
\end{tabular}
\end{table}

```{r, echo=FALSE}

# optional: pretty method names (edit if yours already are pretty)
pretty_method <- c(
  scep = "SCEPTRE",
  crisp = "crispat",
  pert = "pertpy",
  cleanser = "CLEANSER"
)

cells_per_guide <- imap_dfr(all_assns, function(ds, ds_name) {
  imap_dfr(ds, function(meth_assns, meth_name) {
    tibble(
      dataset = ds_name,
      method  = dplyr::recode(meth_name, !!!pretty_method, .default = meth_name),
      guide   = seq_along(meth_assns),
      n_cells = lengths(meth_assns)
    )
  })
}) |>
  mutate(method = factor(method, levels = c("SCEPTRE","crispat","pertpy","CLEANSER")))

p_cells_per_guide <- ggplot(cells_per_guide, aes(x = method, y = n_cells)) +
  geom_violin(trim = FALSE, scale = "width", width = 0.9) +
  # geom_boxplot(width = 0.12, outlier.shape = NA, alpha = 0.4) +
  facet_wrap(~dataset, ncol = 1) +
  theme_bw(base_size = 9) +
  scale_y_continuous(trans = logp1_trans(base = 10)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Cells assigned per guide",
    x = NULL,
    y = "# cells per guide (log10(x+1) scale)"
  )

p_cells_per_guide

```


```{r, echo=FALSE}
cell_counts_one <- function(meth_assns) {
  lens  <- lengths(meth_assns)
  cells <- unlist(meth_assns, use.names = FALSE)
  guides <- rep.int(seq_along(meth_assns), times = lens)

  tibble(cell = cells, guide = guides) |>
    distinct(cell, guide) |>
    count(cell, name = "n_guides")
}

guides_per_cell <- imap_dfr(all_assns, function(ds, ds_name) {
  imap_dfr(ds, function(meth_assns, meth_name) {
    cell_counts_one(meth_assns) |>
      mutate(
        dataset = ds_name,
        method  = dplyr::recode(meth_name, !!!pretty_method, .default = meth_name)
      )
  })
}) |>
  mutate(method = factor(method, levels = c("SCEPTRE","crispat","pertpy","CLEANSER")))

row1 <- guides_per_cell |>
  filter(dataset == "gasperini_medium") |>
  ggplot(aes(x = n_guides)) +
  geom_histogram(binwidth = 1, boundary = -0.5, closed = "left") +
  # scale_y_continuous(trans = logp1_trans(10)) +
  # scale_x_continuous(trans = logp1_trans(10)) +
  facet_grid(dataset ~ method, scales = "free") +
  theme_bw(base_size = 9) +
  labs(
    title = "Distribution of guides per cell",
    # x = "# guides per cell",
    y = "# cells"
  )
row2 <- guides_per_cell |>
  filter(dataset == "replogle_medium") |>
  ggplot(aes(x = n_guides)) +
  geom_histogram(binwidth = 1, boundary = -0.5, closed = "left") +
    # scale_y_log10()+
  # scale_y_continuous(trans = logp1_trans(10)) +
  facet_grid(dataset ~ method, scales = "free_y") +
  theme_bw(base_size = 9) +
  labs(
    # title = "Distribution of guides per cell",
    x = "# guides per cell",
    y = "# cells"
  )
cowplot::plot_grid(row1, row2, ncol=1)

```

```{r, echo=FALSE}
# If your all_assns method names are still short, map them to pretty ones:
pretty_method <- c(scep = "SCEPTRE", crisp = "crispat", pert = "pertpy", cleanser = "CLEANSER")

summ_one_method <- function(meth_assns) {
  # robust to duplicates within a guide
  ug <- lapply(meth_assns, unique)

  avg_cells_per_grna <- mean(lengths(ug))

  total_memberships <- sum(lengths(ug))                         # sum_g |cells_g|
  unique_cells <- length(unique(unlist(ug, use.names = FALSE))) # |union_g cells_g|
  avg_grnas_per_cell <- total_memberships / unique_cells

  tibble(
    avg_cells_per_grna = avg_cells_per_grna,
    avg_grnas_per_cell = avg_grnas_per_cell
  )
}

summary_counts <- imap_dfr(all_assns, function(ds, ds_name) {
  imap_dfr(ds, function(meth_assns, meth_name) {
    summ_one_method(meth_assns) |>
      mutate(
        dataset = ds_name,
        method  = dplyr::recode(meth_name, !!!pretty_method, .default = meth_name)
      )
  })
}) |>
  mutate(
    method = factor(method, levels = c("SCEPTRE","crispat","pertpy","CLEANSER"))
  ) |>
  arrange(dataset, method) |>
  mutate(
    avg_cells_per_grna = round(avg_cells_per_grna, 1),
    avg_grnas_per_cell = round(avg_grnas_per_cell, 2)
  )

# LaTeX table
summary_counts |>
  transmute(
    Dataset = dataset,
    Method  = method,
    `Avg cells per gRNA` = avg_cells_per_grna,
    `Avg gRNAs per cell` = avg_grnas_per_cell
  ) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Average assignment sizes: cells per gRNA and gRNAs per cell.",
    align = c("l","l","r","r"),
    escape = TRUE   # keeps underscores safe in dataset names
  ) |>
  kable_styling(latex_options = "hold_position", font_size = 9) |>
  collapse_rows(columns = 1, valign = "top")

```



```{r, echo=FALSE}

# Usage:
output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  "run_all_medium_parts"
)

trace_data <- read_trace_multipart(output_base_fp, parts = parts)



# Create combined faceted plot
trace_data_long <- trace_data |>
  filter(!is.na(runtime_minutes) | !is.na(peak_memory_gb)) |>
  pivot_longer(
    cols = c(runtime_minutes, peak_memory_gb),
    names_to = "metric",
    values_to = "value"
  ) |>
  filter(!is.na(value)) |>
  mutate(
    metric = recode(metric,
                    "runtime_minutes" = "Runtime (minutes)",
                    "peak_memory_gb" = "Peak Memory (GB)")
  )

combined_plot_scatter <- ggplot(
  trace_data_long,
  aes(x = method, y = value, color = dataset)
) +
  geom_point(
    position = position_dodge(width = 0.6),
    size = 3
  ) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  scale_y_log10() +
  theme_bw() +
  labs(
    title = "Computational Performance by Method and Dataset",
    x = "Method",
    y = "Value (log10 scale)",
    color = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(combined_plot_scatter)



# Summary table
trace_summary <- trace_data |>
  dplyr::select(method, dataset, runtime_minutes, peak_memory_gb) |>
  arrange(method, dataset)

# print(trace_summary)
```

```{r, echo=FALSE, eval=FALSE}
method_renamer <- c(sceptre = "SCEPTRE", crispat = "crispat", cleanser = "CLEANSER", pertpy = "pertpy")
tab <- trace_summary |>
  mutate(method = method_renamer[as.character(method)] |> factor(levels = method_renamer)) |>
  arrange(dataset, method) |>
  mutate(
    runtime_minutes = round(runtime_minutes, 2),
    peak_memory_gb  = round(peak_memory_gb, 2)
  )

# row ranges per dataset (for group_rows)
grp <- tab |>
  count(dataset, name = "n") |>
  mutate(
    end = cumsum(n),
    start = end - n + 1
  )

tab |>
  select(-dataset) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Real data: Computational Performance",
    col.names = c("Method", "Runtime (min)", "Peak memory (GB)"),
    align = c("l", "r", "r")
  ) |>
  kable_styling(latex_options = "hold_position") %>%
  {
    out <- .
    for (i in seq_len(nrow(grp))) {
      out <- group_rows(out, grp$dataset[i], grp$start[i], grp$end[i])
    }
    out
  }
```

\begin{table}[!h]
\centering
\caption{Real data: Computational Performance}
\centering
\begin{tabular}[t]{lrr}
\toprule
Method & Runtime (min) & Peak memory (GB)\\
\midrule
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{gasperini\_medium}}\\
\hspace{1em}SCEPTRE & \textbf{10.63} & \textbf{0.29}\\
\hspace{1em}crispat & 103.03 & 1.40\\
\hspace{1em}CLEANSER & 217.55 & 54.10\\
\hspace{1em}pertpy & 533.00 & 7.70\\
\addlinespace[0.3em]
\multicolumn{3}{l}{\textbf{replogle-rd7\_medium}}\\
\hspace{1em}SCEPTRE & \textbf{12.43} & \textbf{0.30}\\
\hspace{1em}crispat & 107.92 & 1.40\\
\hspace{1em}CLEANSER & 211.62 & 49.60\\
\hspace{1em}pertpy & 660.48 & 8.40\\
\bottomrule
\end{tabular}
\end{table}

```{r, echo=FALSE}

read_sceptre_pipeline_trace <- function(
  dataset_name,
  base_dir = file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "guide_assignment/outputs")
) {
  trace_file <- file.path(base_dir, dataset_name, "sceptre-pipeline/tracing/trace.tsv")

  if (!file.exists(trace_file)) {
    warning("Sceptre-pipeline trace not found for ", dataset_name)
    return(NULL)
  }

  trace_df <- read.table(
    trace_file,
    header = TRUE,
    sep = "\t",
    stringsAsFactors = FALSE,
    comment.char = "",
    quote = "",
    check.names = FALSE
  )

  # Processes of interest
  assign_grnas_tasks <- trace_df %>% dplyr::filter(process == "assign_grnas")
  combine_task       <- trace_df %>% dplyr::filter(process == "combine_assign_grnas")

  if (nrow(assign_grnas_tasks) == 0) {
    warning("No assign_grnas tasks found in trace for ", dataset_name)
    return(NULL)
  }

  # Robust time parser (handles fractional seconds like ...48.879)
  parse_time <- function(x) as.POSIXct(x, format = "%Y-%m-%d %H:%M:%OS")

  # Overall runtime window: from first assign_grnas start to combine completion (or last assign_grnas)
  first_start <- min(parse_time(assign_grnas_tasks$start), na.rm = TRUE)

  combine_complete <- if (nrow(combine_task) > 0) {
    parse_time(combine_task$complete[1])
  } else {
    max(parse_time(assign_grnas_tasks$complete), na.rm = TRUE)
  }

  total_runtime_sec <- as.numeric(difftime(combine_complete, first_start, units = "secs"))
  total_runtime_min <- total_runtime_sec / 60

  # Memory parsing helper
  parse_memory_gb <- function(mem_str) {
    if (is.na(mem_str) || mem_str == "-") return(NA_real_)

    num <- as.numeric(gsub("\\s*[A-Za-z]+$", "", mem_str))
    unit <- gsub("^[0-9.]+\\s*", "", mem_str)

    if (unit == "GB") return(num)
    if (unit == "MB") return(num / 1024)
    if (unit == "KB") return(num / (1024^2))
    return(NA_real_)
  }

  # Memory metrics across assign_grnas + combine
  all_peak_memory <- sapply(c(assign_grnas_tasks$peak_rss, combine_task$peak_rss), parse_memory_gb)
  max_single_node_memory_gb <- max(all_peak_memory, na.rm = TRUE)

  # --- Parallelization metrics: average/peak workers and average cores ---

  # Consider assign_grnas + combine tasks as "workers"
  tasks_for_parallel <- dplyr::bind_rows(assign_grnas_tasks, combine_task)

  starts   <- parse_time(tasks_for_parallel$start)
  finishes <- parse_time(tasks_for_parallel$complete)

  ok <- !is.na(starts) & !is.na(finishes) & (finishes >= starts)
  starts <- starts[ok]
  finishes <- finishes[ok]

  # Wall time (seconds) over which we average
  wall_sec <- total_runtime_sec
  if (!is.finite(wall_sec) || wall_sec <= 0) {
    warning("Non-positive wall time computed for ", dataset_name)
    return(NULL)
  }

  # Total worker-seconds = sum of task durations
  task_sec <- as.numeric(difftime(finishes, starts, units = "secs"))
  avg_workers <- sum(task_sec) / wall_sec

  # Average requested cores (time-weighted), if cpus column exists
  avg_cores <- NA_real_
  if ("cpus" %in% names(tasks_for_parallel)) {
    cpus <- suppressWarnings(as.numeric(tasks_for_parallel$cpus[ok]))
    if (all(is.na(cpus))) {
      avg_cores <- NA_real_
    } else {
      cpus[is.na(cpus)] <- 1
      avg_cores <- sum(task_sec * cpus) / wall_sec
    }
  }

  # Peak concurrent workers via event sweep
  events <- data.frame(
    time = c(starts, finishes),
    delta = c(rep(1L, length(starts)), rep(-1L, length(finishes)))
  )
  # if start==finish ties, process ends before starts to avoid transient overcount
  events <- events[order(events$time, events$delta), ]

  cur <- 0L
  peak_workers <- 0L
  for (d in events$delta) {
    cur <- cur + d
    if (cur > peak_workers) peak_workers <- cur
  }

  # Number of tasks (nodes) observed
  n_nodes <- nrow(assign_grnas_tasks) + nrow(combine_task)

  data.frame(
    method = "sceptre-pipeline",
    dataset = dataset_name,
    runtime_minutes = total_runtime_min,
    max_single_node_memory_gb = max_single_node_memory_gb,
    avg_workers = avg_workers,
    peak_workers = peak_workers,
    avg_cores = avg_cores,
    n_nodes = n_nodes,
    stringsAsFactors = FALSE
  )
}

# Read sceptre-pipeline data for both datasets
sceptre_pipeline_data <- bind_rows(
  read_sceptre_pipeline_trace("gasperini"),
  read_sceptre_pipeline_trace("replogle-rd7")
)

sceptre_pipeline_long <- sceptre_pipeline_data %>%
  pivot_longer(
    cols = c(runtime_minutes, max_single_node_memory_gb, avg_workers),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = recode(metric,
                    "runtime_minutes" = "Runtime (minutes)",
                    "max_single_node_memory_gb" = "Max Single-Node Memory (GB)",
                    "avg_workers" = "Avg. Concurrent Workers") |>
      factor(levels = c("Runtime (minutes)",
                        "Max Single-Node Memory (GB)",
                        "Avg. Concurrent Workers"))
  )

combined_plot <- ggplot(sceptre_pipeline_long,
                        aes(x = dataset, y = value, fill = dataset)) +
  geom_bar(stat = "identity") +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  theme_bw() +
  labs(
    title = "Sceptre-Pipeline Performance by Dataset",
    x = "Dataset",
    y = "Value",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

combined_plot
```

```{r, echo=FALSE}
dataset_name <- "replogle-rd7"
file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs", dataset_name,
  "sceptre-pipeline/tracing/trace.tsv"
) |>
  read_tsv(show_col_types = FALSE) |>
  plot_workers_per_time_scep_pipeline(dataset_name)

dataset_name <- "gasperini"
file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs", dataset_name,
  "sceptre-pipeline/tracing/trace.tsv"
) |>
  read_tsv(show_col_types = FALSE) |>
  plot_workers_per_time_scep_pipeline(dataset_name)
```

TODO:

- array jobs to make these more overlapping?

- add datasets: barnyard? Fancier simulation?

- run on full datasets instead of medium? rename medium as "downsampled"? maybe do random downsample instead of 1:k? 

- add methods: geomux?

- add assessments: held-out or resampling based? sensitivity to downsampling? F1 vs Jaccard?


<!-- # NOTES FROM fishash PAPER -->

<!-- - barnyard dataset w/ partial ground truth -->

<!-- - different performance for sceptre in theirs vs ours (and other methods too) -->

<!-- - fancier simulation -->

<!-- - different metric for agreement -->

<!-- - add more methods -- geomux? -->

