---
title: "grna assignment results"
output:
  pdf_document: default
  html_document: default
date: "2026-01-10"
---

```{r, echo=FALSE, include=FALSE}
library(sceptre)
library(tidyverse)
source("~/.Rprofile")

logp1_trans <- function(base = 10) {
  trans <- function(x) log(x + 1, base)
  inv   <- function(x) base^x - 1  # correct inverse
  
  # Major breaks: at base^k - 1
  brks <- function(limits) {
    rng <- limits + 1
    scales::log_breaks(base = base)(rng) - 0
  }
  
  scales::trans_new(
    name = paste0("logp1-", base),
    transform = trans,
    inverse   = inv,
    breaks    = brks,
    domain = c(0, Inf)
  )
}
```


outline

- methods

- datasets (real and simulated)

- results: simulation ground truth analysis and real data

    
    
# 1. Methods and datasets

For this part of the project, we are assessing five methods (two of which are versions of sceptre) on two real datasets and one simulated dataset. 

The methods:

1. pertpy's `assign_mixture_model()`: Bayesian Poisson-Gaussian mixture model fit via MCMC with CUDA+JAX support

2. crispat's `ga_poisson_gauss()`: a Poisson-Gaussian mixture fit where the MAP estimates are obtained from a variational procedure

3. cleanser: Bayesian NB-NB mixture fit via MCMC with stan (their `direct-capture` version)

4. sceptre's mixture model: latent variable Poisson GLM fit via EM variant

5. new sceptre-pipeline



The datasets:

a. Gasperini 2019 (high moi)

b. Replogle 2022 (low moi)

c. simulations


## 1a. Simulated data

We obtained a simulated low MOI dataset in the following way:

Inputs:

- a gRNA expression vector from Replogle, denoted $Y$

- cell covariates $z_i$: in this case, the actual `grna_n_nonzero` and `grna_n_umis` from the sceptre covariate dataframe

- the probability a simulated cell is perturbed, denoted $\varphi$

- intercept adjustments $\gamma_0$ and $\gamma_1$ for unperturbed and perturbed cells, respectively, and a new NB dispersion parameter $\theta_0$

Model and simulate:

- do a NB regression of $Y$ on the covariates $Z$ to get their effects.

- draw a new $Y^\star$ from a NB mixture, with each cell's mean being determined by the linear predictor from the NB regression, except also with the intercept adjustments. The overdispersion is $\theta_0$ no matter the perturbation status



More precisely, let $Y = (Y_1, \dots, Y_n) \in \mathbb N^{n}$ represent the observed UMI counts for our chosen actual gRNA. For each cell $i$, let $z_i$ give our cell covariates, and collect these into a matrix $Z \in \mathbb R^{n\times p}$. Do a NB regression of $Y$ on $Z$ to get $\hat\beta$. If we just wanted to simulate from this model, we could do $Y^\star_i\mid z \stackrel{\perp \!\!\perp}\sim \text{NB}(e^{z^T\hat\beta}, \hat\theta)$.

But for our case we need to have some cells be perturbed, so we are sampling from a 2-component NB mixture. Let $X_1,\dots, X_n \stackrel{\text{iid}}\sim \text{Ber}(\varphi)$ be our perturbation indicators, so $X_i = 1$ means cell $i$ is perturbed. 

Let 
$$
\log \mu_i^\star = z_i^T\hat\beta + \gamma_0 (1-X_i) + \gamma_1 X_i
$$
so the effects for each covariate are the same, but the intercept is adjusted to be either $\hat\beta_0 + \gamma_0$ or $\hat\beta_0 + \gamma_1$, depending on which component we are in. We then simulate

$$
Y_i^\star \mid z_i, X_i\stackrel{\perp \!\!\perp}\sim \text{NB}(\mu_i^\star, \theta_{0})
$$
where $\theta_0$ is our externally provided overdispersion.

We can't use $\hat\theta$ here, the overdispersion as estimated by the NB regression, because that is marginal over the perturbation indicators. There is much less overdispersion when we know which component each cell is from. The larger $\gamma_1-\gamma_0$ is, the bigger the perturbation effect size.

For this paper, we use row `58` of the Replogle-rd7 gRNA matrix, which has gRNA id `304_ANAPC15_P1P2_ENSG00000110200`. We regress the resulting UMI counts on `~ 1 + log(grna_n_nonzero+1) + log(grna_n_umis+1)`.

Below is a comparison of the actual UMI counts and a simulated dataset: 
```{r, echo=FALSE, fig.width=10, fig.height=7, out.width="\\linewidth", fig.align="center"}
y_row_idx <- 58

# get actual UMI counts
actual_data_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data/replogle-rd7/sceptre-pipeline/"
)
grna_odm <- ondisc::initialize_odm_from_backing_file(paste0(actual_data_fp, "grna.odm"))
y <- grna_odm[y_row_idx, ]

# getting simulated data
sim_data_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data/replogle-rd7_simulated"
)
X <- readRDS(file.path(sim_data_fp, "true_perturbation_status.rds"))[1,]
y_sim <- readRDS(file.path(sim_data_fp, "sceptre/grna_matrix.rds"))[1,]


plt1 <- data.frame(
  umi = c(y, y_sim),
  group = rep(c("Real", "Sim"), each = length(y))
) |>
  ggplot(aes(x = umi, fill = group)) +
  geom_histogram(bins=30) +
  facet_wrap(~group) +
  theme_bw()  +
  scale_y_continuous(trans = logp1_trans(10)) +
  scale_x_continuous(trans = logp1_trans(10)) +
  ggtitle(bquote("Real vs Sim:" ~ gamma[0] ~ "= -4;" ~ gamma[1] ~ "= 6;" ~ theta[0] ~ "= 4.5;" ~ phi ~ "= 0.001")) +
  labs(x = "UMI", fill = "Type")

plt2 <- data.frame(
  umi = y_sim,
  pert_status = ifelse(X == 1, "Pert.", "Not pert.")
) |>
  ggplot(aes(x = umi, fill = pert_status)) +
  geom_histogram(bins=30) +
  facet_wrap(~pert_status) +
  theme_bw()  +
  scale_y_continuous(trans = logp1_trans(10)) +
  scale_x_continuous(trans = logp1_trans(10)) +
  ggtitle("Simulated UMI counts by perturbation status") +
  labs(x = "Simulated UMI", fill = "Pert. status")

cowplot::plot_grid(plt1, plt2, nrow = 2)
```





# 2. Results

## a. simulated data: comparison to ground truth

```{r, echo=FALSE, include=FALSE}
dataset_name <- "replogle-rd7_simulated_medium"
run_name <- "run_all_sims_medium"

input_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data",
  dataset_name
)

output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  run_name
)

ground_truth <- readRDS(file.path(input_base_fp, "true_perturbation_status.rds"))
sims <- readRDS(file.path(input_base_fp, "sceptre/grna_matrix.rds"))
y_sim <- sims[1,]

assn_list <- vector("list", length = nrow(ground_truth)) |>
    setNames(rownames(ground_truth))

# getting a list form for the ground truth
true_assign <- assn_list
for(grna_name in names(true_assign)) {
  true_assign[[grna_name]] <- which(ground_truth[grna_name, ] == 1)
}

process_sceptre <- function(scep_output_fp, assn_list) {
  scep_assign_raw <- readRDS(scep_output_fp)
  
  scep_assign <- assn_list
  # for(i in seq_along(scep_assign)) {
  for(grna_name in names(scep_assign)) {
    scep_assign[[grna_name]] <- which(scep_assign_raw[grna_name,])
  }
  scep_assign
}

process_crispat <- function(crisp_output_fp, assn_list) {
  crisp_assign_raw <- read.csv(crisp_output_fp)
  unordered_cell_list <- split(crisp_assign_raw$cell_id, crisp_assign_raw$grna_id)
  crisp_assign <- assn_list
  for(grna_name in names(unordered_cell_list)) {
    crisp_assign[[grna_name]] <- unordered_cell_list[[grna_name]]
  }
  crisp_assign
}

process_pert <- function(pert_output_fp, assn_list) {
  pert_assign_raw <- read.csv(pert_output_fp)
  
  pert_assign <- assn_list

  for(grna_name in names(pert_assign)) {
    pert_assign[[grna_name]] <- pert_assign_raw$cell_id[grepl(pattern = grna_name, x = pert_assign_raw$grna_id)]
  }
  pert_assign
}

process_clns <- function(clns_output_fp, assn_list) {
  clns_assign_raw <- read.csv(clns_output_fp)
  unordered_cell_list <- split(clns_assign_raw$cell_id, clns_assign_raw$grna_id)
  clns_assign <- assn_list
  for(i in names(unordered_cell_list)) {
    grna_name <- rownames(ground_truth)[[as.integer(i)]]
    clns_assign[[grna_name]] <- unordered_cell_list[[i]]
  }
  clns_assign
}

scep_assign <- process_sceptre(paste0(output_base_fp, "/assignment_matrix_sceptre_", dataset_name, ".rds"), assn_list = assn_list)
crisp_assign <- process_crispat(paste0(output_base_fp, "/assignments_crispat_", dataset_name, ".csv"), assn_list = assn_list)
pert_assign <- process_pert(paste0(output_base_fp, "/assignments_pertpy_", dataset_name, ".csv"), assn_list = assn_list)
clns_assign <- process_clns(paste0(output_base_fp, "/assignments_cleanser_", dataset_name, ".csv"), assn_list = assn_list)

# warning("!!! hard-coding 'medium' for cleanser !!!")
all_assns <- list(scep = scep_assign, crisp = crisp_assign, pert = pert_assign, clns = clns_assign)

jaccard <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  if (union == 0) return(NA_real_)
  return(intersection / union)
}

all_sims <- lapply(
  all_assns,
  function(assn) Map(jaccard, assn, true_assign) |> as.numeric()
)
```

We ran on the `medium` dataset due to cleanser's memory usage.

```{r, fig.cap="For each of the 25 iid simulated datasets, we get the similarity between the true perturbation status indicators and the estimated perturbation status. Cleanser totally fails; pertpy recovers some (if more MCMC samples are taken, pertpy does better); and sceptre and crispat nearly perfectly recover the ground truth.", echo=FALSE}
method_renamer <- c(clns="cleanser", crisp="crispat", scep="sceptre", pert="pertpy")
data.frame(
  sims = unlist(all_sims),
  method = rep(names(all_sims), each = nrow(ground_truth))
) |>
  mutate(method = method_renamer[method]) |>
  ggplot(aes(x = sims, fill = method)) +
  geom_histogram(bins = 50) +
  facet_wrap(~method) +
  theme_bw() +
  labs(x = "Jaccard Similarity", title = paste0("Similarities by method for ", dataset_name))

```



```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)

# Function to parse Nextflow time format (e.g., "1h 43m 2s", "10m 38s", "5s")
parse_nextflow_time <- function(time_str) {
  if (is.na(time_str) || time_str == "-") {
    return(NA_real_)
  }
  
  # Extract hours, minutes, seconds
  hours <- stringr::str_extract(time_str, "\\d+(?=h)") |> as.numeric()
  minutes <- stringr::str_extract(time_str, "\\d+(?=m)") |> as.numeric()
  seconds <- stringr::str_extract(time_str, "[\\d.]+(?=s)") |> as.numeric()
  
  # Convert to total minutes
  total_minutes <- 0
  if (!is.na(hours)) total_minutes <- total_minutes + hours * 60
  if (!is.na(minutes)) total_minutes <- total_minutes + minutes
  if (!is.na(seconds)) total_minutes <- total_minutes + seconds / 60
  
  return(total_minutes)
}

# Function to parse memory (e.g., "293.2 MB", "1.4 GB", "50.1 GB")
parse_memory <- function(mem_str) {
  if (is.na(mem_str) || mem_str == "-") {
    return(NA_real_)
  }
  
  # Extract number and unit
  num <- stringr::str_extract(mem_str, "[\\d.]+") |> as.numeric()
  unit <- stringr::str_extract(mem_str, "[A-Z]+")
  
  # Convert to GB
  if (unit == "MB") {
    return(num / 1024)
  } else if (unit == "GB") {
    return(num)
  } else if (unit == "KB") {
    return(num / (1024^2))
  } else {
    return(NA_real_)
  }
}

# Function to read and combine trace files from multiple parts
read_trace_multipart <- function(output_base_fp, parts) {
  all_traces <- list()
  
  for (part in parts) {
    trace_file <- file.path(output_base_fp, part, "trace.tsv")
    
    if (file.exists(trace_file)) {
      # cat("Reading trace from:", part, "\n")
      
      trace_df <- read.table(
        trace_file,
        header = TRUE,
        sep = "\t",
        stringsAsFactors = FALSE,
        comment.char = "",
        quote = "",
        check.names = FALSE
      )
      
      trace_df$part <- part
      all_traces[[part]] <- trace_df
    } else {
      warning("Trace file not found in part: ", part)
    }
  }
  
  # Combine all traces
  combined <- bind_rows(all_traces)
  
  # Process the combined data
  combined_processed <- combined |>
    mutate(
      method = tolower(gsub("_ASSIGN", "", process)),
      dataset = tag,
      runtime_minutes = sapply(realtime, parse_nextflow_time),
      peak_memory_gb = sapply(peak_rss, parse_memory)
    ) |>
    # Remove failed runs (where both metrics are NA)
    filter(!is.na(runtime_minutes) | !is.na(peak_memory_gb)) |>
    # If there are duplicates (same method+dataset), keep the one with data
    arrange(method, dataset, desc(runtime_minutes)) |>
    group_by(method, dataset) |>
    slice(1) |>
    ungroup() |>
    select(part, method, dataset, runtime_minutes, peak_memory_gb,
           realtime, peak_rss, `%cpu`, `%mem`) |>
    arrange(method, dataset)
  
  return(combined_processed)
}

output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  "run_all_sims_medium"
)

# trace_data <- read_trace_multipart(output_base_fp, parts = c("part1", "part2"))
trace_data <- read_trace_multipart(output_base_fp, parts = c(""))

# View the data
# print(trace_data)

# Create runtime plot
runtime_plot <- trace_data |>
  filter(!is.na(runtime_minutes)) |>
  ggplot(aes(x = method, y = runtime_minutes, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    title = "Runtime by Method and Dataset",
    x = "Method",
    y = "Runtime (minutes)",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# print(runtime_plot)

# Create memory plot
memory_plot <- trace_data |>
  filter(!is.na(peak_memory_gb)) |>
  ggplot(aes(x = method, y = peak_memory_gb, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    title = "Peak Memory Usage by Method and Dataset",
    x = "Method",
    y = "Peak Memory (GB)",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# print(memory_plot)

# Create combined faceted plot
trace_data_long <- trace_data |>
  filter(!is.na(runtime_minutes) | !is.na(peak_memory_gb)) |>
  pivot_longer(
    cols = c(runtime_minutes, peak_memory_gb),
    names_to = "metric",
    values_to = "value"
  ) |>
  filter(!is.na(value)) |>
  mutate(
    metric = recode(metric,
                    "runtime_minutes" = "Runtime (minutes)",
                    "peak_memory_gb" = "Peak Memory (GB)")
  )

combined_plot <- ggplot(trace_data_long,
                        aes(x = method, y = value, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  theme_bw() +
  labs(
    title = "Computational Performance by Method and Dataset",
    x = "Method",
    y = "Value",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(combined_plot)

# Summary table
trace_summary <- trace_data |>
  select(method, dataset, runtime_minutes, peak_memory_gb) |>
  arrange(method, dataset)

```





# 2. real data 
  

```{r, echo=FALSE, include=FALSE}
# rm(list=ls())
find_file_in_parts <- function(base_dir, parts, filename) {
  for (part in parts) {
    filepath <- file.path(base_dir, part, filename)
    if (file.exists(filepath)) {
      return(filepath)
    }
  }
  return(NULL)  # Not found in any part
}

# Modified version that works with multi-part runs
make_assn_lists_multipart <- function(dataset_name, output_base_fp, parts = NULL) {
  
  # If parts is NULL, assume single directory (backward compatible)
  if (is.null(parts)) {
    parts <- c("")  # Empty string means output_base_fp itself
  }
  
  process_crispat <- function(assn_list_template) {
    crisp_output_fp <- find_file_in_parts(
      output_base_fp, parts,
      paste0("assignments_crispat_", dataset_name, ".csv")
    )
    
    if (is.null(crisp_output_fp)) {
      warning("No crispat assignments detected. Returning all-empty list.")
      return(assn_list_template)
    }
    
    cat("   crispat found in:", dirname(crisp_output_fp), "\n")
    crisp_assign_raw <- read.csv(crisp_output_fp) |>
      mutate(cell_id = gsub("CELL_", "", cell_id) |> as.integer())
    unordered_cell_list <- split(crisp_assign_raw$cell_id, crisp_assign_raw$grna_id)
    crisp_assign <- assn_list_template
    crisp_assign[names(unordered_cell_list)] <- unordered_cell_list
    crisp_assign
  }
  
  process_pert <- function(assn_list_template) {
    pert_output_fp <- find_file_in_parts(
      output_base_fp, parts,
      paste0("assignments_pertpy_", dataset_name, ".csv")
    )
    
    if (is.null(pert_output_fp)) {
      warning("No pertpy assignments detected. Returning all-empty list.")
      return(assn_list_template)
    }
    
    cat("   pertpy found in:", dirname(pert_output_fp), "\n")
    pert_assign_raw <- read.csv(pert_output_fp) |>
      filter(grna_id != "negative") |>
      mutate(cell_id = gsub("CELL_", "", cell_id) |> as.integer())
    pert_assign <- assn_list_template
    
    dt <- as.data.table(pert_assign_raw)
    
    # 1) Split each cell's "g1+g2+g3" into multiple columns
    spl <- dt[, tstrsplit(grna_id, "\\+", fixed = FALSE)]
    
    # 2) Go long: one row per (cell_id, grna)
    long <- melt(
      cbind(dt[, .(cell_id)], spl),
      id.vars = "cell_id",
      value.name = "grna",
      na.rm = TRUE
    )[, .(cell_id, grna)]
    
    # (optional) drop duplicates if the same guide appears twice for a cell
    setkey(long, grna, cell_id)
    long <- unique(long)
    
    # 3) Aggregate: list of cells per gRNA
    grna_cells_list <- long[, .(cells = list(cell_id)), by = grna]
    grna_cells_list <- setNames(grna_cells_list$cells, grna_cells_list$grna)
    pert_assign[names(grna_cells_list)] <- grna_cells_list
    
    pert_assign
  }
  
  process_clns <- function(assn_list_template, posterior_thresh = 0.8) {
    clns_output_fp <- find_file_in_parts(
      output_base_fp, parts,
      paste0("assignments_cleanser_", dataset_name, ".csv")
    )
    
    if (is.null(clns_output_fp)) {
      warning("No cleanser assignments detected. Returning all-empty list.")
      return(assn_list_template)
    }
    
    cat("   cleanser found in:", dirname(clns_output_fp), "\n")
    clns_assign_raw <- read.csv(clns_output_fp) |>
      filter(posterior >= posterior_thresh) |>
      mutate(grna_id = names(assn_list_template)[grna_id])
    unordered_cell_list <- split(clns_assign_raw$cell_id, clns_assign_raw$grna_id)
    clns_assign <- assn_list_template
    clns_assign[names(unordered_cell_list)] <- unordered_cell_list
    clns_assign
  }
  
  # 0. load sceptre assignment matrix
  scep_output_fp <- find_file_in_parts(
    output_base_fp, parts,
    paste0("assignment_matrix_sceptre_", dataset_name, ".rds")
  )
  
  if (is.null(scep_output_fp)) {
    stop("Sceptre assignment matrix not found in any part!")
  }
  
  cat("   sceptre found in:", dirname(scep_output_fp), "\n")
  scep_assn_mat <- readRDS(scep_output_fp)
  
  # this list is used to keep all methods in the same order
  grna_names <- rownames(scep_assn_mat)
  assn_list_template <- vector("list", length = length(grna_names)) |>
    setNames(grna_names)
  
  # 1. process scep
  scep_assn <- lapply(grna_names, function(grna) which(scep_assn_mat[grna,])) |>
    setNames(grna_names)
  cat("   sceptre processed.\n")
  
  # 2. process other methods
  crisp_assn <- process_crispat(assn_list_template)
  cat("   crispat processed.\n")
  
  pertpy_assn <- process_pert(assn_list_template)
  cat("   pertpy processed.\n")
  
  cleanser_assn <- process_clns(assn_list_template)
  cat("   cleanser processed.\n")
  
  all_assns <- list(
    scep = scep_assn,
    crisp = crisp_assn,
    pert = pertpy_assn,
    cleanser = cleanser_assn
  )
  
  return(all_assns)
}



source("~/.Rprofile")
library(data.table)

output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  "run_all_medium_parts"
)
repl_assns <- make_assn_lists_multipart(
  "replogle-rd7_medium",
  output_base_fp,
  parts = c("part1", "part2")
)
gasp_assns <- make_assn_lists_multipart(
  "gasperini_medium",
  output_base_fp,
  parts = c("part1", "part2")
)
```


```{r, echo=FALSE}
# make_sim_plots <- function(results) {
#   methods <- names(results)
#   
#   # gRNAs (align by name, don't assume order)
#   grnas <- Reduce(intersect, lapply(results, names))
#   
#   # helper for NULL coalesce (base R)
#   `%||%` <- function(x, y) if (is.null(x)) y else x
#   
#   # Pre-unique cell vectors once (helps correctness + a bit of speed)
#   results_u <- lapply(results, function(L) lapply(L[grnas], function(x) unique(x %||% integer(0))))
#   
#   
#   
#   jaccard <- function(a, b) {
#     a <- unique(a %||% character(0))
#     b <- unique(b %||% character(0))
#     denom <- length(union(a, b))
#     if (denom == 0L) return(NA_real_)
#     length(intersect(a, b)) / denom
#   }
#   
#   # Build a long data.frame with one "value" vector per panel
#   panel_df <- do.call(rbind, lapply(methods, function(mi) {
#     do.call(rbind, lapply(methods, function(mj) {
#       if (mi == mj) {
#         vals <- vapply(grnas, function(g) length(results_u[[mi]][[g]]), integer(1))
#         data.frame(row = mi, col = mj, value = vals, type = "count")
#       } else {
#         vals <- vapply(grnas, function(g) jaccard(results_u[[mi]][[g]], results_u[[mj]][[g]]), numeric(1))
#         data.frame(row = mi, col = mj, value = vals, type = "jaccard")
#       }
#     }))
#   }))
#   
#   panel_df$row <- factor(panel_df$row, levels = methods)
#   panel_df$col <- factor(panel_df$col, levels = methods)
#   
#   ggplot(panel_df, aes(x = value)) +
#     geom_histogram(bins = 50) +
#     facet_wrap(row ~ col, scales = "free") +
#     theme_bw(base_size = 11) +
#     labs(x = NULL, y = "Number of gRNAs")
# }



`%||%` <- function(x, y) if (is.null(x)) y else x
jaccard <- function(a, b) {
  a <- unique(a %||% character(0))
  b <- unique(b %||% character(0))
  denom <- length(union(a, b))
  if (denom == 0L) return(NA_real_)
  length(intersect(a, b)) / denom
}
  

make_pairwise_sims <- function(results) {
  k <- 1
  sim_names <- c()
  sims <- matrix(0, length(results[[1]]), choose(length(results), 2))
  for(i in 1:(length(results)-1)) {
    for(j in (i+1):length(results)) {
      sims[,k] <- Map(jaccard, results[[i]], results[[j]]) |> as.numeric()
      k <- k + 1
      sim_names <- c(sim_names, paste0(names(results)[i], " vs ", names(results)[j]))
    }
  }
  sims <- as.data.frame(sims) |> setNames(sim_names)
  sims
}


lapply(repl_assns, function(l) sapply(l, length)) |>
  do.call(what = cbind) |>
  GGally::ggpairs(title = "num cells assigned per grna for replogle-rd7_medium")

lapply(gasp_assns, function(l) sapply(l, length)) |>
  do.call(what = cbind) |>
  GGally::ggpairs(title = "num cells assigned per grna for gasperini_medium")

make_pairwise_sims(repl_assns) |> 
  pivot_longer(cols = everything(), names_to = "methods", values_to = "jaccard") |>
  ggplot(aes(x = jaccard)) +
  geom_histogram(bins=30) +
  facet_wrap(~methods) +
  theme_bw() +
  labs(title = "grna assignment similarities for replogle-rd7_medium")





make_pairwise_sims(gasp_assns) |> 
  pivot_longer(cols = everything(), names_to = "methods", values_to = "jaccard") |>
  ggplot(aes(x = jaccard)) +
  geom_histogram(bins=30) +
  facet_wrap(~methods) +
  theme_bw() +
  labs(title = "grna assignment similarities for gasperini_medium")
```


















<!-- ```{r} -->
<!-- results <- repl_assns#[c("scep", "crisp", "cleanser")] # no cleanser for this one -->

<!-- methods <- names(results) -->

<!-- # gRNAs (align by name, don't assume order) -->
<!-- grnas <- Reduce(intersect, lapply(results, names)) -->

<!-- # helper for NULL coalesce (base R) -->
<!-- `%||%` <- function(x, y) if (is.null(x)) y else x -->

<!-- # Pre-unique cell vectors once (helps correctness + a bit of speed) -->
<!-- results_u <- lapply(results, function(L) lapply(L[grnas], function(x) unique(x %||% integer(0)))) -->



<!-- jaccard <- function(a, b) { -->
<!--   a <- unique(a %||% character(0)) -->
<!--   b <- unique(b %||% character(0)) -->
<!--   denom <- length(union(a, b)) -->
<!--   if (denom == 0L) return(NA_real_) -->
<!--   length(intersect(a, b)) / denom -->
<!-- } -->

<!-- # Build a long data.frame with one "value" vector per panel -->
<!-- panel_df <- do.call(rbind, lapply(methods, function(mi) { -->
<!--   do.call(rbind, lapply(methods, function(mj) { -->
<!--     if (mi == mj) { -->
<!--       vals <- vapply(grnas, function(g) length(results_u[[mi]][[g]]), integer(1)) -->
<!--       data.frame(row = mi, col = mj, value = vals, type = "count") -->
<!--     } else { -->
<!--       vals <- vapply(grnas, function(g) jaccard(results_u[[mi]][[g]], results_u[[mj]][[g]]), numeric(1)) -->
<!--       data.frame(row = mi, col = mj, value = vals, type = "jaccard") -->
<!--     } -->
<!--   })) -->
<!-- })) -->

<!-- panel_df$row <- factor(panel_df$row, levels = methods) -->
<!-- panel_df$col <- factor(panel_df$col, levels = methods) -->

<!-- ggplot(panel_df, aes(x = value)) + -->
<!--   geom_histogram(bins = 50) + -->
<!--   facet_wrap(row ~ col, scales = "free") + -->
<!--   theme_bw(base_size = 11) + -->
<!--   labs(x = NULL, y = "Number of gRNAs") -->

<!-- ``` -->





```{r, echo=FALSE}

# Usage:
output_base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs",
  "run_all_medium_parts"
)

trace_data <- read_trace_multipart(output_base_fp, parts = c("part1", "part2"))
# trace_data <- read_trace_multipart(output_base_fp, parts = c(""))

# View the data
# print(trace_data)

# Create runtime plot
runtime_plot <- trace_data |>
  filter(!is.na(runtime_minutes)) |>
  ggplot(aes(x = method, y = runtime_minutes, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    title = "Runtime by Method and Dataset",
    x = "Method",
    y = "Runtime (minutes)",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# print(runtime_plot)

# Create memory plot
memory_plot <- trace_data |>
  filter(!is.na(peak_memory_gb)) |>
  ggplot(aes(x = method, y = peak_memory_gb, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  theme_bw() +
  labs(
    title = "Peak Memory Usage by Method and Dataset",
    x = "Method",
    y = "Peak Memory (GB)",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# print(memory_plot)

# Create combined faceted plot
trace_data_long <- trace_data |>
  filter(!is.na(runtime_minutes) | !is.na(peak_memory_gb)) |>
  pivot_longer(
    cols = c(runtime_minutes, peak_memory_gb),
    names_to = "metric",
    values_to = "value"
  ) |>
  filter(!is.na(value)) |>
  mutate(
    metric = recode(metric,
                    "runtime_minutes" = "Runtime (minutes)",
                    "peak_memory_gb" = "Peak Memory (GB)")
  )

combined_plot <- ggplot(trace_data_long,
                        aes(x = method, y = value, fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  facet_wrap(~metric, scales = "free_y", ncol = 1) +
  theme_bw() +
  labs(
    title = "Computational Performance by Method and Dataset",
    x = "Method",
    y = "Value",
    fill = "Dataset"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(combined_plot)

# Summary table
trace_summary <- trace_data |>
  select(method, dataset, runtime_minutes, peak_memory_gb) |>
  arrange(method, dataset)

# print(trace_summary)
```



```{r, echo=FALSE}
  read_sceptre_pipeline_trace <- function(dataset_name, base_dir = "~/data/projects/sceptre3/benchmarking/guide_assignment/outputs") {
    trace_file <- file.path(base_dir, dataset_name, "sceptre-pipeline/tracing/trace.tsv")

    if (!file.exists(trace_file)) {
      warning("Sceptre-pipeline trace not found for ", dataset_name)
      return(NULL)
    }

    # cat("Reading sceptre-pipeline trace from:", dataset_name, "\n")

    trace_df <- read.table(
      trace_file,
      header = TRUE,
      sep = "\t",
      stringsAsFactors = FALSE,
      comment.char = "",
      quote = "",
      check.names = FALSE
    )

    # Aggregate assign_grnas tasks (run in parallel)
    assign_grnas_tasks <- trace_df %>%
      filter(process == "assign_grnas")

    # Get combine task (runs after all assign_grnas)
    combine_task <- trace_df %>%
      filter(process == "combine_assign_grnas")

    # Calculate overall metrics
    # Runtime: from first assign_grnas start to combine completion
    first_start <- min(as.POSIXct(assign_grnas_tasks$start, format = "%Y-%m-%d %H:%M:%S"))
    combine_complete <- if(nrow(combine_task) > 0) {
      as.POSIXct(combine_task$complete[1], format = "%Y-%m-%d %H:%M:%S")
    } else {
      max(as.POSIXct(assign_grnas_tasks$complete, format = "%Y-%m-%d %H:%M:%S"))
    }

    total_runtime_sec <- as.numeric(difftime(combine_complete, first_start, units = "secs"))
    total_runtime_min <- total_runtime_sec / 60

    # Memory parsing helper
    parse_memory_gb <- function(mem_str) {
      if (is.na(mem_str) || mem_str == "-") return(NA_real_)

      # Extract number and unit
      num <- as.numeric(gsub("\\s*[A-Za-z]+$", "", mem_str))
      unit <- gsub("^[0-9.]+\\s*", "", mem_str)

      # Convert to GB
      if (unit == "GB") return(num)
      if (unit == "MB") return(num / 1024)
      if (unit == "KB") return(num / (1024^2))
      return(NA_real_)
    }

    # Memory metrics
    all_peak_memory <- sapply(c(assign_grnas_tasks$peak_rss, combine_task$peak_rss), parse_memory_gb)

    # Max single-node memory
    max_single_node_memory_gb <- max(all_peak_memory, na.rm = TRUE)

    # Total memory across all nodes
    total_memory_gb <- sum(all_peak_memory, na.rm = TRUE)

    # Number of nodes
    n_nodes <- nrow(assign_grnas_tasks) + nrow(combine_task)

    # Return summary row
    data.frame(
      method = "sceptre-pipeline",
      dataset = dataset_name,
      runtime_minutes = total_runtime_min,
      max_single_node_memory_gb = max_single_node_memory_gb,
      total_memory_gb = total_memory_gb,
      n_nodes = n_nodes,
      stringsAsFactors = FALSE
    )
  }

  # Read sceptre-pipeline data for both datasets
  sceptre_pipeline_data <- bind_rows(
    read_sceptre_pipeline_trace("gasperini"),
    read_sceptre_pipeline_trace("replogle-rd7")
  )

  # View the data
  # print(sceptre_pipeline_data)

  # Runtime plot
  runtime_plot <- sceptre_pipeline_data %>%
    ggplot(aes(x = dataset, y = runtime_minutes, fill = dataset)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = paste0(n_nodes, " nodes")),
              vjust = -0.5, size = 3.5) +
    theme_bw() +
    labs(
      title = "Sceptre-Pipeline Runtime by Dataset",
      x = "Dataset",
      y = "Wall-clock Runtime (minutes)",
      fill = "Dataset"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  # print(runtime_plot)

  # Memory comparison plot - both metrics side by side
  memory_comparison <- sceptre_pipeline_data %>%
    pivot_longer(
      cols = c(max_single_node_memory_gb, total_memory_gb),
      names_to = "memory_metric",
      values_to = "memory_gb"
    ) %>%
    mutate(
      memory_metric = recode(memory_metric,
                            "max_single_node_memory_gb" = "Max Single-Node",
                            "total_memory_gb" = "Total (Sum)")
    )

  memory_plot <- ggplot(memory_comparison,
                        aes(x = dataset, y = memory_gb, fill = memory_metric)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
    theme_bw() +
    labs(
      title = "Sceptre-Pipeline Memory Usage by Dataset",
      x = "Dataset",
      y = "Peak Memory (GB)",
      fill = "Memory Metric"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  # print(memory_plot)

  # Combined faceted plot with all metrics
  sceptre_pipeline_long <- sceptre_pipeline_data %>%
    pivot_longer(
      cols = c(runtime_minutes, max_single_node_memory_gb, total_memory_gb),
      names_to = "metric",
      values_to = "value"
    ) %>%
    mutate(
      metric = recode(metric,
                     "runtime_minutes" = "Runtime (minutes)",
                     "max_single_node_memory_gb" = "Max Single-Node Memory (GB)",
                     "total_memory_gb" = "Total Memory (GB)") |>
        factor(levels = c("Runtime (minutes)", "Max Single-Node Memory (GB)", "Total Memory (GB)") )
    )

  combined_plot <- ggplot(sceptre_pipeline_long,
                          aes(x = dataset, y = value, fill = dataset)) +
    geom_bar(stat = "identity") +
    facet_wrap(~metric, scales = "free_y", ncol = 1) +
    theme_bw() +
    labs(
      title = "Sceptre-Pipeline Performance by Dataset",
      x = "Dataset",
      y = "Value",
      fill = "Dataset"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(combined_plot)
```

