---
title: "defense-assignment-analysis"
output: html_document
date: "2026-04-11"
---



```{r, echo=FALSE}
library(Matrix)
library(tidyverse)
library(data.table)

rm(list=ls())
source("~/.Rprofile")

theme_presentation <- function(base_size = 16) {
  theme_bw(base_size = base_size) %+replace% 
    theme(
      # Increase title sizes
      plot.title = element_text(size = rel(1.4), face = "bold", margin = margin(b = 10)),
      plot.subtitle = element_text(size = rel(1.1), margin = margin(b = 10)),
      
      # Axis text and titles
      axis.title = element_text(size = rel(1.1)),
      axis.text = element_text(size = rel(0.9), color = "black"),
      
      # Legend positioning and size
      legend.position = "bottom",
      legend.text = element_text(size = rel(0.9)),
      legend.title = element_text(size = rel(1.0), face = "bold"),
      
      # Facet label (strip) sizes - critical for your multi-panel plots
      strip.text = element_text(size = rel(1.1), face = "bold", margin = margin(5, 5, 5, 5)),
      strip.background = element_rect(fill = "gray95", color = "black"),
      
      # Clean up gridlines for projection
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "gray90", linewidth = 0.3)
    )
}

get_all_guides_per_dataset <- function(dataset_name) {
  curr_path_to_data = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data",
    dataset_name,
    "sceptre/grna_matrix.rds"
  )
  grna_matrix = read_rds(curr_path_to_data)
  rownames(grna_matrix)
}

make_assn_data_structure <- function(dataset_names, methods) {
  # all assignments go here: 1st grouped by dataset, then by method, then by guide
  assns_list <- vector("list", length(dataset_names)) |>
    setNames(names(dataset_names))
  for(i in seq_along(dataset_names)) {
    assns_list[[i]] <- vector("list", length(methods)) |>
      setNames(methods)
    
    curr_guides = get_all_guides_per_dataset(dataset_name = dataset_names[i])
    guide_to_assns_template <- vector("list", length(curr_guides)) |>
      setNames(curr_guides)
    
    for(j in seq_along(methods)) {
      assns_list[[i]][[j]] <- guide_to_assns_template
    }
  }
  assns_list
}


get_run_with_dataset_and_method <- function(run_names, dataset_name, method_name, base_data_dir) {
  which_run_has_this_dataset_and_method <- sapply(run_names, function(run_name) {
    curr_run_dir = file.path(base_data_dir, run_name)
    curr_run_files = dir(curr_run_dir)
    
    file_found <- grepl(method_name, curr_run_files, fixed=TRUE) & 
      grepl(dataset_name, curr_run_files, fixed=TRUE)
    
    if(sum(file_found) > 1) {
      stop("multiple files matched for '", dataset_name, "' and ", method_name, " in '", run_name, "'\n")
    }
    any(file_found)
  })
  if(sum(which_run_has_this_dataset_and_method) > 1) {
    stop("multiple runs matched for '", dataset_name, "' and ", method_name, "\n")
  } else if(sum(which_run_has_this_dataset_and_method) == 0) {
    stop("no runs matched for '", dataset_name, "' and ", method_name, "\n")
  }
  run_names[which_run_has_this_dataset_and_method]
}


load_all_assns <- function(run_names, dataset_names, methods) {
  base_data_dir = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/outputs"
  )
  
  all_assns <- make_assn_data_structure(dataset_names = dataset_names, methods = methods)
  
  for(dataset_idx in seq_along(dataset_names)) {
    dataset_name = dataset_names[dataset_idx]
    dataset_name_short = names(dataset_names)[dataset_idx]
    
    path_to_cell_names = file.path(
      .get_config_path("LOCAL_BENCHMARKING_DIR"),
      "guide_assignment/input_data",
      dataset_name, "sceptre/grna_matrix.rds"
    )
    cell_names_in_order = read_rds(path_to_cell_names) |> colnames()
    
    for(method_name in methods) {
      
      run_with_this_dataset = get_run_with_dataset_and_method(
        run_names = run_names,
        dataset_name = dataset_name,
        method_name = method_name,
        base_data_dir=base_data_dir
      )
      curr_run_dir = file.path(
        base_data_dir, run_with_this_dataset
      )
      curr_files = dir(curr_run_dir)
      
      curr_assns_fname = curr_files[
        grepl(dataset_name, curr_files, fixed=TRUE) &
          grepl(method_name, curr_files, fixed=TRUE)
      ]
      
      curr_assns = load_assns(
        path_to_assns = file.path(curr_run_dir, curr_assns_fname),
        method_name = method_name,
        all_guides = names(all_assns[[dataset_name_short]][[method_name]]),
        cell_names_in_order = cell_names_in_order
      )
      if(!all(names(curr_assns) == names(all_assns[[dataset_name_short]][[method_name]]))) {
        stop("mismatch between the guide names returned when ", dataset_name, " and ", method_name, " is loaded.")
      }
      
      all_assns[[dataset_name_short]][[method_name]] = curr_assns
    }
  }
  all_assns
}


load_assns <- function(path_to_assns, method_name, all_guides, cell_names_in_order) {
  if(method_name == "sceptre") {
    load_assns_sceptre(path_to_assns = path_to_assns, all_guides = all_guides, cell_names_in_order = cell_names_in_order)
  } else if(method_name == "crispat") {
    load_assns_crispat(path_to_assns = path_to_assns, all_guides = all_guides)
  } else if(method_name == "pertpy") {
    load_assns_pertpy(path_to_assns = path_to_assns, all_guides = all_guides)
  } else if(method_name == "cleanser") {
    load_assns_cleanser(path_to_assns = path_to_assns, all_guides = all_guides, cell_names_in_order = cell_names_in_order)
  # }  else if(method_name == "scepnb") {
  #   load_assns_scepnb(path_to_assns = path_to_assns, all_guides = all_guides, cell_names_in_order = cell_names_in_order)
  }   else {
    stop("method '", method_name, "' not recognized.")
  }
}


# path_to_assns = "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/outputs/run_all_3guides_100k/assignment_matrix_sceptre_replogle-rd7_assign_nguides=100_ncells=100k.rds"

# THIS RETURNS NAMES
load_assns_sceptre <- function(path_to_assns, all_guides, cell_names_in_order) {
  scep_assn_mat = read_rds(path_to_assns)
  guide_to_assns = vector("list", length(all_guides)) |>
    setNames(all_guides)
  for(guide in all_guides) {
    guide_to_assns[[guide]] = cell_names_in_order[which(scep_assn_mat[guide,])]
  }
  guide_to_assns
}

# load_assns_scepnb <- function(path_to_assns, all_guides, cell_names_in_order) {
#   scep_assn_mat = read_rds(path_to_assns)
#   guide_to_assns = vector("list", length(all_guides)) |>
#     setNames(all_guides)
#   for(guide in all_guides) {
#     guide_to_assns[[guide]] = cell_names_in_order[which(scep_assn_mat[guide,] == 1)]
#   }
#   guide_to_assns
# }

# path_to_crispat = "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/outputs/run_all_3guides_100k/assignments_crispat_replogle-rd7_assign_nguides=100_ncells=100k.csv"
# all_guides = rownames(read_rds("/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7_assign_nguides=100_ncells=100k/sceptre/grna_matrix.rds"))
# cell_names_in_order  = colnames(read_rds("/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7_assign_nguides=100_ncells=100k/sceptre/grna_matrix.rds"))

# RETURNS NAMES
load_assns_crispat <- function(path_to_assns, all_guides) {
  crisp_assign_raw <- read.csv(path_to_assns) 
      # mutate(cell_idx = gsub("cell_idx_", "", cell_id) |> as.integer())
  unordered_cell_list <- split(crisp_assign_raw$cell_id, crisp_assign_raw$grna_id)
  
  guide_to_assns = vector("list", length(all_guides)) |>
    setNames(all_guides)
  
  guide_to_assns[names(unordered_cell_list)] <- unordered_cell_list
  guide_to_assns
}


load_assns_pertpy <- function(path_to_assns, all_guides) {

    pert_assign_raw <- read.csv(path_to_assns) |>
      filter(grna_id != "negative")
    
    guide_to_assns = vector("list", length(all_guides)) |>
      setNames(all_guides)
    
    dt <- as.data.table(pert_assign_raw)
    
    # 1) Split each cell's "g1+g2+g3" into multiple columns
    spl <- dt[, tstrsplit(grna_id, "\\+", fixed = FALSE)]
    
    # 2) Go long: one row per (cell_id, grna)
    long <- melt(
      cbind(dt[, .(cell_id)], spl),
      id.vars = "cell_id",
      value.name = "grna",
      na.rm = TRUE
    )[, .(cell_id, grna)]
    
    # (optional) drop duplicates if the same guide appears twice for a cell
    setkey(long, grna, cell_id)
    long <- unique(long)
    
    # 3) Aggregate: list of cells per gRNA
    grna_cells_list <- long[, .(cells = list(cell_id)), by = grna]
    grna_cells_list <- setNames(grna_cells_list$cells, grna_cells_list$grna)
    guide_to_assns[names(grna_cells_list)] <- grna_cells_list
    
    guide_to_assns
}
# 
# path_to_assns = "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/outputs/run_all_3guides_100k/assignments_cleanser_replogle-rd7_assign_nguides=100_ncells=100k.csv"
load_assns_cleanser <- function(path_to_assns, all_guides, cell_names_in_order, posterior_thresh = 0.8) {
  # print(all_guides)
  clns_assign_raw <- read_csv(path_to_assns, show_col_types = FALSE) |> as.data.frame()
  
  # all_guides_ <<- all_guides
  # cell_names_in_order_ <<- cell_names_in_order
  #   print(head(clns_assign_raw))
  #   print(sum(clns_assign_raw$posterior > .8))
 clns_assign_raw = clns_assign_raw |>
    filter(posterior >= posterior_thresh) |>
    mutate(grna_id = all_guides[grna_id], cell_name = cell_names_in_order[cell_id])

  guide_to_assns = vector("list", length(all_guides)) |>
    setNames(all_guides)
  
  unordered_cell_list <- split(clns_assign_raw$cell_name, clns_assign_raw$grna_id)
  
  guide_to_assns[names(unordered_cell_list)] <- unordered_cell_list
  guide_to_assns  
}


# 
# 
# 
# test_scep = load_assns_sceptre(
#   "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/outputs/run_all_3guides_100k/assignment_matrix_sceptre_replogle-rd7_assign_nguides=400_ncells=100k.rds",
#     all_guides_,
#   cell_names_
# )
# 
# test_crisp = load_assns_crispat(
#   "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/outputs/run_all_3guides_100k/assignments_crispat_replogle-rd7_assign_nguides=400_ncells=100k.csv",
#   all_guides_
# )
# 
# jac <- function(a,b) {
#   numerator = length(intersect(a,b))
#   denominator = length(union(a,b))
#   if(denominator == 0) {
#     return(NA)
#   }
#   numerator / denominator
# }
# 
# sims = sapply(all_guides_, function(guide) jac(test_scep[guide], test_crisp[guide]))
# hist(sims |> unlist())

```


```{r, echo=FALSE}
run_names <- c("run_all_3guides_100k", "run_all_nguides=800_100k", "rerun_pertpy_more_samples")

dataset_names_repl <- c(
  ng100 = "replogle-rd7_assign_nguides=100_ncells=100k",
  ng200 = "replogle-rd7_assign_nguides=200_ncells=100k",
  ng400 = "replogle-rd7_assign_nguides=400_ncells=100k",
  ng800 = "replogle-rd7_assign_nguides=800_ncells=100k"
)
dataset_names_gasp <- c(
  ng100 = "gasperini_assign_nguides=100_ncells=100k",
  ng200 = "gasperini_assign_nguides=200_ncells=100k",
  ng400 = "gasperini_assign_nguides=400_ncells=100k",
  ng800 = "gasperini_assign_nguides=800_ncells=100k"
)
methods = c("sceptre", "crispat", "cleanser", "pertpy")

method_renamer = c(
  sceptre="SCEPTRE",
  cleanser="CLEANSER",
  crispat="crispat",
  pertpy="pertpy"
)

dataset_renamer = c(
  ng100="100 guides",
  ng200="200 guides",
  ng400="400 guides",
    ng800="800 guides"
)

sparsity_repl = file.path(
      .get_config_path("LOCAL_BENCHMARKING_DIR"),
      "guide_assignment/input_data",
      "replogle-rd7",
      "sceptre-pipeline/grna_summary_stats.csv"
    ) |>
  read_csv(show_col_types = FALSE) |>
  mutate(sparsity_ranking = rank(grna_nonzero))
  

sparsity_gasp = file.path(
      .get_config_path("LOCAL_BENCHMARKING_DIR"),
      "guide_assignment/input_data",
      "gasperini",
      "sceptre-pipeline/grna_summary_stats.csv"
    ) |>
  read_csv(show_col_types = FALSE) |>
  mutate(sparsity_ranking = rank(grna_nonzero))

repl_assns <- load_all_assns(run_names=run_names, dataset_names=dataset_names_repl, methods=methods)

gasp_assns <- load_all_assns(run_names=run_names, dataset_names=dataset_names_gasp, methods=methods)




```



ok... now what


```{r, echo=FALSE}
precision <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  denom <- length(set1)
  if (denom == 0) return(NA_real_)
  return(intersection / denom)
}

recall <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  denom <- length(set2)
  if (denom == 0) return(NA_real_)
  return(intersection / denom)
}
F1 <- function(set1, set2) {
  inter <- length(intersect(set1, set2))
  denom <- length(set1) + length(set2)
  if (denom == 0) return(NA_real_)  # or return(1) if you want "both empty => perfect match"
  2 * inter / denom
}
jaccard <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  if (union == 0) return(NA_real_)
  return(intersection / union)
}


list_prettier <- function(assns_list, method_renamer = NULL, dataset_renamer = NULL) {
  if(!is.null(method_renamer)) {
    for(i in seq_along(assns_list)) {
      curr_names <- names(assns_list[[i]])
      names(assns_list[[i]]) <- method_renamer[curr_names]
    }
  }
  if(!is.null(dataset_renamer)) {
    names(assns_list) <- dataset_renamer[names(assns_list)]
  }
  assns_list
}

plot_method_comparisons <- function(
    assns_list, 
    metric_type = c("F1", "precision", "recall", "jaccard"),
    source_data,
        metric_title_name,
    metric_y_axis_name = NULL,
    method_renamer = NULL,
    dataset_renamer = NULL
    ) {
  metric_type <- match.arg(metric_type)
  
  # Map string to the metric functions you defined
  metric_fn <- switch(metric_type,
    "F1"        = F1,
    "precision" = precision,
    "recall"    = recall,
    "jaccard" = jaccard
  )
  
  assns_list <- list_prettier(
    assns_list,
    method_renamer=method_renamer,
    dataset_renamer=dataset_renamer
  )
  datasets <- names(assns_list)
  methods  <- names(assns_list[[1]])
  
  # Generate all unique pairwise combinations: choose(4, 2) = 6
  method_pairs <- combn(methods, 2, simplify = FALSE)
  
  # Flatten the nested list into a long-format data frame
  results <- map_df(datasets, function(ds) {
    map_df(method_pairs, function(pair) {
      m1 <- pair[1]
      m2 <- pair[2]
      
      # Number of gRNAs
      num_grnas <- length(assns_list[[ds]][[m1]])
      
      # Vectorized calculation over gRNAs
      scores <- map_dbl(seq_len(num_grnas), function(k) {
        metric_fn(assns_list[[ds]][[m1]][[k]], 
                  assns_list[[ds]][[m2]][[k]])
      })
      
      data.frame(
        dataset    = ds,
        comparison = paste(m1, "vs", m2),
        score      = scores
      )
    })
  })

  # Generate the Violin Plot
  ggplot(results, aes(x = dataset, y = score, fill = comparison)) +
    # The 'violin' shows the full density of the scores
    geom_violin(
      trim         = TRUE, 
      alpha        = 0.6, 
      scale        = "width",
      linewidth    = 0.2,
      position     = position_dodge(width = 0.8)
    ) +
    labs(
      title    = paste(source_data, "pairwise gRNA Assignment Agreement:", metric_type),
      subtitle = "Distribution across all gRNAs per dataset",
      y        = metric_y_axis_name,
      x        = "Dataset",
      fill     = "Method Comparison"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      legend.position    = "bottom",
      axis.text.x        = element_text(face = "bold")
    ) +
    # Adjust y-axis to standard metric bounds
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    guides(fill = guide_legend(nrow = 2))
}

plot_method_metrics <- function(
    assns_list, 
    metric_type = c("cardinality"), 
    source_data,
        metric_title_name,
    metric_y_axis_name = NULL,
    log_scale = FALSE,
    method_renamer = NULL,
    dataset_renamer = NULL
) {
  metric_type <- match.arg(metric_type)
  
  metric_fn <- switch(metric_type,
    "cardinality" = length
  )
  if(is.null(metric_y_axis_name)) {
    metric_y_axis_name = metric_type
  }
  
  assns_list <- list_prettier(
    assns_list,
    method_renamer=method_renamer,
    dataset_renamer=dataset_renamer
  )
  
  
  
  datasets <- names(assns_list)
  methods  <- names(assns_list[[1]])
  
  # Flatten list into a data frame
  results <- map_df(datasets, function(ds) {
    map_df(methods, function(m) {
      
      # Calculate metric for each gRNA
      scores <- map_dbl(assns_list[[ds]][[m]], metric_fn)
      
      data.frame(
        dataset = ds,
        method  = m,
        score   = scores
      )
    })
  })

  # Initialize Plot
  plt <- ggplot(results, aes(x = dataset, y = score, fill = method)) +
    geom_violin(
      trim      = TRUE, 
      alpha     = 0.7, 
      scale     = "width",
      linewidth = 0.4, 
      position  = position_dodge(width = 0.8)
    ) +
    labs(
      title    = paste(source_data, "Evaluation:", metric_title_name),
      subtitle = paste("Distribution of scores per gRNA", if(log_scale) "(Log1p Scale)" else ""),
      y        = if(log_scale) paste("log1p(", metric_y_axis_name, ")") else metric_y_axis_name,
      x        = "Dataset",
      fill     = "Method"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "bottom",
      axis.title         = element_text(face = "bold")
    ) +
    scale_fill_brewer(palette = "Set2")

  # Conditionally add log scale
  if (log_scale) {
# 1. Determine the range of your data to set sensible breaks
    max_val <- max(results$score, na.rm = TRUE)
    max_break <- floor(log10(max(1, max_val))) + 1
    # 2. Create a sequence of powers of 10 that covers your data
    # This covers 0, 1, 10, 100, 1000, etc.
    my_breaks <- c(0, 10^(0:max_break), (10^(2:max_break)) / 2) |> sort()
    
    plt <- plt + scale_y_continuous(
      trans  = "log1p",
      breaks = my_breaks,
      labels = scales::label_comma()
    )
  }

  return(plt)
}



# Custom Geom for Split Violins
GeomSplitViolin <- ggproto("GeomSplitViolin", GeomViolin, 
  draw_group = function(self, data, ..., draw_quantiles = NULL) {
    data <- transform(data, xminv = x - violinwidth * (x - xmin), xmaxv = x + violinwidth * (xmax - x))
    grp <- data[1, "group"]
    newdata <- plyr::arrange(transform(data, x = if (grp %% 2 == 1) xminv else xmaxv), if (grp %% 2 == 1) y else -y)
    newdata <- rbind(newdata[1, ], newdata, newdata[nrow(newdata), ], newdata[1, ])
    newdata[c(1, nrow(newdata) - 1, nrow(newdata)), "x"] <- round(newdata[1, "x"])
    if (length(draw_quantiles) > 0 & !scales::zero_range(range(data$y))) {
      stopifnot(all(draw_quantiles >= 0 & draw_quantiles <= 1))
      quantiles <- ggplot2:::create_quantile_segment_frame(data, draw_quantiles)
      aesthetics <- data[rep(1, nrow(quantiles)), setdiff(names(data), c("x", "y")), drop = FALSE]
      aesthetics$x <- quantiles$x
      aesthetics$y <- quantiles$y
      aesthetics$group <- quantiles$group
      aesthetics$xend <- quantiles$xend
      aesthetics$yend <- quantiles$yend
      dot_elements <- list(...)
      segments <- do.call(grid::segmentsGrob, c(list(quantiles$x, quantiles$y, quantiles$xend, quantiles$yend), dot_elements))
      grid::grobTree(GeomPolygon$draw_panel(newdata, ...), segments)
    } else {
      GeomPolygon$draw_panel(newdata, ...)
    }
  }
)

geom_split_violin <- function(mapping = NULL, data = NULL, stat = "ydensity", position = "identity", ..., 
                              draw_quantiles = NULL, trim = TRUE, scale = "area", na.rm = FALSE, 
                              show.legend = NA, inherit.aes = TRUE) {
  layer(data = data, mapping = mapping, stat = stat, geom = GeomSplitViolin, 
        position = position, show.legend = show.legend, inherit.aes = inherit.aes, 
        params = list(trim = trim, scale = scale, draw_quantiles = draw_quantiles, na.rm = na.rm, ...))
}
plot_method_metrics_by_sparsity <- function(
    dataset_list, 
    sparsity_df, 
    metric_type = c("cardinality"), 
    source_data_name, 
    metric_title_name,
    log_scale = FALSE,
    method_renamer = NULL
) {
  metric_type <- match.arg(metric_type)
  metric_fn <- switch(metric_type, "cardinality" = length)
  
  # 1. Prettify
  if (!is.null(method_renamer)) {
    curr_names <- names(dataset_list)
    names(dataset_list) <- ifelse(curr_names %in% names(method_renamer), 
                                  method_renamer[curr_names], 
                                  curr_names)
  }
  
  # 2. Process Data
  results <- map_df(names(dataset_list), function(m) {
    data.frame(
      method = m,
      grna   = names(dataset_list[[m]]),
      score  = map_dbl(dataset_list[[m]], metric_fn)
    )
  }) %>%
    left_join(sparsity_df, by = "grna") %>%
    filter(!is.na(sparsity_label)) %>%
    mutate(sparsity_label = factor(sparsity_label, levels = c("low", "high")))

  # 3. Plot
  plt <- ggplot(results, aes(x = method, y = score, fill = sparsity_label)) +
    geom_split_violin(trim = TRUE, alpha = 0.7,scale = "width", adjust = 1.5) + # Using the custom geom here
    labs(
      title    = paste(source_data_name, "Evaluation:", metric_title_name),
      subtitle = "Split violins: Low (Left) vs High (Right) Sparsity",
      y        = if(log_scale) paste("log1p(", metric_type, ")") else metric_type,
      x        = "Method",
      fill     = "Sparsity"
    ) +
    theme_minimal() +
    scale_fill_manual(values = c("low" = "#66c2a5", "high" = "#fc8d62"))

  # 4. Scale handling
  if (log_scale) {
    max_val <- max(results$score, na.rm = TRUE)
    brks <- c(0, 10^(0:floor(log10(max(1, max_val)))))
    plt <- plt + scale_y_continuous(trans = "log1p", breaks = brks, labels = scales::label_comma())
  }

  return(plt)
}
plot_method_comparisons_by_sparsity <- function(
    dataset_list,         # One dataset (the inner list of methods)
    sparsity_df,          # The dataframe with 'grna' and 'sparsity_label'
    metric_type = c("F1", "precision", "recall", "jaccard"),
    source_data_name,     # String for title
    metric_title_name,
    metric_y_axis_name = NULL,
    method_renamer = NULL
) {
  metric_type <- match.arg(metric_type)
  metric_fn <- switch(metric_type,
    "F1"        = F1,
    "precision" = precision,
    "recall"    = recall,
    "jaccard"   = jaccard
  )
  
  if(is.null(metric_y_axis_name)) metric_y_axis_name = metric_type

  # 1. Prettify method names
  if (!is.null(method_renamer)) {
    curr_names <- names(dataset_list)
    names(dataset_list) <- ifelse(curr_names %in% names(method_renamer), 
                                  method_renamer[curr_names], 
                                  curr_names)
  }
  
  methods <- names(dataset_list)
  method_pairs <- combn(methods, 2, simplify = FALSE)
  
  # 2. Generate Pairwise Scores and Join with Sparsity
  results <- map_df(method_pairs, function(pair) {
    m1 <- pair[1]
    m2 <- pair[2]
    
    # We use the names of the gRNAs from m1 to join with sparsity_df
    grna_names <- names(dataset_list[[m1]])
    
    scores <- map_dbl(seq_along(dataset_list[[m1]]), function(k) {
      metric_fn(dataset_list[[m1]][[k]], 
                dataset_list[[m2]][[k]])
    })
    
    data.frame(
      comparison = paste(m1, "\nvs\n", m2), # \n keeps the x-axis clean
      grna       = grna_names,
      score      = scores
    )
  }) %>%
    left_join(sparsity_df, by = "grna") %>%
    filter(!is.na(sparsity_label)) %>%
    mutate(sparsity_label = factor(sparsity_label, levels = c("low", "high")))
  
  results %>%
  group_by(sparsity_label) %>%
  summarise(
    mean = mean(score),
    sd = sd(score),
    zeros = sum(score == 0),
    ones = sum(score == 1)
  ) |> print()

  # 3. Create the Split Violin Plot
  plt <- ggplot(results, aes(x = comparison, y = score, fill = sparsity_label)) +
    geom_split_violin(trim = TRUE, alpha = 0.7, scale = "width", adjust = 1.5) +
    labs(
      title    = paste(source_data_name, "Pairwise Agreement:", metric_title_name),
      subtitle = "Split by guide sparsity: low or high",
      y        = metric_y_axis_name,
      x        = "Method Comparison",
      fill     = "Sparsity Group"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      legend.position    = "bottom",
      axis.text.x        = element_text(size = 9) # Smaller text for 6 pairs
    ) +
    scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
    scale_fill_manual(values = c("low" = "#66c2a5", "high" = "#fc8d62"))

  return(plt)
}
```


first, let's get summary statistics

use biggest
```{r}
dataset = "ng800"

# get_summaries = function(assns_list, dataset, return_all=FALSE) {
#   # avg num cells per guide
# 
#   
#   cell_counts <- assns_list[[dataset]] |>
#   # 1. Convert nested list to a long data frame
#   enframe(name = "method", value = "guides") |>
#   unnest_longer(guides, indices_to = "guide_id") |>
#   unnest_longer(guides, values_to = "cell_id") |>
#   
#   # 2. Group by method and cell to count guides per cell
#   group_by(method, cell_id) |>
#   summarise(num_guides = n(), .groups = "drop") |>
#   
#   # 3. Pivot wider to see all methods side-by-side
#   pivot_wider(names_from = method, values_from = num_guides, values_fill = 0)
#   
#   out = list(
#     cells_per_guide = lapply(assns_list[[dataset]], function(assns) {
#   sapply(assns, length) |> mean()
# }),
# guides_per_cell = dplyr::select(cell_counts, -cell_id) |> colMeans()
# )
#   if(return_all) {
#     out[["cell_counts"]] = cell_counts
#   }
#   out
# }

# get_summaries(repl_assns, "ng800")
# oo = get_summaries(gasp_assns, "ng800", T)
# oo$cell_counts |>
#   select(-cell_id) |>
#   colMeans()





# need all cells
# cell_names = file.path(
#   .get_config_path("LOCAL_BENCHMARKING_DIR"),
#   "guide_assignment/input_data",
#   dataset_names_repl[[dataset]],
#   "sceptre/grna_matrix.rds"
# ) |> read_rds() |>
#   colnames()




grna_mat_repl = file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data",
  dataset_names_repl[[dataset]],
  "sceptre/grna_matrix.rds"
) |> read_rds() 

grna_mat_gasp = file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data",
  dataset_names_gasp[[dataset]],
  "sceptre/grna_matrix.rds"
) |> read_rds() 


# grna_mat = grna_mat_gasp
# assns_list = gasp_assns[[dataset]]
grna_mat = grna_mat_repl
assns_list = repl_assns[[dataset]]

thresh <- 10 # Example threshold
tail_heaviness <- sapply(rownames(grna_mat), function(guide) {
  counts <- grna_mat[guide, ]
  tail_counts <- counts[counts >= thresh]
  if(length(tail_counts) == 0) return(0)
  mean(tail_counts)
})



# 2. Get Number of Assignments per Guide for each method
# This assumes your assns_list is: List of Methods -> List of Guides -> Vector of Cells
n_assns_per_method <- lapply(assns_list, function(method_assns) {
  sapply(method_assns, length)
})

# 3. Correlation Check
# Compare each method's assignment counts to the tail heaviness
cor_results <- sapply(n_assns_per_method, function(n_assn) {
  # Align by guide names
  common_guides <- intersect(names(n_assn), names(tail_heaviness))
  cor((n_assn[common_guides]), tail_heaviness[common_guides], method = "pearson")
})

print(cor_results)

plot_df <- do.call(rbind, lapply(names(n_assns_per_method), function(mname) {
  data.frame(
    guide = names(n_assns_per_method[[mname]]),
    n_assn = n_assns_per_method[[mname]],
    tail_heavy = tail_heaviness[names(n_assns_per_method[[mname]])],
    method = mname
  )
}))



plot_df |>
  mutate(
    method = factor(method_renamer[method], levels = method_renamer)
  ) |>
ggplot( aes(x = (tail_heavy), y = (n_assn), color = method)) +
  geom_point(alpha = 0.4) +
  # geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~method) +
  theme_bw() +
  scale_y_continuous(
    trans = "log1p", 
    breaks = c(0, 1, 10, 100, 1000)
  ) +
  labs(title = "Replogle: Sensitivity to Tail Heaviness",
       x = paste0("Mean of UMIs above ", thresh),
       y = "Num. assignments / guide") +
  theme_presentation()

ggsave("~/thesis-defense-figures/assign-tail-sensitivity.png", width = 10, height = 7, dpi = 300)


```


plot 1: metrics as a function of guide

```{r}
# 1. Function to calculate mean Jaccard between two methods
get_mean_jaccard <- function(m1_list, m2_list) {
  # Align guides present in both methods
  common_guides <- intersect(names(m1_list), names(m2_list))
  
  scores <- sapply(common_guides, function(g) {
    set_a <- m1_list[[g]]
    set_b <- m2_list[[g]]
    
    inter <- length(intersect(set_a, set_b))
    uni   <- length(union(set_a, set_b))
    
    # Logic: If both methods agree a guide has 0 cells, that's a perfect match (1)
    if (uni == 0) return(1) 
    return(inter / uni)
  })
  
  return(mean(scores))
}

# assns_list = repl_assns[["ng800"]]
assns_list = gasp_assns[["ng800"]]
# 2. Build the Matrix
method_names <- names(assns_list)
per_guide_matrix <- sapply(method_names, function(m1) {
  sapply(method_names, function(m2) {
    get_mean_jaccard(assns_list[[m1]], assns_list[[m2]])
  })
})

print(round(per_guide_matrix, 3))
```

```{r}
# library(corrplot)
# library(RColorBrewer)
# 
# # 1. Define the methods and initialize an asymmetric matrix
# methods <- c("SCEPTRE", "crispat", "CLEANSER", "pertpy")
# n <- length(methods)
# comb_mat <- matrix(NA, nrow = n, ncol = n)
# colnames(comb_mat) <- rownames(comb_mat) <- methods
# 
# # 2. Fill Replogle Data (Upper Triangle)
# # Values: (s,cr)=0.86, (s,cl)=0.80, (s,p)=0.79, (cr,cl)=0.91, (cr,p)=0.81, (cl,p)=0.76
# comb_mat[1, 2] <- 0.86; comb_mat[1, 3] <- 0.80; comb_mat[1, 4] <- 0.79
# comb_mat[2, 3] <- 0.91; comb_mat[2, 4] <- 0.81
# comb_mat[3, 4] <- 0.76
# 
# # 3. Fill Gasperini Data (Lower Triangle)
# # Values: (cr,s)=0.98, (cl,s)=0.94, (p,s)=0.39, (cl,cr)=0.94, (p,cr)=0.40, (p,cl)=0.38
# comb_mat[2, 1] <- 0.98; comb_mat[3, 1] <- 0.94; comb_mat[4, 1] <- 0.39
# comb_mat[3, 2] <- 0.94; comb_mat[4, 2] <- 0.40
# comb_mat[4, 3] <- 0.38
# 
# # 4. Define a warm color palette (Yellow -> Orange -> Red)
# warm_palette <- colorRampPalette(brewer.pal(9, "YlOrRd"))(200)
# 
# # 5. Plotting
# # We use two calls to corrplot to handle the labels for each triangle
# # Call 1: Upper Triangle (Replogle)
# corrplot(comb_mat, 
#          method = "color", 
#          type = "upper", 
#          col = warm_palette,
#          addCoef.col = "black",   # Add text labels
#          number.digits = 2,       # Round to 2 decimal places
#          diag = FALSE,            # Remove diagonal
#          tl.col = "black",        # Text label color
#          cl.pos = "n",            # Remove color legend (scale on side)
#          mar = c(0, 0, 2, 0),     # Adjust margins for title
#          title = "Average per-guide Jaccard: Replogle (Upper) vs Gasperini (Lower)")
# 
# # Call 2: Lower Triangle (Gasperini)
# corrplot(comb_mat, 
#          method = "color", 
#          type = "lower", 
#          col = warm_palette,
#          addCoef.col = "black", 
#          number.digits = 2, 
#          diag = FALSE, 
#          add = TRUE,              # Add to the existing plot
#          tl.pos = "n",            # Don't redraw axis labels
#          cl.pos = "n")            # Ensure no scale on side
```

```{r}
# 1. Update naming to your exact specifications
methods_ <- c("SCEPTRE", "crispat", "CLEANSER", "pertpy")
plot_data <- expand.grid(Row = methods_, Col = methods_)
plot_data$value <- NA

# Replogle (Upper)
plot_data$value[plot_data$Row == "SCEPTRE"  & plot_data$Col == "crispat"]  <- 0.86
plot_data$value[plot_data$Row == "SCEPTRE"  & plot_data$Col == "CLEANSER"] <- 0.80
plot_data$value[plot_data$Row == "SCEPTRE"  & plot_data$Col == "pertpy"]   <- 0.79
plot_data$value[plot_data$Row == "crispat"  & plot_data$Col == "CLEANSER"] <- 0.91
plot_data$value[plot_data$Row == "crispat"  & plot_data$Col == "pertpy"]   <- 0.81
plot_data$value[plot_data$Row == "CLEANSER" & plot_data$Col == "pertpy"]   <- 0.76

# Gasperini (Lower)
plot_data$value[plot_data$Row == "crispat"  & plot_data$Col == "SCEPTRE"]  <- 0.98
plot_data$value[plot_data$Row == "CLEANSER" & plot_data$Col == "SCEPTRE"]  <- 0.94
plot_data$value[plot_data$Row == "pertpy"   & plot_data$Col == "SCEPTRE"]  <- 0.39
plot_data$value[plot_data$Row == "CLEANSER" & plot_data$Col == "crispat"]  <- 0.94
plot_data$value[plot_data$Row == "pertpy"   & plot_data$Col == "crispat"]  <- 0.40
plot_data$value[plot_data$Row == "pertpy"   & plot_data$Col == "CLEANSER"] <- 0.38

# 2. Factorize - Reverse Row to put SCEPTRE at the top
plot_data$Row <- factor(plot_data$Row, levels = rev(methods_))
plot_data$Col <- factor(plot_data$Col, levels = methods_)

# 3. Final Plotting
ggplot(plot_data, aes(x = Col, y = Row, fill = value)) +
  geom_tile(color = "white", linewidth = 1.2) +
  # Custom Gradient: Beige (#F5F5DC) to ggplot-Red (#F8766D)
  scale_fill_gradient(low = "#F5F5DC", high = "#F8766D", 
                      na.value = "white", limits = c(0, 1)) +
  # High-contrast bold numbers
  geom_text(aes(label = ifelse(is.na(value), "", sprintf("%.2f", value))), 
            color = "black", size = 6.5, fontface = "bold") +
  coord_fixed() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    axis.text = element_text(size = 14, face = "bold", color = "black"),
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold", margin = margin(b=20))
  ) +
  # Place labels in the diagonal gaps
  # Row 4, Col 1 is SCEPTRE/SCEPTRE slot (Gasperini side)
  # Row 1, Col 4 is pertpy/pertpy slot (Replogle side)
  annotate("text", x = 1, y = 4, label = "Gasperini\n(Lower)", 
           size = 5.5, fontface = "bold.italic", color = "grey25", lineheight = 0.85) +
  annotate("text", x = 4, y = 1, label = "Replogle\n(Upper)", 
           size = 5.5, fontface = "bold.italic", color = "grey25", lineheight = 0.85) +
  labs(title = "Average per-guide Jaccard")

ggsave("~/thesis-defense-figures/assign-avg-jaccard.png", width = 10, height = 7, dpi = 300)

```

```{r, echo=FALSE}


# plot_method_comparisons(repl_assns, "F1")
# plot_method_comparisons(gasp_assns, "F1")
# plot_method_comparisons(repl_assns, "precision")
# plot_method_comparisons(gasp_assns, "precision")
# plot_method_comparisons(repl_assns, "recall")
# plot_method_comparisons(gasp_assns, "recall")
plot_method_comparisons(repl_assns["ng800"], "jaccard", metric_y_axis_name = "Jaccard similarity", source_data = "Replogle",
                    method_renamer = method_renamer, dataset_renamer = dataset_renamer)
plot_method_comparisons(gasp_assns["ng800"], "jaccard", metric_y_axis_name = "Jaccard similarity", source_data = "Gasperini",
                    method_renamer = method_renamer, dataset_renamer = dataset_renamer)
```


```{r, echo=FALSE}
plot_method_metrics(repl_assns, "cardinality", log_scale = TRUE,
                    source_data = "Replogle", metric_title_name = "num. assignments per guide",
                      metric_y_axis_name = "count",
                    method_renamer = method_renamer, dataset_renamer = dataset_renamer)
plot_method_metrics(gasp_assns[3], "cardinality", log_scale = FALSE, source_data = "Gasperini",
                    metric_title_name = "num. assignments per guide", metric_y_axis_name = "count",
                    method_renamer = method_renamer, dataset_renamer = dataset_renamer)
```

```{r, echo=FALSE}
sparsity_df_repl <- sparsity_repl |>
  filter(grna %in% names(repl_assns[["ng400"]][["sceptre"]])) |>
  mutate(
    sparsity_ranking = rank(grna_nonzero),
    bottom_third_nonzero = grna_nonzero <= quantile(grna_nonzero, .33),
    top_third_nonzero = grna_nonzero > quantile(grna_nonzero, .67)
  )
sparsity_label = rep(NA, nrow(sparsity_df_repl))
sparsity_label[sparsity_df_repl$bottom_third_nonzero] = "high"
sparsity_label[sparsity_df_repl$top_third_nonzero] = "low"
sparsity_label = factor(sparsity_label, levels = c("low", "high"))
sparsity_df_repl = sparsity_df_repl |>
  mutate(sparsity_label = sparsity_label)
rm(sparsity_label)

plot_method_metrics_by_sparsity(
  dataset_list = repl_assns[["ng400"]],     
    sparsity_df = sparsity_df_repl,      
    metric_type = c("cardinality"), 
    source_data_name = "Replogle",     # String for title
    metric_title_name = "num. assignments per guide",
                      # metric_y_axis_name = "count",
                    method_renamer = method_renamer,
  log_scale = TRUE
)

plot_method_comparisons_by_sparsity(
  dataset_list = repl_assns[["ng400"]],
  sparsity_df = sparsity_df_repl,
  metric_type = "jaccard",
  metric_y_axis_name = "Jaccard similarity", source_data = "Replogle",
                    method_renamer = method_renamer,
  metric_title_name = "Jaccard"
)

# 
# # testing this
# df_name = "ng400"
# curr_df = sparsity_df_repl |>
#   filter(grna %in% names(repl_assns[[df_name]][["sceptre"]])) |>
#   mutate(scep_v_crisp = 0)
# for(i in 1:nrow(curr_df)) {
#   curr_df$scep_v_crisp[i] = jaccard(
#     repl_assns[[df_name]]$sceptre[[curr_df$grna[i]]],
#     repl_assns[[df_name]]$crispat[[curr_df$grna[i]]]
#   )
# }
# hist(curr_df$scep_v_crisp[as.character(curr_df$sparsity_label) == "low"])
# plot(curr_df$scep_v_crisp ~ log1p(curr_df$grna_nonzero))
```


```{r, echo=FALSE}
sparsity_df_gasp <- sparsity_gasp |>
  filter(grna %in% names(gasp_assns[["ng400"]][["sceptre"]])) |>
  mutate(
    sparsity_ranking = rank(grna_nonzero),
    bottom_third_nonzero = grna_nonzero <= quantile(grna_nonzero, .33),
    top_third_nonzero = grna_nonzero > quantile(grna_nonzero, .67)
  )
sparsity_label = rep(NA, nrow(sparsity_df_gasp))
sparsity_label[sparsity_df_gasp$bottom_third_nonzero] = "high"
sparsity_label[sparsity_df_gasp$top_third_nonzero] = "low"
sparsity_label = factor(sparsity_label, levels = c("low", "high"))
sparsity_df_gasp = sparsity_df_gasp |>
  mutate(sparsity_label = sparsity_label)
rm(sparsity_label)

plot_method_metrics_by_sparsity(
  dataset_list = gasp_assns[["ng400"]],     
    sparsity_df = sparsity_df_gasp,      
    metric_type = c("cardinality"), 
    source_data_name = "Gasperini",     # String for title
    metric_title_name = "num. assignments per guide",
                      # metric_y_axis_name = "count",
                    method_renamer = method_renamer,
  log_scale = FALSE
)

plot_method_comparisons_by_sparsity(
  dataset_list = gasp_assns[["ng400"]],
  sparsity_df = sparsity_df_gasp,
  metric_type = "jaccard",
  metric_y_axis_name = "Jaccard similarity", source_data = "Gasperini",
                    method_renamer = method_renamer,
  metric_title_name = "Jaccard"
)

```



does var correlate? can i find my simu findings here?






ok let's do compute 

```{r, echo=FALSE}

load_trace_files_in_memory_assignment <- function(run_names, path_to_outputs) {
  lapply(run_names, function(run_name) {
    read_tsv(
      file.path(path_to_outputs, run_name, "trace.tsv"),
      show_col_types = FALSE
    )
  }) |>
    do.call(what = rbind)  |>
    tidyr::extract(
      tag,
      into = c("nguides"),
      regex = "nguides=([0-9]+)",
      remove = FALSE,
      convert = TRUE
    ) |>
    mutate(
      source_data = strsplit(tag, "_") %>%
        sapply(`[`, 1) |>
        gsub(pattern="-rd7", replacement="", fixed=TRUE)
    ) |>
    transmute(
      method = gsub("_assign", "", process, ignore.case = TRUE) |> tolower(),
      peak_rss, realtime,
      num_cells =  sub(".*ncells=([^_]+).*", "\\1", tag), nguides, source_data
    ) |>
    mutate(
      peak_rss_gb = case_when(
        str_detect(peak_rss, regex("\\bGB\\b", ignore_case = TRUE)) ~ parse_number(peak_rss),
        str_detect(peak_rss, regex("\\bMB\\b", ignore_case = TRUE)) ~ parse_number(peak_rss) / 1024,
        TRUE ~ NA_real_
      )
    ) |>
    mutate(
      h = parse_number(str_extract(realtime, "\\d+(?=h)")),
      m = parse_number(str_extract(realtime, "\\d+(?=m)")),
      s = parse_number(str_extract(realtime, "\\d+(?=s)")),
      h = coalesce(h, 0),
      m = coalesce(m, 0),
      s = coalesce(s, 0),
      realtime_sec = 3600*h + 60*m + s,
      realtime_min = realtime_sec / 60
    ) %>%
    select(-h, -m, -s) 
}
path_to_outputs = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/outputs"
  )
trace_inmem = load_trace_files_in_memory_assignment(run_names, path_to_outputs =path_to_outputs )


assoc_datasets <- list(
  replogle = "replogle-rd7_comp_ngenes=1000_ntargets=100_gene_thresh=7",
  gasperini = "gasperini_comp_ngenes=1000_ntargets=100_gene_thresh=7"
)
load_trace_files_pipeline_assignment <- function(assoc_datasets) {
  
  parse_mem_to_gb <- function(x) {
    x <- str_trim(x)
    
    value <- as.numeric(str_extract(x, "^[0-9.]+"))
    unit  <- str_extract(x, "[A-Za-z]+$") |> toupper()
    
    mult <- case_when(
      unit == "B"  ~ 1 / 1024^3,
      unit == "KB" ~ 1 / 1024^2,
      unit == "MB" ~ 1 / 1024,
      unit == "GB" ~ 1,
      unit == "TB" ~ 1024,
      TRUE ~ NA_real_
    )
    
    # hacky wrapper in case any unitless 0's got in there, which become NA
    ifelse(is.na(value * mult), 0, value * mult)
  }
  
  
  
  lapply(seq_along(assoc_datasets), function(i) {
    curr_trace = path_to_an_assoc_run = file.path(
      .get_config_path("LOCAL_BENCHMARKING_DIR"),
      "association/computational/outputs/sceptre-pipeline",
      assoc_datasets[i],
      "tracing/trace.tsv"
    ) |>
      read_tsv(show_col_types = FALSE) |>
      mutate(method = "pipeline", source_data = names(assoc_datasets)[i],
             start = as.POSIXct(start, tz = "America/New_York"),
          complete = as.POSIXct(complete, tz = "America/New_York"),
          submit = as.POSIXct(submit, tz = "America/New_York"),
          
          peak_rss_in_GB = parse_mem_to_gb(peak_rss)) |>
      filter(process == "assign_grnas")
    
    start_time <- min(curr_trace$submit)
    
    corrected_endtimes = curr_trace$submit + (curr_trace$complete - curr_trace$start)
    end_time = max(corrected_endtimes)
      
      # get max memory for any worker
      peak_rss_in_gb_for_a_worker = curr_trace |>
        pull(peak_rss_in_GB) |>
        max()
      
      data.frame(
        source_data = names(assoc_datasets)[i],
        method="pipeline",
        assign_elapsed_time_in_sec = as.numeric(end_time - start_time, units = "secs"),
        max_peak_rss_in_gb_for_assign_workers = peak_rss_in_gb_for_a_worker,
        num_assign_analysis_workers = nrow(curr_trace)
      )
    
  }) |>
    do.call(what = rbind)
    
}
trace_pipe = load_trace_files_pipeline_assignment(assoc_datasets) |>
  left_join(
    data.frame(
      source_data = c("replogle", "gasperini"),
      nguides = c(2666,13078),
      stringsAsFactors = FALSE
    ),
    by="source_data"
  )


trace_pipe
trace_inmem

trace_all = rbind(
  trace_pipe |>
    dplyr::select(method, peak_rss_gb = max_peak_rss_in_gb_for_assign_workers,
                  realtime_sec = assign_elapsed_time_in_sec,
                  nguides, source_data),
  trace_inmem |>
    dplyr::select(method, peak_rss_gb ,
                  realtime_sec,
                  nguides, source_data)
)

```


```{r}
plot_runtime <- function(df, title_text, source_tag) {
  
  # 1. Master Palette definition (Internal or Global)
  method_cols_ <- c(
    "SCEPTRE"  = "#F8766D",
    "crispat"  = "#7CAE00",
    "CLEANSER" = "#00BFC4",
    "pertpy"   = "#C77CFF",
    "pipeline" = "#FF0000"
  )
  
  # 2. Helper for H:M:S formatting
  fmt_hms <- function(x) {
    x <- round(x)
    sprintf("%d:%02d:%02d", x %/% 3600, (x %% 3600) %/% 60, x %% 60)
  }
  
  # 3. Dynamic Time Breaks
  all_breaks <- c(30, 60, 120, 300, 600, 1800, 3600, 7200, 14400, 28800)
  current_max <- max(df$realtime_sec, na.rm = TRUE)
  
  # 4. The Plot Pipeline
  df |>
    dplyr::filter(source_data == !!source_tag) |>
    ggplot(aes(x = nguides, y = realtime_sec, 
               color = method, group = method)) +
    geom_line(linewidth = 0.8, alpha = 0.7) +
    geom_point(size = 4) +
    # Scale logic
    scale_y_log10(breaks = all_breaks[all_breaks <= current_max * 1.2], 
                  labels = fmt_hms) +
    scale_x_log10() +
    # Manual Scales (Built-in)
    scale_color_manual(values = method_cols_) +
    # scale_shape_manual(values = c("TRUE" = 17, "FALSE" = 1), guide = "none") +
    # Formatting
    labs(
      x = "Number of guides (log scale)",
      y = "Runtime (H:M:S, log scale)",
      color = "Method",
      title = title_text
    ) +
    theme_bw() +
    theme_presentation() + # Assumes this is defined in your environment
    theme(
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}
method_renamer_comp <- method_renamer
method_renamer_comp["pipeline"] <- "pipeline"
trace_all |>
  mutate(
    method = factor(method_renamer_comp[method], levels = method_renamer_comp),
    scep = method %in% c("SCEPTRE", "pipeline")
  ) |>
plot_runtime(title = "Assignment: runtime for Replogle", source_tag = "replogle")
ggsave("~/thesis-defense-figures/repl-assign-runtime.png", width = 10, height = 7, dpi = 300)


trace_all |>
  mutate(
    method = factor(method_renamer_comp[method], levels = method_renamer_comp),
    scep = method %in% c("SCEPTRE", "pipeline")
  ) |>
plot_runtime(title = "Assignment: runtime for Gasperini", source_tag = "gasperini")
ggsave("~/thesis-defense-figures/gasp-assign-runtime.png", width = 10, height = 7, dpi = 300)

```


```{r, echo=FALSE}
plot_memory <- function(df, title_text, source_tag) {
  
  # 1. Master Palette (Ensuring consistency across all slide types)
  method_cols_ <- c(
    "SCEPTRE"  = "#F8766D",
    "crispat"  = "#7CAE00",
    "CLEANSER" = "#00BFC4",
    "pertpy"   = "#C77CFF",
    "pipeline" = "#FF0000" # Logical Orange
  )
  
  # 2. Filter and Plot
  df |>
    dplyr::filter(source_data == !!source_tag) |>
    ggplot(aes(x = nguides, y = peak_rss_gb, 
               color = method, group = method)) +
    geom_line(linewidth = 0.8, alpha = 0.7) +
    geom_point(size = 4) +
    
    # 3. Log Scales
    scale_y_log10(
      breaks = c(0.5, 1, 2, 4, 8, 16, 32, 64, 128),
      labels = scales::label_number(suffix = " GB") # Adds "GB" directly to axis
    ) +
    scale_x_log10() +
    
    # 4. Manual Scales (Synced with Runtime plot)
    scale_color_manual(values = method_cols_) +
    # scale_shape_manual(values = c("TRUE" = 17, "FALSE" = 1), guide = "none") +
    
    # 5. Formatting
    labs(
      x = "Number of guides (log scale)",
      y = "Peak memory (log scale)",
      color = "Method",
      title = title_text
    ) +
    theme_bw() +
    theme_presentation() + 
    theme(
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
}
trace_all |>
  mutate(
    method = factor(method_renamer_comp[method], levels = method_renamer_comp),
    scep = method %in% c("SCEPTRE", "pipeline")
  ) |>
plot_memory(title = "Assignment: memory for Replogle", source_tag = "replogle")
ggsave("~/thesis-defense-figures/repl-assign-mem.png", width = 10, height = 7, dpi = 300)

trace_all |>
  mutate(
    method = factor(method_renamer_comp[method], levels = method_renamer_comp),
    scep = method %in% c("SCEPTRE", "pipeline")
  ) |>
  plot_memory( title = "Assignment: memory for Gasperini", source_tag = "gasperini")
ggsave("~/thesis-defense-figures/gasp-assign-mem.png", width = 10, height = 7, dpi = 300)

```



can i make an extra nextflow worker usage plot?

```{r}

# 1. Load the data
# Nextflow trace files use '\t' and usually don't have quotes
curr_trace_raw <- read_delim("~/data/projects/sceptre3/benchmarking/association/computational/outputs/sceptre-pipeline/replogle-rd7_comp_ngenes=10000_ntargets=1600_gene_thresh=7/tracing/trace.tsv", delim = "\t")

# 2. Filter and Prepare the Timeline
# We need to know when each process started and when it finished
concurrency_data <- curr_trace_raw %>%
  filter(grepl("run_association_analysis", process)) %>%
  # Convert character timestamps to actual Date-Time objects
  mutate(
    start_time = as.POSIXct(start, format="%Y-%m-%d %H:%M:%OS"),
    end_time   = as.POSIXct(complete, format="%Y-%m-%d %H:%M:%OS")
  ) %>%
  # Create a "long" format: one row for 'start' (+1) and one for 'end' (-1)
  select(start_time, end_time) %>%
  pivot_longer(cols = everything(), names_to = "event_type", values_to = "time") %>%
  mutate(change = ifelse(event_type == "start_time", 1, -1)) %>%
  # Sort by time to calculate the running sum
  arrange(time) %>%
  mutate(active_processes = cumsum(change))

# 3. Plotting
ggplot(concurrency_data, aes(x = time, y = active_processes)) +
  geom_step(color = "#00BFC4", linewidth = 1) +  # geom_step is best for discrete changes
  # geom_area(fill = "#00BFC4", alpha = 0.2) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Worker Usage Over Time",
    subtitle = "Process: run_association_analysis",
    x = "Time (Wall Clock)",
    y = "Number of Concurrent Workers"
  ) +
  scale_x_datetime(date_labels = "%H:%M", date_breaks = "1 hour") +
  theme_presentation(base_size = 25)

ggsave("~/thesis-defense-figures/worker-usage.png", width = 10, height = 7, dpi = 300)

```




extra figures and whatnot for my defense presentation

```{r}
rm(list=ls())
source("~/.Rprofile")
library(tidyverse)

data_sources <- c("replogle-rd7", "gasperini")

umis_list <- list()

for(i in seq_along(data_sources)) {
  data_source = data_sources[i]
  path_to_data <- file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data", data_source,
    "sceptre-pipeline"
  )
  
  grna_odm <- ondisc::initialize_odm_from_backing_file(
    file.path(path_to_data, "grna.odm")
  )
  
  grna_summary_stats = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data",
    data_source,
    "sceptre-pipeline/grna_summary_stats.csv"
  ) |>
    read_csv(show_col_types = FALSE) |>
    mutate(sparsity_ranking = rank(grna_nonzero, ties="first") ) |>
    arrange(sparsity_ranking) %>%
    filter(sparsity_ranking %in% c(50, nrow(.) - 50))

    
  
  umis_list[[i]] = data.frame(
    sparse = grna_odm[grna_summary_stats$grna[grna_summary_stats$sparsity_ranking == 50],],
    dense = grna_odm[grna_summary_stats$grna[grna_summary_stats$sparsity_ranking >50], ],
    data_source = data_source
  )
}

umis <- do.call(rbind, umis_list) |>
  pivot_longer(cols = !data_source, names_to = "Expression", values_to = "UMI count") |>
  mutate(Expression = ifelse(Expression == "dense", "high", "low"),
         data_source = ifelse(data_source == "replogle-rd7", "Replogle", "Gasperini"))

plot_guides <- function(umis, title) {
  max_val <- max(umis[["UMI count"]], na.rm = TRUE)
  max_break <- floor(log10(max(1, max_val))) + 1

  my_breaks <- c(0, 10^(1:max_break))#, (10^(2:max_break)) / 2) |> sort()
  
  umis |>
    ggplot(aes(x = `UMI count`, fill = data_source)) +
    geom_histogram(bins=50) +
    # facet_wrap(data_source ~ Expression, labeller = label_bquote(cols = "Data: " * .(data_source)))+
    facet_wrap(~data_source + Expression, 
             labeller = labeller(
               Expression = label_both,
               data_source = label_value
             ), scales="free_x") +
    theme_bw() +
    scale_y_continuous(
      trans  = "log1p",
      breaks = my_breaks,
      labels = scales::label_comma()
    ) +
        scale_x_continuous(
      trans  = "log1p",
      breaks = my_breaks,
      labels = scales::label_comma()
    ) +
    labs(title=title)
}
plot_guides(umis, title="Example guide RNA distributions") +
  theme_presentation() +
  theme(legend.position = "none") 
ggsave("~/thesis-defense-figures/guide-dist.png", width = 10, height = 7, dpi = 300)

```

```{r}
rm(list=ls())
source("~/.Rprofile")
library(tidyverse)

data_sources <- c("replogle-rd7", "gasperini")

cell_umis_list <- list()
# cell_metadata <- list()
for(i in seq_along(data_sources)) {
  data_source = data_sources[i]
  path_to_data <- file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data", data_source,
    "sceptre-pipeline"
  )
  
  scep <- sceptre::read_ondisc_backed_sceptre_object(
    sceptre_object_fp = file.path(path_to_data, "sceptre_object.rds"),
    response_odm_file_fp = file.path(path_to_data, "response.odm"),
    grna_odm_file_fp = file.path(path_to_data, "grna.odm")
  )
  
  cell_stats <- data.frame(
    grna_n_nonzero = scep@covariate_data_frame$grna_n_nonzero,
        grna_n_umis = scep@covariate_data_frame$grna_n_umis,
    cell_idx = 1:nrow(scep@covariate_data_frame)
  ) |>
    arrange(grna_n_nonzero)
  
  idx_to_keep <- c(cell_stats$cell_idx[10000], tail(cell_stats$cell_idx, 1000)[1])
  # scep@covariate_data_frame[idx_to_keep,]
  
  grna_odm <- ondisc::initialize_odm_from_backing_file(
    file.path(path_to_data, "grna.odm")
  )
  
  low_ <- high_ <- integer(nrow(grna_odm))
  for(j in 1:nrow(grna_odm)) {
    curr_ <- grna_odm[j,]
    low_[j] = curr_[idx_to_keep[1]]
    high_[j] = curr_[idx_to_keep[2]]
  }
  cell_umis_list[[i]] <- data.frame(low = low_, high=high_, low_idx = idx_to_keep[1], high_idx = idx_to_keep[2], data_source=  data_source)
}




cell_umis <- do.call(rbind, cell_umis_list) |>
  dplyr::select(-low_idx, -high_idx) |>
  pivot_longer(cols = !data_source, names_to = "Expression", values_to = "UMI count") |>
  mutate(data_source = ifelse(data_source == "replogle-rd7", "Replogle", "Gasperini"))

plot_cells <- function(umis, title) {
  max_val <- max(umis[["UMI count"]], na.rm = TRUE)
  max_break <- floor(log10(max(1, max_val))) + 1

  my_breaks_x <- c(0, 5, 10^(0:max_break)) |> sort()#, (10^(2:max_break)) / 2) |> sort()
  my_breaks_y <- c(0,  10^(1:max_break)) |> sort()
  umis |>
    ggplot(aes(x = `UMI count`, fill = data_source)) +
    geom_histogram(bins=50) +
    # facet_wrap(data_source ~ Expression, labeller = label_bquote(cols = "Data: " * .(data_source)))+
    facet_wrap(~data_source + Expression, 
             labeller = labeller(
               Expression = label_both,
               data_source = label_value
             ), scales="free_x") +
    theme_bw() +
    scale_y_continuous(
      trans  = "log1p",
      breaks = my_breaks_y,
      labels = scales::label_comma()
    ) +
        scale_x_continuous(
      trans  = "log1p",
      breaks = my_breaks_x,
      labels = scales::label_comma()
    ) +
    labs(title=title)
}
plot_cells(cell_umis, title="gRNA UMI counts for 4 example cells") +
   theme_presentation() +
  theme(legend.position = "none") 
ggsave("~/thesis-defense-figures/cell-dist.png", width = 10, height = 7, dpi = 300)

```






## SIMULATIONS

```{r}

# i copied the pertpy from this one "rerun_pertpy_sims_4mu_50k_cleanser_params" to the one i'm using
run_names_sims <- c("run_sims_4mu_50k_cleanser_params")


# i manually renamed from   sims = "replogle-rd7_sims_from_cleanser_50k_4_mu"
# because having a method name in the dataset name broke my code
dataset_names_sims <- c(
  sims = "replogle-rd7_sims_from_clns_50k_4_mu"
)

methods = c("sceptre", "crispat", "cleanser", "pertpy")

method_renamer = c(
  sceptre="SCEPTRE",
    crispat="crispat",
  cleanser="CLEANSER",
  pertpy="pertpy"
)
stopifnot((methods) == names(method_renamer))  # these need to agree for factor making!

# dataset_renamer = c(
#   ng100="100 guides",
#   ng200="200 guides",
#   ng400="400 guides",
#     ng800="800 guides"
# )

# gotta process this one a bit more
sims_assns <- load_all_assns(run_names=run_names_sims, dataset_names=dataset_names_sims, methods=methods)



```


```{r}
# plot_method_comparisons(sims_assns, "jaccard", metric_y_axis_name = "Jaccard similarity", source_data = "sims",
#                     method_renamer = method_renamer)
```


ok let's get ground truth comparison
```{r}

true_perts = read_rds(
  "~/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7_sims_from_clns_50k_4_mu/true_pert_matrix.rds"
)
sim_data = read_rds(
  "~/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7_sims_from_clns_50k_4_mu/sceptre/grna_matrix.rds"
)


make_sim_df <- function(sims_assns, methods, true_perts, sim_data, mupow_seq = 0:3, nguides_per = 100, metric_func) {
  
  results <- list()
  for(mupow in mupow_seq) {
    
    guide_id = 1:nguides_per
    guides_to_consider = paste0("params_cleanser_mupow=", mupow, "_grna_", guide_id)
    
    curr_dataset_metrics = vector("list", length(methods)) |> setNames(methods)
    
    for(method_ in methods) {
      curr_dataset_metrics[[method_]] = vector("list", length(guides_to_consider)) |> setNames(guides_to_consider)
      
      for(guide_ in guides_to_consider) {
        
        true_pos = true_perts[guide_,] == 1
        pos_umi = sim_data[guide_,] > 0
        
        metric_val  = metric_func(
          sims_assns$sims[[method_]][[guide_]], 
          colnames(true_perts)[true_pos & pos_umi]
        )
        
        curr_dataset_metrics[[method_]][[guide_]] = metric_val # ifelse(is.na(jacc), 0, jacc)
      }
    }
    curr_results = data.frame(
      guide = guides_to_consider,
      mupow = mupow,
      guide_id
    )
    
    for(method_ in methods) {
      curr_results[[paste0(method_)]] = curr_dataset_metrics[[method_]][curr_results$guide] |> unlist()
    }
    results <- c(results, list(curr_results))
  }
  do.call(rbind, results)
}





sim_results_jacc = make_sim_df(sims_assns = sims_assns, methods = methods, true_perts = true_perts, sim_data = sim_data, mupow_seq = 0:3, nguides_per = 100, metric_func = jaccard)

sim_results_jacc |>
  dplyr::select(-guide) |>
  pivot_longer(cols=!mupow & !guide_id, names_to="method", values_to="jaccard") |>
  ggplot(aes(x = jaccard, fill = method)) +
  geom_histogram(bins=50) +
  facet_wrap(~mupow) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 1))

sim_results_prec = make_sim_df(sims_assns = sims_assns, methods = methods, true_perts = true_perts, sim_data = sim_data, mupow_seq = 0:3, nguides_per = 100, metric_func = precision)

sim_results_prec |>
  dplyr::select(-guide) |>
  pivot_longer(cols=!mupow & !guide_id, names_to="method", values_to="precision") |>
  ggplot(aes(x = precision, fill = method)) +
  geom_histogram(bins=50) +
  facet_wrap(~mupow) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 1))


sim_results_recall = make_sim_df(sims_assns = sims_assns, methods = methods, true_perts = true_perts, sim_data = sim_data, mupow_seq = 0:3, nguides_per = 100, metric_func = recall)

sim_results_recall |>
  dplyr::select(-guide) |>
  pivot_longer(cols=!mupow & !guide_id, names_to="method", values_to="recall") |>
  ggplot(aes(x = recall, fill = method)) +
  geom_histogram(bins=50) +
  facet_wrap(~mupow) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 1))


```




let's get a nice front-page visualization
INCLUDED IN DEFENSE
```{r}
combine_all_sim_results <- function(sim_results_jacc, sim_results_prec, sim_results_recall) {
  avg_metric_df <- sim_results_jacc |>
  group_by(mupow) |>
  summarize(sceptre = mean(sceptre), pertpy = mean(pertpy),
            cleanser = mean(cleanser), crispat = mean(crispat)) |>
  pivot_longer(!mupow, values_to="jacc", names_to  ="method") |>
  left_join(
    sim_results_prec |>
  group_by(mupow) |>
  summarize(sceptre = mean(sceptre), pertpy = mean(pertpy),
            cleanser = mean(cleanser, na.rm=TRUE), crispat = mean(crispat)) |>
  pivot_longer(!mupow, values_to="prec", names_to  ="method"),
  by = c("mupow", "method")
    
  ) |>
  left_join(
    sim_results_recall |>
  group_by(mupow) |>
  summarize(sceptre = mean(sceptre), pertpy = mean(pertpy),
            cleanser = mean(cleanser, na.rm=TRUE), crispat = mean(crispat)) |>
  pivot_longer(!mupow, values_to="recall", names_to  ="method"),
  by = c("mupow", "method")
  )
  
  avg_metric_df
}


avg_metric_df <- combine_all_sim_results(sim_results_jacc, sim_results_prec, sim_results_recall)





mu_breaks = c(100,200,400,800,1600)
metric_renamer <- c(jacc="Jaccard", prec="Precision", recall="Recall")
avg_metric_df |>
  pivot_longer(!mupow & !method, names_to = "metric", values_to = "value") |>
  mutate(
    method = factor(method_renamer[method], levels = method_renamer),
    metric = metric_renamer[metric],
    mu_pert = 971 / (2^mupow)
  ) |>
  ggplot(aes(x = mu_pert, y = value, group = method, color = method)) +
  geom_point() +
  geom_line() +
  facet_wrap(~metric) +
  # theme_bw() +
  scale_x_log10(
    breaks = mu_breaks
  ) +
  # theme(legend.position = "bottom") +
  labs(
    x = "Average NegBin mean for perturbed cells",
    y = "Metric value",
    title = "Simulation metrics summary"
  ) +
  theme_presentation()

ggsave("~/thesis-defense-figures/assign-sims-avg-metrics.png", width = 10, height = 7, dpi = 300)


sim_results_jacc |>
  dplyr::select(-guide) |>
  pivot_longer(cols=!mupow & !guide_id, names_to="method", values_to="jaccard") |>
  filter(mupow %in% c(0,3)) |>
  mutate(
    method = factor(method_renamer[method], levels = method_renamer),
    mu_pert = round(971 / (2^mupow) / 10)*10
  ) |>
  # filter(method != "Crispat") |>
  ggplot(aes(x = jaccard, fill = method)) +
  geom_histogram(bins=50) +
  # geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_wrap(~mu_pert, labeller = label_bquote(cols = "Perturbed mean" == .(mu_pert))) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 1)) +
  labs(
    x = "Jaccard similarity to ground truth",
    title = "Individual simulation results"
  ) +
  theme_presentation()
ggsave("~/thesis-defense-figures/assign-sims-hists.png", width = 10, height = 7, dpi = 300)

```


investigations to back up my key points in "~/thesis-defense-figures/assign-sims-avg-metrics.png"

1. pertpy's poor precision for small $\mu_1$
- is it making too many predictions? yes
- if that happens, it's because gaussian absorbs too many

2. sceptre's poor precision for large $\mu_1$

3. cleanser's failures. does it not assign anything? or just all wrong?
```{r}
sims_assns$sims$pertpy[paste0("params_cleanser_mupow=3_grna_",1:100)] |> sapply(length) |> hist()
sims_assns$sims$crispat[paste0("params_cleanser_mupow=3_grna_",1:100)] |> sapply(length) |> hist()


```


let's try to get example data that shows this
- i want the one sceptre does badly on, the really flat one
- i want to show a really "nice" large one
- a really nice "small" one
- and maybe the real guide?




ok i want a table showing the correlation between 
(1) log(# assigned) and jacc
(2) mean(y[z==1]) and jacc

```{r}
cor_df = data.frame(
  pert_means = Matrix::rowSums(sim_data * true_perts) / Matrix::rowSums(true_perts),
  guide = rownames(true_perts)
)

for(method_name in methods) {
  cor_df[[paste0(method_name, "_num_assns_log1p")]] = sims_assns$sims[[method_name]][rownames(true_perts)] |> sapply(length) |> log1p()
}

cor_df <- cor_df |>
  left_join(
    sim_results_jacc, by = "guide"
  )


cors_jacc_and_num_assn = sapply(methods, function(method_) {
  (dplyr::select(cor_df, starts_with(method_)) |> cor())[1,2]
})


cors_jacc_and_pert_means = sapply(methods, function(method_) {
  cor(cor_df[[method_]], pert_means)
})


table_df <- data.frame(
  cors_jacc_and_num_assn, cors_jacc_and_pert_means
)

plot(cor_df$sceptre ~ cor_df$pert_means)
plot(cor_df$pertpy ~ cor_df$pert_means)

lm(sceptre ~ sceptre_num_assns_log1p + log1p(pert_means), data=cor_df) |> summary()
lm(crispat ~ crispat_num_assns_log1p + pert_means, data=cor_df) |> summary()


```





```{r}
sim_results_jacc |>
  filter(mupow == 2) |>
  arrange(sceptre) |>
  head()



guides_to_plot <- c(
  scep_bad = "params_cleanser_mupow=0_grna_46",
  scep_good = "params_cleanser_mupow=0_grna_40",
  pertpy_bad = "params_cleanser_mupow=3_grna_54",
  sceptre_meh = "params_cleanser_mupow=2_grna_78"
)

sims_assns$sims$sceptre[[guides_to_plot[["sceptre_meh"]]]] |> length()

true_perts[guides_to_plot[["scep_good"]],] |> table()
(sim_data[guides_to_plot[["scep_bad"]],] == 0) |> table()

sum(sim_data[guides_to_plot[["scep_good"]],] <= 3 &
      true_perts[guides_to_plot[["scep_good"]],] == 1)
sum(true_perts[guides_to_plot[["scep_good"]],] == 1)

# let's compute the fraction of true perts that got <= 3
fracs = sapply(seq_len(nrow(true_perts)), function(i) {
  mean(sim_data[i,][true_perts[i,] == 1] <= 3)
})

pert_sds = sapply(seq_len(nrow(true_perts)), function(i) {
  sd(sim_data[i,][true_perts[i,] == 1])
})
pert_means = sapply(seq_len(nrow(true_perts)), function(i) {
  mean(sim_data[i,][true_perts[i,] == 1])
})

num_scep_preds = sims_assns$sims$sceptre[rownames(true_perts)] |> sapply(length)

ss = sim_results_jacc |>
  left_join(
    data.frame(fracs=fracs, pert_sds=pert_sds, pert_means=pert_means, num_scep_preds = num_scep_preds, guide=rownames(true_perts)), by="guide"
  ) |>
  mutate(
    overdisp = pert_means^2/(pert_sds^2 - pert_means)
  )

ss |>
  dplyr::select()

ss |>
  ggplot(aes(x = pert_means, y = log(num_scep_preds))) +
  geom_point()


umi_df_list <- lapply(seq_along(guides_to_plot), function(i) {
  guide_ = guides_to_plot[i]
  data.frame(
    grna_sim = sim_data[guide_,],
    is_pert = true_perts[guide_,],
    group = names(guides_to_plot)[i]
  )
})
  # do.call(what=rbind)



# ok let's get correlation between log(num preds) and jaccard
num_assign_dat <- lapply(seq_along(sims_assns$sims), function(i) {
  data.frame(
    num_assigns = sims_assns$sims[[i]][rownames(true_perts)] |> sapply(length),
    method = names(sims_assns$sims)[i]
  ) %>%
    mutate(guide_id = rownames(.))
}) |>
  do.call(what=rbind) |>
  pivot_wider(names_from = method, values_from = num_assigns, id_cols = guide_id)


aa = left_join(
  sim_results_jacc, num_assign_dat, by = join_by(guide == guide_id)
) 

cor(aa$sceptre.x, log(aa$sceptre.y))
cor(aa$crispat.x, log(aa$crispat.y))
cor(aa$pertpy.x, log(aa$pertpy.y))
cor(aa$cleanser.x, log1p(aa$cleanser.y))

plot(aa$cleanser.x[aa$cleanser.y > 0], log1p(aa$cleanser.y[aa$cleanser.y > 0]))
plot(aa$cleanser.x ~ log1p(aa$cleanser.y))
plot(aa$crispat.x~  log1p(aa$crispat.y))
plot(aa$pertpy.x~  log1p(aa$pertpy.y))
plot(aa$sceptre.x~  log1p(aa$sceptre.y))




pltlist = lapply(umi_df_list, function(df) {
  make_umi_hist_of_sims(
  grna_sim = df$grna_sim,
  is_pert = df$is_pert,
  title_sim =df$group[1]
)
})


cowplot::plot_grid(plotlist=pltlist)
```


















ok now let's visualize each one


```{r}


# helper: log1p transform in base 10
logp1 <- function(x, base = 10) {
  log(x + 1, base = base)
}

# helper: build mixed exact / exponentially growing bins
# bins are:
#   0, 1, ..., k_exact as singleton bins
# then widths 2, 4, 8, 16, ...
make_mixed_bin_info <- function(max_x, k_exact = 10L) {
  stopifnot(length(max_x) == 1, is.finite(max_x), max_x >= 0)
  
  upper <- 0:k_exact
  
  width <- 2L
  while (tail(upper, 1L) < max_x) {
    upper <- c(upper, tail(upper, 1L) + width)
    width <- width * 2L
  }
  
  lower <- c(0L, head(upper, -1L) + 1L)
  
  label <- ifelse(
    lower == upper,
    as.character(upper),
    paste0(lower, "-", upper)
  )
  
  data.frame(
    bin_id = seq_along(label),
    lower = lower,
    upper = upper,
    label = label,
    xmin = seq_along(label) - 0.45,
    xmax = seq_along(label) + 0.45,
    stringsAsFactors = FALSE
  )
}

# helper: assign each count to one mixed bin
bin_with_info <- function(x, bin_info) {
  cut(
    x,
    breaks = c(-1, bin_info$upper),
    labels = bin_info$label,
    right = TRUE
  )
}

# helper: choose readable y-axis breaks on the original count scale
make_y_breaks <- function(max_count) {
  breaks <- c(
    0, 1, 2, 3, 5,
    10, 20, 30, 50,
    100, 200, 300, 500,
    1000, 2000, 3000, 5000,
    10000, 20000, 30000, 50000,
    100000, 200000, 300000, 500000,
    1000000, 2000000, 3000000, 5000000,
    10000000
  )
  breaks[breaks <= max_count]
}

# helper: exact binned counts for one vector
make_binned_counts <- function(x, bin_info) {
  stopifnot(all(is.finite(x)), all(x >= 0), all(x == floor(x)))
  
  bin <- factor(
    bin_with_info(x, bin_info),
    levels = bin_info$label
  )
  
  out <- data.frame(bin = bin) |>
    dplyr::count(bin, .drop = FALSE, name = "count") |>
    dplyr::mutate(bin = as.character(bin)) |>
    dplyr::left_join(bin_info, by = c("bin" = "label")) |>
    dplyr::arrange(bin_id)
  
  stopifnot(sum(out$count) == length(x))
  out
}

# helper: declutter x-axis labels
# label all singleton bins up to k_exact, then every `tail_every`-th tail bin, plus the last bin
# make_x_axis_spec <- function(bin_info, k_exact, tail_every = 2L) {
#   is_exact <- bin_info$lower == bin_info$upper & bin_info$upper <= k_exact
#   tail_idx <- which(!is_exact)
#   
#   keep <- is_exact
#   if (length(tail_idx) > 0) {
#     keep[tail_idx[seq(1, length(tail_idx), by = tail_every)]] <- TRUE
#     keep[max(tail_idx)] <- TRUE
#   }
#   
#   list(
#     breaks = bin_info$bin_id[keep],
#     labels = bin_info$label[keep]
#   )
# }
# helper: declutter x-axis labels
# exact_every: keep every n-th label in the 0:k_exact range
# tail_every: keep every n-th label in the exponentially growing range
make_x_axis_spec <- function(bin_info, k_exact, exact_every = 1L, tail_every = 2L) {
  is_exact <- bin_info$lower == bin_info$upper & bin_info$upper <= k_exact
  exact_idx <- which(is_exact)
  tail_idx <- which(!is_exact)
  
  # Initialize keep vector as all FALSE
  keep <- rep(FALSE, nrow(bin_info))
  
  # 1. Thin out the exact bins (e.g., 0, 1, 2, 3...)
  if (length(exact_idx) > 0) {
    keep[exact_idx[seq(1, length(exact_idx), by = exact_every)]] <- TRUE
    # Optional: always keep the last exact bin (k_exact)
    keep[max(exact_idx)] <- TRUE
  }
  
  # 2. Thin out the tail bins (e.g., 11-12, 13-16...)
  if (length(tail_idx) > 0) {
    keep[tail_idx[seq(1, length(tail_idx), by = tail_every)]] <- TRUE
    # Always keep the very last bin
    keep[max(tail_idx)] <- TRUE
  }
  
  list(
    breaks = bin_info$bin_id[keep],
    labels = bin_info$label[keep]
  )
}
# real-data plot
make_plt1a <- function(grna_real, bin_info, k_exact = 10L, tail_every = 2L, title = "real") {
  real_counts <- make_binned_counts(grna_real, bin_info)
  
  y_breaks <- make_y_breaks(max(real_counts$count))
  y_breaks_t <- logp1(y_breaks)
  x_axis <- make_x_axis_spec(bin_info, k_exact = k_exact, tail_every = tail_every)
  
  plot_df <- real_counts |>
    dplyr::mutate(
      ymin = 0,
      ymax = logp1(count)
    )
  
  ggplot2::ggplot(plot_df) +
    ggplot2::geom_rect(
      ggplot2::aes(
        xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax
      ),
      fill = "grey70",
      color = "black"
    ) +
    ggplot2::scale_x_continuous(
      breaks = x_axis$breaks,
      labels = x_axis$labels,
      expand = ggplot2::expansion(add = 0.6)
    ) +
    ggplot2::scale_y_continuous(
      breaks = y_breaks_t,
      labels = scales::label_number(big.mark = ",")(y_breaks),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::labs(
      title = title,
      x = "gRNA count bin",
      y = "cell count"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
    )
}

make_plt1b <- function(grna_sim, is_pert, bin_info, k_exact = 10L, tail_every = 2L, title = "sim") {
  stopifnot(length(grna_sim) == length(is_pert))
  stopifnot(all(is.finite(grna_sim)), all(grna_sim >= 0), all(grna_sim == floor(grna_sim)))
  
  sim_df <- data.frame(
    umi = grna_sim,
    pert_status = factor(
      ifelse(is_pert == 1, "pert", "not pert"),
      levels = c("not pert", "pert")
    )
  )
  
  sim_df$bin <- factor(
    bin_with_info(sim_df$umi, bin_info),
    levels = bin_info$label
  )
  
  sim_counts <- sim_df |>
    dplyr::count(bin, pert_status, .drop = FALSE, name = "count") |>
    dplyr::mutate(bin = as.character(bin)) |>
    dplyr::left_join(bin_info, by = c("bin" = "label")) |>
    dplyr::arrange(bin_id, pert_status) |>
    dplyr::group_by(bin, bin_id, lower, upper, xmin, xmax) |>
    dplyr::mutate(
      y0 = dplyr::lag(cumsum(count), default = 0),
      y1 = cumsum(count),
      ymin = logp1(y0),
      ymax = logp1(y1)
    ) |>
    dplyr::ungroup()
  
  stopifnot(sum(sim_counts$count) == length(grna_sim))
  if ("0" %in% sim_counts$bin) {
    stopifnot(
      sum(sim_counts$count[sim_counts$bin == "0"]) == sum(grna_sim == 0)
    )
  }
  
  y_breaks <- make_y_breaks(max(sim_counts$y1))
  y_breaks_t <- logp1(y_breaks)
  x_axis <- make_x_axis_spec(bin_info, k_exact = k_exact, tail_every = tail_every)
  
  ggplot2::ggplot(sim_counts) +
    ggplot2::geom_rect(
      ggplot2::aes(
        xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax,
        fill = pert_status
      ),
      color = "black"
    ) +
    ggplot2::scale_x_continuous(
      breaks = x_axis$breaks,
      labels = x_axis$labels,
      expand = ggplot2::expansion(add = 0.6)
    ) +
    ggplot2::scale_y_continuous(
      breaks = y_breaks_t,
      labels = scales::label_number(big.mark = ",")(y_breaks),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::labs(
      title = title,
      x = "gRNA count bin",
      y = NULL,
      fill = NULL
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
      axis.title.y = ggplot2::element_blank(),
      legend.position = c(0.98, 0.98),
      legend.justification = c(1, 1),
      legend.background = ggplot2::element_rect(fill = "white", color = "black"),
      legend.key = ggplot2::element_rect(fill = "white", color = NA)
    )
}
# wrapper: build shared bins from both datasets and return both plots
make_umi_hist_of_sims <- function(grna_sim, is_pert, k_exact = 10L, tail_every = 2L, title_sim = "sim") {
  stopifnot(length(grna_sim) == length(is_pert))
  stopifnot(all(grna_sim >= 0))
  stopifnot(all(grna_sim == floor(grna_sim)))
  
  bin_info <- make_mixed_bin_info(
    max_x = max(grna_sim),
    k_exact = k_exact
  )
  
  plt1b <- make_plt1b(
    grna_sim = grna_sim,
    is_pert = is_pert,
    bin_info = bin_info,
    k_exact = k_exact,
    tail_every = tail_every,
    title = title_sim
  )
  
  plt1b

}



```

INCLUDED IN THESIS
```{r}

make_plt1b_for_defense <- function(grna_sim, is_pert, bin_info, k_exact = 10L, tail_every = 2L, title = "sim", make_y_breaks=NULL) {
  stopifnot(length(grna_sim) == length(is_pert))
  stopifnot(all(is.finite(grna_sim)), all(grna_sim >= 0), all(grna_sim == floor(grna_sim)))
  
  sim_df <- data.frame(
    umi = grna_sim,
    pert_status = factor(
      ifelse(is_pert == 1, "Pert.", "Ambient"),
      levels = c("Ambient", "Pert.")
    )
  )
  
  sim_df$bin <- factor(
    bin_with_info(sim_df$umi, bin_info),
    levels = bin_info$label
  )
  
  sim_counts <- sim_df |>
    dplyr::count(bin, pert_status, .drop = FALSE, name = "count") |>
    dplyr::mutate(bin = as.character(bin)) |>
    dplyr::left_join(bin_info, by = c("bin" = "label")) |>
    dplyr::arrange(bin_id, pert_status) |>
    dplyr::group_by(bin, bin_id, lower, upper, xmin, xmax) |>
    dplyr::mutate(
      y0 = dplyr::lag(cumsum(count), default = 0),
      y1 = cumsum(count),
      ymin = logp1(y0),
      ymax = logp1(y1)
    ) |>
    dplyr::ungroup()
  
  stopifnot(sum(sim_counts$count) == length(grna_sim))
  if ("0" %in% sim_counts$bin) {
    stopifnot(
      sum(sim_counts$count[sim_counts$bin == "0"]) == sum(grna_sim == 0)
    )
  }
  
  if(is.null(make_y_breaks)) {
    make_y_breaks = function(max_count) {
  breaks <- c(
    0, 1, 2,  5,
    10, 20, 30, 50,
    100, 200, 300, 500,
    1000, 2000, 3000, 5000,
    10000, 20000, 30000, 50000,
    100000, 200000, 300000, 500000,
    1000000, 2000000, 3000000, 5000000,
    10000000
  )
  breaks[breaks <= max_count]
}
  }

  
  y_breaks <- make_y_breaks(max(sim_counts$y1))
  y_breaks_t <- logp1(y_breaks)
  x_axis <- make_x_axis_spec(bin_info, k_exact = k_exact, tail_every = tail_every, exact_every = 2L)
  
  ggplot2::ggplot(sim_counts) +
    ggplot2::geom_rect(
      ggplot2::aes(
        xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax,
        fill = pert_status
      ),
      color = "black"
    ) +
    ggplot2::scale_x_continuous(
      breaks = x_axis$breaks,
      labels = x_axis$labels,
      expand = ggplot2::expansion(add = 0.6)
    ) +
    ggplot2::scale_y_continuous(
      breaks = y_breaks_t,
      labels = scales::label_number(big.mark = ",")(y_breaks),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::labs(
      title = title,
      x = "gRNA count bin",
      y = NULL,
      fill = NULL
    ) +
    ggplot2::theme_bw() +
    theme_presentation() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
      axis.title.y = ggplot2::element_blank(),
      legend.position = c(0.98, 0.98),
      legend.justification = c(1, 1),
      legend.background = ggplot2::element_rect(fill = "white", color = "black"),
      legend.key = ggplot2::element_rect(fill = "white", color = NA)
    )
}
# make_umi_hist_of_sims_for_defense <- function(grna_sim, is_pert, k_exact = 10L, tail_every = 2L, title_sim = "sim", make_y_breaks=NULL) {
#   stopifnot(length(grna_sim) == length(is_pert))
#   stopifnot(all(grna_sim >= 0))
#   stopifnot(all(grna_sim == floor(grna_sim)))
#   
#   bin_info <- make_mixed_bin_info(
#     max_x = max(grna_sim),
#     k_exact = k_exact
#   )
#   
#   plt1b <- make_plt1b_for_defense(
#     grna_sim = grna_sim,
#     is_pert = is_pert,
#     bin_info = bin_info,
#     k_exact = k_exact,
#     tail_every = tail_every,
#     title = title_sim,
#     make_y_breaks=make_y_breaks
#   )
#   
#   plt1b
# 
# }
make_umi_hist_of_sims_for_defense <- function(grna_sim, is_pert, k_exact = 10L, 
                                              tail_every = 2L, title_sim = "sim", 
                                              make_y_breaks = NULL, 
                                              force_max_x = NULL) {
  stopifnot(length(grna_sim) == length(is_pert))
  stopifnot(all(grna_sim >= 0))
  stopifnot(all(grna_sim == floor(grna_sim)))
  
  # Determine the upper bound for bins
  # If force_max_x is provided, use that; otherwise, use the data max
  actual_max <- max(grna_sim)
  bin_limit <- if (!is.null(force_max_x)) max(force_max_x, actual_max) else actual_max
  
  bin_info <- make_mixed_bin_info(
    max_x = bin_limit,
    k_exact = k_exact
  )
  
  plt1b <- make_plt1b_for_defense(
    grna_sim = grna_sim,
    is_pert = is_pert,
    bin_info = bin_info, # Uses the shared/forced bin structure
    k_exact = k_exact,
    tail_every = tail_every,
    title = title_sim,
    make_y_breaks = make_y_breaks
  )
  
  plt1b
}
make_umi_hist_of_sims_for_defense(
  sim_data[1,], true_perts[1,], title_sim = "Simulated gRNA UMI counts"
) 
ggsave("~/thesis-defense-figures/grna-dist-with-components.png", width = 10, height = 7, dpi = 300)



```


next i want to show how it behaves 
```{r}

make_y_breaks = function(max_count) {
  breaks <- c(
    0, 
    10, 
    100, 
    1000, 
    10000, 
    100000, 
    1000000, 
    10000000
  )
  breaks[breaks <= max_count]
}
mupows=c(0,3)
pert_mean = c(970,120)
guides_to_plot = paste0("params_cleanser_mupow=", mupows, "_grna_", 5)
pltlist = list()
for(i in seq_along(mupows)) {
  pltlist[[i]] = make_umi_hist_of_sims_for_defense(
  sim_data[guides_to_plot[i],], true_perts[guides_to_plot[i],], title_sim = paste0("Pert mean = ", pert_mean[i]),
  make_y_breaks=make_y_breaks, force_max_x = 15000
) 
}
cowplot::plot_grid(plotlist = pltlist)
ggsave("~/thesis-defense-figures/example-sims.png", width = 10, height = 7, dpi = 300)

```



```{r}


mupow_seq = 0:3
guide_id = 1
# guides_to_plot = paste0("params_cleanser_mupow=", mupow_seq, "_grna_", 1)

df_list = list()
for(i in seq_along(mupow_seq)) {
  guide_ = paste0("params_cleanser_mupow=", mupow_seq[i], "_grna_", guide_id)
  df_list <- c(df_list, list(data.frame(
    ysim = sim_data[guide_,],
    z = true_perts[guide_,],
    mupow = mupow_seq[i]
  )))
  
}

pltlist <- lapply(df_list, function(sim_df) {
  
  if(sim_df$mupow[1] == 0) {
    title_ = "CLEANSER posterior"
  } else {
    title_ = bquote(mu[pert] ~ "/" ~ 2^.(sim_df$mupow[1]))
  }
  make_umi_hist_of_sims(
  grna_sim = sim_df$ysim,
  is_pert = sim_df$z,
  title_sim =title_
)
})

cowplot::plot_grid(plotlist=pltlist)



```


what goes wrong for the sceptre ones that don't do well with mupow=0?


```{r}
# sim_results |>
#   filter(mupow == 0) |>
#   filter(sceptre == min(sceptre))

curr_sims_sorted = sim_results |>
  filter(mupow == 0) |>
  arrange(sceptre)

# let's plot for 1,50,100

guides_ = curr_sims_sorted$guide[c(1,25, 75, 100)]

pltlist = lapply(seq_along(guides_), function(i) {
  guide_ = guides_[i]
      make_umi_hist_of_sims(
  grna_sim = sim_data[guide_,],
  is_pert = true_perts[guide_, ],
  title_sim = paste0("Jacc. rank: ", c(1,25, 75, 100)[i])
      )
})

cowplot::plot_grid(plotlist = pltlist)

print(guides_)
sim_results_jacc |>
  filter(guide %in% guides_)
sim_results_prec |>
  filter(guide %in% guides_)
sim_results_recall |>
  filter(guide %in% guides_)

# i think sceptre 

```

