---
title: "gRNA assignment benchmarking — dissertation figures"
output: html_document
date: "2026-07-14"
---

Figure pipeline for the gRNA-assignment benchmarking chapter (adapted from
`assignment-for-defense.Rmd`).

**All helper functions, palettes, the print theme, and the `save_fig()` writer live
in `assignment-helpers.R`.** Run the `setup` chunk once and every figure chunk below
then stands alone — no need to run several cells to see one plot. Figures are written
as PDFs to the `thesis-figures/` folder alongside this writeup.

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("assignment-helpers.R")

# Quantile levels for the Fig 1 low/high panels — single knob for all of §1.
q_eda <- c(.01, .25, .75, .99)

sim <- load_sim_bundle()   # $sims_assns, $true_perts, $sim_data
sim_scores <- sim_per_guide_scores(sim)
```

# §1 — EDA: illustrating the two datasets

Give the reader a feel for the raw gRNA-count data before any method touches it.
The two axes of difficulty are **sparsity** (most guides are near-absent in most
cells) and **ambient noise / heavy tails** (a low background of counts everywhere).
We show this from two angles: the distribution of counts *within a guide* (across
cells) and *within a cell* (across guides).

## Fig 1a — Guides at low vs. high sparsity/abundance

Two panels per dataset — a low-quantile (10th) and a high-quantile (90th) panel.
Guides are selected **separately** by each statistic: each panel overlays the guide
at that quantile of `grna_nonzero` (**Nonzero UMIs**) with the guide at that
quantile of `grna_total` (**Total UMIs**), colored by which statistic selected it.
(Jointly-extreme guides like "high nonzero, low total" don't exist and aren't
useful.) One figure per dataset.



<!-- ```{r fig-1a-replogle, fig.width=6.5, fig.height=3.2} -->
<!-- p_guide_repl <- plot_overlay_sweep_hist( -->
<!--   build_guide_overlay_df("replogle-rd7", q = q_eda), -->
<!--   title = "Replogle: guides by varying sparsity and total expression" -->
<!-- ) + -->
<!--   theme(axis.text=element_text(size=6)) -->
<!-- p_guide_repl -->
<!-- save_fig(p_guide_repl, "eda-guide-sweep-replogle", width = 5, height = 3) -->
<!-- ``` -->

<!-- ```{r fig-1a-gasperini, fig.width=6.5, fig.height=3.2} -->
<!-- p_guide_gasp <- plot_overlay_sweep_hist( -->
<!--   build_guide_overlay_df("gasperini", q = q_eda), -->
<!--   title = "Gasperini: guides by varying sparsity and total expression" -->
<!-- ) -->
<!-- p_guide_gasp -->
<!-- save_fig(p_guide_gasp, "eda-guide-sweep-gasperini", width = 6.5, height = 3.2) -->
<!-- ``` -->




```{r}
p_guide_both = plot_overlay_sweep_hist_combined(
  build_guide_overlay_df, q = c(.049, .95),
  title = "Guides by varying sparsity and total",
  features = "total"
) + 
  labs(x = bquote("Cell UMI count (linear then log"[2] ~"scale)"),
       title = "Exploring the data: four gRNAs by varying total expression",
       y = "Number of cells (log scale)") +
  theme(axis.text=element_text(size=6))
p_guide_both
save_fig(p_guide_both, "eda-guide-sweep-both", width = 5, height = 5)
```






## Fig 1b — Cells at low vs. high sparsity/abundance

Exact mirror of Fig 1a, for cells. Each panel (a quantile level) overlays the cell
at that quantile of per-cell `grna_n_nonzero` (**Nonzero UMIs** — number of guides
detected) with the cell at that quantile of `grna_n_umis` (**Total UMIs**), colored
by which statistic selected it. One figure per dataset.

*Note: each figure loops over every guide in the odm to pull the picked cells'
entries, so these two chunks are the slow part of §1. If it drags, add `cache=TRUE`
to the chunk options.*

<!-- ```{r fig-1b-replogle, fig.width=6.5, fig.height=3.2} -->
<!-- p_cell_repl <- plot_overlay_sweep_hist( -->
<!--   build_cell_overlay_df("replogle-rd7", q = q_eda), -->
<!--   title = "Replogle: cells by varying sparsity and total", -->
<!--   y_lab = "Number of guides" -->
<!-- ) -->
<!-- p_cell_repl -->
<!-- save_fig(p_cell_repl, "eda-cell-sweep-replogle", width = 6.5, height = 3.2) -->
<!-- ``` -->

<!-- ```{r fig-1b-gasperini, fig.width=6.5, fig.height=3.2} -->
<!-- p_cell_gasp <- plot_overlay_sweep_hist( -->
<!--   build_cell_overlay_df("gasperini", q = q_eda), -->
<!--   title = "Gasperini: cells by varying sparsity and total", -->
<!--   y_lab = "Number of guides" -->
<!-- ) -->
<!-- p_cell_gasp -->
<!-- save_fig(p_cell_gasp, "eda-cell-sweep-gasperini", width = 6.5, height = 3.2) -->
<!-- ``` -->




```{r}
p_cell_both = plot_overlay_sweep_hist_combined(
  build_cell_overlay_df, q = c(.049, .95),
  title = "Guides by varying sparsity and total",
  features = "total"
) + 
  labs(x = bquote("gRNA UMI count (linear then log"[2] ~"scale)"),
       title = "Exploring the data: four cells by varying total expression",
       y = "Number of gRNAs (log scale)") +
  theme(axis.text=element_text(size=6))
p_cell_both
save_fig(p_cell_both, "eda-cell-sweep-both", width = 5, height = 5)
```



## Fig 1c — Cell-level covariates (2x2 hexbins)

The joint structure of the four per-cell count covariates, read straight from the
ondisc-backed sceptre object's covariate data frame. Each panel is a hexbin of one
covariate pair (strip reads "y vs. x"). Top row shares x = number of nonzero gRNAs;
bottom row shares x = number of expressed genes. One figure per dataset.


<!-- ```{r fig-1c-replogle, cache=TRUE, fig.width=6.5, fig.height=5.6} -->
<!-- p_cov_grid_repl <- plot_covariate_grid( -->
<!--   build_cell_covariate_df("replogle-rd7"), "Replogle" -->
<!-- ) -->
<!-- p_cov_grid_repl -->
<!-- save_fig(p_cov_grid_repl, "eda-covariate-grid-replogle", width = 6.5, height = 5.6) -->
<!-- ``` -->

<!-- ```{r fig-1c-gasperini, cache=TRUE, fig.width=6.5, fig.height=5.6} -->
<!-- p_cov_grid_gasp <- plot_covariate_grid( -->
<!--   build_cell_covariate_df("gasperini"), "Gasperini" -->
<!-- ) -->
<!-- p_cov_grid_gasp -->
<!-- save_fig(p_cov_grid_gasp, "eda-covariate-grid-gasperini", width = 6.5, height = 5.6) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- p_cov_pair_gasp <- plot_covariate_pair( -->
<!--   build_cell_covariate_df("gasperini"), "Gasperini" -->
<!-- ) -->
<!-- p_cov_pair_gasp -->
<!-- save_fig(p_cov_pair_gasp, "eda-covariate-pair-gasperini", width = 6.5, height = 5.6) -->
<!-- ``` -->


## Fig 1d — CLEANSER low-pass depth (hexbin)

CLEANSER's per-cell "low-pass" depth — the sum of the cell's gRNA counts that are
$\leq 2$, the signal-free ambient scaling factor it conditions on — against total
gRNA UMIs per cell. One figure per dataset.

*Note: the low-pass depth loops over every guide in the odm, so these two chunks are
slow (~10s Replogle, ~45s Gasperini) — `cache=TRUE` keeps them one-time.*


<!-- ```{r fig-1d-replogle, cache=TRUE, fig.width=5.5, fig.height=4.4} -->
<!-- p_lowpass_repl <- plot_low_pass_hex( -->
<!--   build_grna_covariate_df("replogle-rd7"), "Replogle" -->
<!-- ) -->
<!-- p_lowpass_repl -->
<!-- save_fig(p_lowpass_repl, "eda-low-pass-hex-replogle", width = 5.5, height = 4.4) -->
<!-- ``` -->

<!-- ```{r fig-1d-gasperini, cache=TRUE, fig.width=5.5, fig.height=4.4} -->
<!-- p_lowpass_gasp <- plot_low_pass_hex( -->
<!--   build_grna_covariate_df("gasperini"), "Gasperini" -->
<!-- ) -->
<!-- p_lowpass_gasp -->
<!-- save_fig(p_lowpass_gasp, "eda-low-pass-hex-gasperini", width = 5.5, height = 4.4) -->
<!-- ``` -->



# sceptre bug

<!-- ```{r} -->
<!-- source("assignment-helpers.R") -->
<!-- ``` -->


```{r, cache=TRUE}
sim_metrics_df <- load_sim_bugfix_metrics()   # scores all 3 datasets (a bit slow; cache it)
plot_bug = plot_sim_bugfix_metrics(
  sim_metrics_df,
  title = "      Evaluation of SCEPTRE bug\n      on simulated guides",
  metric_order = c("jaccard", "precision", "recall"),
  methods = c("SCEPTRE", "Fixed Poisson"),
  y_lim_min = .5
) 
plot_bug
save_fig(plot_bug, "sceptre-bug-sims", width = 5.5, height = 3.5)
```




combining sim guides + sceptre bug performance
```{r}
combo_plot = cowplot::plot_grid(
  p_example_sims + theme(axis.text=element_text(size=6), plot.title = element_text(size=14)),
  plot_bug + theme(axis.text=element_text(size=6), plot.title = element_text(size=6)),
  nrow=1, labels = c("(A)", "(B)"))
combo_plot
save_fig(combo_plot, "sims-and-sceptre-bug-evals", width = 9, height = 6)
```


<!-- ```{r} -->
<!-- sim_metrics_df <- load_sim_bugfix_metrics()   # scores all 3 datasets (a bit slow; cache it) -->
<!-- plot_bug = plot_sim_bugfix_metrics_t( -->
<!--   sim_metrics_df, -->
<!--   title = "Evaluation of SCEPTRE bug on simulated guides", -->
<!--   metric_order = c("precision", "recall"), -->
<!--   methods = c("SCEPTRE", "Fixed Poisson") -->
<!-- )  -->
<!-- plot_bug -->
<!-- save_fig(plot_bug, "sceptre-bug-sims", width = 5.5, height = 5.5) -->
<!-- ``` -->














# §2 — Simulations with ground truth

With no real ground truth for gRNA assignment, we simulate from CLEANSER's model:
sequencing-depth scaling factors from Replogle, perturbation status
$Z_i \sim \text{Ber}(\pi)$, and counts $Y_i \mid Z_i \sim \text{NegBin}(L_i\mu_{Z_i},
\theta_{Z_i})$. Four perturbation strengths are generated ($\mu_1 = 971/2^j$ for
$j = 0,1,2,3$), all sharing the same $L_i$. Each method is then scored against the
known perturbation matrix.

# ```{r sim-load, cache=TRUE}
# sim <- load_sim_bundle()   # $sims_assns, $true_perts, $sim_data
# ```

## Fig 2a — Example simulated guides (ambient vs perturbed)

Four example simulated guides spanning the burst and overdispersion settings. Top
row: the two medium-overdispersion settings (large vs. small burst) at
$\mu_\text{pert} = 1000$. Bottom row: the large-overdispersion, no-burst setting at
a strong ($\mu_\text{pert} = 970$) and a weak ($\mu_\text{pert} = 121$)
perturbation. Each cell's count is colored by its ground-truth status: the ambient
component (teal) is the low-count background present in every cell, while the
perturbed component (red) shifts to higher counts — clearly separated at large
$\mu_\text{pert}$, but collapsing toward the ambient pile at small $\mu_\text{pert}$.

```{r fig-2a, fig.width=6.5, fig.height=5.2}
IN_BASE <- file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "guide_assignment/input_data")

# 2x2 of example simulated guides. Top row: the two burst settings at the largest
# mu (1000). Bottom row: the no-burst setting at two mu levels (strong 970, weak
# 121). mu is added as a third facet line to mark the choice.
mu_of <- function(ds, p) { t <- sim_bugfix_datasets[[ds]]; t$mu_pert[match(p, t$p)] }
mk_2a <- function(key, p) {
  parts <- strsplit(key, "|", fixed = TRUE)[[1]]; ds <- parts[1]; np <- parts[2]
  list(dataset = ds, guide = sprintf("sim_%s_%s_1", np, p),
       tag = paste0(sim_row_renamer[[key]], "\nµ = ", round(mu_of(ds, p))))
}
picks_2a <- list(
  mk_2a("sims_sum_2np_3p|NPhighvar",  "P3"),
  mk_2a("sims_sum_2np_3p|NPlowvar",   "P3"),
  mk_2a("sims_sum_repeat_old|NPpois", "Plarge"),
  mk_2a("sims_sum_repeat_old|NPpois", "Psmall")
)
df_2a <- purrr::map_dfr(picks_2a, function(p) {
  gm <- readRDS(file.path(IN_BASE, p$dataset, "sceptre", "grna_matrix.rds"))
  tp <- readRDS(file.path(IN_BASE, p$dataset, "true_pert_matrix.rds"))
  tibble(umi = as.integer(gm[p$guide, ]),
         pert_status = factor(ifelse(tp[p$guide, ] == 1, "Pert.", "Ambient"),
                              levels = c("Ambient", "Pert.")),
         panel = p$tag)
}) |> mutate(panel = factor(panel, levels = vapply(picks_2a, `[[`, "", "tag")))

p_example_sims <- plot_sim_component_hist(
  df_2a, title = "Simulated gRNA counts:\nambient vs. perturbed"
) + facet_wrap(~ panel, nrow = 2) + theme(axis.text.x = element_text(size = 6), 
                                          axis.text.y = element_text(size = 7))
p_example_sims
save_fig(p_example_sims, "sim-example-guides", width = 6.5, height = 5.5)
```

## Fig 2b — Mean metrics vs. perturbation strength

The central simulation result. Each method is scored per guide against ground truth
(truly-perturbed cells with a nonzero UMI), and we plot the mean **Jaccard /
Precision / Recall** at each perturbation strength ($\mu_1 = 971/2^j$). Recall stays
high for all methods (they capture most true positives); the separation is in
precision at low $\mu_1$, where over-assignment hurts.

# ```{r sim-scores, cache=TRUE}
# # Per-(guide, method) Jaccard/Precision/Recall — the expensive step; reused by 2c.
# sim_scores <- sim_per_guide_scores(sim)
# ```

```{r fig-2b, fig.width=6.5, fig.height=3}
p_sim_metrics <- plot_sim_avg_metrics(
  sim_avg_metrics(sim_scores),
  title = "Performance on simulated data"
) +
  labs(x = expression("Perturbed cell average mean"))
p_sim_metrics
save_fig(p_sim_metrics, "sim-avg-metrics", width = 6.5, height = 3)
```

## Fig 2c — Per-guide Jaccard, strong vs. weak perturbation

Beyond the means: the full spread of per-guide Jaccard for each method at the
strongest and weakest perturbation. At $\mu_1 = 971$ most guides sit near 1; at
$\mu_1 = 121$ the distributions slide down and spread out, and any mass piling up at
0 marks guides a method failed on outright.

```{r fig-2c, fig.width=6.5, fig.height=3}
p_sim_jaccard <- plot_sim_jaccard_hist(
  sim_scores,
  mupows = c(0, 3),   # largest (mu=971) and smallest (mu=121) only
  title = "Per-guide Jaccard to ground truth"
) +
  theme(aspect.ratio=1) +
  facet_wrap(~panel)
p_sim_jaccard
save_fig(p_sim_jaccard, "sim-jaccard-hists", width = 6, height = 4)
```



```{r}
combo_sim_metrics = cowplot::plot_grid(
  p_sim_metrics + theme(aspect.ratio=1),
  p_sim_jaccard,
  nrow=2,
  labels = c("(A)", "(B)")
)
combo_sim_metrics
save_fig(combo_sim_metrics, "sim-avg-metrics-and-jaccard-hists", width = 6, height = 4)
```


## Fig 2d — Assignments per guide vs. tail heaviness

The same view used on the real data (§3), here on the simulations: for each guide,
the number of cells a method assigns vs. the guide's tail mean (mean UMI among
counts $\geq$ 10). Methods that key on the heavy tail assign more cells as the tail
grows; the sensitivity differs sharply by method.

```{r fig-2d, fig.width=6.5, fig.height=5}
p_sim_tail <- plot_tail_sensitivity(
  build_tail_sensitivity_df(sim$sim_data, sim$sims_assns$sims, thresh = 10),
  title = "Simulations: assignments vs. tail heaviness",
  thresh = 10,
  colour_by = "mu"
) + facet_wrap(~ method, nrow = 1) 
p_sim_tail
save_fig(p_sim_tail, "sim-tail-sensitivity", width = 7, height = 3)
```

## Table 2 — Calling thresholds by method and perturbation strength

For each method and perturbation level, averaged over the 100 guides: the smallest
UMI called perturbed (min over assigned cells) and the largest UMI left unperturbed
(max over unassigned cells). Where the "largest unperturbed" exceeds the "smallest
perturbed," the method's decision boundary is non-monotone in count — a sign it is
leaning on covariates rather than a hard UMI threshold.


<!-- ```{r table-2-thresholds, cache=TRUE} -->
<!-- sim_thresh_tbl <- sim_threshold_table(sim) -->
<!-- knitr::kable(sim_thresh_tbl, digits = 1) -->
<!-- ``` -->


## Fig 2e — Calling boundary by method

Table 2 made graphical. For each method (x-axis), two boxplots across the 100
guides: the largest UMI it leaves unperturbed and the smallest it calls perturbed.
Whiskers span the full range, so each box is the five-number summary (min, Q1,
median, Q3, max). Faceted by perturbation strength (2x2). Where a method's two
boxes overlap, its decision boundary is not a clean count cutoff.

<!-- ```{r fig-2e, fig.width=6.5, fig.height=5} -->
<!-- sim_thresh_pg <- sim_threshold_per_guide(sim) -->
<!-- p_sim_boundary <- plot_sim_threshold_box( -->
<!--   sim_thresh_pg, -->
<!--   title = "Calling boundary by method and perturbation strength" -->
<!-- ) + -->
<!--   scale_y_log10() -->
<!-- p_sim_boundary -->
<!-- save_fig(p_sim_boundary, "sim-threshold-box", width = 6.5, height = 5) -->
<!-- ``` -->

# §3 — Real data: agreement between methods

No ground truth here, so instead of scoring against a known answer we ask how much
the methods agree with each other, and what drives their assignments. The
real-data assignments are loaded once below (800-guide downsample of each dataset,
100k cells).

```{r real-load, cache=TRUE}
real <- load_real_bundle(nguides = 800L)  # $assns, $grna_mat, per dataset
```

## Fig 3a — Average pairwise Jaccard between methods

For each pair of methods, the mean per-guide Jaccard of their assigned-cell sets.
One dataset per triangle — Replogle above the diagonal, Gasperini below — so a
single matrix shows both. Warmer cells = closer agreement.

```{r fig-3a, fig.width=5, fig.height=4.6}
p_jaccard_heat <- plot_jaccard_heatmap(
  build_jaccard_heatmap_df(real, upper = "replogle-rd7", lower = "gasperini"),
  upper_label = "Replogle",
  lower_label = "Gasperini",
  title = "Average per-guide Jaccard between methods"
) +
  theme(axis.text=element_text(size=8), plot.title=element_text(size=10))
p_jaccard_heat
save_fig(p_jaccard_heat, "real-jaccard-heatmap", width = 4, height = 4)
```

## Fig 3b — Assignments per guide vs. tail heaviness (Replogle)

The real-data analogue of Fig 2d. For each guide, the number of cells a method
assigns vs. the guide's tail mean (mean UMI among counts $\geq$ 10). Methods that
key on the heavy tail assign more cells as the tail grows.

<!-- ```{r fig-3b, fig.width=6.5, fig.height=5} -->
<!-- p_tail_repl <- plot_tail_sensitivity( -->
<!--   build_tail_sensitivity_df(real$grna_mat[["replogle-rd7"]], -->
<!--                             real$assns[["replogle-rd7"]], thresh = 10), -->
<!--   title = "Replogle: number of assignments vs. tail heaviness", -->
<!--   thresh = 10 -->
<!-- ) -->
<!-- p_tail_repl -->
<!-- save_fig(p_tail_repl, "real-tail-sensitivity-replogle", width = 6.5, height = 5) -->
<!-- ``` -->

<!-- ## Fig 3c — Assignments per guide vs. tail heaviness (Gasperini) -->

<!-- Same as Fig 3b, for Gasperini. -->

<!-- ```{r fig-3c, fig.width=6.5, fig.height=5} -->
<!-- p_tail_gasp <- plot_tail_sensitivity( -->
<!--   build_tail_sensitivity_df(real$grna_mat[["gasperini"]], -->
<!--                             real$assns[["gasperini"]], thresh = 10), -->
<!--   title = "Gasperini: number of assignments vs. tail heaviness", -->
<!--   thresh = 10 -->
<!-- ) -->
<!-- p_tail_gasp -->
<!-- save_fig(p_tail_gasp, "real-tail-sensitivity-gasperini", width = 6.5, height = 5) -->
<!-- ``` -->



## Fig 3d — Tail heaviness, both datasets overlaid

Fig 3b and 3c on one figure: both datasets in the same panel per method (colour =
dataset, Replogle blue / Gasperini orange). x is `log1p` since Replogle's tails are
far heavier than Gasperini's.

```{r fig-3d, fig.width=6.5, fig.height=5.5}
p_tail_both <- plot_tail_sensitivity_combined(
  build_tail_sensitivity_combined_df(real, thresh = 10),
  title = "Number of assignments vs. tail heaviness",
  thresh = 10
)
p_tail_both
save_fig(p_tail_both, "real-tail-sensitivity-both", width = 6.5, height = 5.5)
```


```{r}
combo_real <- cowplot::plot_grid(
  p_jaccard_heat,
  p_tail_both,
  nrow=1
)
combo_real
```



## Fig 3e — Guides where a method went wrong (UMI histograms)

Four illustrative guides, one per failure mode, each a UMI histogram of its nonzero
counts (coloured by dataset). (a) a Gasperini guide crispat's mixture model fits
degenerately (0 assigned); (b) a Replogle guide whose heavy tail makes SCEPTRE
over-assign; (c) a Replogle guide CLEANSER leaves empty despite a heavy tail; (d) a
Gasperini guide pertpy leaves empty. Titles are passed via `titles`.



```{r fig-3e, fig.width=9, fig.height=6.5}
problem_picks <- list(
  list(dataset = "Gasperini", guide = "ATAAAGCCTGTTATTACCAA"),          # crispat = 0
  list(dataset = "Replogle",  guide = "9570_UFL1_P1P2_ENSG00000014123"),# sceptre over-assigns
  list(dataset = "Replogle",  guide = "8221_SNRNP48_P1P2_ENSG00000168566"), # cleanser = 0
  list(dataset = "Gasperini", guide = "TTGTCTCGACTCTCACCATA")           # pertpy = 0
)
p_problem_guides <- plot_guide_umi_grid(
  real, problem_picks,
  titles = c("(a) Gasperini: crispat assigns 0",
             "(b) Replogle: SCEPTRE over-assigns",
             "(c) Replogle: CLEANSER assigns 0 ",
             "(d) Gasperini: pertpy assigns 0")
)
p_problem_guides
save_fig(p_problem_guides, "real-problem-guides", width = 9, height = 6.5)
```



# §4 — Computational benchmarking

Runtime and peak memory of each method as the guide count grows. The four methods
run in-memory across a guide-count sweep (circles); the SCEPTRE pipeline runs
distributed over workers on the full dataset (triangle, one point). The
`comp-load` chunk reads all the Nextflow trace files once.

```{r comp-load, cache=TRUE}
trace_all <- load_comp_bundle()  # method, peak_rss_gb, realtime_sec, nguides, source_data
```

## Fig 4 — Combined (2×2: dataset × runtime/memory)

```{r fig-4, fig.width=9, fig.height=7.5}
p_comp_grid <- plot_comp_grid(trace_all)
p_comp_grid
save_fig(p_comp_grid, "assign-compute-grid", width = 9, height = 7.5)
```

## Fig 4′ — Combined, faceted (dataset × metric)

A tidier alternative to Fig 4: dataset as facet columns, runtime and memory
stacked as rows. Each metric keeps its own y-axis with the exact ticks and labels
of the standalone figures (H:M:S for runtime, GB for memory).

```{r fig-4-faceted, fig.width=7.5, fig.height=6.2}
p_comp_grid_faceted <- plot_comp_grid_faceted(trace_all, title = "gRNA assignment: runtime and memory benchmarking")
p_comp_grid_faceted
save_fig(p_comp_grid_faceted, "assign-compute-grid-faceted", width = 8, height = 6.75)
```

## Fig 4a — Runtime (Replogle)

```{r fig-4a, fig.width=6.5, fig.height=4.4}
p_runtime_repl <- plot_comp_runtime(trace_all, "Replogle: assignment runtime", "replogle")
p_runtime_repl
save_fig(p_runtime_repl, "assign-runtime-replogle", width = 6.5, height = 4.4)
```

## Fig 4b — Memory (Replogle)

```{r fig-4b, fig.width=6.5, fig.height=4.4}
p_memory_repl <- plot_comp_memory(trace_all, "Replogle: assignment memory", "replogle")
p_memory_repl
save_fig(p_memory_repl, "assign-memory-replogle", width = 6.5, height = 4.4)
```

## Fig 4c — Runtime (Gasperini)

```{r fig-4c, fig.width=6.5, fig.height=4.4}
p_runtime_gasp <- plot_comp_runtime(trace_all, "Gasperini: assignment runtime", "gasperini")
p_runtime_gasp
save_fig(p_runtime_gasp, "assign-runtime-gasperini", width = 6.5, height = 4.4)
```

## Fig 4d — Memory (Gasperini)

```{r fig-4d, fig.width=6.5, fig.height=4.4}
p_memory_gasp <- plot_comp_memory(trace_all, "Gasperini: assignment memory", "gasperini")
p_memory_gasp
save_fig(p_memory_gasp, "assign-memory-gasperini", width = 6.5, height = 4.4)
```
