---
title: "Bug compare, no circularity — glmpoisallgrnafix: poisbug vs poisfix on simulations"
output: html_document
---

Cross the two axes of the guide-assignment model on the three simulated datasets:

- **offset** (colour): `glmpois` (shared-X offset, keeps the `grna_n_umis`
  circularity) vs `glmpoisallgrnafix` (circularity-corrected gRNA covariates +
  uncorrected response covariates).
- **mixture M-step** (linetype + point shape): `poisbug` = additive
  `g_mus_pert1 = g_mus_pert0 + gamma` vs `poisfix` = multiplicative
  `g_mus_pert1 = g_mus_pert0 * exp(gamma)`.

The `glmpois` methods come from run `run_bugfix_sims`; the `glmpoisallgrnafix`
methods from `run_grnafix_sims_compare_bug`.

Ground truth per guide = cells that are perturbed (`true_pert_matrix == 1`) AND
observed nonzero (`grna_matrix > 0`). We score per-guide precision / recall /
jaccard against that, then average over guides at each perturbation strength.

Plot layout mirrors `plot_sim_bugfix_metrics` from the thesis helpers: one row per
simulation (dataset x noise-tail `np`), one column per metric, a line per method
across `mu_pert` (log2 x-axis; rows keep independent x-scales since `mu_pert`
ranges differ by orders of magnitude across datasets).

**The plot chunk depends ONLY on the `setup` chunk.** Run `setup` plus the plot.

```{r setup, message=FALSE, warning=FALSE}
rm(list=ls())
suppressPackageStartupMessages({
  library(tidyverse)
  library(Matrix)
})

## --- config --------------------------------------------------------------
BM  <- "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment"
IN  <- file.path(BM, "input_data")

# Every (offset x mixture) method, its script name, and the run it lives in.
#   offset:  glmpois            = shared-X offset, has the grna_n_umis circularity
#            glmpoisallgrnafix  = circularity-corrected grna covars (+ response covars)
#   mixture: poisbug            = additive (bugged) M-step
#            poisfix            = multiplicative (fixed) M-step
# colour  = offset choice; linetype + point shape = mixture choice.
METHOD_SPECS <- tribble(
  ~method,                      ~run,                          ~offset,                        ~mixture,
  "glmpois_poisbug",            "run_bugfix_sims",             "glmpois (circularity)",        "poisbug (additive)",
  "glmpois_poisfix",            "run_bugfix_sims",             "glmpois (circularity)",        "poisfix (multiplicative)",
  "glmpoisallgrnafix_poisbug",  "run_grnafix_sims_compare_bug","glmpoisallgrnafix (no circ.)", "poisbug (additive)",
  "glmpoisallgrnafix_poisfix",  "run_grnafix_sims_compare_bug","glmpoisallgrnafix (no circ.)", "poisfix (multiplicative)"
)
OFFSET_LEVELS  <- c("glmpois (circularity)", "glmpoisallgrnafix (no circ.)")
MIXTURE_LEVELS <- c("poisbug (additive)", "poisfix (multiplicative)")

# Each dataset's P-label -> mu_pert (perturbation strength) mapping, and the order
# rows should appear in. (Same maps used in bugfix-sims-metrics.R / thesis.)
DATASETS <- list(
  sims_sum_2np_3p = tibble(
    p = c("P1", "P2", "P3"), mu_pert = c(500, 750, 1000)
  ),
  sims_sum_1np_3p_disp = tibble(
    p = c("Plowdisp", "Pmeddisp", "Phighdisp"), mu_pert = c(101000, 1001000, 10001000)
  ),
  sims_sum_repeat_old = tibble(
    p = c("Psmall", "Pmed", "Plarge"), mu_pert = c(970 / 8, 970 / 4, 970)
  )
)

## --- metric helpers ------------------------------------------------------
set_metrics <- function(pred, truth) {
  pred <- unique(pred); truth <- unique(truth)
  n_inter <- length(intersect(pred, truth))
  n_pred  <- length(pred); n_truth <- length(truth)
  n_union <- length(union(pred, truth))
  tibble(
    precision = if (n_pred  == 0) NA_real_ else n_inter / n_pred,
    recall    = if (n_truth == 0) NA_real_ else n_inter / n_truth,
    jaccard   = if (n_union == 0) NA_real_ else n_inter / n_union
  )
}
part_k     <- function(x, k) vapply(strsplit(x, "_", fixed = TRUE),
                                    function(z) if (length(z) >= k) z[[k]] else NA_character_,
                                    character(1))
mean_or_na <- function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)
sd_or_na   <- function(x) if (sum(!is.na(x)) <= 1) NA_real_ else sd(x, na.rm = TRUE)

## --- per-guide metrics for one dataset x method --------------------------
# assignment matrix: outputs/<run>/assignment_matrix_script_<method>_<dataset>.rds
# ground truth: perturbed (true_pert_matrix == 1) AND observed nonzero (grna > 0).
metrics_for <- function(dataset, method, run) {
  in_ds       <- file.path(IN, dataset)
  grna_matrix <- readRDS(file.path(in_ds, "sceptre", "grna_matrix.rds"))
  true_perts  <- readRDS(file.path(in_ds, "true_pert_matrix.rds"))
  guide_names <- rownames(grna_matrix)

  assn <- readRDS(file.path(BM, "outputs", run,
                            paste0("assignment_matrix_script_", method, "_", dataset, ".rds")))
  stopifnot(all(guide_names == rownames(assn)))

  pred_assns <- lapply(guide_names, function(g) as.integer(which(assn[g, ])))
  true_assns <- lapply(guide_names, function(g)
    as.integer(which(true_perts[g, ] == 1 & grna_matrix[g, ] > 0)))
  names(pred_assns) <- names(true_assns) <- guide_names

  map_dfr(guide_names, function(g)
    set_metrics(pred_assns[[g]], true_assns[[g]]) |>
      mutate(dataset = dataset, method = method, grna_id = g,
             np = part_k(g, 2), p = part_k(g, 3), .before = 1))
}

## --- load & score all datasets x methods, average per (np, mu_pert) ------
sim_metrics <- imap_dfr(DATASETS, function(p_to_mean, dataset) {
  pmap_dfr(METHOD_SPECS, function(method, run, offset, mixture)
    metrics_for(dataset, method, run)) |>
    left_join(p_to_mean, by = "p")
}) |>
  left_join(select(METHOD_SPECS, method, offset, mixture), by = "method") |>
  pivot_longer(c(precision, recall, jaccard), names_to = "metric", values_to = "value") |>
  group_by(dataset, np, p, mu_pert, method, offset, mixture, metric) |>
  summarize(mean = mean_or_na(value), sd = sd_or_na(value),
            n = sum(!is.na(value)), .groups = "drop") |>
  mutate(
    ci   = 2 * sd / sqrt(n),
    ymin = pmax(0, mean - ci), ymax = pmin(1, mean + ci),
    # row key: one simulation = dataset x noise-tail; column = metric
    sim_label = paste(dataset, np, sep = " | "),
    offset    = factor(offset,  levels = OFFSET_LEVELS),
    mixture   = factor(mixture, levels = MIXTURE_LEVELS),
    metric    = factor(metric, levels = c("jaccard", "precision", "recall"))
  )

# order sim_label rows by dataset order in DATASETS, then np within dataset
sim_lab_levels <- sim_metrics |>
  distinct(dataset, np, sim_label) |>
  mutate(dataset = factor(dataset, levels = names(DATASETS))) |>
  arrange(dataset, np) |>
  pull(sim_label)
sim_metrics$sim_label <- factor(sim_metrics$sim_label, levels = sim_lab_levels)

print(sim_metrics)
```

## Precision / recall / jaccard vs ground truth across simulations

Rows = simulation (dataset x noise-tail `np`); columns = metric; line per method
across `mu_pert`. Error bars are ±2 SE. Each row is its own plot (independent
log2 x-scale, since `mu_pert` ranges differ by orders of magnitude), stacked with
patchwork and sharing one legend — same layout as `plot_sim_bugfix_metrics`.

```{r plot-metrics, fig.width=9, fig.height=7}
library(patchwork)

# Zoom the (shared) y-axis to where the data actually live, with a little padding,
# instead of the full [0, 1]. Covers the error bars (ymin/ymax), not just means.
y_lo <- max(0, min(sim_metrics$ymin, na.rm = TRUE) - 0.02)
y_hi <- min(1, max(sim_metrics$ymax, na.rm = TRUE) + 0.02)

sims <- levels(sim_metrics$sim_label)
row_plot <- function(s, is_top, is_bottom) {
  d <- filter(sim_metrics, sim_label == s)
  ggplot(d, aes(mu_pert, mean, colour = offset, shape = mixture,
                linetype = mixture, group = interaction(offset, mixture))) +
    geom_line(linewidth = 0.5) +
    geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.04, alpha = 0.7,
                  linetype = "solid") +
    geom_point(size = 1.9) +
    facet_grid(sim_label ~ metric) +
    scale_x_continuous(trans = "log2", breaks = sort(unique(d$mu_pert))) +
    coord_cartesian(ylim = c(y_lo, y_hi)) +
    labs(x = if (is_bottom) expression(mu[pert]) else NULL, y = NULL,
         colour = "offset", shape = "mixture", linetype = "mixture") +
    theme_bw() +
    theme(legend.position = "bottom") +
    (if (!is_top) theme(strip.text.x = element_blank()) else NULL)  # metric labels once
}

rows <- imap(sims, function(s, i)
  row_plot(s, is_top = i == 1, is_bottom = i == length(sims)))

pm <- wrap_plots(rows, ncol = 1, guides = "collect") +
  plot_annotation(
    title = "poisbug vs poisfix, circularity vs no circularity, on simulations"
  ) &
  theme(legend.position = "bottom")
print(pm)
```

ok so even with the circularity removed, we still have the bug essentially dominating

## Replogle 800-guide subsample: #assignments vs upper-tail mean, all 4 models

Real-data view of the same 2x2. For each guide: x = upper-tail mean
`mean(y | y >= 10)`, y = number of cells that method assigned it to. Facets are
rows = offset, cols = mixture. y-scales are shared, so the circularity + poisbug
overprediction (many guides assigned to 1000s of cells) is directly visible
against the other three. Depends only on the `setup` chunk.

```{r plot-replogle-assign, fig.width=8.5, fig.height=7}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"

# Where each model's assignments for this dataset live (different runs).
REPL_SPECS <- tribble(
  ~method,                      ~run,                          ~offset,                        ~mixture,
  "glmpois_poisbug",            "script_real_data",            "glmpois (circularity)",        "poisbug (additive)",
  "glmpois_poisfix",            "script_real_data",            "glmpois (circularity)",        "poisfix (multiplicative)",
  "glmpoisallgrnafix_poisbug",  "run_grnafix_sceptre",         "glmpoisallgrnafix (no circ.)", "poisbug (additive)",
  "glmpoisallgrnafix_poisfix",  "run_grnafix_bugfix_sceptre",  "glmpoisallgrnafix (no circ.)", "poisfix (multiplicative)"
)

grna_repl        <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names_repl <- rownames(grna_repl)
tail_mean_repl   <- apply(grna_repl, 1, function(y) {
  y <- as.numeric(y); t <- y[y >= 10]; if (length(t) == 0) NA_real_ else mean(t)
})

repl_assign <- pmap_dfr(REPL_SPECS, function(method, run, offset, mixture) {
  assn <- readRDS(file.path(BM, "outputs", run,
                            paste0("assignment_matrix_script_", method, "_", DS_REPL, ".rds")))
  stopifnot(all(guide_names_repl == rownames(assn)))
  tibble(grna_id = guide_names_repl, tail_mean = tail_mean_repl,
         n_assign = as.numeric(Matrix::rowSums(assn)),
         offset = offset, mixture = mixture)
}) |>
  mutate(offset  = factor(offset,  levels = OFFSET_LEVELS),
         mixture = factor(mixture, levels = MIXTURE_LEVELS))

pr <- ggplot(repl_assign, aes(tail_mean, n_assign)) +
  geom_point(colour = "grey25", alpha = 0.5, size = 1.1) +
  facet_grid(offset ~ mixture) +
  labs(x = "upper-tail mean  mean(y | y >= 10)", y = "# assignments",
       title = "Replogle 800-guide subsample: #assignments vs upper-tail mean",
       subtitle = "rows = offset choice, cols = mixture M-step") +
  theme_bw()
print(pr)
```

## Pairwise assignment agreement per guide (Jaccard)

For every pair of the 4 methods and every guide `g`, compute the Jaccard overlap
of the two cell-sets each method assigns to `g`:
`|A_i ∩ A_j| / |A_i ∪ A_j|`. One histogram per pair (`choose(4,2) = 6`), over the
800 guides. Guides where **both** methods assign zero cells have an empty union
(Jaccard undefined) and are dropped; the per-pair count is printed at the end.
Depends only on the `setup` chunk.

```{r plot-replogle-pair-jaccard, fig.width=10, fig.height=6}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"
REPL_SPECS <- tribble(
  ~method,                      ~run,                          ~offset,                        ~mixture,
  "glmpois_poisbug",            "script_real_data",            "glmpois (circularity)",        "poisbug (additive)",
  "glmpois_poisfix",            "script_real_data",            "glmpois (circularity)",        "poisfix (multiplicative)",
  "glmpoisallgrnafix_poisbug",  "run_grnafix_sceptre",         "glmpoisallgrnafix (no circ.)", "poisbug (additive)",
  "glmpoisallgrnafix_poisfix",  "run_grnafix_bugfix_sceptre",  "glmpoisallgrnafix (no circ.)", "poisfix (multiplicative)"
)

grna_repl        <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names_repl <- rownames(grna_repl)

# compact per-method label: <circ|nocirc>+<bug|fix>
mlab <- setNames(
  paste0(ifelse(grepl("allgrnafix", REPL_SPECS$method), "nocirc", "circ"), "+",
         ifelse(grepl("poisfix",    REPL_SPECS$method), "fix",    "bug")),
  REPL_SPECS$method)

# per method: guide -> integer vector of assigned cell indices
method_assign <- pmap(REPL_SPECS, function(method, run, offset, mixture) {
  assn <- readRDS(file.path(BM, "outputs", run,
                            paste0("assignment_matrix_script_", method, "_", DS_REPL, ".rds")))
  stopifnot(all(guide_names_repl == rownames(assn)))
  lapply(guide_names_repl, function(g) as.integer(which(assn[g, ]))) |> setNames(guide_names_repl)
})
names(method_assign) <- REPL_SPECS$method

jac <- function(a, b) {
  u <- length(union(a, b)); if (u == 0) NA_real_ else length(intersect(a, b)) / u
}

pairs      <- combn(REPL_SPECS$method, 2, simplify = FALSE)
pair_levels <- vapply(pairs, function(pr) paste(mlab[pr[1]], "vs", mlab[pr[2]]), character(1))
pair_df <- map_dfr(pairs, function(pr) {
  mi <- pr[1]; mj <- pr[2]
  tibble(grna_id = guide_names_repl,
         pair    = paste(mlab[mi], "vs", mlab[mj]),
         jaccard = vapply(guide_names_repl,
                          function(g) jac(method_assign[[mi]][[g]], method_assign[[mj]][[g]]),
                          numeric(1)))
}) |>
  mutate(pair = factor(pair, levels = pair_levels))

pj <- ggplot(pair_df, aes(jaccard)) +
  geom_histogram(bins = 40, boundary = 0, fill = "steelblue", colour = "white") +
  facet_wrap(~ pair, ncol = 3) +
  labs(x = "per-guide Jaccard of assigned cell-sets",
       y = "# guides",
       title = "Replogle 800-guide subsample: pairwise assignment agreement",
       subtitle = "one panel per method pair; Jaccard over each guide's assigned cells") +
  theme_bw()

pair_summary <- pair_df |>
  group_by(pair) |>
  summarize(n_defined      = sum(!is.na(jaccard)),
            n_both_empty   = sum(is.na(jaccard)),
            median_jaccard = median(jaccard, na.rm = TRUE),
            mean_jaccard   = mean(jaccard,   na.rm = TRUE),
            .groups = "drop")

print(pj)
print(pair_summary)
```

## Bug-vs-fix agreement vs upper-tail mean (within each offset)

Scatter of per-guide Jaccard against the guide's upper-tail mean
`mean(y | y >= 10)`, for just the two **within-offset** bug-vs-fix comparisons:
`circ+bug vs circ+fix` and `nocirc+bug vs nocirc+fix`. Colour distinguishes the
two. Shows how the additive-vs-multiplicative M-step disagreement scales with the
guide's tail heaviness, with vs without the offset circularity. Depends only on
the `setup` chunk.

```{r plot-replogle-bugfix-jaccard-vs-tail, fig.width=8, fig.height=5.5}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"
REPL_SPECS <- tribble(
  ~method,                      ~run,                          ~offset,                        ~mixture,
  "glmpois_poisbug",            "script_real_data",            "glmpois (circularity)",        "poisbug (additive)",
  "glmpois_poisfix",            "script_real_data",            "glmpois (circularity)",        "poisfix (multiplicative)",
  "glmpoisallgrnafix_poisbug",  "run_grnafix_sceptre",         "glmpoisallgrnafix (no circ.)", "poisbug (additive)",
  "glmpoisallgrnafix_poisfix",  "run_grnafix_bugfix_sceptre",  "glmpoisallgrnafix (no circ.)", "poisfix (multiplicative)"
)

grna_repl        <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names_repl <- rownames(grna_repl)
tail_mean_repl   <- apply(grna_repl, 1, function(y) {
  y <- as.numeric(y); t <- y[y >= 10]; if (length(t) == 0) NA_real_ else mean(t)
})

# per method: guide -> integer vector of assigned cell indices
method_assign <- pmap(REPL_SPECS, function(method, run, offset, mixture) {
  assn <- readRDS(file.path(BM, "outputs", run,
                            paste0("assignment_matrix_script_", method, "_", DS_REPL, ".rds")))
  stopifnot(all(guide_names_repl == rownames(assn)))
  lapply(guide_names_repl, function(g) as.integer(which(assn[g, ]))) |> setNames(guide_names_repl)
})
names(method_assign) <- REPL_SPECS$method

jac <- function(a, b) {
  u <- length(union(a, b)); if (u == 0) NA_real_ else length(intersect(a, b)) / u
}

# just the two within-offset bug-vs-fix comparisons
COMPARISONS <- tribble(
  ~label,               ~mi,                          ~mj,
  "circ: bug vs fix",   "glmpois_poisbug",            "glmpois_poisfix",
  "nocirc: bug vs fix", "glmpoisallgrnafix_poisbug",  "glmpoisallgrnafix_poisfix"
)

bugfix_df <- pmap_dfr(COMPARISONS, function(label, mi, mj) {
  tibble(grna_id   = guide_names_repl,
         tail_mean = tail_mean_repl,
         comparison = label,
         jaccard = vapply(guide_names_repl,
                          function(g) jac(method_assign[[mi]][[g]], method_assign[[mj]][[g]]),
                          numeric(1)))
}) |>
  mutate(comparison = factor(comparison, levels = COMPARISONS$label))

pbf <- ggplot(bugfix_df, aes(tail_mean, jaccard, colour = comparison)) +
  geom_point(alpha = 0.55, size = 1.3) +
  scale_colour_manual(values = c("circ: bug vs fix"   = "firebrick",
                                 "nocirc: bug vs fix" = "steelblue")) +
  labs(x = "upper-tail mean  mean(y | y >= 10)",
       y = "per-guide Jaccard (bug vs fix)", colour = NULL,
       title = "Replogle 800-guide subsample: bug-vs-fix agreement vs upper-tail mean",
       subtitle = "within-offset poisbug-vs-poisfix Jaccard, with vs without circularity") +
  theme_bw() +
  theme(legend.position = "bottom")
print(pbf)
```

ok so this confirms what i've basically known: there is a lot less agreement in the assignments made to the heavy tailed guides for the circular methods, than for the decirc methods.


ok. so what story am i building?
- the biggest failing we observed in sceptre is actually due to the circular feature messing with the offsets, not due to the bug.
- we haven't actually pinned down the exact failings of the bug itself!

can i still make some argument about how the bug guards us against bad offsets? what do i need to do to make this claim?
- this would be a comparison between circ poisbug and poisfix, since that's where the degenerate offsets come in. Both feature overprediction though. so the bug does not actually help us with the biggest observed consequence of bad offsets.

i need to find differences between how poisbug and poisfix behave, separate from the offsets. so this would be patterns that are apparent regardless of the offset choice. and maybe i should also find these things in gasperini, as a way to also do this when the offset consequences are less.

in `plot-metrics`, i observe that the bug generally does better in either case. and this experiment accidentally mitigates the overprediction due to circularity, to some extent, so this is closer to what i want to do now.


on these data, the bug generally leads to better recall, while precision is high across the board. so better recall means lower dec boundary. Can we say that poisfix leads to an excessively large decision boundary?

## Where nocirc poisbug & poisfix disagree: UMI histograms, by poisbug's call

The 16 guides from the offsets notebook (`overprediction-offsets.Rmd`): the 8
heavy-tail overpredictors and the 8 modest-tail controls, selected by the same
reference method (`glmpois_poisbug`). For each guide, take the cells where the two
**nocirc** methods disagree (assigned by exactly one of
`glmpoisallgrnafix_poisbug` / `glmpoisallgrnafix_poisfix`) and histogram their gRNA
UMI counts. Binning is the linear-then-log2 mixed scheme from thesis fig-2a
(exact bins `0..10`, then doubling-width tail bins), drawn as stacked bars on a
`log10(count+1)` y-axis. Bars are conditioned on **poisbug's** call — `poisbug
pert` vs `poisbug NP` — which is enough, since on the disagreement set poisfix is
always the opposite. Top 8 panels = heavy tail, bottom 8 = modest tail. Depends
only on the `setup` chunk.

```{r plot-nocirc-disagree-umi, fig.width=13, fig.height=8}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"

NOCIRC_BUG <- list(method = "glmpoisallgrnafix_poisbug", run = "run_grnafix_sceptre")
NOCIRC_FIX <- list(method = "glmpoisallgrnafix_poisfix", run = "run_grnafix_bugfix_sceptre")
REF_BUG    <- list(method = "glmpois_poisbug",           run = "script_real_data")

read_assn <- function(spec)
  readRDS(file.path(BM, "outputs", spec$run,
                    paste0("assignment_matrix_script_", spec$method, "_", DS_REPL, ".rds")))

grna_repl   <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names <- rownames(grna_repl)
genes       <- vapply(strsplit(guide_names, "_", fixed = TRUE), function(z) z[[2]], character(1))
tail_mean   <- apply(grna_repl, 1, function(y) {
  y <- as.numeric(y); t <- y[y >= 10]; if (length(t) == 0) NA_real_ else mean(t)
})

## --- reproduce the 16-guide selection from overprediction-offsets.Rmd -----
## (reference = glmpois_poisbug: top-quartile tail -> top 8 by #assign;
##  low-quartile (and < 1000) tail -> bottom 8 by #assign)
ref_assn <- read_assn(REF_BUG)
stopifnot(all(guide_names == rownames(ref_assn)))
ref_df   <- tibble(grna_id = guide_names, tail_mean = tail_mean,
                   n_assign = as.numeric(Matrix::rowSums(ref_assn)))
q_tail       <- quantile(ref_df$tail_mean, 0.75, na.rm = TRUE)
overpred_ids <- ref_df |> filter(tail_mean >= q_tail) |>
  slice_max(n_assign, n = 8, with_ties = FALSE) |> pull(grna_id)
q1_tail      <- min(quantile(ref_df$tail_mean, 0.25, na.rm = TRUE), 1000)
control_ids  <- ref_df |> filter(tail_mean <= q1_tail, tail_mean < 1000) |>
  slice_min(n_assign, n = 8, with_ties = FALSE) |> pull(grna_id)
hist_ids <- c(overpred_ids, control_ids)

## --- disagreement cells + poisbug's call ----------------------------------
assn_bug <- read_assn(NOCIRC_BUG); assn_fix <- read_assn(NOCIRC_FIX)
stopifnot(all(guide_names == rownames(assn_bug)), all(guide_names == rownames(assn_fix)))

facet_lab <- setNames(
  sprintf("%s\n(tail %.0f, #assign %d)",
          genes[match(hist_ids, guide_names)],
          tail_mean[match(hist_ids, guide_names)],
          ref_df$n_assign[match(hist_ids, guide_names)]),
  hist_ids)

df_dis <- map_dfr(hist_ids, function(g) {
  b <- as.logical(assn_bug[g, ]); f <- as.logical(assn_fix[g, ])
  disagree <- xor(b, f)
  tibble(umi    = as.integer(grna_repl[g, disagree]),
         status = factor(ifelse(b[disagree], "poisbug pert", "poisbug NP"),
                         levels = c("poisbug NP", "poisbug pert")),
         panel  = facet_lab[[g]])
}) |>
  mutate(panel = factor(panel, levels = facet_lab[hist_ids]))

## --- mixed linear-then-log2 binning (inlined from thesis fig-2a) ----------
logp1 <- function(x, base = 10) log(x + 1, base = base)
make_mixed_bin_info <- function(max_x, k_exact = 10L) {
  upper <- 0:k_exact; width <- 2L
  while (tail(upper, 1L) < max_x) { upper <- c(upper, tail(upper, 1L) + width); width <- width * 2L }
  lower <- c(0L, head(upper, -1L) + 1L)
  label <- ifelse(lower == upper, as.character(upper), paste0(lower, "-", upper))
  data.frame(bin_id = seq_along(label), lower = lower, upper = upper, label = label,
             xmin = seq_along(label) - 0.45, xmax = seq_along(label) + 0.45,
             stringsAsFactors = FALSE)
}
bin_with_info    <- function(x, bi) cut(x, breaks = c(-1, bi$upper), labels = bi$label, right = TRUE)
make_y_breaks_umi <- function(mx) { b <- c(0,1,5,10,50,100,500,1e3,5e3,1e4,5e4,1e5,5e5); b[b <= mx] }
make_x_axis_spec <- function(bi, k_exact, exact_every = 2L, tail_every = 2L) {
  is_exact <- bi$lower == bi$upper & bi$upper <= k_exact
  ei <- which(is_exact); ti <- which(!is_exact); keep <- rep(FALSE, nrow(bi))
  if (length(ei)) { keep[ei[seq(1, length(ei), by = exact_every)]] <- TRUE; keep[max(ei)] <- TRUE }
  if (length(ti)) { keep[ti[seq(1, length(ti), by = tail_every)]] <- TRUE; keep[max(ti)] <- TRUE }
  list(breaks = bi$bin_id[keep], labels = bi$label[keep])
}

k_exact  <- 10L
bin_info <- make_mixed_bin_info(max(df_dis$umi), k_exact = k_exact)
counts <- df_dis |>
  mutate(bin = factor(bin_with_info(umi, bin_info), levels = bin_info$label)) |>
  group_by(panel, bin, status, .drop = FALSE) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(panel, bin, status) |>
  group_by(panel, bin) |>
  mutate(y1 = cumsum(count), y0 = lag(y1, default = 0)) |>
  ungroup() |>
  mutate(bin = as.character(bin)) |>
  left_join(bin_info, by = c("bin" = "label"))

x_axis   <- make_x_axis_spec(bin_info, k_exact)
y_breaks <- make_y_breaks_umi(max(counts$y1))

pdis <- ggplot(counts) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = logp1(y0), ymax = logp1(y1),
                fill = status), colour = "black", linewidth = 0.1) +
  # shared bins (one global bin_info) AND shared x/y scales -> every panel directly comparable
  facet_wrap(~ panel, ncol = 4) +
  scale_fill_manual(values = c("poisbug NP" = "#00BFC4", "poisbug pert" = "#F8766D")) +
  scale_x_continuous(breaks = x_axis$breaks, labels = x_axis$labels,
                     expand = expansion(add = 0.6)) +
  scale_y_continuous(breaks = logp1(y_breaks), labels = scales::label_comma()(y_breaks),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "Nocirc poisbug vs poisfix: gRNA UMIs where they disagree",
       subtitle = "16 selected guides (top 2 rows heavy tail, bottom 2 modest); stacked by poisbug's call",
       x = "gRNA UMI count (linear-then-log2 bins)", y = "number of cells", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
        legend.position = "bottom", strip.text = element_text(size = 7))
print(pdis)
```

## Nocirc poisbug vs poisfix: per-guide 2x2 confusion matrix

For the same 16 guides, the full cross-tabulation of the two **nocirc** methods'
per-cell calls over all cells: #cells both call NP, poisbug P / poisfix NP,
poisbug NP / poisfix P, and both P. One 2x2 panel per guide (poisbug on the
y-axis, poisfix on the x-axis); tiles are filled on a `log10(count+1)` scale and
labelled with the exact count. The both-NP cell dominates (≈all 100k cells), so
the log fill is what makes the off-diagonal disagreement cells legible. Depends
only on the `setup` chunk.

```{r plot-nocirc-confusion, fig.width=13, fig.height=9}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"

NOCIRC_BUG <- list(method = "glmpoisallgrnafix_poisbug", run = "run_grnafix_sceptre")
NOCIRC_FIX <- list(method = "glmpoisallgrnafix_poisfix", run = "run_grnafix_bugfix_sceptre")
REF_BUG    <- list(method = "glmpois_poisbug",           run = "script_real_data")

read_assn <- function(spec)
  readRDS(file.path(BM, "outputs", spec$run,
                    paste0("assignment_matrix_script_", spec$method, "_", DS_REPL, ".rds")))

grna_repl   <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names <- rownames(grna_repl)
genes       <- vapply(strsplit(guide_names, "_", fixed = TRUE), function(z) z[[2]], character(1))
tail_mean   <- apply(grna_repl, 1, function(y) {
  y <- as.numeric(y); t <- y[y >= 10]; if (length(t) == 0) NA_real_ else mean(t)
})

## --- reproduce the 16-guide selection from overprediction-offsets.Rmd -----
ref_assn <- read_assn(REF_BUG)
stopifnot(all(guide_names == rownames(ref_assn)))
ref_df   <- tibble(grna_id = guide_names, tail_mean = tail_mean,
                   n_assign = as.numeric(Matrix::rowSums(ref_assn)))
q_tail       <- quantile(ref_df$tail_mean, 0.75, na.rm = TRUE)
overpred_ids <- ref_df |> filter(tail_mean >= q_tail) |>
  slice_max(n_assign, n = 8, with_ties = FALSE) |> pull(grna_id)
q1_tail      <- min(quantile(ref_df$tail_mean, 0.25, na.rm = TRUE), 1000)
control_ids  <- ref_df |> filter(tail_mean <= q1_tail, tail_mean < 1000) |>
  slice_min(n_assign, n = 8, with_ties = FALSE) |> pull(grna_id)
hist_ids <- c(overpred_ids, control_ids)

## --- per-guide 2x2 counts over all cells ----------------------------------
assn_bug <- read_assn(NOCIRC_BUG); assn_fix <- read_assn(NOCIRC_FIX)
stopifnot(all(guide_names == rownames(assn_bug)), all(guide_names == rownames(assn_fix)))

facet_lab <- setNames(
  sprintf("%s\n(tail %.0f, #assign %d)",
          genes[match(hist_ids, guide_names)],
          tail_mean[match(hist_ids, guide_names)],
          ref_df$n_assign[match(hist_ids, guide_names)]),
  hist_ids)

conf_df <- map_dfr(hist_ids, function(g) {
  b <- as.logical(assn_bug[g, ]); f <- as.logical(assn_fix[g, ])
  tibble(
    poisbug = factor(c("NP", "NP", "P", "P"),  levels = c("NP", "P")),
    poisfix = factor(c("NP", "P", "NP", "P"),  levels = c("NP", "P")),
    count   = c(sum(!b & !f), sum(!b & f), sum(b & !f), sum(b & f)),
    panel   = facet_lab[[g]])
}) |>
  mutate(panel = factor(panel, levels = facet_lab[hist_ids]))

pconf <- ggplot(conf_df, aes(poisfix, poisbug, fill = log10(count + 1))) +
  geom_tile(colour = "grey40") +
  geom_text(aes(label = formatC(count, format = "d", big.mark = ",")), size = 2.8) +
  facet_wrap(~ panel, ncol = 4) +
  scale_fill_viridis_c(name = expression(log[10](count + 1))) +
  labs(x = "poisfix call", y = "poisbug call",
       title = "Nocirc poisbug vs poisfix: per-guide 2x2 assignment confusion",
       subtitle = "counts over all cells; top 2 rows heavy tail, bottom 2 modest. Fill = log10(count+1).") +
  theme_bw() +
  theme(legend.position = "bottom", strip.text = element_text(size = 7),
        panel.grid = element_blank())
print(pconf)
```


ok so we're getting this picture that we do indeed make more assignments with poisbug than poisfix. on the sims, that's a good thing, and we get more recall. do we think these are good assignments to be making in general? that's where the UMI histograms come into play. do those seem like cells that should be pert? some of them definitely, others are borderline. 

this is getting pretty close to a good enough explanation of the difference on Replogle. a plot like this can give a quantitative sense of whether we like the extra cells we're generally including or not. But also there's CRNKL1 which is totally different!

I need to look at more guides.

## Nocirc poisbug vs poisfix: per-guide disagreement counts, all 800 guides

Across all 800 Replogle guides, the joint distribution of the two off-diagonal
confusion counts from the previous plot: for each guide, x = #cells poisbug calls
P while poisfix calls NP (poisbug-only), y = #cells poisbug calls NP while poisfix
calls P (poisfix-only). Points above the dashed `y = x` are guides where poisfix
adds more than poisbug; points below (the bulk, expected) are where poisbug adds
more. Depends only on the `setup` chunk.

```{r plot-nocirc-disagree-hexbin, fig.width=7.5, fig.height=6}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"

NOCIRC_BUG <- list(method = "glmpoisallgrnafix_poisbug", run = "run_grnafix_sceptre")
NOCIRC_FIX <- list(method = "glmpoisallgrnafix_poisfix", run = "run_grnafix_bugfix_sceptre")

read_assn <- function(spec)
  readRDS(file.path(BM, "outputs", spec$run,
                    paste0("assignment_matrix_script_", spec$method, "_", DS_REPL, ".rds")))

assn_bug <- read_assn(NOCIRC_BUG); assn_fix <- read_assn(NOCIRC_FIX)
stopifnot(all(rownames(assn_bug) == rownames(assn_fix)))

# per-guide off-diagonal counts over all cells (vectorised over the 800 rows)
bug_only <- Matrix::rowSums(assn_bug & !assn_fix)   # poisbug P, poisfix NP
fix_only <- Matrix::rowSums(!assn_bug & assn_fix)   # poisbug NP, poisfix P
disagree_df <- tibble(grna_id = rownames(assn_bug),
                      bug_only = as.numeric(bug_only),
                      fix_only = as.numeric(fix_only))

# log1p axes (handles the many zero-disagreement guides); hexes bin in log1p space
log1p_breaks <- c(0, 1, 3, 10, 30, 100, 300, 1000, 3000, 10000)
ph <- ggplot(disagree_df, aes(bug_only, fix_only)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_hex(bins = 40) +
  scale_fill_viridis_c(trans = "log10", name = "# guides") +
  scale_x_continuous(trans = "log1p", breaks = log1p_breaks,
                     labels = scales::label_comma()(log1p_breaks)) +
  scale_y_continuous(trans = "log1p", breaks = log1p_breaks,
                     labels = scales::label_comma()(log1p_breaks)) +
  coord_fixed() +
  labs(x = "# cells poisbug P & poisfix NP  (poisbug-only)",
       y = "# cells poisbug NP & poisfix P  (poisfix-only)",
       title = "Nocirc poisbug vs poisfix: per-guide disagreement counts",
       subtitle = "all 800 Replogle guides; log1p axes, dashed = y = x") +
  theme_bw()
print(ph)
```

## The poisfix-heavy guides: UMI histograms (fix_only > 10)

The handful of guides from the hexbin above where poisfix adds many more cells
than poisbug (`fix_only > 10`). One panel per such guide, showing the full gRNA
UMI-count distribution over all cells with the same linear-then-log2 mixed binning
as fig-2a, shared bins and shared x/y axes across panels. Each cell is coloured by
**which of the two nocirc methods assigns it** (neither / poisbug only / poisfix
only / both), stacked per bin. Depends only on the `setup` chunk.

```{r plot-poisfix-heavy-umi, fig.width=12, fig.height=5}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"

NOCIRC_BUG <- list(method = "glmpoisallgrnafix_poisbug", run = "run_grnafix_sceptre")
NOCIRC_FIX <- list(method = "glmpoisallgrnafix_poisfix", run = "run_grnafix_bugfix_sceptre")

read_assn <- function(spec)
  readRDS(file.path(BM, "outputs", spec$run,
                    paste0("assignment_matrix_script_", spec$method, "_", DS_REPL, ".rds")))

grna_repl   <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names <- rownames(grna_repl)
genes       <- vapply(strsplit(guide_names, "_", fixed = TRUE), function(z) z[[2]], character(1))

## --- pick the poisfix-heavy guides: fix_only > 10 -------------------------
assn_bug <- read_assn(NOCIRC_BUG); assn_fix <- read_assn(NOCIRC_FIX)
stopifnot(all(guide_names == rownames(assn_bug)), all(guide_names == rownames(assn_fix)))
bug_only <- as.numeric(Matrix::rowSums(assn_bug & !assn_fix))
fix_only <- as.numeric(Matrix::rowSums(!assn_bug & assn_fix))
sel <- guide_names[fix_only > 10]
message(length(sel), " guides with fix_only > 10: ", paste(genes[match(sel, guide_names)], collapse = ", "))

facet_lab <- setNames(
  sprintf("%s\n(poisfix-only %d, poisbug-only %d)",
          genes[match(sel, guide_names)],
          as.integer(fix_only[match(sel, guide_names)]),
          as.integer(bug_only[match(sel, guide_names)])),
  sel)

## umi over ALL cells for each selected guide, tagged by which method assigns it
STATUS_LEVELS <- c("neither", "poisbug only", "poisfix only", "both")
df_umi <- map_dfr(sel, function(g) {
  b <- as.logical(assn_bug[g, ]); f <- as.logical(assn_fix[g, ])
  status <- ifelse(b & f, "both",
            ifelse(b & !f, "poisbug only",
            ifelse(!b & f, "poisfix only", "neither")))
  tibble(umi    = as.integer(grna_repl[g, ]),
         status = factor(status, levels = STATUS_LEVELS),
         panel  = facet_lab[[g]])
}) |>
  mutate(panel = factor(panel, levels = facet_lab[sel]))

## --- mixed linear-then-log2 binning (inlined from thesis fig-2a) ----------
logp1 <- function(x, base = 10) log(x + 1, base = base)
make_mixed_bin_info <- function(max_x, k_exact = 10L) {
  upper <- 0:k_exact; width <- 2L
  while (tail(upper, 1L) < max_x) { upper <- c(upper, tail(upper, 1L) + width); width <- width * 2L }
  lower <- c(0L, head(upper, -1L) + 1L)
  label <- ifelse(lower == upper, as.character(upper), paste0(lower, "-", upper))
  data.frame(bin_id = seq_along(label), lower = lower, upper = upper, label = label,
             xmin = seq_along(label) - 0.45, xmax = seq_along(label) + 0.45,
             stringsAsFactors = FALSE)
}
bin_with_info    <- function(x, bi) cut(x, breaks = c(-1, bi$upper), labels = bi$label, right = TRUE)
make_y_breaks_umi <- function(mx) { b <- c(0,1,5,10,50,100,500,1e3,5e3,1e4,5e4,1e5,5e5); b[b <= mx] }
make_x_axis_spec <- function(bi, k_exact, exact_every = 2L, tail_every = 2L) {
  is_exact <- bi$lower == bi$upper & bi$upper <= k_exact
  ei <- which(is_exact); ti <- which(!is_exact); keep <- rep(FALSE, nrow(bi))
  if (length(ei)) { keep[ei[seq(1, length(ei), by = exact_every)]] <- TRUE; keep[max(ei)] <- TRUE }
  if (length(ti)) { keep[ti[seq(1, length(ti), by = tail_every)]] <- TRUE; keep[max(ti)] <- TRUE }
  list(breaks = bi$bin_id[keep], labels = bi$label[keep])
}

k_exact  <- 10L
bin_info <- make_mixed_bin_info(max(df_umi$umi), k_exact = k_exact)
counts <- df_umi |>
  mutate(bin = factor(bin_with_info(umi, bin_info), levels = bin_info$label)) |>
  group_by(panel, bin, status, .drop = FALSE) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(panel, bin, status) |>
  group_by(panel, bin) |>
  mutate(y1 = cumsum(count), y0 = lag(y1, default = 0)) |>
  ungroup() |>
  mutate(bin = as.character(bin)) |>
  left_join(bin_info, by = c("bin" = "label"))

x_axis   <- make_x_axis_spec(bin_info, k_exact)
y_breaks <- make_y_breaks_umi(max(counts$y1))

pfh <- ggplot(counts) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = logp1(y0), ymax = logp1(y1),
                fill = status), colour = "black", linewidth = 0.1) +
  facet_wrap(~ panel, ncol = 4) +   # shared bins + shared x/y across panels
  scale_fill_manual(values = c("neither" = "grey85", "poisbug only" = "#F8766D",
                               "poisfix only" = "#00BFC4", "both" = "#9467BD")) +
  scale_x_continuous(breaks = x_axis$breaks, labels = x_axis$labels,
                     expand = expansion(add = 0.6)) +
  scale_y_continuous(breaks = logp1(y_breaks), labels = scales::label_comma()(y_breaks),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title = "poisfix-heavy guides (fix_only > 10): gRNA UMI distributions",
       subtitle = "all cells, stacked by which nocirc method assigns the cell; linear-then-log2 bins, shared axes",
       x = "gRNA UMI count (linear-then-log2 bins)", y = "number of cells", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
        legend.position = "bottom", strip.text = element_text(size = 7))
print(pfh)
```

## Guides whose nocirc offset GLM diverged (CRNKL1-type failures)

Flag the guides where the `glmpoisallgrnafix` offset Poisson-GLM did **not**
converge. The failure mode (see CRNKL1): a guide that is spiky **and** dominates
its own cells' gRNA totals, so leave-one-out strips `other_grna_n_umis ≈ 0` off
exactly the high-count cells, forcing the fit onto the near-collinear response
covariates until `mu0` explodes. Diagnostics per flagged guide: `max_y`, the
gRNA total and leftover `other_umis` at that top cell (how self-dominated it is),
the offset deviance and largest |coef|, plus whether each mixture converged and
how many cells it assigned. The offset is shared by poisbug/poisfix, so
`offset_converged` is identical across them. Table printed at the end. Depends
only on the `setup` chunk.

```{r flag-offset-diverged}
DS_REPL <- "replogle-rd7_assign_nguides=800_ncells=100k"
NOCIRC_BUG <- list(method = "glmpoisallgrnafix_poisbug", run = "run_grnafix_sceptre")
NOCIRC_FIX <- list(method = "glmpoisallgrnafix_poisfix", run = "run_grnafix_bugfix_sceptre")

read_extras <- function(spec)
  readRDS(file.path(BM, "outputs", spec$run,
                    paste0("assignment_extras_script_", spec$method, "_", DS_REPL, ".rds")))

grna_repl   <- readRDS(file.path(IN, DS_REPL, "sceptre", "grna_matrix.rds"))
guide_names <- rownames(grna_repl)
genes       <- vapply(strsplit(guide_names, "_", fixed = TRUE), function(z) z[[2]], character(1))
gtot_umis   <- Matrix::colSums(grna_repl)

# guide-level: at each guide's max-count cell, how self-dominated is it?
top_info <- map_dfr(guide_names, function(g) {
  y <- as.numeric(grna_repl[g, ]); j <- which.max(y)
  tibble(grna_id = g, max_y = y[j], gtot_at_top = gtot_umis[j],
         other_umis_at_top = gtot_umis[j] - y[j],
         self_frac_at_top = if (gtot_umis[j] > 0) y[j] / gtot_umis[j] else NA_real_)
})

# per-guide offset + EM diagnostics for one method
diag_method <- function(spec) {
  ex <- read_extras(spec); pg <- ex$per_guide
  tibble(
    grna_id          = names(pg),
    offset_converged = vapply(pg, function(r) isTRUE(r$offset_model_summary$converged), logical(1)),
    offset_deviance  = vapply(pg, function(r) as.numeric(r$offset_model_summary$deviance %||% NA), numeric(1)),
    max_abs_coef     = vapply(pg, function(r) max(abs(r$offset_model_summary$coefficients)), numeric(1)),
    em_converged     = vapply(pg, function(r) isTRUE(r$em_converged), logical(1)),
    n_assigned       = vapply(pg, function(r) as.integer(r$n_assigned), integer(1)))
}

d_bug <- diag_method(NOCIRC_BUG)
d_fix <- diag_method(NOCIRC_FIX)

flagged <- d_bug |>
  transmute(grna_id, gene = genes[match(grna_id, guide_names)],
            offset_converged, offset_deviance, max_abs_coef,
            em_conv_bug = em_converged, n_assn_bug = n_assigned) |>
  left_join(transmute(d_fix, grna_id, em_conv_fix = em_converged, n_assn_fix = n_assigned),
            by = "grna_id") |>
  left_join(top_info, by = "grna_id") |>
  filter(!offset_converged) |>
  arrange(desc(max_y)) |>
  select(gene, max_y, gtot_at_top, other_umis_at_top, self_frac_at_top,
         offset_deviance, max_abs_coef, em_conv_bug, n_assn_bug, em_conv_fix, n_assn_fix)

message(nrow(flagged), " of ", nrow(d_bug), " guides have a non-converged nocirc offset GLM")
print(flagged, n = nrow(flagged))
```


