---
title: "additive-model"
output:
  pdf_document: default
  html_document: default
date: "2026-06-18"
---


KEY RESULTS

I have found a model that feels more principled and basically agrees with SCEPTRE, except it also does better in the situations where SCEPTRE loses precision with heavy tailed guides.

So at this point I do have a model that uniformly improves upon SCEPTRE in all my simulation settings, as well as real data!

It still has a rather questionable decision in it, to do the $\gamma$ vs $e^{\gamma}$ difference, but I will give some mathematical justification. And it works...


# 1. The problem

SCEPTRE uses $e^{o_i} + \gamma$ as the perturbed mean, when it should be $e^{o_i+\gamma}$, and yet performance mysteriously is far better than if we actually did the correct model.

## 1.2 Using a summed model to replicate SCEPTRE's likelihood ratio

We can get this form for the mean via
$$
Y_i \sim \text{Pois}(e^{o_i}) + \text{Ber}(\pi) \text{Pois}(e^{o_i} + e^{\delta}).
$$

This is a special case of the Cellbender model that I discussed on Monday, with no endogenous noise in the NP distribution, and Poisson instead of NB for the perturbed component. So this is a very simple form of a reasonable model that appears elsewhere in the literature.

We then get a likelihood ratio of
$$
\frac{\text{Pois}(y | e^{o} + e^\delta)}{\text{Pois}(y | e^o)}
$$
which is the same additive form that SCEPTRE accidentally uses.

So that's all fine and good... except the MLE of $e^\delta$ will be very large, since it's on the response scale, not the data scale. In practice, this model has all the same problems as the bug-fixed offset model, where the estimate of the mean is so large that we assign far too many values to NP. 

## 1.3 Statistics jail

Now it's time to get unprincipled. In order to replicate SCEPTRE, we need to find a justification to do the following:
$$
\text{use} \quad \frac{\text{Pois}(y | e^{o} + \delta)}{\text{Pois}(y | e^o)} \quad\text{instead of} \quad \frac{\text{Pois}(y | e^{o} + e^\delta)}{\text{Pois}(y | e^o)}
$$

In words, we need a reason to use the log of the estimated mean in our decision boundary instead of just using the mean.


I think there actually is a decent mathematical argument for using this. I'm still thinking through it, but basically it comes down to how $e^\delta$ is a very poor fit for a lot of our data points, and especially when things are heavy tailed. Since we just need a good decision boundary, using $\delta$ is actually pretty sane.


For cells with a very low NP mean $e^{o_i}$, we are effectively comparing to the constant distribution $\text{Pois}(\delta)$. Comparing to $\text{Pois}(e^{\delta})$ performs very poorly, since our classification boundary is typically $\leq 20$, so these points are nowhere close to $e^\delta$.

And for cells with a large $e^{o_i}$, it still performs fine. 

I'm going to do more mathematical analysis once we discuss and confirm that this direction is useful. 


# 2. Simulation results

In this section, I compare 6 models: SCEPTRE, crispat, and 4 custom ones. For the custom ones, the name is of the form `<x>_<y>`, with `<x>` denoting the offset method, and `<y>` denoting the mixture method.

`glmpois_poissumlog` is the additive model that uses $\delta$ instead of $e^{\delta}$, and is the proposed replacement. The offsets are exactly the same as with SCEPTRE. Only the mixture is different.

The other 3 methods all use thresholding to fit the GLM only on the cells with $y_i \leq 1000$. They vary in their mixure:

- `threshglmpois1000_poisbug` is these robustified offsets with the current bugged Pois-Pois mixture used by SCEPTRE

- `threshglmpois1000_poissumlog` is the additive model using the log mean, except also with robustified offsets

- `threshglmpois1000_pois0nb` is the robustified offsets with a Pois-NB mixture, and represents the best method I found that actually uses $exp(o_i + \gamma)$ as the perturbed mean.




```{r, echo=FALSE, cache=TRUE}
rm(list=ls())
library(tidyverse)


source("~/.Rprofile")
source("~/katsevich-lab/code/sceptre3-project-v2/benchmarking/guide-assignment/assignment-analysis/script_runs_analysis_functions.R")
source("~/katsevich-lab/code/sceptre3-project-v2/benchmarking/guide-assignment/sceptre-nb/nb-bench_v3-helpers.R")
load_assns_crispat <- function(path_to_assns, all_guides, cell_name_to_idx) {
  crisp_assign_raw <- read.csv(path_to_assns) |>
     left_join(cell_name_to_idx, by=join_by(cell_id == cell_name)) |>
     dplyr::select(-cell_id)
      # mutate(cell_idx = gsub("cell_idx_", "", cell_id) |> as.integer())
  unordered_cell_list <- split(crisp_assign_raw$cell_idx, crisp_assign_raw$grna_id)
  
  guide_to_assns = vector("list", length(all_guides)) |>
    setNames(all_guides)
  
  guide_to_assns[names(unordered_cell_list)] <- unordered_cell_list
  guide_to_assns
}

load_run <- function(run_name, dataset_name) {
  base_path = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/outputs"
  )
  input_data_path = file.path(
    .get_config_path("LOCAL_BENCHMARKING_DIR"),
    "guide_assignment/input_data",
    dataset_name
  )
  if(file.exists(true_perts_fp <- file.path(input_data_path, "true_pert_matrix.rds"))) {
    true_perts = read_rds(true_perts_fp)
  }
  grna_matrix = file.path(input_data_path, "sceptre/grna_matrix.rds") |>
    read_rds()
  
  
  data_path = file.path(base_path, run_name)
  
  cov_df <- read_csv(file.path(input_data_path, "sceptre/cell_covariates.csv"), show_col_types = FALSE) |>
    mutate(
      grna_n_nonzero_subset = Matrix::colSums(grna_matrix > 0),
      grna_n_umis_subset = Matrix::colSums(grna_matrix)
    )
  
  
  
  
  fnames = dir(data_path)
  
  # 1. load assignment matrices
  assn_mat_fnames = fnames[grepl("^assignment_matrix", fnames) & grepl(dataset_name, fnames, fixed = TRUE)]
  all_assns = list()
  guide_names = rownames(grna_matrix)
  for(i in seq_along(assn_mat_fnames)) {
    curr_fname = assn_mat_fnames[i]
    script_name = gsub("^assignment_matrix_", "", curr_fname) |>
      gsub(pattern = "^script_", replacement="", fixed=TRUE) |>
      gsub(pattern =paste0("_", dataset_name, ".rds"), replacement="", fixed=TRUE)
    
    mat = read_rds(file.path(data_path, curr_fname))
    
    curr_assns = lapply(guide_names, function(guide) which(mat[guide,]) |> as.integer()) |>
      setNames(guide_names) |>
      list() |>
      setNames(script_name)
    
    all_assns = c(all_assns, curr_assns)
    
  }
  
  # 2. load extras from script runs
  extras_fnames = fnames[grepl("^assignment_extras", fnames) & grepl(dataset_name, fnames, fixed = TRUE)]
  extras_all <- lapply(extras_fnames, function(fname) read_rds(file.path(data_path, fname)))
  names(extras_all) <- gsub("assignment_extras_script_", "", extras_fnames) |>
    gsub(pattern = paste0("_", dataset_name, ".rds"), replacement="", fixed=TRUE)
  
  
  # 3. load other methods if there
  
  # i need this to change cell name assignments to cell index assignments
  
  cell_name_to_idx <- data.frame(
    cell_name = colnames(grna_matrix),
    cell_idx = 1:ncol(grna_matrix)
  )
  
  if(length(fname_crispat <- fnames[grepl("crispat", fnames) & grepl(dataset_name, fnames)]) > 0) {
    all_assns$crispat <- load_assns_crispat(file.path(data_path, fname_crispat), guide_names, cell_name_to_idx)
  }
  
  out = list(
    run_name = run_name, dataset_name = dataset_name,
    input_data_path = input_data_path,
    grna_matrix = grna_matrix,
    cov_df = cov_df,
    all_assns = all_assns,
    all_extras = extras_all
  )
  if(exists("true_perts")) {
    out$true_perts = true_perts
    
    out$true_assns = lapply(guide_names, function(guide) {
      which(true_perts[guide,] == 1 & grna_matrix[guide,] > 0) |> as.integer()
      # which(sim_results$true_perts[guide,] == 1 ) |> as.integer()
    }) |>
      setNames(guide_names)
    
    
  }
  out
}


sim_pois0nb_2np_3p <- load_run(
  run_name =  "pois0nb_sims_sum_2np_3p",
  dataset_name = "sims_sum_2np_3p"
)


sim_pois0nb_1np_3p_disp <-  load_run(
  run_name =  "pois0nb_sims_sum_1np_3p_disp",
  dataset_name = "sims_sum_1np_3p_disp"
)



sim_pois0nb_repeat_old <-  load_run(
  run_name =  "pois0nb_sims_repeat_old",
  dataset_name = "sims_sum_repeat_old"
)

methods_to_keep <- c(
      "crispat", "sceptre", "script_threshglmpois1000_pois0nb", "script_threshglmpois1000_poissumlog",
      "script_threshglmpois1000_poisbug", "script_glmpois_poissumlog"
    )

# guide_names = rownames(sim_results$grna_matrix)

# sim_pois0nb_2np_3p$all_assns$thresh_10 <- lapply(rownames(sim_pois0nb_2np_3p$grna_matrix), function(guide) {
#   which(sim_pois0nb_2np_3p$grna_matrix[guide,] > 10)
# }) |>
#   setNames(rownames(sim_pois0nb_2np_3p$grna_matrix))
# sim_pois0nb_1np_3p_disp$all_assns$thresh_10 <- lapply(rownames(sim_pois0nb_1np_3p_disp$grna_matrix), function(guide) {
#   which(sim_pois0nb_1np_3p_disp$grna_matrix[guide,] > 10)
# }) |>
#   setNames(rownames(sim_pois0nb_1np_3p_disp$grna_matrix))

```



Reminder of my current simulation settings, copied from my report on Monday: inspired by Cellbender model, I simulate guides via

$$
Y_i = \text{Pois}(L_i \mu_{\text{exo}}) + \text{Ber}(\pi_{\text{endo}}) \text{Pois}(L_i \mu_{\text{endo}}) + \text{Ber}(\pi_{\text{pert}}) \text{NB}(L_i \mu_{\text{pert}}, \theta_{\text{pert}}).
$$


The idea: 

- the perturbed distribution is the same: NegBin with cell-specific library size rescaling factors 

- but now the NP is a sum of two processes:
  
  - exogeneous noise, representing random stray bits getting in there. $\mu_{\text{exo}}$ is small
  
  - endogenous noise, from droplet contamination, chimeras, etc, giving rare but larger values. $\pi_{\text{endo}}$ is very small but $\mu_{\text{endo}}$ is larger.



I am able to get satisfying looking distributions from this, and in particular I can now simulate distributions where precision is the hard part. Before, with such light tails, precision was often trivial because the NPs were so concentrated near zero.


```{r, echo=FALSE, fig.cap="Example simulated guides from the three simulated datasets under consideration."}
library(tidyverse)
grna_odm_repl = ondisc::initialize_odm_from_backing_file(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7/sceptre-pipeline/grna.odm"
)

repl_guide_stats = read_csv(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/replogle-rd7/sceptre-pipeline/grna_summary_stats.csv",
  show_col_types = FALSE
)

grna_odm_gasp = ondisc::initialize_odm_from_backing_file(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/gasperini/sceptre-pipeline/grna.odm"
)

gasp_guide_stats = read_csv(
  "/home/josep/data/projects/sceptre3/benchmarking/guide_assignment/input_data/gasperini/sceptre-pipeline/grna_summary_stats.csv",
  show_col_types = FALSE
)


get_guides <- function(guide_stats, n, min_idx = 200) {
   guide_stats |>
  arrange(grna_total) |>
  pull(grna) %>%
  `[`(seq(min_idx, nrow(guide_stats) - 50, length = n) |> round())
}

# let's pick a sequence of guides in terms of expression
repl_guides <- get_guides(repl_guide_stats, 4)

gasp_guides <-  get_guides(gasp_guide_stats, 4) 



plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    NP_lowvar_P_low = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPlowvar_P1_1",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPlowvar_P1_1",]
    ),
      NP_lowvar_P_high = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPlowvar_P3_2",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPlowvar_P3_2",]
    ),
    NP_highvar_P_low = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPhighvar_P1_1",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPhighvar_P1_1",]
    ),
      NP_highvar_P_high = list(
      umis_sim = sim_pois0nb_2np_3p$grna_matrix["sim_NPhighvar_P3_1",],
  is_pert = sim_pois0nb_2np_3p$true_perts["sim_NPhighvar_P3_1",]
    )
  ),
   title = "Real vs simulated: sims_sum_2np_3p"
)

plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    NP_highvar_P_lowdisp = list(
      umis_sim = sim_pois0nb_1np_3p_disp$grna_matrix["sim_NPhighvar_Plowdisp_1",],
  is_pert = sim_pois0nb_1np_3p_disp$true_perts["sim_NPhighvar_Plowdisp_1",]
    ),
      NP_highvar_P_meddisp = list(
      umis_sim = sim_pois0nb_1np_3p_disp$grna_matrix["sim_NPhighvar_Pmeddisp_1",],
  is_pert = sim_pois0nb_1np_3p_disp$true_perts["sim_NPhighvar_Pmeddisp_1",]
    ),
    NP_highvar_P_highdisp = list(
      umis_sim = sim_pois0nb_1np_3p_disp$grna_matrix["sim_NPhighvar_Phighdisp_1",],
  is_pert = sim_pois0nb_1np_3p_disp$true_perts["sim_NPhighvar_Phighdisp_1",]
    )
  ),
   title = "Real vs simulated: sims_sum_1np_3p_disp"
)



plot_umi_histogram_real_vs_sim_list(
  umis_real = grna_odm_repl[tail(repl_guides, 1),],
  sim_list = list(
    NP_pois_P_small = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Psmall_1",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Psmall_1",]
    ),
      NP_pois_P_med = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Pmed_74",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Pmed_74",]
    ),
    NP_pois_P_large = list(
      umis_sim = sim_pois0nb_repeat_old$grna_matrix["sim_NPpois_Plarge_26",],
  is_pert = sim_pois0nb_repeat_old$true_perts["sim_NPpois_Plarge_26",]
    )
  ),
   title = "Real vs simulated: sims_repeat_old"
)
```



```{r, echo=FALSE, fig.cap="Comparing sceptre, crispat, and the new methods on data simulated with varying NP tails and perturbed means / overdispersions. SCEPTRE and `gmpois_poissumlog` perform very similarly."}
# for "pois0nb_sims_sum_2np_3p" ~~~~~~~~~~~~~~~~~~~~~~

P_to_mean <- tibble(
  p = c("P1", "P2", "P3"),
  mu_pert = c(500, 750, 1000)
)

guide_meta <- tibble(
  grna_id = rownames(sim_pois0nb_2np_3p$grna_matrix),
  np = extract_part(grna_id, 2),
  p = extract_part(grna_id, 3)
) %>%
  left_join(P_to_mean, by = "p")


## get metrics and avg ~~~~~~~~~~~~~~~~~~~~~~~~~~~~

guide_metric_tbl <- make_guide_metric_tbl(sim_pois0nb_2np_3p) |>
  left_join(guide_meta, by = "grna_id")


avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "np", "p", "mu_pert")
)


plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )

stopifnot(methods_to_keep %in% plot_df$method_name)

## 1. comparing the ones we're most interested in


plot_df |>
  filter(method_name %in% methods_to_keep) |>
  mutate(
    method_name = gsub("^script_", "", method_name)
  ) |>
make_avg_metrics_plot() +
  ggtitle("Metrics for key methods on sims_sum_2np_3p")



## 2. different thresholds

# plot_df |>
#   filter(
#     grepl("threshglmpois", method_name),
#     grepl("pois0nb", method_name)
#   ) |>
# make_avg_metrics_plot() +
#   ggtitle("Metrics for key methods")



```


```{r, echo=FALSE, fig.cap = "Simulations with much bigger and more overdispersed values. We see very similar performance between SCEPTRE and `glmpois_poissumlog`."}
# for "sims_sum_1np_3p_disp" ~~~~~~~~~~~~~~~~~~~~~~

P_to_mean <- tibble(
  p = c("Phighdisp", "Pmeddisp", "Plowdisp"),
  mu_pert = c(10001000, 1001000, 101000)
)

guide_meta <- tibble(
  grna_id = rownames(sim_pois0nb_1np_3p_disp$grna_matrix),
  np = extract_part(grna_id, 2),
  p = extract_part(grna_id, 3)
) %>%
  left_join(P_to_mean, by = "p")


## get metrics and avg ~~~~~~~~~~~~~~~~~~~~~~~~~~~~

guide_metric_tbl <- make_guide_metric_tbl(sim_pois0nb_1np_3p_disp) |>
  left_join(guide_meta, by = "grna_id")


avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "np", "p", "mu_pert")
)

plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )
stopifnot(c(setdiff(methods_to_keep, "sceptre"), "script_glmpois_poisbug")
          %in% plot_df$method_name)

plot_df |>
  filter(method_name %in% c(methods_to_keep, "script_glmpois_poisbug")) |>
  mutate(method_name = ifelse(method_name == "script_glmpois_poisbug", "sceptre", method_name)) |>
    mutate(
    method_name = gsub("^script_", "", method_name)
  ) |>
make_avg_metrics_plot(x_breaks = c(1e3,1e4,1e5,1e6,1e7)) +
  ggtitle("Metrics for more overdispersed data")

```



```{r, echo=FALSE, fig.cap="Repeating old simulations. This is a HIGHLY overdispered pert and very low mean non-pert, so basically every non-pert is <= 3, while the perts are all over the place. SCEPTRE breaks down here but the more principled estimation in glm_poissumlog keeps things ok."}
# for "sims_repeat_old" ~~~~~~~~~~~~~~~~~~~~~~

P_to_mean <- tibble(
  p = c("Plarge", "Pmed", "Psmall"),
  mu_pert = c(970, 970/4, 970/8)
)

guide_meta <- tibble(
  grna_id = rownames(sim_pois0nb_repeat_old$grna_matrix),
  np = extract_part(grna_id, 2),
  p = extract_part(grna_id, 3)
) %>%
  left_join(P_to_mean, by = "p")


## get metrics and avg ~~~~~~~~~~~~~~~~~~~~~~~~~~~~

guide_metric_tbl <- make_guide_metric_tbl(sim_pois0nb_repeat_old) |>
  left_join(guide_meta, by = "grna_id")


avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "np", "p", "mu_pert")
)


plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  )
stopifnot(methods_to_keep %in% plot_df$method_name)


## 1. comparing the ones we're most interested in


plot_df |>
  # filter(! method_name %in% c("script_trimglmpois_poisfix", "script_glmpois_poisbug")) |>
  filter(
    method_name %in% methods_to_keep
  ) |>
      mutate(
    method_name = gsub("^script_", "", method_name)
  ) |>
make_avg_metrics_plot(x_breaks = c(1e3,1e4,1e5,1e6,1e7)) +
  ggtitle("Metrics for old simulations")




```









# 3. Data results


I want to see how much my new methods change vs SCEPTRE. This plot computes precision and recall against SCEPTRE on Replogle.

```{r, echo=FALSE}
repl_results <-  load_run(
  run_name =  "script_real_data",
  dataset_name = "replogle-rd7_assign_nguides=800_ncells=100k"
)

gasp_results <-  load_run(
  run_name =  "script_real_data",
  dataset_name = "gasperini_assign_nguides=800_ncells=100k"
)
```




```{r, echo=FALSE, fig.cap="This plot bins Replogle guides based on `mean(guide[gudie >= 10])`, and computes the average metric in that group. SCEPTRE is the baseline. We see that our new `glm_poissumlog` is very close, until the heaviest tailed guides. The low recall of every method here vs SCEPTRE shows that SCEPTRE makes more positive predictions than they all do."}

methods_to_use <- c("script_glmpois_poisbug", "crispat", "script_glmpois_poissumlog", "script_threshglmpois1000_poissumlog", "script_threshglmpois1000_poisbug")
stopifnot(all(methods_to_use %in% names(repl_results$all_assns)))

guide_metric_tbl = make_guide_metric_tbl_vs_ref(
  all_assns = repl_results$all_assns[methods_to_use],
  grna_matrix = repl_results$grna_matrix,
    ref_method = "script_glmpois_poisbug"
)

avg_metrics_tbl <- summarize_metric_tbl(
  guide_metric_tbl,
  group_vars = c("method_name", "tail_mean_group")
)

make_avg_metrics_plot_ref <- function(df, x_breaks=NULL) {
  if(is.null(x_breaks)) x_breaks <- c(200,400,800,1600)
  ggplot(
    df,
    aes(
      x = tail_mean_group,
      y = mean,
      group = method_name,
      color = method_name,
      shape = method_name
      # linetype = method_name
    )
  ) +
    geom_line() +
    geom_errorbar(
      aes(ymin = ymin, ymax = ymax),
      width = 0.04,
      alpha = 0.7
    ) +
    geom_point(size = 2) +
    facet_grid( ~ metric) +
    # scale_x_continuous(
    #   trans = "log2",
    #   breaks = x_breaks
    # ) +
    scale_shape_manual(
      values = c(16, 17, 15, 3, 7, 8, 4, 5, 6, 9,10,11)
    ) +
    labs(
      x = "Mean UMI count of UMIS >= 10",
      y = "Metric mean",
      color = "Method",
      shape = "Method"
      # linetype = "Method"
    ) +
    theme_bw() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)
    )
}

plot_df <- avg_metrics_tbl %>%
  mutate(
    ci = 2 * sd / sqrt(n_nonmissing),
    ymin = pmax(0, mean - ci),
    ymax = pmin(1, mean + ci)
  ) |>
  mutate(
    method_name = gsub("^script_", "", method_name)
  )

plot_df |>
  make_avg_metrics_plot_ref() +
  ggtitle("Replogle: comparing methods to SCEPTRE") +
    guides(
    color = guide_legend(nrow = 2),
    shape = guide_legend(nrow = 2)
  )
```


```{r, echo=FALSE, fig.cap = "This plot shows that `glmpois_poissumlog` has some of the breakdown that SCEPTRE exhibits, in terms of making more assignments for heavier tailed guides, but nowhere nearly as bad."}
 new_labs <- c(
  crispat = "crispat",
  script_glmpois_poisbug = "sceptre",
  script_threshglmpois1000_poissumlog = "threshglmpois1000_poissumlog",
  script_threshglmpois1000_poisbug = "threshglmpois1000_poisbug",
  script_glmpois_poissumlog = "glmpois_poissumlog"
)

plot_avg_n_assignments_by_group(
  all_assns = repl_results$all_assns[methods_to_use],
  grna_matrix = repl_results$grna_matrix
) +
  scale_y_log10() +
  ggtitle(
    "Replogle: number of assignments per guide tail mean group"
  ) +
  scale_color_discrete(
    name = "Method",
    labels = new_labs
  ) +
  scale_shape_discrete(
    name = "Method",
    labels = new_labs
  ) +
  guides(
    color = guide_legend(nrow = 2),
    shape = guide_legend(nrow = 2)
  )
```



All methods perform very similarly on Gasperini so those plots are not interesting. 
