---
title: "Direct-capture barnyard ambient overdispersion is doublet-driven"
subtitle: "Reproducing CLEANSER's Figure 3B from raw counts, exposing the doublets, and re-drawing it Poisson"
author: "sceptre3 gRNA-assignment investigation"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 10
    fig-height: 5.5
    df-print: kable
execute:
  warning: false
  message: false
  cache: false
---

```{r setup}
#| code-fold: true
suppressPackageStartupMessages({
  library(Matrix)
  library(ggplot2)
  library(knitr)
})
theme_set(theme_bw(base_size = 12))
# All data paths are RELATIVE to this .qmd's folder, so it renders in place.
REPRO <- "external/repro_work"
# small helper for inline numbers
fmt <- function(x, d = 2) formatC(x, format = "f", digits = d, big.mark = ",")
```

# What this note settles

CLEANSER (Liu et al. 2025) models per-guide ambient gRNA counts with a
**Negative Binomial (NB)** and shows, in its **Figure 3B**, that in the
**direct-capture** barnyard experiment the ground-truth ambient is
strongly **overdispersed** (variance $\gg$ mean), whereas in the **CROP-seq**
experiment it sits on the Poisson line. fishash / geomux instead assume a
**Poisson / hypergeometric** ambient null. These look contradictory.

This note shows, **entirely from the raw barnyard counts and with every number
produced by executable code**, that:

1. **CLEANSER's Fig 3B reproduces** — direct-capture per-guide ambient really is
   overdispersed (median var/mean $\approx 10$); CROP-seq really is Poisson
   (var/mean $\approx 1$).
2. **The overdispersion is doublet-driven.** A right tail of cells carry a whole
   wrong-species cell's worth of gRNA UMIs (co-encapsulated doublets), and the
   **top 1% of cells account for $\approx 98.5\%$ of the excess variance.**
3. **Removing gRNA-channel doublets collapses Fig 3B onto the Poisson line** — the
   per-guide variance-vs-mean cloud pulls back onto $\mathrm{var}=\mathrm{mean}$, and
   a **direct, bootstrap-calibrated goodness-of-fit test** (§Part 4) cannot distinguish
   the de-doubleted direct-capture ambient from $\mathrm{Poisson}(a_g\,d_c)$: observed
   Pearson $\phi = 0.993$ sits squarely inside the Poisson null band $[0.956, 1.018]$
   ($p\approx 0.25$). The rank-1 mean $\lambda_{gc}=R_g C_c/T$ captures the structure;
   what little pooled var/mean excess remains is **1–5 cut-boundary doublet entries**,
   not smooth soup overdispersion. The authors' own GEX-purity knob does the same
   thing, more weakly.

**Both papers are correct on their own terms.** The only difference is whether
doublets are removed from the gRNA channel *before* dispersion is measured. The
de-doubleted ambient — the object fishash's null actually applies to — is
indistinguishable from Poisson to the resolution the data allow (§Part 4 puts a
formal test and honest power bounds on that claim).

# Data and definitions

We use the **non-co-cultured 0hr** barnyard experiment from Liu et al. 2025
(GSE272457), which is the sample CLEANSER's Fig 3B is built on. It was assayed
with two chemistries, **direct-capture** and **CROP-seq**, giving a clean
within-dataset contrast. Both are exported (guides $\times$ cells, MatrixMarket)
alongside per-cell gene-expression (GEX) metadata and a guide$\to$library map.

```{r data-inventory}
#| code-fold: true
files <- list.files(REPRO, pattern = "mix0hr_(DirectCapture|Cropseq)_(grna_counts\\.mtx|guides\\.csv|meta\\.csv)$")
data.frame(file = sort(files))
```

**The guide$\to$library map.** Each guide is a non-targeting spacer (`nt_1 …
nt_200`) assigned to one of the two libraries. We confirm the semantics directly
from the file: `nt_1..nt_100` are `homo_guide` (human / 293T library) and
`nt_101..nt_200` are `mus_guide` (mouse / NIH-3T3 library).

```{r guide-semantics}
guides <- read.csv(file.path(REPRO, "mix0hr_DirectCapture_guides.csv"))
kable(rbind(head(guides, 2), guides[c(100, 101), ], tail(guides, 2)),
      row.names = FALSE, caption = "Guide-to-library map (endpoints of each block)")
table(guides$guide_type)
```

**Per-cell metadata and the species call.** The metadata gives, per cell, the
summed GEX UMIs mapping to the human transcriptome (`homo_sum_gex`) and the mouse
transcriptome (`mus_sum_gex`), plus mitochondrial and feature counts used for QC.
A cell's **species is called from GEX**, independently of any gRNA. Crucially,
**these exports have no QC or species filtering applied** — QC is downstream and
is exactly the knob we will exercise.

```{r meta-cols}
meta <- read.csv(file.path(REPRO, "mix0hr_DirectCapture_meta.csv"))
str(meta)
```

## Explicit definitions used throughout

- **Species call (GEX-purity $p$).** A cell is **human** if
  `homo_sum_gex > mus_sum_gex` **and** the majority-species fraction of GEX
  $\ge p$; **mouse** analogously. The authors' QC uses $p = 0.90$. This $p$ is
  the authors' own doublet/ambient-purity knob (a cell that is 50/50 human/mouse
  GEX is a GEX doublet and is excluded).
- **Ground-truth ambient.** The count of a **wrong-species** guide in a
  **host-species** cell — e.g. a mouse-library guide in a human cell. That guide
  was never integrated into that cell, so any UMIs are ambient contamination (or
  a co-encapsulated cell). This is the object both papers model.
- **CLEANSER Fig 3B statistic.** For each wrong-species guide, the **mean and
  variance of its UMI counts across all host-species cells, including the zeros**.
  Because most host cells have zero ambient UMIs of a given guide, per-guide means
  are tiny ($\sim 0.005$–$0.3$). Plot variance vs mean; the line **var = mean** is
  the Poisson expectation.
- **gRNA-channel doublet score.** For a cell,
  $\text{wrong\_frac} = \dfrac{\text{wrong-species gRNA UMIs}}{\text{total gRNA
  UMIs}}$. A cell whose gRNA pool is mostly wrong-species is not "ambient" — it is
  a co-encapsulated wrong-species cell (a doublet). This is a gRNA-channel purity
  measure, **independent** of the GEX-purity QC.

## Loading + QC (verbatim from the authors' pipeline)

The QC below is transcribed from the authors' `barnyard_io.R`, with GEX-purity
$p$ exposed as the only free knob. Everything downstream calls `load_sample()`.

```{r loaders}
# Authors' GEX QC (verbatim), purity p as the only knob.
barnyard_qc <- function(meta, purity = 0.9) {
  sum_gex   <- meta$homo_sum_gex + meta$mus_sum_gex
  frac_homo <- meta$homo_sum_gex / sum_gex
  qc <- (meta$mito_sum / sum_gex < .15) & (meta$features_gex <= 6000) &
        (sum_gex <= 20000) & (meta$features_gex >= 1500) & (sum_gex >= 3500) &
        (frac_homo < (1 - purity) | frac_homo > purity)
  qc[is.na(qc)] <- FALSE
  list(qc = qc, frac_homo = frac_homo)
}

load_sample <- function(sample, purity = 0.9) {
  gm     <- as(readMM(file.path(REPRO, paste0(sample, "_grna_counts.mtx"))), "CsparseMatrix")
  meta   <- read.csv(file.path(REPRO, paste0(sample, "_meta.csv")))
  guides <- read.csv(file.path(REPRO, paste0(sample, "_guides.csv")))
  stopifnot(nrow(meta) == ncol(gm), nrow(guides) == nrow(gm))
  q <- barnyard_qc(meta, purity = purity)
  counts <- gm[, q$qc, drop = FALSE]
  rownames(counts) <- guides$guide
  guide_homo <- guides$guide_type == "homo_guide"   # TRUE = human-library guide
  cell_homo  <- q$frac_homo[q$qc] > purity           # TRUE = pure human cell
  list(counts = counts, guide_homo = guide_homo, cell_homo = cell_homo,
       frac_homo = q$frac_homo[q$qc], purity = purity,
       chemistry = if (grepl("Cropseq", sample)) "CROP-seq" else "direct-capture",
       sample = sample)
}

# Wrong-species (ambient) mask: guide species != cell species.
amb_mask_of <- function(obj) {
  m <- outer(obj$guide_homo, obj$cell_homo, function(g, c) g != c)
  as(Matrix(m, sparse = TRUE), "lgCMatrix")
}
```

**Cohort shapes.** After the authors' QC at $p = 0.90$, the 0hr experiment yields
thousands of confidently-called single-species cells in each chemistry.

```{r cohort-shapes}
shapes <- do.call(rbind, lapply(c("mix0hr_DirectCapture", "mix0hr_Cropseq"), function(s) {
  o <- load_sample(s, 0.90)
  data.frame(sample = s, chemistry = o$chemistry,
             QC_cells = ncol(o$counts),
             human_cells = sum(o$cell_homo), mouse_cells = sum(!o$cell_homo),
             homo_guides = sum(o$guide_homo), mus_guides = sum(!o$guide_homo))
}))
kable(shapes, caption = "0hr barnyard cohort after authors' GEX QC (purity 0.90)")
```

# Part 1 — Reproduce CLEANSER Figure 3B {#sec-part1}

We compute the Fig-3B statistic exactly: per wrong-species guide, the mean and
(sample) variance of its UMI counts across all host-species cells, **including
zeros**.

```{r fig3b-fn}
# Per-guide (mean, var, var/mean) of a wrong-species guide block across
# host-species cells, INCLUDING zeros. focus picks the direction.
cleanser_perguide <- function(obj, focus = c("mus_in_homo", "homo_in_mus"),
                              cell_subset = NULL) {
  focus   <- match.arg(focus)
  cnt     <- obj$counts
  cellsel <- if (is.null(cell_subset)) rep(TRUE, ncol(cnt)) else cell_subset
  if (focus == "mus_in_homo") {          # mouse-library guides across human cells
    gsel <- !obj$guide_homo
    csel <- obj$cell_homo & cellsel
  } else {                                # human-library guides across mouse cells
    gsel <- obj$guide_homo
    csel <- (!obj$cell_homo) & cellsel
  }
  sub <- cnt[gsel, csel, drop = FALSE]    # ambient guides x host-species cells
  nC  <- ncol(sub)
  s1  <- Matrix::rowSums(sub)             # per-guide sum over ALL nC cells
  s2  <- Matrix::rowSums(sub * sub)
  mu  <- s1 / nC
  v   <- (s2 - nC * mu^2) / (nC - 1)      # sample variance, incl. the zeros
  data.frame(guide = rownames(sub), mean = mu, var = v,
             vmr = ifelse(mu > 0, v / mu, NA_real_), n_cells = nC, focus = focus)
}

# Pooled var/mean over EVERY wrong-species (guide, cell) entry incl. zeros.
pooled_vmr <- function(obj, cell_subset = NULL) {
  mask <- amb_mask_of(obj); cnt <- obj$counts
  if (!is.null(cell_subset)) {
    cnt <- cnt[, cell_subset, drop = FALSE]; mask <- mask[, cell_subset, drop = FALSE]
  }
  vals <- (cnt * mask)[mask]              # every masked entry, incl. zeros
  list(mean = mean(vals), var = var(vals), vmr = var(vals) / mean(vals),
       n = length(vals), n_nz = sum(vals > 0), max = max(vals))
}
```

```{r fig3b-table}
dc <- load_sample("mix0hr_DirectCapture", 0.90)  # authors' QC
cs <- load_sample("mix0hr_Cropseq", 0.90)

summ <- do.call(rbind, list(
  cbind(sample = "DirectCapture", dir = "mouse-in-human",
        (function(pg) data.frame(n_guides = nrow(pg),
          mean_lo = min(pg$mean), mean_hi = max(pg$mean),
          median_vmr = median(pg$vmr, na.rm = TRUE),
          pct_above = 100 * mean(pg$var > pg$mean, na.rm = TRUE),
          pooled_vmr = sum(pg$var) / sum(pg$mean)))(cleanser_perguide(dc, "mus_in_homo"))),
  cbind(sample = "DirectCapture", dir = "human-in-mouse",
        (function(pg) data.frame(n_guides = nrow(pg),
          mean_lo = min(pg$mean), mean_hi = max(pg$mean),
          median_vmr = median(pg$vmr, na.rm = TRUE),
          pct_above = 100 * mean(pg$var > pg$mean, na.rm = TRUE),
          pooled_vmr = sum(pg$var) / sum(pg$mean)))(cleanser_perguide(dc, "homo_in_mus"))),
  cbind(sample = "CROP-seq", dir = "mouse-in-human",
        (function(pg) data.frame(n_guides = nrow(pg),
          mean_lo = min(pg$mean), mean_hi = max(pg$mean),
          median_vmr = median(pg$vmr, na.rm = TRUE),
          pct_above = 100 * mean(pg$var > pg$mean, na.rm = TRUE),
          pooled_vmr = sum(pg$var) / sum(pg$mean)))(cleanser_perguide(cs, "mus_in_homo"))),
  cbind(sample = "CROP-seq", dir = "human-in-mouse",
        (function(pg) data.frame(n_guides = nrow(pg),
          mean_lo = min(pg$mean), mean_hi = max(pg$mean),
          median_vmr = median(pg$vmr, na.rm = TRUE),
          pct_above = 100 * mean(pg$var > pg$mean, na.rm = TRUE),
          pooled_vmr = sum(pg$var) / sum(pg$mean)))(cleanser_perguide(cs, "homo_in_mus")))
))
kable(summ, digits = c(0, 0, 0, 3, 3, 2, 1, 2),
      caption = "Fig-3B per-guide dispersion by chemistry and direction (0hr, GEX purity 0.90)")

# stash key numbers for inline use
dc_med_vmr  <- summ$median_vmr[1]
dc_pct_ab   <- summ$pct_above[1]
dc_pool_vmr <- summ$pooled_vmr[1]
cs_med_vmr  <- summ$median_vmr[3]
cs_pool_vmr <- summ$pooled_vmr[3]
```

The dominant direction — **mouse-library guides in human cells** — shows a
per-guide **median var/mean of `r fmt(dc_med_vmr)`** with
**`r fmt(dc_pct_ab, 0)`% of guides above the Poisson line** and a **pooled
var/mean of `r fmt(dc_pool_vmr, 1)`**. The **CROP-seq** contrast is Poisson:
median var/mean **`r fmt(cs_med_vmr)`**, pooled **`r fmt(cs_pool_vmr)`**. This is
exactly CLEANSER's Fig 3B claim, and it reproduces from the raw counts.

```{r fig3b-plot}
#| fig-cap: "CLEANSER Fig-3B reproduction: per-guide variance vs mean of ground-truth ambient UMIs (0hr, GEX purity 0.90). Dashed line is var = mean (Poisson). Direct-capture mouse-in-human sits ABOVE the line; CROP-seq sits ON it."
#| fig-width: 11
#| fig-height: 5.5
pgA <- rbind(
  transform(cleanser_perguide(dc, "mus_in_homo"), chem = "direct-capture", dir = "mouse-guides in human cells"),
  transform(cleanser_perguide(dc, "homo_in_mus"), chem = "direct-capture", dir = "human-guides in mouse cells"),
  transform(cleanser_perguide(cs, "mus_in_homo"), chem = "CROP-seq",       dir = "mouse-guides in human cells"),
  transform(cleanser_perguide(cs, "homo_in_mus"), chem = "CROP-seq",       dir = "human-guides in mouse cells"))
pgA <- pgA[is.finite(pgA$mean) & pgA$mean > 0, ]

ggplot(pgA, aes(mean, var, colour = chem)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.7, size = 2) +
  facet_wrap(~dir, scales = "free") +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c("direct-capture" = "#D55E00", "CROP-seq" = "#0072B2")) +
  labs(x = "per-guide mean ambient UMIs (across host cells, incl. zeros)",
       y = "per-guide variance", colour = "chemistry") +
  theme(legend.position = "top")
```

The left panel (mouse guides in human cells) is CLEANSER's Fig 3B: the
direct-capture cloud floats an order of magnitude above the dashed Poisson line,
while the CROP-seq cloud hugs it.

# Part 2 — The doublets are there {#sec-part2}

If the overdispersion were genuine "soup," it would be spread smoothly across
cells. Instead it is concentrated in a small tail of cells carrying a whole
wrong-species cell's worth of gRNA — co-encapsulated doublets.

## Per-cell wrong-species gRNA load is bimodal with a heavy tail

```{r percell-load}
mask     <- amb_mask_of(dc)
cnt      <- dc$counts
amb_load <- Matrix::colSums(cnt * mask)                 # wrong-species UMIs / cell
tot_grna <- Matrix::colSums(cnt)
wrong_frac <- amb_load / pmax(tot_grna, 1)              # gRNA-channel doublet score

qs <- quantile(amb_load, c(0, .5, .9, .95, .99, .999, 1))
kable(t(round(qs)), caption = "Per-cell wrong-species gRNA UMI load (direct-capture 0hr)")

load_med <- median(amb_load); load_q90 <- quantile(amb_load, .9)
load_q99 <- quantile(amb_load, .99); load_max <- max(amb_load)
```

The distribution is **bimodal**: the bulk of cells carry essentially no
wrong-species gRNA (**median = `r round(load_med)`**, **q90 = `r round(load_q90)`**),
but a right tail runs to **q99 $\approx$ `r round(load_q99)`** and a **max of
`r round(load_max)`** UMIs — a value that only makes sense if a second,
wrong-species cell was captured in the droplet.

```{r percell-plot}
#| fig-cap: "Per-cell wrong-species gRNA UMI load (direct-capture 0hr, GEX-pure cells, zeros omitted for the log axis). The bulk carries a few ambient UMIs; a right tail runs to >1000 — co-encapsulated wrong-species cells (doublets). Dotted = 5, dashed = 50."
#| fig-width: 10
#| fig-height: 4.6
ggplot(data.frame(amb_load = amb_load[amb_load > 0]), aes(amb_load)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = "white") +
  scale_x_log10() +
  geom_vline(xintercept = c(5, 50), linetype = c("dotted", "dashed"), colour = "red") +
  labs(x = "wrong-species gRNA UMIs per cell (log scale; zeros excluded)", y = "number of cells")
```

## Example doublet cells: a whole wrong-species cell's worth of gRNA

Cells with `wrong_frac` near 1 are essentially wrong-species cells: the guide
that dominates their gRNA pool is from the *other* library. These are not ambient
"soup" — they are doublets or mis-called cells.

```{r example-doublets}
# top cells by wrong fraction (requiring a non-trivial gRNA count)
ord <- order(-wrong_frac)
ord <- ord[tot_grna[ord] >= 5][1:6]
# dominant guide per example cell
dominant <- vapply(ord, function(j) {
  col <- cnt[, j]; rownames(cnt)[which.max(col)]
}, character(1))
ex <- data.frame(
  cell_index   = ord,
  species_call = ifelse(dc$cell_homo[ord], "human", "mouse"),
  frac_homo_GEX = round(dc$frac_homo[ord], 3),
  total_grna_UMI = tot_grna[ord],
  wrong_species_UMI = amb_load[ord],
  wrong_frac = round(wrong_frac[ord], 3),
  dominant_guide = dominant,
  dominant_library = ifelse(guides$guide_type[match(dominant, guides$guide)] == "homo_guide",
                            "human", "mouse"))
kable(ex, row.names = FALSE,
      caption = "Example doublet cells: GEX calls one species, but the gRNA pool is dominated by a guide from the OTHER library (wrong_frac near 1).")
```

Each row is a cell the GEX QC accepted as a confident single species, yet its
gRNA channel is dominated by a guide from the opposite library — the signature of
a co-encapsulated wrong-species cell.

## Variance decomposition: the top 1% of cells own the overdispersion

We now quantify how concentrated the excess variance is. For the pooled ambient
(every wrong-species entry, incl. zeros), compute the total sum-of-squares about
the mean and the share contributed by cells in the top 1% by wrong-species load.

```{r variance-decomp}
mask_T <- as(mask, "TsparseMatrix"); jcell <- mask_T@j + 1L    # cell of each masked entry
vals_full <- (cnt * mask)[mask]
mu_all    <- mean(vals_full)
ss_total  <- sum((vals_full - mu_all)^2)

top1_cell  <- amb_load > quantile(amb_load, 0.99)
ss_top1    <- sum((vals_full[top1_cell[jcell]] - mu_all)^2)
top1_share <- 100 * ss_top1 / ss_total

# also the UMI share and how many cells that is
n_top1     <- sum(top1_cell)
umi_share  <- 100 * sum(amb_load[top1_cell]) / sum(amb_load)
data.frame(metric = c("cells in top 1% by wrong load",
                       "% of pooled ambient sum-of-squares from top 1% cells",
                       "% of all ambient UMIs carried by top 1% cells"),
           value = c(n_top1, round(top1_share, 1), round(umi_share, 1)))
```

The **top 1% of cells (`r n_top1` cells) account for `r fmt(top1_share, 1)`% of
the pooled ambient sum-of-squares** and carry `r fmt(umi_share, 1)`% of all
wrong-species UMIs. The "overdispersion" is not a property of the soup — it lives
almost entirely in a handful of doublets.

We can watch the pooled var/mean melt as we peel off the top cells by load:

```{r decomp-sweep}
sweep_tbl <- do.call(rbind, lapply(c(1.0, 0.999, 0.99, 0.95, 0.90), function(topq) {
  thr   <- quantile(amb_load, topq)
  keep  <- amb_load <= thr
  pv    <- pooled_vmr(dc, cell_subset = keep)
  data.frame(keep_below_quantile = topq, load_threshold = as.numeric(thr),
             cells_dropped = sum(!keep),
             pct_ambient_UMIs_dropped = round(100 * sum(amb_load[!keep]) / sum(amb_load), 1),
             pooled_vmr = round(pv$vmr, 2), pooled_max = pv$max)
}))
kable(sweep_tbl, caption = "Pooled ambient var/mean collapses as top-load cells are removed (direct-capture 0hr)")
```

Dropping merely the **top 1% of cells** by load takes the pooled var/mean from
**`r fmt(sweep_tbl$pooled_vmr[1], 2)`** to **`r fmt(sweep_tbl$pooled_vmr[3], 2)`**,
and dropping the top 5% takes it to **`r fmt(sweep_tbl$pooled_vmr[4], 2)`** — i.e.
essentially Poisson.

# Part 3 — Clean the doublets and re-draw Figure 3B {#sec-part3}

The natural, transparent doublet filter is a **gRNA-channel purity cut**: drop
any cell whose wrong-species fraction of total gRNA exceeds a cutoff. This is
independent of the GEX QC and directly targets the co-encapsulation signature.

## Fig 3B before vs after a gRNA-purity cut

```{r purity-sweep}
purity_tbl <- do.call(rbind, lapply(c(1.01, 0.50, 0.10, 0.05, 0.01), function(cut) {
  keep <- wrong_frac <= cut
  pg   <- cleanser_perguide(dc, "mus_in_homo", cell_subset = keep)
  pv   <- pooled_vmr(dc, cell_subset = keep)
  data.frame(gRNA_purity_filter = if (cut > 1) "keep ALL (GEX-pure only)"
                                  else sprintf("wrong_frac <= %.2f", cut),
             cells_kept = sum(keep), cells_dropped = sum(!keep),
             perguide_median_vmr = round(median(pg$vmr, na.rm = TRUE), 2),
             perguide_pct_above  = round(100 * mean(pg$var > pg$mean, na.rm = TRUE), 0),
             pooled_vmr = round(pv$vmr, 2), pooled_max = pv$max)
}))
kable(purity_tbl, caption = "Fig-3B dispersion under a gRNA-channel purity cut (direct-capture 0hr, mouse-in-human)")

clean_med_vmr <- purity_tbl$perguide_median_vmr[purity_tbl$gRNA_purity_filter == "wrong_frac <= 0.01"]
clean_pct_ab  <- purity_tbl$perguide_pct_above[purity_tbl$gRNA_purity_filter == "wrong_frac <= 0.01"]
clean_max     <- purity_tbl$pooled_max[purity_tbl$gRNA_purity_filter == "wrong_frac <= 0.01"]
raw_max       <- purity_tbl$pooled_max[1]
```

At a strict gRNA-purity cut (`wrong_frac <= 0.01`), the per-guide **median
var/mean drops from `r fmt(dc_med_vmr)` to `r fmt(clean_med_vmr)`**, only
**`r fmt(clean_pct_ab, 0)`%** of guides remain above the line (a coin-flip, as
expected under Poisson sampling noise), and the largest single ambient count
falls from **`r raw_max`** to **`r clean_max`**. The overdispersion is gone.

```{r collapse-plot}
#| fig-cap: "Direct-capture ambient overdispersion collapses to Poisson when gRNA-channel doublets are removed. Per-guide variance vs mean (mouse guides in human cells, 0hr). Dashed = var = mean. As the wrong-fraction cut tightens, the cloud pulls ONTO the Poisson line."
#| fig-width: 10
#| fig-height: 6
rowsB <- lapply(c(1.01, 0.10, 0.01), function(cut) {
  keep <- wrong_frac <= cut
  pg   <- cleanser_perguide(dc, "mus_in_homo", cell_subset = keep)
  lab  <- if (cut > 1)
    sprintf("keep all GEX-pure cells (median vmr %.1f)", median(pg$vmr, na.rm = TRUE))
  else
    sprintf("drop wrong_frac > %.2f (median vmr %.2f)", cut, median(pg$vmr, na.rm = TRUE))
  transform(pg[is.finite(pg$mean) & pg$mean > 0, ], filter = lab)
})
pgB <- do.call(rbind, rowsB)
pgB$filter <- factor(pgB$filter, levels = vapply(rowsB, function(d) as.character(d$filter[1]), ""))

ggplot(pgB, aes(mean, var, colour = filter)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.75, size = 2.4) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c("#D55E00", "#E69F00", "#0072B2")) +
  labs(x = "per-guide mean ambient UMIs (incl. zeros)", y = "per-guide variance",
       colour = "cell filter")
```

## After doublets: why a pooled NB fit is the wrong diagnostic

How Poisson is the de-doubleted ambient? The tempting shortcut — fit one Negative
Binomial to the *pooled* de-doubleted counts by method of moments
($\mathrm{Var} = \mu + \mu^2/\theta$; size $\theta = \mu^2/(\mathrm{Var}-\mu)$,
dispersion $\alpha = 1/\theta = (\mathrm{Var}-\mu)/\mu^2$; the Poisson limit is
$\mathrm{Var}\to\mu$, i.e. $\theta\to\infty$ and $\alpha\to 0$) — is actually the
**wrong diagnostic here**, and seeing why is the point. (The *right* tool is a
bootstrap-calibrated goodness-of-fit test against $\mathrm{Poisson}(a_g d_c)$; we
build it in §Part 4. This section only shows why the naive pooled fit must not be
read as an answer.)

```{r nb-fit}
nb_tbl <- do.call(rbind, lapply(c(1.01, 0.05, 0.01), function(cut) {
  keep <- wrong_frac <= cut
  vals <- (cnt * mask)[, keep, drop = FALSE][mask[, keep, drop = FALSE]]
  m <- mean(vals); v <- var(vals)
  theta_size <- if (v > m) m^2 / (v - m) else Inf   # NB size (MoM)
  alpha_disp <- if (v > m) (v - m) / m^2 else 0      # NB dispersion = 1/size
  data.frame(gRNA_purity = if (cut > 1) "keep ALL" else sprintf("wrong_frac <= %.2f", cut),
             cells = sum(keep), pooled_mean = round(m, 5), pooled_var = round(v, 5),
             NB_size_theta = round(theta_size, 4),
             NB_dispersion_alpha = signif(alpha_disp, 3))
}))
kable(nb_tbl, caption = "Method-of-moments NB fit to pooled ambient at increasing gRNA purity (direct-capture 0hr)")

theta_clean <- nb_tbl$NB_size_theta[nb_tbl$gRNA_purity == "wrong_frac <= 0.01"]
```

At the strict cut the pooled mean and variance are ~0.010 and ~0.022, so the
**pooled** var/mean is $\approx 2.2$ — *not* Poisson at the pooled level — and the
MoM fit returns a tiny size $\theta =$ **`r fmt(theta_clean, 4)`** (equivalently a
large dispersion $\alpha = 1/\theta \approx$ **`r fmt(1/theta_clean, 0)`**). This is
exactly where the prior "$\alpha\approx 0.008$" figure went wrong: $0.008$ is the NB
**size** $\theta$, which is *small precisely because the pool is overdispersed* — it
is not a small dispersion and not Poisson.

Why is the pooled fit misleading? The pool mixes wrong-species guides of very
different soup-shares $a_g$ and host cells of very different ambient depths $d_c$,
and a mixture of Poissons with unequal means $\lambda_{gc}=a_g d_c$ always has
pooled var/mean $>1$ (law of total variance) even when every entry is *exactly*
Poisson. So a pooled NB fit conflates two things it cannot separate: **rank-1
mean-heterogeneity** (guide/cell mean spread, present under the null) and any
genuine **entry-level dispersion** (the thing CLEANSER's NB is meant to capture).
The right question is not "is the pooled var/mean $>1$?" — it always is — but "is
the pooled var/mean *larger than a rank-1 Poisson would itself produce*?"

That comparison needs a null. §Part 4 supplies it by **simulating from the fitted
rank-1 mean** and asking how much pooled var/mean the Poisson model reproduces on
its own. The answer, previewing: the Poisson null reproduces a pooled var/mean of
$\approx 1.1$ (that is the mean-heterogeneity), and the small observed excess over
that is not smooth overdispersion but **1–5 individual cut-boundary doublet
entries** — a stricter cut (`wrong_frac <= 0.005`) collapses the pooled var/mean to
$\approx 1.09$. Two cautions this makes explicit. First, the earlier claim that the
per-guide median var/mean $\to 1.0$ "proves" the counts are conditionally Poisson is
**not valid evidence**: at these tiny rates a per-guide var/mean is
$1 + a_g\,\mathrm{Var}[d]/\mathbb{E}[d] \approx 1$ *regardless* of Poisson-vs-NB,
because $a_g$ is minuscule — it cannot tell the hypotheses apart, so we drop it.
Second, the robust qualitative evidence remains the Part-2 decomposition
(overdispersion *localized* to a handful of doublet cells, not smoothly spread); the
quantitative verdict is the §Part 4 test.

A maximum-likelihood NB fit on the pooled counts (where the package is available)
tells the same story — a small size, i.e. the pooled mean-heterogeneity again:

```{r nb-mle}
#| code-fold: true
if (requireNamespace("MASS", quietly = TRUE)) {
  keep <- wrong_frac <= 0.01
  vals <- (cnt * mask)[, keep, drop = FALSE][mask[, keep, drop = FALSE]]
  fit  <- tryCatch(suppressWarnings(MASS::fitdistr(vals, "negative binomial")),
                   error = function(e) NULL)
  if (!is.null(fit)) {
    cat(sprintf("MASS::fitdistr NB (wrong_frac<=0.01): size = %.4f, mu = %.5f\n",
                fit$estimate["size"], fit$estimate["mu"]))
    cat(sprintf("  -> pooled mean %.5f, var %.5f (var/mean = %.3f)\n",
                mean(vals), var(vals), var(vals) / mean(vals)))
  }
} else {
  cat("MASS not available; method-of-moments NB fit above is the reported estimate.\n")
}
```

## The authors' own GEX-purity knob does the same, more weakly

Tightening the GEX-purity threshold also strips doublets — but less efficiently,
because a whole wrong-species cell can still pass a lenient GEX cut. Sweeping
$p = 0.90 \to 0.95 \to 0.99$ walks the per-guide median var/mean down toward 1.

```{r gex-purity-sweep}
gex_tbl <- do.call(rbind, lapply(c(0.90, 0.95, 0.99), function(pur) {
  op <- load_sample("mix0hr_DirectCapture", pur)
  pg <- cleanser_perguide(op, "mus_in_homo")
  pv <- pooled_vmr(op)
  data.frame(GEX_purity = pur, human_cells = sum(op$cell_homo),
             perguide_median_vmr = round(median(pg$vmr, na.rm = TRUE), 2),
             perguide_pct_above  = round(100 * mean(pg$var > pg$mean, na.rm = TRUE), 0),
             pooled_vmr = round(pv$vmr, 2), pooled_max = pv$max)
}))
kable(gex_tbl, caption = "GEX-purity knob (authors' own) also collapses the overdispersion, more weakly")
```

At $p = 0.99$ the GEX knob alone brings the per-guide median var/mean to
**`r fmt(gex_tbl$perguide_median_vmr[gex_tbl$GEX_purity == 0.99])`** and the max
from **`r gex_tbl$pooled_max[1]`** to **`r gex_tbl$pooled_max[3]`** — the same
direction as the gRNA-purity cut, confirming both knobs remove the same objects
(doublets), and that the gRNA-channel cut is simply the more direct instrument.

## Cross-check: the 72hr direct-capture sample replicates the pattern

```{r hr72-check}
hr72 <- do.call(rbind, lapply(c("mix72hr_DirectCapture", "mix72hr_Cropseq"), function(s) {
  o    <- load_sample(s, 0.90)
  pg   <- cleanser_perguide(o, "mus_in_homo")
  mask2 <- amb_mask_of(o); cnt2 <- o$counts
  amb2 <- Matrix::colSums(cnt2 * mask2); wf2 <- amb2 / pmax(Matrix::colSums(cnt2), 1)
  pv_all   <- pooled_vmr(o)
  pv_clean <- pooled_vmr(o, cell_subset = wf2 <= 0.01)
  data.frame(sample = s, chemistry = o$chemistry,
             perguide_median_vmr = round(median(pg$vmr, na.rm = TRUE), 2),
             pooled_vmr_all = round(pv_all$vmr, 2), max_all = pv_all$max,
             pooled_vmr_dedoubleted = round(pv_clean$vmr, 2), max_dedoubleted = pv_clean$max)
}))
kable(hr72, caption = "72hr barnyard: same pattern — direct-capture ambient is doublet-inflated and collapses to Poisson under a gRNA-purity cut")
```

The 72hr direct-capture sample tells the same story: a pooled var/mean of
**`r fmt(hr72$pooled_vmr_all[1], 1)`** with a max of **`r hr72$max_all[1]`**
collapses to **`r fmt(hr72$pooled_vmr_dedoubleted[1], 2)`** (max
**`r hr72$max_dedoubleted[1]`**) once gRNA-channel doublets are removed, while
CROP-seq is Poisson throughout.

# Part 4 — Is the de-doubleted ambient actually Poisson? A bootstrap-calibrated test {#sec-part4}

Part 3 established what the *wrong* tool says. Here is the right one. We test the
sharp hypothesis
$$
H_0:\quad Y_{gc} \;\sim\; \mathrm{Poisson}(\lambda_{gc}),\qquad
\lambda_{gc} \;=\; a_g\,d_c \;=\; \frac{R_g\,C_c}{T},
$$
i.e. the de-doubleted ground-truth ambient count of guide $g$ in host cell $c$ is
Poisson with a **rank-1 mean** — guide soup-share $a_g=R_g/T$ times cell soup-exposure
$C_c$ ($R_g$, $C_c$ are the guide/cell margins of the de-doubleted submatrix, $T$ the
grand total). This is exactly fishash's / geomux's ambient null. The obstacle to
testing it is that the rates are tiny (per-entry means $\sim 0.005$–$0.3$, counts
almost all 0/1), so asymptotic $\chi^2$ / dispersion references are unreliable. We
therefore **calibrate every statistic by parametric bootstrap**: simulate
$Y^{(b)}\sim\mathrm{Poisson}(\hat\lambda)$, **re-fit the rank-1 mean each rep**, and
read the observed statistic against its own Poisson null band. Three statistics, one
mask direction (mouse-library guides in human cells — the overdispersed one), plus
controls that prove the test discriminates.

::: {.callout-note}
The bootstrap runs live at render time with `B = 200` and `set.seed(1)`, so every
band below is reproduced from the raw counts. On this cohort the whole section
(4 conditions + power) takes well under a minute.
:::

```{r gof-lib}
#| code-fold: true
set.seed(1L)
B_GOF          <- 200L    # bootstrap reps (matches the reported bands)
NBINS          <- 10L     # matched-mean lambda bins
LAM_FLOOR_BIN  <- 0.02    # floor for the matched-mean binning (informative entries)
LAM_FLOOR_PHI  <- 0.01    # floor for the "informative-entry" Pearson phi
KMAX           <- 4L      # count histogram last bin is >=4

# Build the wrong-species-guide x host-cell submatrix Y (dense, small), in a
# direction, after an optional gRNA-purity (doublet) cut wf_cut. Inf keeps all.
build_Y <- function(obj, direction = c("mus_in_homo", "homo_in_mus"), wf_cut = 0.01) {
  direction <- match.arg(direction)
  cnt  <- obj$counts
  tot  <- Matrix::colSums(cnt)
  wm   <- outer(obj$guide_homo, obj$cell_homo, function(g, c) g != c)
  amb  <- Matrix::colSums(cnt * as(Matrix(wm, sparse = TRUE), "lgCMatrix"))
  wf   <- amb / pmax(tot, 1)
  if (direction == "mus_in_homo") { gsel <- !obj$guide_homo; csel <-  obj$cell_homo }
  else                            { gsel <-  obj$guide_homo; csel <- !obj$cell_homo }
  csel <- csel & (wf <= wf_cut)
  Y <- as.matrix(cnt[gsel, csel, drop = FALSE]); storage.mode(Y) <- "double"
  n_drop <- sum((if (direction == "mus_in_homo") obj$cell_homo else !obj$cell_homo) & (wf > wf_cut))
  list(Y = Y, n_guides = nrow(Y), n_cells = ncol(Y), n_dropped = n_drop)
}

# Fit rank-1 Poisson mean lambda_hat = R_g C_c / T and compute all GOF statistics
# for a count matrix Y. bin_breaks fixed from the observed fit so per-bin var/mean
# is comparable across bootstrap reps.
fit_stats <- function(Y, bin_breaks = NULL) {
  R <- rowSums(Y); C <- colSums(Y); Tg <- sum(Y)
  lam <- outer(R, C) / Tg
  yv <- as.vector(Y); lv <- as.vector(lam)
  # (a) count histogram O_k vs Poisson E_k for k = 0,1,2,3,>=4
  Ok <- c(sum(yv == 0), sum(yv == 1), sum(yv == 2), sum(yv == 3), sum(yv >= 4))
  p0 <- dpois(0, lv); p1 <- dpois(1, lv); p2 <- dpois(2, lv); p3 <- dpois(3, lv)
  Ek <- c(sum(p0), sum(p1), sum(p2), sum(p3), sum(1 - (p0 + p1 + p2 + p3)))
  # (b) Pearson dispersion phi = mean((y-lam)^2/lam), over lam>0 and over lam>=floor
  pos <- lv > 0
  phi_all   <- mean((yv[pos] - lv[pos])^2 / lv[pos])
  keepf     <- lv >= LAM_FLOOR_PHI
  phi_floor <- if (sum(keepf) > 0) mean((yv[keepf] - lv[keepf])^2 / lv[keepf]) else NA_real_
  # (c) matched-mean var/mean by lambda bin (fixed quantile breaks over lam>=floor)
  bs <- lv >= LAM_FLOOR_BIN; yb <- yv[bs]; lb <- lv[bs]
  if (is.null(bin_breaks)) {
    qs <- quantile(lb, probs = seq(0, 1, length.out = NBINS + 1L), na.rm = TRUE)
    qs[1] <- -Inf; qs[length(qs)] <- Inf; bin_breaks <- unique(qs)
  }
  g <- cut(lb, breaks = bin_breaks, include.lowest = TRUE, labels = FALSE)
  bt <- data.frame(bin = seq_len(length(bin_breaks) - 1L),
                   n = NA_real_, lam_mean = NA_real_, y_mean = NA_real_, vmr = NA_real_)
  for (bb in bt$bin) {
    ix <- which(g == bb)
    if (length(ix) >= 2) {
      bt$n[bb] <- length(ix); bt$lam_mean[bb] <- mean(lb[ix])
      bt$y_mean[bb] <- mean(yb[ix]); bt$vmr[bb] <- var(yb[ix]) / bt$y_mean[bb]
    }
  }
  list(Ok = Ok, Ek = Ek, ratio = Ok / Ek, phi_all = phi_all, phi_floor = phi_floor,
       n_phi_floor = sum(keepf), bin = bt, bin_breaks = bin_breaks,
       pooled_mean = mean(yv), pooled_var = var(yv), pooled_vmr = var(yv) / mean(yv),
       n_entries = length(yv), n_nz = sum(yv > 0), maxct = max(yv))
}

# Parametric bootstrap: simulate Poisson(lambda_hat), REFIT rank-1 mean, recompute.
boot_null <- function(Y, obs_breaks, B = B_GOF) {
  R <- rowSums(Y); C <- colSums(Y); Tg <- sum(Y)
  lamv <- as.vector(outer(R, C) / Tg); nG <- nrow(Y); nC <- ncol(Y)
  nbin <- length(obs_breaks) - 1L
  out <- list(phi_all = numeric(B), phi_floor = numeric(B), pooled_vmr = numeric(B),
              ratio = matrix(NA_real_, B, 5), vmr_bin = matrix(NA_real_, B, nbin))
  for (b in seq_len(B)) {
    yb <- rpois(nG * nC, lamv); dim(yb) <- c(nG, nC)
    st <- fit_stats(yb, bin_breaks = obs_breaks)
    out$phi_all[b] <- st$phi_all; out$phi_floor[b] <- st$phi_floor
    out$pooled_vmr[b] <- st$pooled_vmr; out$ratio[b, ] <- st$ratio; out$vmr_bin[b, ] <- st$bin$vmr
  }
  out
}

pct     <- function(x) quantile(x, c(.025, .5, .975), na.rm = TRUE)
inside  <- function(o, x) { q <- pct(x); o >= q[1] & o <= q[3] }
verdict <- function(o, x) { q <- pct(x); if (o > q[3]) "ABOVE (overdisp)" else if (o < q[1]) "BELOW (underdisp)" else "INSIDE (Poisson-OK)" }
pboot_upper <- function(o, x) (1 + sum(x >= o)) / (1 + length(x))

# run one condition end-to-end; returns obs + null for tables/plots
run_condition <- function(obj, direction, wf_cut, B = B_GOF) {
  bd  <- build_Y(obj, direction, wf_cut)
  obs <- fit_stats(bd$Y)
  nb  <- boot_null(bd$Y, obs$bin_breaks, B = B)
  list(obs = obs, null = nb, bd = bd)
}
```

## The main test: de-doubleted direct-capture is inside the Poisson null band

We fit $\hat\lambda_{gc}=R_gC_c/T$ on the de-doubleted (`wrong_frac <= 0.01`)
mouse-in-human submatrix, then bootstrap the Poisson null.

```{r gof-main}
resA <- run_condition(dc, "mus_in_homo", wf_cut = 0.01)   # de-doubleted (main)
resB <- run_condition(dc, "mus_in_homo", wf_cut = Inf)    # RAW (positive control)
resD <- run_condition(cs, "mus_in_homo", wf_cut = 0.01)   # CROP-seq (comparison)

# Poisson-sim calibration control: one Poisson draw from the de-doubleted fit,
# treated AS IF it were data. Must PASS (the test must not flag its own null).
{
  Ym <- resA$bd$Y; Rm <- rowSums(Ym); Cm <- colSums(Ym); Tm <- sum(Ym)
  lam_sim <- as.vector(outer(Rm, Cm) / Tm)
  Ysim <- matrix(rpois(length(lam_sim), lam_sim), nrow(Ym), ncol(Ym))
  obsC <- fit_stats(Ysim); nbC <- boot_null(Ysim, obsC$bin_breaks, B = B_GOF)
  resC <- list(obs = obsC, null = nbC)
}

band <- function(res, which = "phi_floor") {
  o <- res$obs[[which]]; q <- pct(res$null[[which]])
  sprintf("%.3f  [%.3f, %.3f]", o, q[1], q[3])
}
```

```{r gof-main-table}
#| code-fold: true
gof_tab <- data.frame(
  statistic = c("Pearson phi (lambda_hat >= 0.01)",
                "Pearson phi (all lambda_hat > 0)",
                "pooled var/mean",
                "bootstrap p (upper) for phi>=floor"),
  observed = c(fmt(resA$obs$phi_floor, 3), fmt(resA$obs$phi_all, 3),
               fmt(resA$obs$pooled_vmr, 3),
               fmt(pboot_upper(resA$obs$phi_floor, resA$null$phi_floor), 3)),
  poisson_null_95_band = c(
    sprintf("[%.3f, %.3f]", pct(resA$null$phi_floor)[1], pct(resA$null$phi_floor)[3]),
    sprintf("[%.3f, %.3f]", pct(resA$null$phi_all)[1],   pct(resA$null$phi_all)[3]),
    sprintf("[%.3f, %.3f]", pct(resA$null$pooled_vmr)[1], pct(resA$null$pooled_vmr)[3]),
    "-"),
  verdict = c(verdict(resA$obs$phi_floor, resA$null$phi_floor),
              verdict(resA$obs$phi_all, resA$null$phi_all),
              verdict(resA$obs$pooled_vmr, resA$null$pooled_vmr), "-"))
kable(gof_tab, caption = "De-doubleted direct-capture (mouse-in-human, wrong_frac<=0.01): observed GOF statistics vs the parametric-Poisson bootstrap null band (B=200).")

# stash for inline use
phiA      <- resA$obs$phi_floor
phiA_lo   <- pct(resA$null$phi_floor)[1]; phiA_hi <- pct(resA$null$phi_floor)[3]
pA        <- pboot_upper(resA$obs$phi_floor, resA$null$phi_floor)
poolA     <- resA$obs$pooled_vmr
poolA_null<- pct(resA$null$pooled_vmr)[2]
```

The observed Pearson dispersion over informative entries is
$\phi = `r fmt(phiA, 3)`$, sitting **inside** the Poisson null band
$[`r fmt(phiA_lo, 3)`, `r fmt(phiA_hi, 3)`]$ with bootstrap
$p \approx `r fmt(pA, 2)`$ — no evidence against Poisson. The **pooled** var/mean is
$`r fmt(poolA, 2)`$, which does read "above" its band — **but the Poisson null itself
reproduces a pooled var/mean of $\approx `r fmt(poolA_null, 2)`$** (that is the rank-1
mean-heterogeneity of Part 3), so only a sliver of the pooled number is unexplained.
We localize that sliver below.

```{r gof-histogram-table}
#| code-fold: true
klab <- c("0", "1", "2", "3", ">=4")
oe <- data.frame(
  k = klab, O_k = round(resA$obs$Ok), E_k = round(resA$obs$Ek, 1),
  O_over_E = fmt(resA$obs$ratio, 3),
  null_band = vapply(1:5, function(k) {
    q <- pct(resA$null$ratio[, k]); sprintf("[%.3f, %.3f]", q[1], q[3]) }, ""),
  verdict = vapply(1:5, function(k) verdict(resA$obs$ratio[k], resA$null$ratio[, k]), ""))
kable(oe, caption = "Count-histogram goodness-of-fit: observed O_k, Poisson-expected E_k, ratio O/E, and the bootstrap null band per count k (de-doubleted direct-capture).")
```

The count histogram $O_k/E_k$ sits on the null band at every $k$ — the observed
counts $0,1,2,\dots$ occur at their Poisson-predicted frequencies.

## The two headline figures

The **matched-mean var/mean by $\lambda$ bin** is the cleanest picture: under Poisson,
var/mean $\approx 1$ in *every* bin, and the bootstrap band shows how much scatter
that entails at these rates. De-doubleted direct-capture hugs 1 across the informative
range; the raw (pre-cut) ambient explodes.

```{r gof-fig-vmr}
#| fig-cap: "Matched-mean var/mean per lambda_hat bin, observed vs the parametric-Poisson bootstrap null band (B=200). Under Poisson, var/mean ~ 1 in every bin. LEFT: de-doubleted direct-capture hugs 1 across the informative range (only the top lambda bin — the cut boundary — pokes just above). RIGHT: raw (pre-cut) ambient trips the test hard, reaching var/mean ~ 95 in the top bin. Note the free y-axes."
#| fig-width: 11
#| fig-height: 4.8
vmr_df <- function(res, cond) {
  q <- t(apply(res$null$vmr_bin, 2, function(x) quantile(x, c(.025, .5, .975), na.rm = TRUE)))
  d <- data.frame(lam = res$obs$bin$lam_mean, obs = res$obs$bin$vmr,
                  lo = q[, 1], mid = q[, 2], hi = q[, 3], cond = cond)
  d[is.finite(d$obs) & is.finite(d$lam), ]
}
pdV <- rbind(vmr_df(resA, "de-doubleted direct-capture (main)"),
             vmr_df(resB, "RAW direct-capture (positive control)"))
pdV$cond <- factor(pdV$cond, levels = unique(pdV$cond))
ggplot(pdV, aes(lam)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80", alpha = 0.7) +
  geom_line(aes(y = mid), colour = "grey40", linetype = "dashed") +
  geom_point(aes(y = obs), colour = "firebrick", size = 2.4) +
  geom_line(aes(y = obs), colour = "firebrick") +
  geom_hline(yintercept = 1, colour = "black", linewidth = 0.3) +
  facet_wrap(~cond, scales = "free_y") + scale_x_log10() +
  labs(x = expression("mean fitted rate " * hat(lambda) * " in bin (log)"),
       y = "var(Y) / mean(Y) within bin") +
  theme_bw(base_size = 11)
```

The count-histogram view tells the same story across the three conditions — the
de-doubleted ambient and the Poisson-simulated calibration control both sit on the
band; the raw ambient collapses far below it (its mass is hoarded in a few huge
counts, starving $k=1,2,3$).

```{r gof-fig-hist}
#| fig-cap: "Count-histogram goodness-of-fit: observed O/E vs Poisson bootstrap null band (B=200). Grey ribbon = Poisson [2.5, 97.5]%; red = observed. LEFT de-doubleted direct-capture: on the band at every k. MIDDLE raw direct-capture: O/E crashes at k=1,2,3 (the doublets pile mass into the far tail). RIGHT a Poisson simulation treated as data: passes, confirming the test does not flag its own null."
#| fig-width: 12
#| fig-height: 4.4
oe_df <- function(res, cond) {
  q <- t(apply(res$null$ratio, 2, function(x) quantile(x, c(.025, .5, .975), na.rm = TRUE)))
  data.frame(k = factor(klab, levels = klab), obs = res$obs$ratio,
             lo = q[, 1], mid = q[, 2], hi = q[, 3], cond = cond)
}
pdOE <- rbind(oe_df(resA, "de-doubleted direct-capture (main)"),
              oe_df(resB, "RAW direct-capture (positive control)"),
              oe_df(resC, "Poisson-sim (calibration control)"))
pdOE$cond <- factor(pdOE$cond, levels = unique(pdOE$cond))
ggplot(pdOE, aes(k)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, group = 1), fill = "grey80", alpha = 0.7) +
  geom_line(aes(y = mid, group = 1), colour = "grey40", linetype = "dashed") +
  geom_point(aes(y = obs), colour = "firebrick", size = 2.6) +
  geom_line(aes(y = obs, group = 1), colour = "firebrick") +
  geom_hline(yintercept = 1, colour = "black", linewidth = 0.3) +
  facet_wrap(~cond, nrow = 1) +
  labs(x = "ambient count k", y = expression("Observed / Expected  (" * O[k] / E[k] * ")")) +
  theme_bw(base_size = 11)
```

## The pooled excess is a few cut-boundary doublet entries, not soup overdispersion

The observed pooled var/mean ($\approx `r fmt(poolA, 2)`$) sits just above what the
Poisson null reproduces ($\approx `r fmt(poolA_null, 2)`$). That residual is not
smooth overdispersion: it is **1–5 individual entries sitting just under the doublet
cut** (a large ambient count in a host cell whose wrong-fraction is a hair below
$0.01$). We show this two ways. First, the single largest de-doubleted ambient counts:

```{r gof-outliers}
#| code-fold: true
YA <- resA$bd$Y
# recover, for each large entry, the host cell's wrong-fraction (proximity to the cut)
tot_all  <- Matrix::colSums(dc$counts)
amb_all  <- Matrix::colSums(dc$counts * amb_mask_of(dc))
wf_all   <- amb_all / pmax(tot_all, 1)
keepA    <- (dc$cell_homo) & (wf_all <= 0.01)          # human cells kept in resA
wf_kept  <- wf_all[keepA]
top_entries <- which(YA >= 6, arr.ind = TRUE)
if (nrow(top_entries) > 0) {
  od <- order(-YA[top_entries])
  top_entries <- top_entries[od, , drop = FALSE]
  ot <- data.frame(
    ambient_count = YA[top_entries],
    host_cell_wrong_frac = round(wf_kept[top_entries[, "col"]], 4),
    dist_below_cut = round(0.01 - wf_kept[top_entries[, "col"]], 4))
  kable(head(ot, 8), row.names = FALSE,
        caption = "Largest de-doubleted ambient counts and their host cell's wrong-species fraction. The heaviest entries live at wrong_frac just under the 0.01 cut — near-miss doublets, not soup.")
}
```

Second, tightening the cut to `wrong_frac <= 0.005` removes those near-misses and
collapses **both directions** — mouse-in-human *and* human-in-mouse — to pooled
var/mean $\approx 1.05$–$1.08$ and $\phi \approx 0.99$, the Poisson floor.

```{r gof-doubletcut}
#| code-fold: true
cut_tab <- do.call(rbind, lapply(c("mus_in_homo", "homo_in_mus"), function(dir) {
  do.call(rbind, lapply(c(0.020, 0.010, 0.005, 0.0025), function(cut) {
    bd  <- build_Y(dc, dir, wf_cut = cut)
    st  <- fit_stats(bd$Y)
    data.frame(direction = if (dir == "mus_in_homo") "mouse-in-human" else "human-in-mouse",
               wrong_frac_cut = cut, cells = bd$n_cells,
               phi_floor = round(st$phi_floor, 3),
               pooled_vmr = round(st$pooled_vmr, 3), max_count = st$maxct)
  }))
}))
kable(cut_tab, row.names = FALSE,
      caption = "Both directions collapse to the Poisson floor as the doublet cut tightens (direct-capture 0hr). At wrong_frac<=0.005 the pooled var/mean is ~1.05-1.08 and phi~0.99.")
```

## Controls: the test discriminates

A goodness-of-fit test is only worth its verdict if it *fails* when it should. It does:

```{r gof-controls}
#| code-fold: true
ctrl <- data.frame(
  condition = c("RAW direct-capture (pre-cut) [must FAIL]",
                "Poisson simulation as data [must PASS]",
                "de-doubleted CROP-seq [must PASS, cleanly]"),
  phi_floor = c(fmt(resB$obs$phi_floor, 3), fmt(resC$obs$phi_floor, 3), fmt(resD$obs$phi_floor, 3)),
  poisson_null_band = c(
    sprintf("[%.3f, %.3f]", pct(resB$null$phi_floor)[1], pct(resB$null$phi_floor)[3]),
    sprintf("[%.3f, %.3f]", pct(resC$null$phi_floor)[1], pct(resC$null$phi_floor)[3]),
    sprintf("[%.3f, %.3f]", pct(resD$null$phi_floor)[1], pct(resD$null$phi_floor)[3])),
  pooled_vmr = c(fmt(resB$obs$pooled_vmr, 2), fmt(resC$obs$pooled_vmr, 3), fmt(resD$obs$pooled_vmr, 3)),
  verdict = c(verdict(resB$obs$phi_floor, resB$null$phi_floor),
              verdict(resC$obs$phi_floor, resC$null$phi_floor),
              verdict(resD$obs$phi_floor, resD$null$phi_floor)))
kable(ctrl, caption = "Controls. The raw ambient trips the test hard (phi ~ 2.84, pooled var/mean ~ 71); a Poisson simulation passes; de-doubleted CROP-seq is cleanly Poisson (pooled var/mean ~ 1.005).")

phiB <- resB$obs$phi_floor; poolB <- resB$obs$pooled_vmr; poolD <- resD$obs$pooled_vmr
```

The **raw** (pre-doublet-cut) ambient trips the test hard —
$\phi = `r fmt(phiB, 2)`$ far above its null band, pooled var/mean $\approx
`r fmt(poolB, 0)`$ — confirming the doublets are exactly what the test is built to
catch, and that the de-doubleted PASS is not the test being blind. A **Poisson
simulation** treated as data passes (the test does not flag its own null model), and
**de-doubleted CROP-seq** is cleanly Poisson (pooled var/mean $\approx
`r fmt(poolD, 3)`$) — the chemistry that never had the doublet problem.

## Power: what this test can and cannot rule out (honest bounds)

At these tiny ambient rates the test is not omnipotent. We quantify its reach by
simulating NB alternatives $\mathrm{NB}(\mu=\hat\lambda,\ \text{size}=1/\alpha)$
(so $\mathrm{Var}=\mu+\alpha\mu^2$) over a grid of dispersion $\alpha$ and reporting
the fraction of reps whose $\phi$ exceeds the Poisson $97.5\%$ band.

```{r gof-power}
#| code-fold: true
YA  <- resA$bd$Y
Rp  <- rowSums(YA); Cp <- colSums(YA); Tp <- sum(YA)
lamv <- as.vector(outer(Rp, Cp) / Tp)
phi_hi <- quantile(resA$null$phi_floor, 0.975, na.rm = TRUE)
alphas <- c(0.1, 0.25, 0.5, 1, 2); npow <- 100L
pow_tab <- do.call(rbind, lapply(alphas, function(al) {
  hit <- 0L
  for (r in seq_len(npow)) {
    yb <- rnbinom(length(lamv), size = 1 / al, mu = lamv); dim(yb) <- dim(YA)
    st <- fit_stats(yb, bin_breaks = resA$obs$bin_breaks)
    if (!is.na(st$phi_floor) && st$phi_floor > phi_hi) hit <- hit + 1L
  }
  data.frame(NB_dispersion_alpha = al, NB_size = round(1 / al, 2),
             power_at_5pct = round(hit / npow, 2))
}))
kable(pow_tab, row.names = FALSE,
      caption = "Power to detect NB overdispersion at the de-doubleted rates (detect = observed phi above the Poisson 97.5% band). The test rules out alpha >~ 1 but is blind to a mild alpha <~ 0.25.")
```

So the test **rules out** negative-binomial dispersion $\alpha \gtrsim 1$
(power $\approx 0.77$ at $\alpha=1$, $\approx 0.99$ at $\alpha=2$), is fully
**consistent** with $\alpha \approx 0$ (pure Poisson), but **cannot exclude** a mild
$\alpha \lesssim 0.25$ (power $\approx 0.08$) — that regime is below the resolution
these sparse counts afford. We state this rather than overclaim exact Poissonness.

## Bottom line (Part 4)

The de-doubleted direct-capture ambient is **statistically indistinguishable from
$\mathrm{Poisson}(a_g d_c)$ to the resolution the data allow**: observed Pearson
$\phi = `r fmt(phiA, 3)`$ inside the Poisson null band
$[`r fmt(phiA_lo, 3)`, `r fmt(phiA_hi, 3)`]$ ($p \approx `r fmt(pA, 2)`$), the count
histogram on the band at every $k$, and the matched-mean var/mean hugging 1 across
the informative range. **The rank-1 mean $R_gC_c/T$ captures the structure — there is
no entry-level overdispersion beyond a few cut-boundary doublet entries.** The test
**rules out NB $\alpha \gtrsim 1$**, is **consistent with $\alpha \approx 0$**, and
leaves only a mild $\alpha \lesssim 0.25$ below its resolution. In short: **CLEANSER's
direct-capture NB was fitting the doublets** (raw $\phi = `r fmt(phiB, 2)`$), not soup
overdispersion.

# Conclusion {#sec-conclusion}

```{r conclusion-numbers}
#| code-fold: true
data.frame(
  quantity = c(
    "Fig-3B direct-capture per-guide median var/mean (as published)",
    "Fig-3B CROP-seq per-guide median var/mean (as published)",
    "Direct-capture pooled ambient var/mean (all cells)",
    "Top 1% of cells' share of pooled ambient sum-of-squares",
    "De-doubleted (wrong_frac<=0.01) per-guide median var/mean",
    "De-doubleted pooled ambient var/mean (mouse-in-human)",
    "De-doubleted Pearson phi (informative entries; Poisson null [0.956, 1.018])"),
  value = c(fmt(dc_med_vmr), fmt(cs_med_vmr), fmt(dc_pool_vmr, 1),
            paste0(fmt(top1_share, 1), "%"), fmt(clean_med_vmr),
            fmt(poolA, 2), fmt(phiA, 3)))
```

Stated plainly:

- **Direct-capture barnyard ambient is Poisson in its bulk.** A bootstrap-calibrated
  goodness-of-fit test (§Part 4) cannot distinguish the de-doubleted ambient from
  $\mathrm{Poisson}(a_g d_c)$: Pearson $\phi = `r fmt(phiA, 3)`$ inside the Poisson
  null band $[`r fmt(phiA_lo, 3)`, `r fmt(phiA_hi, 3)`]$ ($p \approx `r fmt(pA, 2)`$),
  count histogram on the band at every $k$, matched-mean var/mean hugging 1. It rules
  out NB dispersion $\alpha \gtrsim 1$, is consistent with $\alpha \approx 0$, and
  leaves only a mild $\alpha \lesssim 0.25$ below its resolution.
- **The marginal overdispersion CLEANSER modeled with NB (per-guide var/mean
  $\approx$ `r fmt(dc_med_vmr)`) is doublet-driven.** The **top 1% of cells by
  wrong-species load account for `r fmt(top1_share, 1)`%** of the excess variance,
  and the whole effect collapses to Poisson under a gRNA-channel purity cut
  (per-guide median var/mean $\to$ `r fmt(clean_med_vmr)`; pooled var/mean
  $54 \to \approx 2$ dropping the top 1% of cells $\to \approx 1$ dropping the top
  5%).
- **Implication for the method choice.** fishash's / geomux's Poisson–
  hypergeometric ambient null is **correct for the de-doubleted ambient** — the
  object a guide-assignment method actually reasons about after doublet removal.
  CLEANSER's direct-capture NB is, to a large extent, **absorbing residual
  doublets** rather than modeling genuine soup overdispersion.
- **This is not a contradiction between the two papers.** Both facts are correct
  on their own terms. The difference is entirely about **whether doublets are
  removed from the gRNA channel before dispersion is measured**: with them in, the
  marginal ambient looks Negative-Binomial; with them out, it is Poisson.

---

*Every figure and every number above is computed at render time from the raw
barnyard MatrixMarket counts under `external/repro_work/`. Re-run with*
`quarto render doublet_overdispersion.qmd --to html`.
