---
title: "sims for defense final"
output: html_document
date: "2026-04-11"
---




idea: fit a NB-NB using cleanser to get baseline parameters for a NB-NB mixure. Then simulate from this, except use a normalized grna_total to add a covariate effect.

- this effect will be constant for each cell we simulate, so if we stack all sims in one matrix, then the methods that can learn it will have a chance to. 



```{r}
rm(list=ls())
library(tidyverse)
source("~/.Rprofile")


# helper: log1p transform in base 10
logp1 <- function(x, base = 10) {
  log(x + 1, base = base)
}

# helper: build mixed exact / exponentially growing bins
# bins are:
#   0, 1, ..., k_exact as singleton bins
# then widths 2, 4, 8, 16, ...
make_mixed_bin_info <- function(max_x, k_exact = 10L) {
  stopifnot(length(max_x) == 1, is.finite(max_x), max_x >= 0)
  
  upper <- 0:k_exact
  
  width <- 2L
  while (tail(upper, 1L) < max_x) {
    upper <- c(upper, tail(upper, 1L) + width)
    width <- width * 2L
  }
  
  lower <- c(0L, head(upper, -1L) + 1L)
  
  label <- ifelse(
    lower == upper,
    as.character(upper),
    paste0(lower, "-", upper)
  )
  
  data.frame(
    bin_id = seq_along(label),
    lower = lower,
    upper = upper,
    label = label,
    xmin = seq_along(label) - 0.45,
    xmax = seq_along(label) + 0.45,
    stringsAsFactors = FALSE
  )
}

# helper: assign each count to one mixed bin
bin_with_info <- function(x, bin_info) {
  cut(
    x,
    breaks = c(-1, bin_info$upper),
    labels = bin_info$label,
    right = TRUE
  )
}

# helper: choose readable y-axis breaks on the original count scale
make_y_breaks <- function(max_count) {
  breaks <- c(
    0, 1, 2, 3, 5,
    10, 20, 30, 50,
    100, 200, 300, 500,
    1000, 2000, 3000, 5000,
    10000, 20000, 30000, 50000,
    100000, 200000, 300000, 500000,
    1000000, 2000000, 3000000, 5000000,
    10000000
  )
  breaks[breaks <= max_count]
}

# helper: exact binned counts for one vector
make_binned_counts <- function(x, bin_info) {
  stopifnot(all(is.finite(x)), all(x >= 0), all(x == floor(x)))
  
  bin <- factor(
    bin_with_info(x, bin_info),
    levels = bin_info$label
  )
  
  out <- data.frame(bin = bin) |>
    dplyr::count(bin, .drop = FALSE, name = "count") |>
    dplyr::mutate(bin = as.character(bin)) |>
    dplyr::left_join(bin_info, by = c("bin" = "label")) |>
    dplyr::arrange(bin_id)
  
  stopifnot(sum(out$count) == length(x))
  out
}

# helper: declutter x-axis labels
# label all singleton bins up to k_exact, then every `tail_every`-th tail bin, plus the last bin
make_x_axis_spec <- function(bin_info, k_exact, tail_every = 2L) {
  is_exact <- bin_info$lower == bin_info$upper & bin_info$upper <= k_exact
  tail_idx <- which(!is_exact)
  
  keep <- is_exact
  if (length(tail_idx) > 0) {
    keep[tail_idx[seq(1, length(tail_idx), by = tail_every)]] <- TRUE
    keep[max(tail_idx)] <- TRUE
  }
  
  list(
    breaks = bin_info$bin_id[keep],
    labels = bin_info$label[keep]
  )
}

# real-data plot
make_plt1a <- function(grna_real, bin_info, k_exact = 10L, tail_every = 2L, title = "real") {
  real_counts <- make_binned_counts(grna_real, bin_info)
  
  y_breaks <- make_y_breaks(max(real_counts$count))
  y_breaks_t <- logp1(y_breaks)
  x_axis <- make_x_axis_spec(bin_info, k_exact = k_exact, tail_every = tail_every)
  
  plot_df <- real_counts |>
    dplyr::mutate(
      ymin = 0,
      ymax = logp1(count)
    )
  
  ggplot2::ggplot(plot_df) +
    ggplot2::geom_rect(
      ggplot2::aes(
        xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax
      ),
      fill = "grey70",
      color = "black"
    ) +
    ggplot2::scale_x_continuous(
      breaks = x_axis$breaks,
      labels = x_axis$labels,
      expand = ggplot2::expansion(add = 0.6)
    ) +
    ggplot2::scale_y_continuous(
      breaks = y_breaks_t,
      labels = scales::label_number(big.mark = ",")(y_breaks),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::labs(
      title = title,
      x = "gRNA count bin",
      y = "cell count"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
    )
}

make_plt1b <- function(grna_sim, is_pert, bin_info, k_exact = 10L, tail_every = 2L, title = "sim") {
  stopifnot(length(grna_sim) == length(is_pert))
  stopifnot(all(is.finite(grna_sim)), all(grna_sim >= 0), all(grna_sim == floor(grna_sim)))
  
  sim_df <- data.frame(
    umi = grna_sim,
    pert_status = factor(
      ifelse(is_pert == 1, "pert", "not pert"),
      levels = c("not pert", "pert")
    )
  )
  
  sim_df$bin <- factor(
    bin_with_info(sim_df$umi, bin_info),
    levels = bin_info$label
  )
  
  sim_counts <- sim_df |>
    dplyr::count(bin, pert_status, .drop = FALSE, name = "count") |>
    dplyr::mutate(bin = as.character(bin)) |>
    dplyr::left_join(bin_info, by = c("bin" = "label")) |>
    dplyr::arrange(bin_id, pert_status) |>
    dplyr::group_by(bin, bin_id, lower, upper, xmin, xmax) |>
    dplyr::mutate(
      y0 = dplyr::lag(cumsum(count), default = 0),
      y1 = cumsum(count),
      ymin = logp1(y0),
      ymax = logp1(y1)
    ) |>
    dplyr::ungroup()
  
  stopifnot(sum(sim_counts$count) == length(grna_sim))
  if ("0" %in% sim_counts$bin) {
    stopifnot(
      sum(sim_counts$count[sim_counts$bin == "0"]) == sum(grna_sim == 0)
    )
  }
  
  y_breaks <- make_y_breaks(max(sim_counts$y1))
  y_breaks_t <- logp1(y_breaks)
  x_axis <- make_x_axis_spec(bin_info, k_exact = k_exact, tail_every = tail_every)
  
  ggplot2::ggplot(sim_counts) +
    ggplot2::geom_rect(
      ggplot2::aes(
        xmin = xmin, xmax = xmax,
        ymin = ymin, ymax = ymax,
        fill = pert_status
      ),
      color = "black"
    ) +
    ggplot2::scale_x_continuous(
      breaks = x_axis$breaks,
      labels = x_axis$labels,
      expand = ggplot2::expansion(add = 0.6)
    ) +
    ggplot2::scale_y_continuous(
      breaks = y_breaks_t,
      labels = scales::label_number(big.mark = ",")(y_breaks),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::labs(
      title = title,
      x = "gRNA count bin",
      y = NULL,
      fill = NULL
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1),
      axis.title.y = ggplot2::element_blank(),
      legend.position = c(0.98, 0.98),
      legend.justification = c(1, 1),
      legend.background = ggplot2::element_rect(fill = "white", color = "black"),
      legend.key = ggplot2::element_rect(fill = "white", color = NA)
    )
}
# wrapper: build shared bins from both datasets and return both plots
make_umi_hists <- function(grna_real, grna_sim, is_pert, k_exact = 10L, tail_every = 2L, title_real = "real", title_sim = "sim", return_plots = FALSE) {
  stopifnot(length(grna_sim) == length(is_pert))
  stopifnot(all(grna_real >= 0), all(grna_sim >= 0))
  stopifnot(all(grna_real == floor(grna_real)), all(grna_sim == floor(grna_sim)))
  
  bin_info <- make_mixed_bin_info(
    max_x = max(c(grna_real, grna_sim)),
    k_exact = k_exact
  )
  
  plt1a <- make_plt1a(
    grna_real = grna_real,
    bin_info = bin_info,
    k_exact = k_exact,
    tail_every = tail_every,
    title = title_real
  )
  
  plt1b <- make_plt1b(
    grna_sim = grna_sim,
    is_pert = is_pert,
    bin_info = bin_info,
    k_exact = k_exact,
    tail_every = tail_every,
    title = title_sim
  )
  
  plots <- list(
    plt1a = plt1a,
    plt1b = plt1b
  )
  if(return_plots) {
    return(plots)
  } else {
    print(cowplot::plot_grid(plt1a, plt1b, nrow = 1, align = "h"))
  }
  invisible()
}





source_data <- "replogle-rd7"
# source_data <- "gasperini"
path_to_data <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data", source_data,
  "sceptre-pipeline"
)
cat("Using", source_data, "\n")
grna_odm <- ondisc::initialize_odm_from_backing_file(
  file.path(path_to_data, "grna.odm")
)
cat("grna.odm :", nrow(grna_odm), "guides ;", ncol(grna_odm), "cells.\n")

scep <- sceptre::read_ondisc_backed_sceptre_object(
  sceptre_object_fp = file.path(path_to_data, "sceptre_object.rds"),
  response_odm_file_fp = file.path(path_to_data, "response.odm"),
  grna_odm_file_fp = file.path(path_to_data, "grna.odm")
)

cell_covariates = scep@covariate_data_frame

grna_summary_stats <- read_csv(file.path(path_to_data, "grna_summary_stats.csv"), show_col_types = FALSE)
  # mutate(grna=rownames(grna_odm))



# guides_to_use = grna_summary_stats |>
#   filter(
#     grna_nonzero > quantile(grna_nonzero, .89),
#     grna_nonzero < quantile(grna_nonzero, .91)
#   ) 

guides_to_use = grna_summary_stats |>
  filter(
    grna_nonzero > quantile(grna_nonzero, .95),
    grna_nonzero < quantile(grna_nonzero, .951 )
  ) 

set.seed(1)
guides_to_use |>
  sample_n(1)

# guide = c("replogle-rd7" = "2685_EPS8L1_P2_ENSG00000131037", "gasperini" = "CCGTAGCTCCAGAGAAACAG")[source_data]
guide = c("replogle-rd7" = "306_ANAPC1_P1P2_ENSG00000153107", "gasperini" = "CCGTAGCTCCAGAGAAACAG")[source_data]
y = grna_odm[guide]

data.frame(
  umi=grna_odm[guide,]
) |>
  # filter(umi > 0) |>
  ggplot(aes(x = umi)) +
  geom_histogram(bins=50) +
  theme_bw() +
  scale_y_continuous(transform = scales::pseudo_log_trans()) +
    scale_x_continuous(transform = scales::pseudo_log_trans()) +
    ggtitle(guide)
cat("Using", source_data, "\n")
```


```{r}
guides <- c( "306_ANAPC1_P1P2_ENSG00000153107", "2685_EPS8L1_P2_ENSG00000131037", "2_AAAS_P1P2_ENSG00000094914")
data.frame(
  guide_id = rep(1:3, each = ncol(grna_odm)),
  umi = c(grna_odm[guides[1],], grna_odm[guides[2],], grna_odm[guides[3],])
) |>
  mutate(guide_id = factor(guide_id)) |>
  ggplot(aes(x = umi, fill = guide_id)) +
  geom_histogram(bins=30) +
  theme_bw() +
  scale_y_continuous(transform = scales::pseudo_log_trans()) +
    scale_x_continuous(transform = scales::pseudo_log_trans()) +
  ggtitle("Replogle guides of varying sequence depth")



```


```{r}
sim_data <- function(params, n = NULL, cell_scaling = NULL, seed=1, require_pert_positive=TRUE) {
  set.seed(seed)
  if(is.null(n) && is.null(cell_scaling)) {
    stop("at least one of `n` and `cell_scaling` must be provided.")
  }
  if(!is.null(n) && !is.null(cell_scaling)) {
    if(length(n) != length(cell_scaling)) {
      stop("`n` and `cell_scaling` must have the same length.")
    }
  }
  if(is.null(cell_scaling)) {
    cell_scaling <- 1
  }
  if(is.null(n)) {
    n <- length(cell_scaling)
  }
  # if(any(cell_scaling == 0)) {
  #   stop("`cell_scaling` has exact 0s.")
  # }
  
  z <- rbinom(n = n, size = 1, prob = params["pi"])
  is_pert = z == 1
  num_pert = sum(is_pert)
  num_nonpert = n - num_pert
  cell_scaling_pert = cell_scaling[is_pert]
  
  # 1. simulate perturbed cells from NB; optionally conditioned on being positive
  ysim_pos = MASS::rnegbin(
    n = num_pert,
    mu = params["mu_pert"] * cell_scaling_pert,
    theta = params["theta_pert"]
  )
  if(require_pert_positive) {
    pert_is_zero = ysim_pos == 0 & cell_scaling_pert > 0
    while(sum(pert_is_zero) > 0) {
      y_new <- MASS::rnegbin(
        n = sum(pert_is_zero),
        mu = params["mu_pert"] * cell_scaling_pert[pert_is_zero],
        theta = params["theta_pert"]
      )
      ysim_pos[pert_is_zero] <- y_new
      pert_is_zero = ysim_pos == 0 & cell_scaling_pert > 0
    }
  }
  
  # 2. simulate non-pert from NB; optionally, zero inflate
  ysim_neg <- MASS::rnegbin(
    n = num_nonpert,
    mu = params["mu_np"] * cell_scaling[z == 0],
    theta = params["theta_np"]
  )
  # if TRUE, zero inflate
  # phi gives the overall probabilty of being zero
  if("phi" %in% names(params)) { 
    # this is found by solving phi = P(Y=0) for the prob to zero out
    # we have P(Y=0) = P(Y=0|Z=0)P(Z=0)
    # = P(NB = 0 OR zero out) (1 - pi) 
    prob_nonpert_nb_is_zero = exp(params["theta_np"] * log(
      params["theta_np"] / (params["theta_np"] + params["mu_np"])
    ))
    prob_to_zero_out <- (params["phi"] / (1 - params["pi"]) - prob_nonpert_nb_is_zero) / 
      (1 - prob_nonpert_nb_is_zero)
    to_zero_out <- rbinom(num_nonpert, 1, prob_to_zero_out)
    ysim_neg[to_zero_out == 1] <- 0
  }
  
  # collect
  ysim <- numeric(n)
  ysim[z == 0] <- ysim_neg
  ysim[z == 1] <- ysim_pos
    
  list(ysim=ysim, z=z)
}
```


```{r}
# params_from_cleanser = c(
#   pi = .00558,
#   mu_pert = 1128.62,
#   theta_pert = 1.0251,
#   mu_np = 0.074444,
#   theta_np = 2.36718
# )

# params_well_sep <- c(
#   pi = .001,
#   mu_pert = 1000,
#   theta_pert = .75,
#   mu_np = 0.03,
#   theta_np = .1
# )
params_well_sep <- c(
  pi = .002,
  mu_pert = 800,
  theta_pert = .5,
  mu_np = 0.03,
  theta_np = .1
)

V0 <- params_well_sep["mu_pert"] + params_well_sep["mu_pert"]^2 / params_well_sep["theta_pert"]

mu_pert_seq <- seq(params_well_sep["mu_pert"], params_well_sep["mu_np"] * 1.5, length.out = 20)  # shrink toward ambient
theta_pert_seq <- mu_pert_seq^2 / (V0 - mu_pert_seq)
i=14  # 6
# params_overlap <- c(
#   pi = .002,
#   mu_pert = mu_pert_seq[i],
#   theta_pert = theta_pert_seq[i],
#   mu_np = 0.03,
#   theta_np = .1
# )
params_overlap <- c(
  pi = .002,
  mu_pert = mu_pert_seq[i],
  theta_pert = theta_pert_seq[i],
  mu_np = 0.03,
  theta_np = .1
)

seed_ = 8
set.seed(seed_+1)
n = 50000
locs = sample(ncol(grna_odm), n)
grna_nonzero_locs = cell_covariates$grna_n_nonzero[locs]
grna_nonzero_normed_locs = grna_nonzero_locs / mean(grna_nonzero_locs)
y_curr = y[locs]

sims <- sim_data(params = params_well_sep, cell_scaling = grna_nonzero_normed_locs, seed=seed_+2)

make_umi_hists(grna_real = y_curr, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_curr), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_curr == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)
```



```{r}
# params_list = list(
#   most_sep = c(
#     pi = .002,
#     mu_pert = 800,
#     theta_pert = .5,
#     mu_np = 0.03,
#     theta_np = .1
#   ),
#   some_sep = c(
#     pi = .002,
#     mu_pert = 600,
#     theta_pert = .25,
#     mu_np = 0.03,
#     theta_np = .1
#   ),
#   least_sep = c(
#     pi = .002,
#     mu_pert = 250,
#     theta_pert = 0.05,
#     mu_np = 0.03,
#     theta_np = .1
#   )
# )

# pi_ = 0.002
# params_list = list(
#   a1 = c(
#     pi = pi_,
#     mu_pert = 800,
#     theta_pert = .75,
#     mu_np = 0.03,
#     theta_np = .1
#   ),
#   # some_sep = c(
#   #   pi = .002,
#   #   mu_pert = 600,
#   #   theta_pert = .25,
#   #   mu_np = 0.03,
#   #   theta_np = .1
#   # ),
#   # some_sep = c(
#   #   pi = pi_,
#   #   mu_pert = 250,
#   #   theta_pert = 0.05,
#   #   mu_np = 0.03,
#   #   theta_np = .1
#   # ),
#   # least_sep = c(
#   #   pi = pi_,
#   #   mu_pert = 50,
#   #   theta_pert = 0.005 ,
#   #   mu_np = 0.03,
#   #   theta_np = .1
#   # )
#       a2 = c(
#     pi = pi_,
#     mu_pert = 400,
#     theta_pert = .75,
#     mu_np = 0.03,
#     theta_np = .1
#   ),
#     a3 = c(
#     pi = pi_,
#     mu_pert = 200,
#     theta_pert = .75,
#     mu_np = 0.03,
#     theta_np = .1
#   ),
#   a4 = c(
#     pi = pi_,
#     mu_pert = 100,
#     theta_pert = .75 ,
#     mu_np = 0.03,
#     theta_np = .1
#   ),
#     a5 = c(
#     pi = pi_,
#     mu_pert = 50,
#     theta_pert = .75 ,
#     mu_np = 0.03,
#     theta_np = .1
#   )
# )


## these are from cleanser
params_base <- c(
  pi = 0.004,
  mu_pert = 971.437,
  theta_pert = 0.126,
  mu_np = 0.066,
  theta_np = 2.26
)

## modifying
# params_base["pi"] = 0.001


## sparsity adjusted
# params_base <- c(
#   pi = 0.001,
#   mu_pert = 971.437,
#   theta_pert = 0.126,
#   mu_np = 0.066,
#   theta_np = 2.26
# )

# params_base <- c(pi = 0.00052, mu_pert = 2323.33019, theta_pert = 0.8487, mu_np = 0.021, 
# theta_np = 0.06091)

## GLM fit depending on if it's above or below the threshold
# params_base <- c(
#   pi = 0.001,
#   mu_pert = exp(7.098504),
#   theta_pert = 0.8800,
#   mu_np = exp(-4.6753847),
#   theta_np = 0.15536
# )

## GLM fit with increased overdispersion
# params_base <- c(
#   pi = 0.002,
#   mu_pert = exp(7.098504),
#   theta_pert = 0.8800,
#   mu_np = exp(-4.6753847) * 3,
#   theta_np = 0.15536 
# )

## GLM fit with no thresh for NP 
# params_base <- c(
#   pi = 0.002,
#   mu_pert = exp(7.098504),
#   theta_pert = 0.8800,
#   mu_np = exp(-0.830237),
#   theta_np = 0.0034521 
# )


mu_pert_seq <- params_base["mu_pert"] / 2^(0:4)
params_list <- lapply(mu_pert_seq, function(mu) {
  curr_params = params_base
  curr_params["mu_pert"] = mu
  curr_params
}) |>
  # set_names("params = CLEANSER posterior mean")
  set_names(letters[1L:length(mu_pert_seq)])



n = 50000
seed_base <- 1
pltlist <- list()
for(param_idx in seq_along(params_list)) {
  seed_ = param_idx + seed_base
  set.seed(seed_+1)

  locs = sample(ncol(grna_odm), n)
  grna_nonzero_locs = cell_covariates$grna_n_nonzero[locs]
  grna_nonzero_normed_locs = grna_nonzero_locs / mean(grna_nonzero_locs) * 2
  y_curr = y[locs]
  
  sims <- sim_data(params = params_list[[param_idx]], cell_scaling = grna_nonzero_normed_locs, seed=seed_+2, require_pert_positive = FALSE)
  print(c(max(sims$ysim), mean(sims$ysim[sims$z == 1] == 0)))
  plts <- make_umi_hists(grna_real = y_curr, grna_sim = sims$ysim, is_pert = sims$z, return_plots = TRUE, title_sim = names(params_list)[param_idx])
  if(param_idx == 1) {
    pltlist[["real"]] = plts$plt1a
    pltlist[["sim1"]] = plts$plt1b
  } else {
    pltlist[[paste0("sim", param_idx)]] = plts$plt1b
  }
}

cowplot::plot_grid(plotlist = pltlist)

```


```{r}


est_params_via_thresholding <- function(y, thresh, L = NULL) {
  
  if(!is.null(L)) {
    y <- y[L > 0]
    L <- L[L > 0]
    y <- y / L
  } 
  y_np <- y[y <= thresh]
  y_p <- y[y > thresh]
  
  params <- c(
    pi = mean(y > thresh),
    mu_pert = mean(y_p),
    theta_pert = mean(y_p)^2 / (var(y_p) - mean(y_p)),
    mu_np = mean(y_np),
    theta_np = mean(y_np)^2 / (var(y_np) - mean(y_np))
  )
  params
}
params_base <- est_params_via_thresholding(y, 10, L = cell_covariates$grna_n_nonzero / mean(cell_covariates$grna_n_nonzero))


df = data.frame(
  y = y,
  x = cell_covariates$grna_n_nonzero 
) |>
  filter(x > 0)
mod_p = MASS::glm.nb(y ~ log(x), data = df[df$y > 10,], ); mod_p |> summary()
mod_np = MASS::glm.nb(y ~ log(x), data = df[df$y <= 10,], ); mod_np |> summary()
mod_all = MASS::glm.nb(y ~ log(x), data = df); mod_all |> summary()



params_glm <- c(
  pi = mean(df$y > 10),
  mu_pert = exp(5.63783),
  theta_pert = 0.9199,
  mu_np  =exp(-6.67555 ),
  theta_np = 0.16719
)

# sim_data_glm <- function(pi, mod_pert, mod_nonpert, y, x, seed=1) {


fit_glm_sims <- function(y, x, thresh) {
  
  get_mod_params <- function(mod, param_name_tag) {
    coefs_ <- coef(mod) |> as.numeric()  # strip names
    params <- c(
      beta0 = coefs_[1],
      log1p_x = coefs_[2],
      theta = summary(mod)$theta
    )
    names(params) <- paste0(names(params), "_", param_name_tag)
    params
  }
  
  df = data.frame(
    y = y,
    x = x
  )
  nsim = nrow(df)
   mod_pert = MASS::glm.nb(y ~ log1p(x), data = df[df$y > thresh,])
  mod_nonpert = MASS::glm.nb(y ~ log1p(x), data = df[df$y <= thresh,])
  
  
  
  params <- c(
    pi = mean(y > thresh),
    get_mod_params(mod_pert, "pert"),
     get_mod_params(mod_nonpert, "nonpert")
  )
  params
}

make_sims_glm <- function(glm_sim_params, xsim) {
  
  nsim = length(xsim)
  z = rbinom(nsim, 1, glm_sim_params["pi"])
  
  mu_pert = exp(glm_sim_params["beta0_pert"] + log1p(xsim[z == 1]) * glm_sim_params["log1p_x_pert"])
  mu_nonpert = exp(glm_sim_params["beta0_nonpert"] + log1p(xsim[z == 0]) * glm_sim_params["log1p_x_nonpert"])
  
  
  ysim = integer(nsim)
  ysim[z == 1] <- MASS::rnegbin(
    sum(z == 1),
    mu_pert,
    glm_sim_params["theta_pert"]
  )
    ysim[z == 0] <- MASS::rnegbin(
    sum(z == 0),
    mu_nonpert,
    glm_sim_params["theta_nonpert"]
  )
  list(ysim=ysim, z=z)
}
glm_sim_params_ <- fit_glm_sims(y = grna_odm[guide,], x = cell_covariates$grna_n_nonzero, thresh = 10)





n = 200000
locs_glm = sample(ncol(grna_odm), n)

sims_glm = make_sims_glm(glm_sim_params_, xsim = cell_covariates$grna_n_nonzero[locs_glm])

# sims_glm$ysim[sims_glm$z == 0] <- 2^(rpois(sum(sims_glm$z == 0), .07))

make_umi_hists(grna_real = grna_odm[guide,][locs_glm], grna_sim = sims_glm$ysim, is_pert = sims_glm$z)

```






```{r}
make_sims_resamp <- function(y, below_this_is_np, above_this_is_pert, pi, nsim) {
  y_pert = y[y >= above_this_is_pert]
  y_np <- y[y <= below_this_is_np]
  
  
  z <- rbinom(nsim, 1, pi)
  ysim = integer(nsim)
  ysim[z == 1] = sample(y_pert, sum(z), TRUE)
  ysim[z == 0] = sample(y_np, sum(1 - z), TRUE)
  
  list(ysim=ysim, z=z)
}


# sims_resamp = make_sims_resamp(y, below_this_is_np = 10, above_this_is_pert = 3, .001, 400000)
# make_umi_hists(grna_real = grna_odm[guide,][locs], grna_sim = sims_resamp$ysim, is_pert = sims_resamp$z, title_sim = "Resampling based on thresholds")


below_this_is_np = 10
above_this_is_pert_seq <- c(3,6,9,12,15)
pltlist <- list()
for(i in seq_along(above_this_is_pert_seq)) {
  sims_resamp = make_sims_resamp(y, below_this_is_np = 10, above_this_is_pert = above_this_is_pert_seq[i], .001, 400000)
  plts <- make_umi_hists(grna_real = grna_odm[guide,][locs], grna_sim = sims_resamp$ysim, is_pert = sims_resamp$z, title_sim = bquote(min[p] ~ "=" ~ .(above_this_is_pert_seq[i]) * "," ~  max[np] ~ "=" ~ .(below_this_is_np)), return_plots = TRUE)
  
  if(i == 1) {
    pltlist <- plts
  } else {
    pltlist <- c(pltlist, plts$plt1b)
  }
}
# names(pltlist)[-1] <- paste0()

cowplot::plot_grid(plotlist = pltlist)






```







ok, now i need to make my datasets.

my goal: i want to do 3 main-effects-only analyses.

- 3 points for overlap
- 3 points for sparsity
- 3 points for batch effect



idea:
- sparsity is .1%, 1%, 10%
- then 1% is my "middle" value. We expect 200 true positives, which feels like enough to learn from? hmmm.. well, no, actually, i'd like more.

maybe i shouldn't try to do a nice exponential grid...

what if i do this: .5%, 2.5%, 5%
then 2.5% is my middle point with 500 perturbed cells. 


hmmm... ok actually i think i really like .1%, 1%, 10%. 

so let's try to use 1% for my other datasets as the baseline. 200 samples is decent still. if only 1.5% of cells are non-zero, and most perturbed ones are, 200 of them should stick out very clearly!


middle point dataset:
- overlap = medium
- sparsity = 1%
- batch effect = 0

varying sparsity:
- low, to .1%
- high, to 10%

varying overlap:
- low, to near perfect separation
- high, to lots of overlap

varying batch:
- low = some effect
- large = big effect

--> only 7 different datasets if i can get this all to work nicely enough.


FIGURING OUT BASELINE

for now, let's ONLY be concerned with overlap.

```{r}
set.seed(123)
n_sim = 100000
locs = sample(which(cell_covariates$grna_n_nonzero > 0), n_sim)
x_downsamp = cell_covariates$grna_n_nonzero[locs]
x_downsamp_normed = x_downsamp / mean(x_downsamp)
y_downsamp = y[locs]
```


for these, i vary both P and NP groups
```{r}
# params_baseline <- c(
#   pi = .01,
#   mu_pert = 800,
#   theta_pert = .75,
#   mu_np = 0.01,
#   theta_np = .01
# )

params_minimal_overlap <- c(
  pi = .005,
  mu_pert = 800,
  theta_pert = 1,
  mu_np = 0.01,
  theta_np = .01
)

# this might be too much overlap?
params_some_overlap <- c(
  pi = .005,
  mu_pert = 500,
  theta_pert = .5,
  mu_np = 0.01,
  theta_np = .0075
)

params_most_overlap <- c(
  pi = .005,
  mu_pert = 100,
  theta_pert = .05,
  mu_np = 0.01,
  theta_np = .005
)

sims <- sim_data(params = params_most_overlap, cell_scaling = x_downsamp_normed, seed=6)



make_umi_hists(grna_real = y_downsamp, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_downsamp), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_downsamp == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)
```



but biologically, it makes more sense to only vary the perturbed group.

```{r}
grna_nonzero = cell_covariates$grna_n_nonzero
x_nonzero <- grna_nonzero[grna_nonzero > 0]
x_nonzero_normed <- x_nonzero / mean(x_nonzero)
y_n_nonzero <- y[grna_nonzero > 0]

# bootstrap
seed_ = 7
# set.seed(seed_ + 1)
# y_n_nonzero = sample(y_n_nonzero, length(y_n_nonzero), TRUE)
# here is a different way to get the distribution for the NP group
# mod = MASS::glm.nb(y ~ 1, data = data.frame(y = y_downsamp[y_downsamp < 5]))

# mod = MASS::glm.nb(y ~ log(x), data = data.frame(y = y_downsamp[y_downsamp < 10], x = x_downsamp[y_downsamp < 10]))
# summary(mod)

params_minimal_overlap <- c(
  pi = .005,
  mu_pert = 800,
  theta_pert = 1,
  mu_np = 0.01,
  theta_np = 0.1
  # phi = .9875
)

params_kinda_getting_it <- c(
  pi = .005,
  mu_pert = 800,
  theta_pert = 1,
  mu_np = 0.1,
  theta_np = 0.0075,
  phi = .994  # from mean(y == 0)
)


sims <- sim_data(params = params_minimal_overlap, cell_scaling = x_nonzero_normed, seed=seed_)



make_umi_hists(grna_real = y_n_nonzero, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_n_nonzero), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_n_nonzero == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)
```




well, i'm an idiot: i was ignoring the bayesian part of this and just doing NB samples. it's heavier tailed than that!!

let's try doing a poisson-log-normal 
```{r}
grna_nonzero = cell_covariates$grna_n_nonzero
x_nonzero <- grna_nonzero[grna_nonzero > 0]
x_nonzero_normed <- x_nonzero / mean(x_nonzero)
y_n_nonzero <- grna_odm["306_ANAPC1_P1P2_ENSG00000153107",][grna_nonzero > 0]


cleanser_samples = read.table("/home/josep/scratch/fit-mixture-models/cleanser-out/samples.txt", sep="\t", header = TRUE) |>
  filter(guide.id == 1)  |> # this is my chosen guide
  dplyr::select(-guide.id) |>
  as_tibble()

n_post <- nrow(cleanser_samples)

L <- x_nonzero_normed
set.seed(6)
nsim <- length(L)

# Sample one posterior draw per cell
idx <- sample(n_post, size = nsim, replace = TRUE)

r_s        <- 0.001 # cleanser_samples$r[idx]
mu_pert_s  <- cleanser_samples$mu[idx]
disp_pert_s <- cleanser_samples$Disp[idx]
mu_np_s    <- cleanser_samples$n_nbMean[idx] #/ 15
disp_np_s  <- cleanser_samples$n_nbDisp[idx] 


# can i try sampling from a heavier tailed dist for the disperion?



z <- rbinom(n = nsim, size = 1, prob = r_s)

ysim <- MASS::rnegbin(
  n = nsim,
  mu = ifelse(z == 1, mu_pert_s, mu_np_s) * L,
  theta = ifelse(z == 1, disp_pert_s, disp_np_s)
)

sims = list(ysim=ysim, z=z)

make_umi_hists(grna_real = y_n_nonzero, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_n_nonzero), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_n_nonzero == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)

```



ok let's try running cleanser with a real L on just this guide

```{r}
L <- grna_nonzero / mean(grna_nonzero)
counts <- grna_odm["306_ANAPC1_P1P2_ENSG00000153107"] # around 95% for n_nonzero
nonzero <- which(counts > 0)

df <- data.frame(
  count = as.integer(counts[nonzero]),
  L     = L[nonzero]
)
write.csv(df, "~/scratch/fit-mixture-models/guide-and-L-more-expressed.csv", row.names = FALSE)
```


```{r}
guide = "306_ANAPC1_P1P2_ENSG00000153107"
y = grna_odm[guide,]
L <- cell_covariates$grna_n_nonzero / mean(cell_covariates$grna_n_nonzero)

# y <- y[L > 0]



cleanser_samples = read.table("/home/josep/scratch/fit-mixture-models/cleanser-out/guide-with-real-L-more-expressed-samples.csv", sep="\t", header = TRUE) |>
  filter(guide.id == 1)  |> # this is my chosen guide
  dplyr::select(-guide.id) |>
  as_tibble()

# hist(cleanser_samples$r)

n_post <- nrow(cleanser_samples)

# L <- x_nonzero_normed  # i'm using same L
set.seed(5)
nsim <- length(L)

# Sample one posterior draw per cell
idx <- sample(n_post, size = nsim, replace = TRUE)

r_s        <- 0.001 # cleanser_samples$r[idx]
mu_pert_s  <-  1000 # cleanser_samples$mu[idx]
disp_pert_s <- .5 # cleanser_samples$Disp[idx]
mu_np_s    <- .03 # cleanser_samples$n_nbMean[idx]
disp_np_s  <- .1 # cleanser_samples$n_nbDisp[idx]

# r_s        <- 0.001 # rep(mean(cleanser_samples$r), nsim)
# mu_pert_s  <- rep(mean(cleanser_samples$mu), nsim)
# disp_pert_s <- rep(mean(cleanser_samples$Disp), nsim)
# mu_np_s    <- cleanser_samples$n_nbMean[idx]
# disp_np_s  <- 1/cleanser_samples$n_nbDisp[idx]

z <- rbinom(n = nsim, size = 1, prob = r_s)

ysim <- MASS::rnegbin(
  n = nsim,
  mu = ifelse(z == 1, mu_pert_s, mu_np_s) * L,
  theta = ifelse(z == 1, disp_pert_s, disp_np_s)
)

sims = list(ysim=ysim, z=z)

y_curr = y
make_umi_hists(grna_real = y_curr, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_curr), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_curr == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)
```























```{r}
set.seed(6)
idx <- sample(nrow(cleanser_samples), size = 1)
# params_draw <- cleanser_samples[idx, ]

params_draw <- list(
  r = 0.005, 
)

z <- rbinom(n = nsim, size = 1, prob = params_draw$r)

ysim <- MASS::rnegbin(
  n = nsim,
  mu = ifelse(z == 1, params_draw$mu, params_draw$n_nbMean) * L,
  theta = ifelse(z == 1, params_draw$Disp, params_draw$n_nbDisp)
)
sims = list(ysim=ysim, z=z)

make_umi_hists(grna_real = y_n_nonzero, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_n_nonzero), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_n_nonzero == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)
```



hmmmm.... i don't like this

let's try NB-log-N myself

```{r}


pi <- .005  # just fix this

mu_pert <- exp(rnorm(nsim, log(50), log(3)))
theta_pert = exp(rnorm(nsim, log(3), log(2)))

mu_np <- exp(rnorm(nsim, log(1.075), log(1.1)))
theta_np = exp(rnorm(nsim, log(1.1), log(1.1)))


z <- rbinom(nsim, 1, pi)

ysim <- MASS::rnegbin(
  n = nsim,
  mu = ifelse(z == 1, mu_pert, mu_np) * L,
  theta = ifelse(z == 1, theta_pert, theta_np)
)
sims = list(ysim=ysim, z=z)

make_umi_hists(grna_real = y_n_nonzero, grna_sim = sims$ysim, is_pert = sims$z)

cat(
  " num. pert =", sum(sims$z), "\n",
  " num. pert below max unpert:", sum(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])),
  "(", round(100 * mean(sims$ysim[sims$z == 1] < max(sims$ysim[sims$z == 0])), 2 ), "%)", "\n",
  " Real: max =", max(y_n_nonzero), "; Sim: max =", max(sims$ysim), "\n",
  "Real: % 0 =", round(100 * mean(y_n_nonzero == 0), 2), "Sim: % 0 =", round(100 * mean(sims$ysim == 0), 2), "\n"
)

```

