---
title: "simulation"
output: html_document
date: "2025-11-06"
---


```{r}
source("~/.Rprofile")
library(tidyverse)

base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data/gasperini/sceptre-pipeline/"
)


```

manual exploration

what are my covariates? (as labeled here)
- prep_batch
- grna_n_umis
- grna_n_nonzero

```{r}
library(ondisc)
library(sceptre)
scep <- read_ondisc_backed_sceptre_object(
  sceptre_object_fp = paste0(base_fp, "sceptre_object.rds"),
  response_odm_file_fp = paste0(base_fp, "response.odm"),
  grna_odm_file_fp = paste0(base_fp, "grna.odm")
)

scep@covariate_data_frame |> head()


grna_odm <- ondisc::initialize_odm_from_backing_file(paste0(base_fp, "grna.odm"))
```

determining which gRNA(s) to use as our target. Maybe give a pool of them and average effect sizes or something, to try to integrate out the dependence on this particular gRNA?
```{r, echo = FALSE, include = FALSE, eval = FALSE}

# getting helper function for loading the full matrix
source(file.path(
  .get_config_path("LOCAL_SCEPTRE3_REPO_DIR"),
  "benchmarking/guide-assignment/data-preprocessing/convert_odm_to_h5ad.R"
))
grna_matrix <- odm_to_R(grna_odm)

rowsums <- numeric(nrow(grna_matrix))
for(i in 1:nrow(grna_matrix)) {
  rowsums[i] <- sum(grna_matrix[i,])
}

# getting a highly active but not maximally active gRNA
grnas_to_use <- rowsums > quantile(rowsums, .9) & rowsums < quantile(rowsums, .99)
set.seed(12321)
which(grnas_to_use) |> sample(5, replace = FALSE) |> sort()
# the resulting idx are: 5034  7971  9796 10437 12669
```


```{r}

df <- scep@covariate_data_frame |>
  transmute(
    grna_n_nonzero, grna_n_umis, batch = ifelse(prep_batch == "prep_batch_2", 1, 0),
   y = grna_odm[5034,] # a highly but not maximally expressed gRNA, determined by previous code chunk
  )



# are these bad ... ? or should we use a transform that can handle 0?
(locs <- list(
which_n_nonzero = which(df$grna_n_nonzero == 0),
which_n_umis = which(df$grna_n_umis == 0)
))
df <- df |>
  dplyr::filter(grna_n_nonzero > 0, grna_n_umis > 0)
# both same

```


fit model:

our model is
$$
Y_i \sim \text{NB}_\theta(\mu_i) \quad \log \mu_i = \beta_0 + \beta_1 b_i  +\beta_2 \log \ell_i + \beta_3 \log m_i
$$
and 
$$
\operatorname{Var}[Y_i] = \mu_i + \mu^2/\theta
$$
```{r}
library(MASS)
# filtered to non-zero covariates so log defined everywhere
mod <- glm.nb(y ~ batch + log(grna_n_nonzero) + log(grna_n_umis), data = df)
summary(mod)
```



ok now we can sample
```{r}

num_samp = 50
gamma <- log(3)
sample_from_mod <- function(mod, cell_covs, num_samp, gamma = log(3), num_grnas = 1, prob_perturbed=0.02, seed=123) {
  set.seed(seed)

  # 1. get simulated cell covariates
  locs <- sample(x = nrow(cell_covs), size = num_samp, replace = TRUE)
  cell_covs_new <- cell_covs[locs,]
  log_mu <- predict(mod, newdata=cell_covs_new) 
  
  
  # 2. make each new grna
  for(i in seq_len(num_grnas)) {
    cell_is_pertubed <- rbinom(num_samp, 1, prob_perturbed)
    log_mu_new <- log_mu + gamma * cell_is_pertubed
    Y_new <- MASS::rnegbin(num_samp, mu = exp(log_mu_new), theta = mod$theta)
    cell_covs_new[paste0("sim_grna_", i)] <- Y_new
    cell_covs_new[paste0("is_perturbed_sim_grna_", i)] <- cell_is_pertubed
  }

  return(cell_covs_new)
}

set.seed(132)
n <- 1000
p <- 5
betas <- c(2, runif(p, -5, 5))
x <- cbind(1, matrix(rnorm(n*p),n,p))
log_mu <- x %*% betas
z <- MASS::rnegbin(n, mu = exp(log_mu), theta = 3.1)
fit <- MASS::glm.nb(z ~ x - 1)
fit$theta
```


do a test run with sceptre on this data

```{r}
num_grnas <- 5
grna_simulated <- sample_from_mod(
  mod = mod, df = df, num_samp = nrow(df), gamma = log(3),
  num_grnas = num_grnas
)

sim_grna_matrix <- grna_simulated[,paste0("sim_grna_", seq_len(num_grnas))] |>
  t() |>
  as.matrix()

fake_response_matrix <- matrix(0, nrow = 1, ncol = ncol(sim_grna_matrix), dimnames = list("fake_gene", colnames(sim_grna_matrix)))

cov_df <- dplyr::transmute(grna_simulated, batch, grna_n_nonzero_ground_truth = grna_n_nonzero,
                           grna_n_umis_ground_truth = grna_n_umis) 
fake_grna_target_df <- data.frame(
  grna_id = rownames(sim_grna_matrix),
  grna_target = c("a","a","b","b","non-targeting"),
  stringsAsFactors = FALSE
)

scep_sim <- import_data(
  response_matrix = fake_response_matrix,
  grna_matrix = sim_grna_matrix,
  grna_target_data_frame = fake_grna_target_df,
  extra_covariates = cov_df,
  moi = "high"
) |>
  set_analysis_parameters(formula = ~1)

# TODO why are some of these zero? shouldn't those have been removed?
has_zero_nonzero <- any(scep_sim@covariate_data_frame$grna_n_nonzero_ground_truth == 0)
has_zero_umis <- any(scep_sim@covariate_data_frame$grna_n_umis_ground_truth == 0)

nonzero_term <- if(has_zero_nonzero) "log(grna_n_nonzero_ground_truth + 1)" else "log(grna_n_nonzero_ground_truth)"
umis_term <- if(has_zero_umis) "log(grna_n_umis_ground_truth + 1)" else "log(grna_n_umis_ground_truth)"

formula_str <- paste0("~ batch +", nonzero_term, " + ", umis_term)
assign_grnas_formula_object <- as.formula(formula_str)

scep_sim <- assign_grnas(scep_sim, method = "mixture", formula_object = assign_grnas_formula_object)

out <- get_grna_assignments(scep_sim)

plot(scep_sim)
```


ok we are going to treat the covariates as fixed, and keep the sample size what it is 

```{r}
make_fake_data <- function(mod, cell_covs, perturbation_effect_sizes = log(3), num_grnas_per_effect = 1, prob_perturbed=0.02, seed=123) {
  set.seed(seed)

  # lets us loop over perturb effects 
  if(length(perturbation_effect_sizes) == 1 && is.numeric(perturbation_effect_sizes)) {
    perturbation_effect_sizes = list(perturbation_effect_sizes)
  }
  n <- nrow(cell_covs)
  
  # 1. get simulated cell covariates
  log_mu <- predict(mod, newdata=cell_covariates) 
  results <- list()
  for(perturb_idx in seq_along(perturbation_effect_sizes)) {
    curr_perturb_effect <- perturbation_effect_sizes[[perturb_idx]]
    
    # TODO make this more elegant later
    is_perturbed <- matrix(rbinom(n * num_grnas_per_effect, 1, prob_perturbed), nrow = n)
    y_new <- matrix(0, nrow = n, ncol = num_grnas_per_effect)
    colnames(is_perturbed) <- paste0("is_perturbed_effect_", perturb_idx, "_grna_", seq_len(num_grnas_per_effect))
    colnames(y_new) <- paste0("sim_grna_effect_", perturb_idx, "_grna_", seq_len(num_grnas_per_effect))
    
    for(grna_idx in seq_len(num_grnas_per_effect)) {
      y_new[,grna_idx] <- MASS::rnegbin(n, mu = exp(log_mu + curr_perturb_effect * is_perturbed[,grna_idx]), theta=mod$theta)
    }
    results[[perturb_idx]] <- cbind(is_perturbed, y_new)
  }
  out <- cbind(cell_covs, do.call(cbind, results))
  
  return(out)
}



xx <- make_fake_data(mod=mod, cell_covs=cell_covs, perturbation_effect_sizes = list(log(3), log(2)), num_grnas_per_effect = 3)
head(xx)

make_grna_matrix <- function(fake_data) {
  fake_data[,grepl("sim_grna_", names(fake_data))] |>
    t()
}

grna_mat <- make_grna_matrix(xx)
```




```{r}
simulate_grna_matrix <- function(grna_targets, ) {
  # outline
  # 1. train the models
  # 2. for each model: for each gamma: for each 1:num_grnas_per_gamma
  #   sample a new response
  # 3. assemble into a sparse matrix, write
  
  
}
```

