---
title: "sim-assign-analysis"
output: html_document
date: "2025-11-13"
---


```{r}
source("~/.Rprofile")
library(tidyverse)

run_name <- "gasperini_simulated_mid_rate_big_effects"

base_dir <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  paste0("guide_assignment/outputs/", run_name, "/")
)

cell_covs <- read.csv(
  paste0(.get_config_path("LOCAL_BENCHMARKING_DIR"), "/guide_assignment/input_data/", run_name, "/sceptre/covariate_data_frame.csv"),
  row.names="X"
)

# takes sceptre assignment matrix and gets a list
# with one element per gRNA, and each element contains
# the indices of the cells assigned to that gRNA
# TODO rewrite this to be more efficient and directly work with sparse matrix form
sparse_matrix_to_list_named <- function(mat) {
  result <- lapply(1:nrow(mat), function(i) {
    rownames(cell_covs)[which(mat[i, ])]
  })
  names(result) <- rownames(mat)
  return(result)
}



# for crispat, we have our output as a csv like
# 'data.frame':	7143379 obs. of  2 variables:
#  $ cell_id: chr  "CELL_394" "CELL_759" "CELL_814" "CELL_911" ...
#  $ grna_id: chr  "AAGCCCGGCCAGCCGACCGG" "AAGCCCGGCCAGCCGACCGG" "AAGCCCGGCCAGCCGACCGG" "AAGCCCGGCCAGCCGACCGG" ...
# so we split on grna_id, and turn that into the integer cell index by processing
# "CELL_j" to j
csv_to_list <- function(df) {
  # Group by gRNA
  split(df$cell_id, df$grna_id)
}

```



```{r}
# dataset_names <- c("gasperini_simulated")
dataset_names <- run_name

## doing sceptre ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# NOTE: since sceptre has the full assignment matrices, i use this to get the grna ids.


sceptre_assignments_raw <- lapply(dataset_names, function(fname) {
  fpath <- file.path(base_dir, paste0("assignment_matrix_sceptre_", fname, ".rds"))
  readRDS(fpath)
}) |>
  setNames(dataset_names)

sceptre_assignments <- lapply(sceptre_assignments_raw, sparse_matrix_to_list_named)
sceptre_assignments[[run_name]][[1]][1:5]  # should be string cell names
grna_names <- lapply(sceptre_assignments_raw, rownames)

## doing crispat ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# outline:
# 1. load the assignment csvs and process them to have fast grna_id lookup
# 2. make an empty list of lists that has every grna_id in it, using the sceptre assignment matrices
# 3. fill that with the results of step (1) so that we know grna_id order and names
#     line up with different methods


# 1. loading the assignment CSVs
crispat_assignments_raw <- lapply(dataset_names, function(fname) {
  file.path(base_dir, paste0("assignments_crispat_", fname, ".csv")) |>
    read.csv() |>
    csv_to_list()
}) |>
  setNames(dataset_names)

# 2. making an empty data struc to hold them all with no missing names
crispat_assignments <- lapply(dataset_names, function(name) {
  lapply(grna_names[[name]], function(i) integer(0)) |>
    setNames(grna_names[[name]])
}) |>
  setNames(dataset_names)

# 3. filling the empty data structure
for(name in dataset_names) {
  for(grna_name in names(crispat_assignments_raw[[name]])) {
    crispat_assignments[[name]][[grna_name]] <- crispat_assignments_raw[[name]][[grna_name]]
  }
}

crispat_assignments[[run_name]][[1]][1:5]  # should be string cell names
```


```{r}
jaccard <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  if (union == 0) return(NA_real_)
  return(intersection / union)
}



sims <- lapply(dataset_names, function(name) {
  sapply(grna_names[[name]], function(grna) jaccard(sceptre_assignments[[name]][[grna]], crispat_assignments[[name]][[grna]]))
}) |>
  setNames(dataset_names)

```

```{r}
library(tidyverse)
theme_set(theme_bw(base_size = 12))

# Tidy similarities: one row per gRNA per dataset
df_sims <- bind_rows(lapply(names(sims), function(nm) {
  tibble(dataset = nm, similarity = sims[[nm]])
})) %>%
  filter(!is.na(similarity))

# Plot: two panels (one per dataset)
p <- ggplot(df_sims, aes(x = similarity)) +
  geom_histogram(bins = 50) +
  facet_wrap(dataset ~ ., ncol = 1, scales = "free") +
  labs(
    title = "Jaccard similarites for each gRNA",
    x = "Jaccard similarity",
    y = "Count"
  ) +
  coord_cartesian(xlim = c(0, 1))

print(p)
```




# ground truth comparison

```{r}
ground_truth_mat <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  paste0("guide_assignment/input_data/", run_name, "/true_perturbation_status.rds")
) |>
  readRDS()
ground_truth <- lapply(1:nrow(ground_truth_mat), function(i) {
  colnames(ground_truth_mat)[which(ground_truth_mat[i,] == 1)]
})
names(ground_truth) <- rownames(ground_truth_mat)

sims_true <- lapply(grna_names[[run_name]], function(name) {
  data.frame(crispat = jaccard(ground_truth[[name]], crispat_assignments[[run_name]][[name]]),
             sceptre = jaccard(ground_truth[[name]], sceptre_assignments[[run_name]][[name]]))
}) |>
  do.call(what = rbind)
sims_true |> head()

hist(sims_true$crispat)
hist(sims_true$sceptre)

plot(sims_true$crispat, sims_true$sceptre)
abline(0,1,col=2)
```



how to analyze? let's start with a single gRNA
```{r}
curr_grna_name <- "GTATTGCTGCCTCCTCGCCG"
mask <- grepl(curr_grna_name, names(ground_truth))

curr_truth <- ground_truth[mask]
curr_truth_mat <- ground_truth_mat[mask, ]
effect_sizes <- strsplit(names(curr_truth), "_") |> sapply(`[`, 4) |> as.numeric()

# now let's see how well we did

results <- lapply(1:length(curr_truth), function(i) {
  curr_grna <- names(curr_truth)[i]
  data.frame(
    grna = curr_grna,
    crispat = jaccard(curr_truth[[curr_grna]], crispat_assignments[[run_name]][[curr_grna]]),
    sceptre = jaccard(curr_truth[[curr_grna]], sceptre_assignments[[run_name]][[curr_grna]]),
    effect_size = effect_sizes[i]
  )
}) |>
  do.call(what = rbind) |>
  mutate(target = curr_grna_name)

head(results)

results |>
  dplyr::select(crispat, sceptre, effect_size) |>
  ggplot(aes(x = crispat, y = sceptre, color = effect_size)) +
  geom_point() +
  geom_abline(slope=1,intercept=0,color="red", linetype="dashed") +
  labs(x = "jaccard(crispat, truth)", y = "jaccard(sceptre, truth)",
       title = paste0("Ground truth similarities for target = ", curr_grna_name))

mean(results$crispat < results$sceptre)

```


i bet there are big differences in how many we assign vs ground truth

```{r}
curr_grna_name <- "GTATTGCTGCCTCCTCGCCG"
mask <- grepl(curr_grna_name, names(ground_truth))

curr_truth <- ground_truth[mask]
effect_sizes <- strsplit(names(curr_truth), "_") |> sapply(`[`, 4) |> as.numeric()
curr_truth_mat <- ground_truth_mat[mask, ]


curr_scep <- sceptre_assignments[[run_name]][names(curr_truth)]
curr_crisp <- crispat_assignments[[run_name]][names(curr_truth)]

num_scep_assigned <- sapply(curr_scep, length)
num_crisp_assigned <- sapply(curr_crisp, length)

plotdat <- data.frame(
  num_scep_assigned = num_scep_assigned,
  num_crisp_assigned = num_crisp_assigned,
  effect_size = effect_sizes, 
  true_num_perturbed = rowSums(curr_truth_mat)
)

pdat2 <- data.frame(
  num_assigned = c(num_scep_assigned, num_crisp_assigned),
  truth = rep(rowSums(curr_truth_mat), times = 2),
  method = rep(c("sceptre", "crispat"), each = nrow(curr_truth_mat)),
  effect_size = c(effect_sizes,effect_sizes)
)
pdat2 |>
  ggplot(aes(x = truth, y = num_assigned, color = effect_size)) +
  geom_point() +
  facet_wrap(~method) +
  labs(x = "Number of cells expressing this gRNA",
       y = "Number of cells assigned to this gRNA",
       title = "Actual vs Estimated #cells per gRNA")
# pdat2 |>
#   ggplot(aes(x = truth, y = num_assigned, color = method)) +
#   geom_point() +
#   labs(x = "Number of cells expressing this gRNA",
#        y = "Number of cells assigned to this gRNA",
#        title = "Actual vs Estimated #cells per gRNA")
# hist(num_scep_assigned)
# hist(num_crisp_assigned)
# 
# plot(num_scep_assigned ~ num_crisp_assigned, col = effect_sizes + 1)

```

so these are really low...

what's real moi vs what we tend to get in practice? MOI = number of gRNAs a particular cell expresses --> colSums of ground_truth_mat.

```{r}
moi_per_cell <- colSums(curr_truth_mat)
hist(moi_per_cell)
hist(rbinom(1e5, 500, .02), 50)  # yep it's just this : 100*5 iid draws of Bern(0.02) 
```

so this doesn't tell us anything important ... ?

why are the jaccards so small? are we classifying way too few?

```{r}

    # We'll just reuse the same matrices for fair GPU vs CPU comparison

```



what's going on here... why does the number we assign not seem to change by dataset? and the similarities are SO small...

```{r}
input_data_base_dir <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data"
)

grna_odm <- ondisc::initialize_odm_from_backing_file(file.path(input_data_base_dir, "gasperini/sceptre-pipeline/grna.odm"))
grna_gasp <- Matrix::readMM(file.path(input_data_base_dir, "gasperini/cleanser/grna_matrix.mtx")) |>
  `rownames<-`(rownames(grna_odm))

dim(grna_gasp)

grna_sim <- Matrix::readMM(file.path(input_data_base_dir, run_name, "cleanser/grna_matrix.mtx")) |>
  `rownames<-`(grna_names[[run_name]])
dim(grna_sim)
# let's look at row and col sums vs my simulated matrix

# counting the fraction of non-zero cells per gRNA
grna_sparsity_gasp <- tabulate(grna_gasp@i + 1L, nbins = nrow(grna_gasp))
grna_sparsity_sim <- apply(grna_sim, 1, function(v) sum(v > 0))

plotdat <- data.frame(
  num_non_zero = c(grna_sparsity_gasp, grna_sparsity_sim),
  source = rep(c("real", "sim"), c(nrow(grna_gasp), nrow(grna_sim)))
) 

plotdat |>
  ggplot(aes(x = num_non_zero, fill = source)) +
  geom_histogram(bins = 30) +
  facet_wrap(~source, scales = "free") +
  theme_bw()
  

plotdat |>
  ggplot(aes(x = num_non_zero, fill = source)) +
  geom_histogram(bins = 30) +
  facet_wrap(~source, scales = "free_y") +
  theme_bw()
```

so: we have extremely homogeneous numbers of cells with non-zero expressions in our gRNAs. Well, that's ok. Each gRNA is supposed to be pretty similar. So that's not concerning. This is a lot like a p-fold repetition of a single gRNA. 


what's the real concern? that there is no relationship between effect size and assignment results, like similarity or number assigned. 

something else we can look at: the fraction of non-zero gRNAs for each cell --> column function.
```{r}
library(Matrix)

## count >0 per column (works on sparse, no densification)
grna_gasp_C <- as(grna_gasp, "dgCMatrix")            # cheap, O(nnz), also collapses duplicates
# pos_per_col <- Matrix::colSums(grna_gasp_C > 0) / ncol(grna_gasp)

## if you really mean “nonzero (≠0)” counts per column:
cell_sparsity_gasp <- diff(grna_gasp_C@p) / ncol(grna_gasp)


cell_sparsity_sim <- apply(grna_sim, 2, function(v) mean(v > 0))


plot(cell_sparsity_gasp ~ cell_sparsity_sim)
abline(0,1,col="red")

```

so our cells are much sparser on average than those in Gasperini. 

so we need to increase the rate of 1s in each column. ?
well, no... we need to increase the rate of 0 in the data. 

how do we look marginally...? what about plotting a histogram of the quantiles of each simulated grna vs the real one?

```{r}
unique_grnas <- strsplit(rownames(grna_sim), "_") |> sapply(`[`, 2) |> unique()
# TODO
# how does the distribution of our realized Y compare to actual?

hist(grna_sim["sim_CGTCTGCTCCATAGGCAGCG_effect_2.000_grna_1", ] |> asinh(), 50)
hist(grna_gasp["CGTCTGCTCCATAGGCAGCG", ] |> asinh(), 50)
mean(grna_gasp["CGTCTGCTCCATAGGCAGCG", ] > 0)
mean(grna_sim["sim_CGTCTGCTCCATAGGCAGCG_effect_2.000_grna_1", ] > 0)


hist(grna_sim["sim_GTATTGCTGCCTCCTCGCCG_effect_2.000_grna_1", ] |> asinh(), 50)
hist(grna_gasp["GTATTGCTGCCTCCTCGCCG", ] |> asinh(), 50)
mean(grna_gasp["GTATTGCTGCCTCCTCGCCG", ] > 0)
mean(grna_sim["sim_GTATTGCTGCCTCCTCGCCG_effect_2.000_grna_1", ] > 0)


```


what's P(Y=0) vs P(not perturbed)? 

any chance theta is being used wrong? should 1/theta be used? is that tiny theta a problem? does our model fit poorly?


```{r}
grna_sparsity_true <- apply(ground_truth_mat, 1, function(v) sum(v > 0))
plot(grna_sparsity_true ~ grna_sparsity_sim)
abline(0,1,col=2)

mod <- lm(grna_sparsity_true ~ grna_sparsity_sim)
summary(mod)
abline(mod, col=4)

```

so in general we have decently more non-zero Y_i than X_i, but no correlation. Makes sense. 

Is P(Y=0) simple to relate to X? 

rnegbin(
          n = num_cells,
          mu = exp(log_mu + effect_size * is_perturbed),
          theta = mod$theta
        )

so we should have $\mu = r(1-p)/p$ and $\theta=r$ for wikipedia parameterization. Then

$$
P(Y = 0) = p^r = \left(\frac{\theta}{\mu+\theta}\right)^\theta
$$
so tiny $\theta$ will mean we get very concentrated on zero... that is intended behavior. I guess this is how the negbin handles zero inflation!

```{r}
hist(grna_sim["sim_GTATTGCTGCCTCCTCGCCG_effect_6.000_grna_1", ] |> asinh(), 50)
# due to covariates, we can't LLN this probability right now

mod <- readRDS(
  paste0(.get_config_path("LOCAL_BENCHMARKING_DIR"), "/guide_assignment/input_data/", run_name, "/model_info.rds")
)

means <- cell_covs |> mutate(log_mu = mod$sim_GTATTGCTGCCTCCTCGCCG$coefficients["(Intercept)","Estimate"] + mod$sim_GTATTGCTGCCTCCTCGCCG$coefficients["batch","Estimate"] * batch + 
 mod$sim_GTATTGCTGCCTCCTCGCCG$coefficients["log(grna_n_nonzero + 1)","Estimate"] * log(true_grna_n_nonzero + 1)+ 
   mod$sim_GTATTGCTGCCTCCTCGCCG$coefficients["log(grna_n_umis + 1)","Estimate"] *  log(true_grna_n_umis + 1) + 
   2 * ground_truth_mat["sim_GTATTGCTGCCTCCTCGCCG_effect_6.000_grna_1", ]) |>
  pull(log_mu) |>
  exp()

plot(x = log(1 + grna_sim["sim_GTATTGCTGCCTCCTCGCCG_effect_6.000_grna_1", ]), y = log(means))

```




let's get the model we used to make this:
```{r}
grna_odm <- ondisc::initialize_odm_from_backing_file(paste0(base_fp, "grna.odm"))

grnas_to_use <- c(5034, 7971)#, 9796, 10437, 12669)

rownames(grna_odm)[grnas_to_use]
stopifnot(rownames(grna_odm)[grnas_to_use[1]] == "GTATTGCTGCCTCCTCGCCG")

grna_matrix_rows <- lapply(grnas_to_use, function(i) grna_odm[i,]) |>
  do.call(what = rbind) |>
  `dimnames<-`(list(
    rownames(grna_odm)[grnas_to_use],
    rownames(cell_covs)
  ))

mod_df <- cell_covs |>
  mutate(y = grna_matrix_rows["GTATTGCTGCCTCCTCGCCG", ])
mod1 <- MASS::glm.nb(y ~ batch + log(true_grna_n_nonzero + 1) + log(true_grna_n_umis + 1), data = mod_df)
summary(mod1)
# so tiny theta --> very overdispersed

```


