---
title: "smaller_n_sims"
output: html_document
date: "2026-01-25"
---


My current simulations use Replogle's sample size of ~600k. But that puts a computational burden that we don't need. Let's try doing a smaller n.


idea:
- still based on replogle, but now we downsample cell covariates
- should we use a gRNA that sceptre assigns a lot of cells to?



```{r}
rm(list=ls())
source("~/.Rprofile")
library(tidyverse)
library(sceptre)
library(ondisc)

source(file.path(
  .get_config_path("LOCAL_SCEPTRE3_REPO_DIR"),
  "benchmarking/guide-assignment/grna-simulator/grna-simulator.R"
))

plot_results <- function(grna_real, grna_sim, is_pert, sim_name) {
  
  logp1_trans <- function(base = 10) {
    trans <- function(x) log(x + 1, base)
    inv   <- function(x) base^x - 1  # correct inverse
    
    # Major breaks: at base^k - 1
    brks <- function(limits) {
      rng <- limits + 1
      scales::log_breaks(base = base)(rng) - 0
    }
    
    scales::trans_new(
      name = paste0("logp1-", base),
      transform = trans,
      inverse   = inv,
      breaks    = brks,
      domain = c(0, Inf)
    )
  }
  
  # parse sim name
  # sim_name_pieces <- strsplit(sim_name, "_")[[1]]
  # keep <- sim_name_pieces[which(sim_name_pieces == "probpert"):(which(sim_name_pieces == "grna")-1)]
  # sim_name_plot <- paste0(keep[c(1,3,5,7)], "=", keep[c(2,4,6,8)], collapse=" ") 
  
  plt1 <- data.frame(
    umi = c(grna_real, grna_sim),
    group = rep(c("real", "sim"), each = length(grna_real))
  ) |>
    ggplot(aes(x = umi, fill = group)) +
    geom_histogram(bins=50) +
    facet_wrap(~group) +
    theme_bw()  +
    scale_y_continuous(trans = logp1_trans(10)) +
    scale_x_continuous(trans = logp1_trans(10)) +
    ggtitle(sim_name)
  
  plt2 <- data.frame(
    umi = grna_sim,
    pert_status = ifelse(is_pert == 1, "pert", "not pert")
  ) |>
    ggplot(aes(x = umi, fill = pert_status)) +
    geom_histogram(bins=50) +
    facet_wrap(~pert_status) +
    theme_bw()  +
    scale_y_continuous(trans = logp1_trans(10)) +
    scale_x_continuous(trans = logp1_trans(10)) +
    ggtitle("Simulated UMI counts by pert status")
  
  # printing some numeric comparisons
  comparison_df <- data.frame(
    real = c(max(grna_real),   mean(grna_real == 0),    var(grna_real),    mean(grna_real)),
    sim = c(max(grna_sim),     mean(grna_sim==0),      var(grna_sim),      mean(grna_sim))
  ) |>
    `rownames<-`(c("max", "frac_0", "var", "mean"))
  print(comparison_df)
  
  cowplot::plot_grid(plt1, plt2, nrow = 2)
}

source_data <- "replogle-rd7"

stopifnot(source_data %in% c("gasperini", "replogle-rd7"))

# 1. loading the data and preparing everything ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
base_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data", source_data, "sceptre-pipeline/"
)


scep <- read_ondisc_backed_sceptre_object(
  sceptre_object_fp = paste0(base_fp, "sceptre_object.rds"),
  response_odm_file_fp = paste0(base_fp, "response.odm"),
  grna_odm_file_fp = paste0(base_fp, "grna.odm")
)
if(source_data == "gasperini") {
  cell_covariates <- scep@covariate_data_frame |>
    transmute(
      grna_n_nonzero, grna_n_umis,
      batch = ifelse(prep_batch == "prep_batch_2", 1, 0)
    ) 
} else if(source_data == "replogle-rd7") {
  cell_covariates <- scep@covariate_data_frame |>
    transmute(
      grna_n_nonzero, grna_n_umis
    ) 
}

# determining which gRNAs to run this on [no need to run again]
grna_odm <- ondisc::initialize_odm_from_backing_file(paste0(base_fp, "grna.odm"))

assignments_fp <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs", source_data, "sceptre-pipeline/"
)
assn_mat <- readRDS(file.path(assignments_fp,  "grna_assignment_matrix.rds" ))
dim(assn_mat)

setequal(rownames(assn_mat), rownames(grna_odm)) |>
  stopifnot()
```



i want to get target gRNAs that are both highly expressed but also have lots of cells assigned from sceptre. So let's get a data.frame with the gRNA covariates, plus the number of cells sceptre assigns.


hmmm lol ok so that's the transpose. I want to look at guide-level summaries. 

# 1. determining which guide to use [can be skipped now that a guide is chosen]

```{r}


guide_covs <- data.frame(
  num_nonzero_umi = numeric(nrow(grna_odm)),
  tot_umi = 0,
  num_assigned_cells = 0
)
rownames(guide_covs) <- rownames(grna_odm)
for(i in 1:nrow(grna_odm)) {
  guide_covs[i,"num_nonzero_umi"] <- sum(grna_odm[i,] > 0)
  guide_covs[i,"tot_umi"] <- sum(grna_odm[i,])
}

for(guide in rownames(assn_mat)) {
  guide_covs[guide, "num_assigned_cells"] <- sum(assn_mat[guide, ])
}

head(guide_covs)
```

ok now how do we pick which ones to use?

i want one with a lot of expression so that when i do a 1/6th sample, i still get a lot of perturbed cells.

```{r}
guide_covs |>
  transmute(
    log_num_nonzero_umi = log(num_nonzero_umi),
    log_tot_umi = log(tot_umi),
    log_num_assigned_cells = log(num_assigned_cells)
  ) |>
  GGally::ggpairs()
```

ok so these three are pretty strongly jointly correlated.

To that end, then, I'm going to sample from among the guides that sceptre determined to have the most expressed cells.

```{r}
guide_covs |>
  dplyr::filter(num_assigned_cells > quantile(num_assigned_cells, .95),
                num_assigned_cells < quantile(num_assigned_cells, .99)) |>
  arrange(desc(num_nonzero_umi))

# from this sort, 8645_TACC3_P1_ENSG00000013810 is the top
```


ok we're going to roll with this one for now: `8645_TACC3_P1_ENSG00000013810`

# 2. simulating using my chosen guide

```{r}
guide = "8645_TACC3_P1_ENSG00000013810"
# guide_covs[guide, ]

```

this is still only 0.2\% expressed, but that's more than before. 

now for fitting the model: do i want to change this at all?



```{r}

# num_cells <- 100000
# expressed_cells <- which(assn_mat[guide,])
# non_expressed_cells <-  which(!assn_mat[guide,])
# 
# set.seed(5675)
# cell_idx <- c(
#   expressed_cells,
#   sample(non_expressed_cells, num_cells - length(expressed_cells))
# ) |> sort()
# 
# y <- grna_odm[guide, ][cell_idx]
# model_covs <- cell_covariates[cell_idx, ]

y_all <- grna_odm[guide, ]

mod <- glm.nb(y ~ 1 + log(grna_n_nonzero+1) + log(grna_n_umis+1), 
              data = cbind(y = y_all, cell_covariates))
summary(mod)
log_mu_all <- predict(mod, newdata = cell_covariates)
```

ok now i've got my model fit on this subsample.

Do i actually want my covariates to be fit on this sample ... ? i learned a different relationship! 

So, two options:
- fit the model on everything to learn the relationship on the whole dataset. In this case, the relationship i'm learning is basically just for the non-perturbed cells as that's by FAR the majority of the dataset
- if instead i fit the model on this downsample, then the perturbed cells are a much bigger fraction. And we get this reversal.

I am leaning towards fitting the model on everything.

ok, let's do that and proceed with downsampled AFTER.
```{r}
num_cells <- 100000
expressed_cells <- which(assn_mat[guide,])
non_expressed_cells <-  which(!assn_mat[guide,])

set.seed(5675)
cell_idx <- c(
  expressed_cells,
  sample(non_expressed_cells, num_cells - length(expressed_cells))
) |> sort()

y <- y_all[cell_idx]
log_mu <- log_mu_all[cell_idx]
```


does it make sense to simulate `is_perturbed` independently from the covariates / `log_mu` ... ? probably not... 

what about a logistic regression to model perturbation instead ... ?

why am  i doing a negbin regression? That's because i want to understand the effects of these covariates on the actual UMI counts. So that when i simulate new UMI counts, the simulated data makes sense with the covariates that i provide, and to give some heterogeny.

Does simulating with no regard to the covariates make sense? No, there definitely is a relationship there. So i should ALSO understand how these covariates relate to P(perturbed)

i don't actually know that, but let's use the estimated ones. Do i want this on my subset, or in general? Let's do in general first.

```{r}
data_is_pert <- cell_covariates |> mutate(y = assn_mat[guide, ])
mod_is_pert <- glm(y ~ 1 + log(grna_n_nonzero+1) + log(grna_n_umis+1), 
                   data = data_is_pert, family = binomial())
cell_pert_probs_all_logit <- fitted(mod_is_pert) |> qlogis()
summary(mod_is_pert)
```

interesting -- only `log(grna_n_umis + 1)` is significant. So, in light of this, how do i want to sample perturbation indicators? From this model? well, that won't give me control over the perturbation rate. Can i do that with an intercept adjustment?

```{r}
# cell_pert_probs_logit <- cell_pert_probs_all_logit[cell_idx]
# head(cell_pert_probs_logit)
# 
# is_perturbed <- rbinom(length(cell_pert_probs_logit), size = 1, plogis(cell_pert_probs_logit)) |>
#   as.logical()
# table(is_perturbed)
# mean(assn_mat[guide,])
# mean(is_perturbed)
# 
# mutate(cell_covariates[cell_idx, ], cell_pert_probs_logit = cell_pert_probs_logit, is_perturbed = is_perturbed) |>
#   dplyr::filter(is_perturbed) |>
#   sample_n(10)
```


ok so this gives me perturbation indicators that match the observed rate. If i want to simply bump up the number of expressed cells, that would be an intercept adjustment, which i can do by adding a constant to `cell_pert_probs`

```{r}
set.seed(43858)
cell_pert_probs_logit <- cell_pert_probs_all_logit[cell_idx]
head(cell_pert_probs_logit)

bump_rate <- 2
is_perturbed <- rbinom(length(cell_pert_probs_logit), size = 1, plogis(cell_pert_probs_logit + bump_rate)) |>
  as.logical()
table(is_perturbed)
mean(assn_mat[guide,])
mean(is_perturbed)

mutate(cell_covariates[cell_idx, ], cell_pert_probs_logit = cell_pert_probs_logit  +bump_rate, is_perturbed = is_perturbed, log_mu) |> 
  dplyr::filter(is_perturbed) |>
  sample_n(10)
```

ok now i want to simulate from a negative binomial using all of these pieces.

```{r}
set.seed(194)

non_pert_effect <- -5
pert_effect <- 5.5
theta_pert <- 10
theta_non_pert <- .1


table(is_perturbed)
# Simulate counts: add effect to perturbed and non-perturbed cells
# log_mu_pert <- log_mu + pert_effect * is_perturbed + non_pert_effect * (1 - is_perturbed)


simulated_counts <- numeric(num_cells)
simulated_counts[is_perturbed] <- rnegbin(
  n = sum(is_perturbed),
  mu = exp( log_mu[is_perturbed] + pert_effect),
  theta = theta_pert
)

simulated_counts[!is_perturbed] <- rnegbin(
  n = sum(!is_perturbed),
  mu = exp(log_mu[!is_perturbed] + non_pert_effect),
  theta = theta_non_pert
)



plot_results(
  grna_real = y,
  grna_sim = simulated_counts,
  is_pert = is_perturbed,
  sim_name = bquote(gamma[0] ~ "=" ~ .(non_pert_effect) * ";" 
                    ~ gamma[1] ~ "=" ~ .(pert_effect) * ";"
                    ~ theta[0] ~ "=" ~ .(theta_non_pert)* ";"
                    ~ theta[1] ~ "=" ~ .(theta_pert))
)

ysim <- simulated_counts
data.frame(
  real = c(max(y),   mean(y==0),    var(y),    mean(y)),
  sim = c(max(ysim), mean(ysim==0), var(ysim), mean(ysim))
) |>
  `rownames<-`(c("max", "frac_0", "var", "mean"))
```



let's try to make a simple function for doing this so i can test it

```{r}

make_simulated_umi_counts <- function(umi_counts, num_guides_to_simulate, cell_covariates, sceptre_assignments, formula, num_cells, sim_params, guide_name, clip_to_real_max = TRUE, seed=NULL) {
  
  if(!is.null(seed)) {
    set.seed(seed)
  }
  
  mod_data <- cell_covariates |> mutate(umi_counts, sceptre_assignments)
  max_real_umi <- max(umi_counts)
  # 1. determine which cells we are using
  expressed_cells <- which(sceptre_assignments)
  non_expressed_cells <- setdiff(1:length(sceptre_assignments), expressed_cells)
  
  cell_idx_subsampled <- c(
    expressed_cells,
    sample(non_expressed_cells, num_cells - length(expressed_cells))
  ) |> sort()
  
  cat("Beginning to fit models: ")
  
  # 2. get the effect of the covariates on the mean of the UMI counts
  fmla_umi <- as.formula(paste("umi_counts ~", as.character(formula)[2]))
  mod_umi <- glm.nb(fmla_umi, data = mod_data)
  log_mean_umi <- fitted(mod_umi)[cell_idx_subsampled] |> log()
  
  # 3. get the effect of the covariates on the perturbation probability
  fmla_pert <- as.formula(paste("sceptre_assignments ~", as.character(formula)[2]))
  mod_pert <- glm(fmla_pert, data = mod_data, family = binomial())
  logit_prob_pert <- fitted(mod_pert)[cell_idx_subsampled] |> qlogis()
  
  cat("done.\n")
  
  # 4. for each guide we simulate, do the following:
  #   4a. simulate perturbation indicators for each cell
  #   4b. simulate UMI counts for each cell using those

  perturbation_indicators <- simulated_umi_counts <- matrix(0, num_guides_to_simulate, num_cells)
  
  cat("Beginning to simulate data: ")
  
  for(sim_idx in 1:num_guides_to_simulate) {
    # 4a. simulate perturbation indicators for our subsampled cells, using `mod_pert`
    is_perturbed <- rbinom(
      num_cells, size = 1,
      plogis(logit_prob_pert + sim_params$logit_prob_pert_increase)
    ) |>
      as.logical()
    # if(sim_idx == 1) {
    #   cat("The first simulated guide has", sum(is_perturbed), "perturbed cells.\n")
    # }
    perturbation_indicators[sim_idx,] <- is_perturbed
    
    
    # 4b. simulate UMI counts, sampling from a 2 component NB mixture with means
    # coming from `mod_umi` 
    log_mean_umi_with_pert <- log_mean_umi + sim_params$pert_effect * is_perturbed +
      sim_params$non_pert_effect * (1 - is_perturbed)
    thetas <- ifelse(is_perturbed, sim_params$theta_pert, sim_params$theta_non_pert)
    simulated_counts <- rnegbin(
      num_cells,
      mu = exp(log_mean_umi_with_pert),
      theta = thetas
    )
    if(clip_to_real_max) simulated_counts <- pmin(simulated_counts, max_real_umi)
    simulated_umi_counts[sim_idx,] <- simulated_counts
  }
  cat("done.\n")
  # 5. collect and return results
  dimnames_ <- list(
    paste0(guide_name, "_sim_", 1:num_guides_to_simulate),
    paste0("cell_idx_", cell_idx_subsampled)
  )

  sim_results <- list(
    umi_counts = umi_counts,
    sceptre_assignments = sceptre_assignments,
    mod_umi = mod_umi,
    mod_pert = mod_pert,
    cell_idx_subsampled = cell_idx_subsampled,
    sim_params = sim_params,
    clip_to_real_max = clip_to_real_max,
    formula = formula,
    perturbation_indicators = perturbation_indicators |> Matrix::Matrix(sparse=TRUE) |> `dimnames<-`(dimnames_),
    simulated_umi_counts = simulated_umi_counts |> Matrix::Matrix(sparse=TRUE) |> `dimnames<-`(dimnames_)
  )
  return(sim_results)
}


guide <- "8645_TACC3_P1_ENSG00000013810"

# umi_counts <- grna_odm[guide,]
# sceptre_assignments = assn_mat[guide,]
# formula = ~ 1 + log(grna_n_nonzero+1) + log(grna_n_umis+1)
# 
# 
# sim_params = list(
#   logit_prob_pert_increase = 2,
#   non_pert_effect = -5,
#   pert_effect = 5.5,
#   theta_pert = 10,
#   theta_non_pert = .1
# )

# umi_counts, cell_covariates, sceptre_assignments, formula, num_cells, sim_params, seed=NULL

sim_results <- make_simulated_umi_counts(
  umi_counts = grna_odm[guide,],
  num_guides_to_simulate = 25,
  cell_covariates = cell_covariates,
  sceptre_assignments = assn_mat[guide,],
  formula = ~ 1 + log(grna_n_nonzero+1) + log(grna_n_umis+1),
  num_cells = 100000,
  sim_params = list(
    logit_prob_pert_increase = 2,
    non_pert_effect = -5,
    pert_effect = 5.25,
    theta_pert = 10,
    theta_non_pert = .1
  ),
  guide_name = guide,
  seed = 123
)


idx <- 1
plot_results(
  grna_real = sim_results$umi_counts[sim_results$cell_idx_subsampled],
  grna_sim = sim_results$simulated_umi_counts[idx,],
  is_pert = sim_results$perturbation_indicators[idx,],
  sim_name = "test"
)
```

ok now i gotta save this

```{r}
library(reticulate)
library(SingleCellExperiment)
library(zellkonverter)


write_to_h5ad <- function(mat, path_to_write) {
  env_name <- "r-anndata"
  curr_envs <- conda_list()$name
  if(!env_name %in% curr_envs) {
    conda_create("r-anndata", packages = c("python=3.12"))
    py_install(c("numpy", "scipy", "h5py", "anndata"),
               envname = "r-anndata", pip = TRUE)
  }
  use_condaenv("r-anndata", required = TRUE)
  py_config()
  
  sce <- SingleCellExperiment(assays = list(counts = mat))
  writeH5AD(sce, path_to_write)
}

run_name <- "replogle-rd7_simulated_100k_0.015-pert"

output_dir <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  paste0("guide_assignment/input_data/", run_name)
)
dir.create(output_dir, recursive=TRUE, showWarnings=FALSE)

# 3.0: saving model info and true perturbation status
saveRDS(sim_results$perturbation_indicators, file = file.path(output_dir, "true_perturbation_status.rds"))
saveRDS(sim_results, file = file.path(output_dir, "all_sim_results.rds"))


# 3.1: crispat
crispat_dir <- file.path(output_dir, "crispat")
dir.create(crispat_dir, recursive=TRUE, showWarnings=FALSE)
print(paste("  Writing crispat to", crispat_dir))
write_to_h5ad(sim_results$simulated_umi_counts, file.path(crispat_dir, "grna_matrix.h5ad"))

# 3.2: pertpy
pertpy_dir <- file.path(output_dir, "pertpy")
dir.create(pertpy_dir, recursive=TRUE, showWarnings=FALSE)
print(paste("  Writing pertpy to", pertpy_dir))
write_to_h5ad(sim_results$simulated_umi_counts, file.path(pertpy_dir, "grna_matrix.h5ad"))

# 3.3: cleanser
cleanser_dir <- file.path(output_dir, "cleanser")
dir.create(cleanser_dir, recursive=TRUE, showWarnings=FALSE)
print(paste("  Writing cleanser to", cleanser_dir))
Matrix::writeMM(sim_results$simulated_umi_counts, file.path(cleanser_dir, "grna_matrix.mtx"))

# 3.4: sceptre
sceptre_dir <- file.path(output_dir, "sceptre")
dir.create(sceptre_dir, recursive=TRUE, showWarnings=FALSE)
print(paste("  Writing sceptre gRNA matrix and other files to", sceptre_dir))
saveRDS(sim_results$simulated_umi_counts, file.path(sceptre_dir, "grna_matrix.rds"))

# 3.4.1: grna_target_data_frame
guide_target <- scep@grna_target_data_frame |>
  dplyr::filter(grna_id == guide) |> pull(grna_target)
sim_grna_target_df <- data.frame(
  grna_id = rownames(sim_results$simulated_umi_counts),
  grna_target = guide_target
)

# making NT gRNAs for sceptre
if(!any(sim_grna_target_df$grna_target == "non-targeting")) {
  sim_grna_target_df$grna_target[1:2] <- "non-targeting"
  warning("no NTs detected in `sim_grna_target_df` so first 2 are set to `non-targeting`.")
}

write.csv(sim_grna_target_df, file.path(sceptre_dir, "grna_target_data_frame.csv"), row.names=FALSE)

# 3.4.2: response_matrix
fake_response_mat <- Matrix::sparseMatrix(
  i = integer(0),  # No non-zero entries
  j = integer(0),
  x = numeric(0),
  dims = c(1L, ncol(sim_results$simulated_umi_counts)),
  dimnames = list(
    "FAKE_GENE_1",  # Single fake gene
    colnames(sim_results$simulated_umi_counts)  # Use same cell names as grna_matrix
  )
)
saveRDS(fake_response_mat, file.path(sceptre_dir, "response_matrix.rds"))


# 3.4.3: covariate_data_frame
cell_covariates[sim_results$cell_idx_subsampled, ] |>
  dplyr::rename(true_grna_n_nonzero = grna_n_nonzero, true_grna_n_umis = grna_n_umis) |>
  write.csv(file.path(sceptre_dir, "covariate_data_frame.csv"), row.names=TRUE)

# 3.4.4 saving formula object for assign_grnas()
saveRDS(sim_results$formula, file.path(sceptre_dir, "assign_grnas_formula.rds"))

cat("Command to transfer to HPC:\n")
cat(paste0("scp -r ", output_dir, " ", .get_config_path("REMOTE_BENCHMARKING_DIR"), "guide_assignment/input_data/", run_name, "\n"))
```

