---
title: "association"
output:
  html_document: default
  pdf_document: default
date: "2026-01-17"
---


```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(magrittr)
```

Methods

1. sceptre


2. mixscale

3. FR-Perturb

idea: first get perturbation vectors. 

Algo:

1. for each cell $x_i$, get the $k$ nearest NT cells in PCA space, all with the same biological replicate value as $x_i$. Denote $y_{i1}, \dots, y_{ik}$.

2. the \emph{perturbation vector} is defined as $p_i = x_i - \operatorname{Avg}(y_ij)$

Controls for biological replicate at least, as the nearest NT neighbors all have the same replicate. And tries to mitigate against more general systematic effects. 

Let $t$ be a target, and $P_t$ is the perturbation vectors for the cells expressing a guide for that target. $Y$ will be the NTs. In mixscale, we first do a DE test for every gene other than $t$ (mitigate selection bias), and only proceed if there are at least 5 marginally DE genes. We then only use those genes' measurements for our $p_i \in P_t$. Thus every coordinate of the $p_i$ in $P_t$ is the expression of a gene that is DE for this target $t$, and the dimension of $p_i$ is at least 5 (and at most 100 by default).

So we expect the $p_i \in P_t$ to be clustered with a center different in every coordinate from $Y$, since the mean is significantly different in every coordinate. 

We then get the \emph{perturbation scores} $s_i$ by projecting on the line going from $P_t$'s center to $Y$'s center. The $s_i$ give us a one-dimensional measure for how perturbed a cell is. 

We finish by getting the \emph{standardized perturbation scores} $\tilde s_i$ by z-transforming the $s_i$ with the mean and std dev of the NT cells' $s_i$. The end result is that the $\tilde s_i$ for NT cells should be centered around 0, while the perturbed $\tilde s_i$ should be clustered "significantly" far away.  

To actually DE test: let $\mathbf y \in \mathbb R^{N+M}$ give the response gene expressions for all $N$ cells targeting this gene $t$, and for all $M$ NT cells.

I'm going to write this by ignoring cell line. In the paper they control for it. 

Define $S\in\mathbb R^{N+M}$ by $S_{i} = \tilde s_i$ if cell $i$ targets $t$, and $0$ otherwise. Also define $\mathbf c \in \mathbb{R}^{N+M}$ where $\mathbf c_i$ is the $\log_{10}$ total expression for cell $i$. We then do a NB regression of $\mathbf y$ on $S$ and $\mathbf c$, with the coefficient $\beta$ of $S$ giving our weighted DE test. 


Question: can we do an on-target test without testing every target-gene pair? 

Pick gene 1. Get all cells targeting gene 1, call it $X$. Is there a difference on average between the gene 1 expressions for these cells $X$, and for the NT cells $Y$?

- case 1: there are at least 5 other genes that are DE between these two groups of cells 

  - compute the standardized perturbation scores 
  
  model $\log \operatorname{E}[\mathbf y] = \beta S+ \gamma \mathbf c$


- case 2: Fewer than 5 genes (excluding gene 1) are DE between $X$ and $Y$

  - model $\log \operatorname{E}[\mathbf y] = \beta \mathbf 1_{\text{ this cell targets }t} + \gamma \mathbf c$
  
  - we can compute this directly once we've picked our target $t$, so this case is efficient

Pick a target $t$, do a marginal DE test on all other genes (LOO). Suppose at least 5 are significant (if not, $S$ is just the targeted indicator) We have to compute the $\tilde s_i$ for $X_t$, the cells targeting $t$,

Datasets:


```{r, eval=FALSE, include=FALSE}
# not run
dataset_names <- c("replogle-rd7_large", "gasperini_pos_control")
for(dataset in dataset_names) {
  base_path <- file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/input_data", dataset)

  response_mat <- read_rds(file.path(base_path, "sceptre/response_matrix.rds"))
  # cell_info <- read_csv(file.path(base_path, "cell_info.csv"), show_col_types = FALSE)
  grna_target_df <- read_csv(file.path(base_path, "sceptre/grna_target_data_frame.csv"), show_col_types = FALSE)
  
  cat("Dataset name:", dataset, "\n")
  cat("Number of cells:", ncol(response_mat), "; number of genes:", nrow(response_mat),
      "; number of unique guides:",  length(unique(grna_target_df$grna_id)), "\n")
}
```

Dataset name: replogle-rd7_large                                                                                                                                                                             
Number of cells: 210110 ; number of genes: 1619 ; number of unique guides: 1813 

Dataset name: gasperini_pos_control                                                                                                                                                                          
Number of cells: 185171 ; number of genes: 377 ; number of unique guides: 855 

1. subsampled replogle 

- run sceptre-pipeline on the full replogle-rd7. method = mixture, with the default formula used, so we regressed on response_n_nonzero, response_n_umis, grna_n_nonzero, and grna_n_umis (all log or log1p as needed). `batch` was not used because it has 56 levels. 

Then, to get the desired subset of the response matrix:

- Filter to low MOI cells: Keep only cells expressing exactly one gRNA

  - Select positive control cells:
  
    - Iterate through response genes in dataset order
    
    - For each gene, collect all cells expressing gRNAs targeting that gene
    
    - Skip genes with fewer than min_num_cells_per_target cells (default: 20, used for "large")
    
    - Continue until reaching num_pc_cells positive control cells (e.g., 200,000 for "large")
    
  - Add non-targeting cells: Include cells expressing non-targeting gRNAs up to num_nt_cells (e.g., 10,000 for "large"), up to all of them
  
  - Extract expression data: Subset the response matrix to selected genes and cells
  
  - Quality filtering: Remove cells with zero expression across all genes, then remove any gRNAs left with zero cells
  
  - Export in multiple formats: Write data for sceptre (binary indicator matrix), mixscale (named vectors), and
  FR-Perturb (h5ad)


The resulting `response_matrix` resembles a block diagonal one, with ambient noise off the diagonal, as the first gene's targeting cells are the first cells in, and so on, until the final columns of NT cells. 

2. subsampled gasperini

- run sceptre-pipeline on the full gasperini data. method = mixture, with the default formula used, so we regressed on response_n_nonzero, response_n_umis, grna_n_nonzero, grna_n_umis (all log or log1p as needed), and batch (2 levels)

- Identify on-target genes: Find genes that are both targeted by gRNAs and present in the response matrix

  - Collect cells with on-target gRNAs:
  
    - For each on-target gene, get all gRNAs targeting that gene
    
    - Collect all cells expressing any of these gRNAs (no filtering to single gRNA per cell - this is high MOI)
    
  - Add non-targeting cells: Include all cells expressing non-targeting gRNAs
  
  - Create cell-level assignments: Build long-form dataframe mapping each cell to its expressed gRNAs (cells can have
  multiple gRNAs)
  
  - Extract expression data: Subset response matrix to on-target genes and selected cells
  
  - Compute gRNA metrics: Calculate grna_n_nonzero_subset and grna_n_umis_subset from the gRNA UMI counts for the cell
  subset
  
  - Export in multiple formats:
  
    - sceptre: Binary indicator matrix, cell covariates including prep_batch
    
    - FR-Perturb: h5ad file with colon-delimited multi-perturbation column (e.g., "GENE1:GENE2" for cells with multiple
  gRNAs), includes prep_batch covariate
  

```{r, echo = FALSE, include = FALSE}
rm(list=ls())


revlog_trans <- function (base = exp(1)) 
{
    trans <- function(x) {
        -log(x, base)
    }
    inv <- function(x) {
        base^(-x)
    }
    scales::trans_new(name = paste("revlog-", base, sep = ""), 
        transform = trans, inverse = inv, breaks = scales::log_breaks(base = base), 
        domain = c(1e-250, Inf))
}

library(tidyverse)
library(scales)

# run_name <- "run_all_medium"
# dataset_name = "replogle-rd7_medium"
# run_name <- "run_all_large"
# dataset_name = "replogle-rd7_large"
# run_name <- "run_all_gasperini"
# dataset_name = "gasperini_pos_control"

# datasets <- list(
#   gasperini = list(run_name = "run_all_gasperini", dataset_name = "gasperini_pos_control"),
#   replogle_large = list(run_name = "run_all_large", dataset_name = "replogle-rd7_large")
# )
# datasets <- list(
#   gasperini = list(run_name = "run_all_but_mixscale_skew-t", dataset_name = "gasperini_pos_control"),
#   replogle_large = list(run_name = "run_all_but_mixscale_skew-t", dataset_name = "replogle-rd7_pos_control_large")
# )
datasets <- list(
  gasperini_skew = list(run_name = "run_all_but_mixscale_skew-t", dataset_name = "gasperini_pos_control"),
  # replogle_medium_skew = list(run_name = "run_all_but_mixscale_skew-t", dataset_name = "replogle-rd7_pos_control_medium"),
  replogle_large_skew = list(run_name = "run_all_but_mixscale_skew-t", dataset_name = "replogle-rd7_pos_control_large"),
  
  gasperini_resamp = list(run_name = "run_all_but_mixscale_resamp", dataset_name = "gasperini_pos_control"),
  # replogle_medium_resamp = list(run_name = "run_all_but_mixscale_resamp", dataset_name = "replogle-rd7_pos_control_medium"),
  replogle_large_resamp = list(run_name = "run_all_but_mixscale_resamp", dataset_name = "replogle-rd7_pos_control_large"),
  # replogle_large = list(run_name = "run_all_but_mixscale_skew-t", dataset_name = "replogle-rd7_pos_control_large"),
  
  replogle_medium = list(run_name = "test_medium", dataset_name = "replogle-rd7_pos_control_medium")
)
# datasets <- list(
#   replogle_medium = list(run_name = "test_medium", dataset_name = "replogle-rd7_pos_control_medium")
# )


base_fp <- file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/outputs")

load_associations <- function(base_fp, dataset_name) {
  scep = read.csv(paste0(base_fp, "/association_on_target_sceptre_", dataset_name, ".csv"))
  if(file.exists(mixscale_fp <- paste0(base_fp, "/association_on_target_mixscale_", dataset_name, ".csv"))) {
    mix = read.csv(mixscale_fp)
  } else {
    mix = NULL
  }
  
  
  # FR-Perturb: p-values and logFC (base e, not base 2, so we convert) 
  frpert_pvals_raw <- read.table(paste0(base_fp, "/frperturb/frperturb_results_", dataset_name, "_pvals.txt"))
  stopifnot("non.targeting" %in% colnames(frpert_pvals_raw), !"non-targeting" %in% rownames(frpert_pvals_raw))
  frpert_lfc_raw <- read.table(paste0(base_fp, "/frperturb/frperturb_results_", dataset_name, "_LFCs.txt"))
  stopifnot(rownames(frpert_pvals_raw) == rownames(frpert_lfc_raw))
  frpert <- data.frame(
    target = rownames(frpert_pvals_raw),
    p_value_frpert = sapply(rownames(frpert_pvals_raw), function(target) frpert_pvals_raw[target,target]),
    # convert from natural log to base 2
    log2fc_frpert = sapply(rownames(frpert_pvals_raw), function(target) frpert_lfc_raw[target,target]) / log(2)
  )

  out = list(scep=scep, mix=mix, frpert = frpert)
  out[!sapply(out, is.null)]
}

collect_results <- function(base_fp, dataset_name) {
  assocs = load_associations(base_fp, dataset_name)
  mixscale_used <- "mix" %in% names(assocs)
  results <- dplyr::select(assocs$scep, log2fc_scep = log_2_fold_change, p_value_scep = p_value, target = grna_target) |>
    full_join(assocs$frpert, by = "target")
  if(mixscale_used) {
    results <- results |> 
      full_join(
        dplyr::select(assocs$mix, target=gene_ID, log2fc_mix = log2FC, p_value_mix = p_weight, mixscale_method = DE_method),
        by = "target"
      ) 
  }
  results
}

# results <- collect_results(base_fp, dataset_name)

all_results <- lapply(datasets, function(run_params) {
  run_name = run_params$run_name
  dataset_name = run_params$dataset_name
  fp <- file.path(base_fp, run_name)
  collect_results(fp, dataset_name)
}) |>
  setNames(names(datasets))

```



```{r, echo = FALSE}
library(dplyr)
library(ggplot2)
library(cowplot)
library(scales)


make_scatter <- function(df, xcol, ycol, xname, yname, use_revlog = FALSE) {
  p <- ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
    geom_point(alpha = 0.4, size = 0.6) +
    theme_bw() +
    geom_abline(slope = 1, intercept = 0, col = "grey", linetype="dashed") +
    labs(
      x = paste0(xname, " p-value (−log10)"),
      y = paste0(yname, " p-value (−log10)")
    )
  if(use_revlog) {
    p <- p + scale_x_continuous(trans = revlog_trans(10)) +
    scale_y_continuous(trans = revlog_trans(10)) 
  }
  p
}

plot_pairwise_p_values <- function(results, dataset_name, eps = 1e-250, remove_fr_perturb_outliers = FALSE) {
  dfp <- results %>%
    dplyr::select(p_value_scep, p_value_frpert) 

  mixscale_used <- "p_value_mix" %in% colnames(results)
  
  if(mixscale_used) {
    dfp$p_value_mix <- results$p_value_mix
  }
  
  dfp <- dfp %>%
    mutate(across(everything(), ~ pmax(.x, eps)))
  
  if(remove_fr_perturb_outliers) {
    dfp <- dfp |> dplyr::filter(p_value_frpert > eps)
  }
  
  cmplt <- complete.cases(dfp)
  if(!all(cmplt)) {
    dfp <- dfp[cmplt,]
    cat(sum(!cmplt), "rows removed due to incomplete cases.\n")
  }
  
  pltlist <- if(mixscale_used) {
    list(
      p1 = make_scatter(dfp, "p_value_scep",   "p_value_mix",    "SCEPTRE", "Mixscale", use_revlog = TRUE),
      p2 = make_scatter(dfp, "p_value_scep",   "p_value_frpert", "SCEPTRE", "FR-Pert", use_revlog = TRUE),
      p3 = make_scatter(dfp, "p_value_mix",    "p_value_frpert", "Mixscale","FR-Pert", use_revlog = TRUE)
    )
  } else {
    list(p1 = make_scatter(dfp, "p_value_scep",   "p_value_frpert", "SCEPTRE", "FR-Pert", use_revlog = TRUE))
  }
  
  
  title <- cowplot::ggdraw() +
    cowplot::draw_label(
      paste("Pairwise comparison of on-target p-values for", dataset_name),
      fontface = "bold",
      x = 0.5, hjust = 0.5
    ) +
    theme(plot.margin = margin(0, 0, 0, 0))
  
  grid <- cowplot::plot_grid(plotlist = pltlist, nrow = ifelse(mixscale_used, 2, 1), ncol=ifelse(mixscale_used, 2, 1), align = "hv")
  
  cowplot::plot_grid(
    title,
    grid,
    ncol = 1,
    rel_heights = c(0.12, 1)
  )
}
plot_pairwise_p_values(results = all_results$replogle_medium, dataset_name = "replogle_medium skew")
# plot_pairwise_p_values(results = all_results$replogle_medium_resamp, dataset_name = "replogle_medium resamp")
# plot_pairwise_p_values(results = all_results$replogle_large, dataset_name = "replogle_large")
plot_pairwise_p_values(results = all_results$gasperini_skew, dataset_name = "gasperini_pos_control skew")
plot_pairwise_p_values(results = all_results$gasperini_resamp, dataset_name = "gasperini_pos_control resamp")
plot_pairwise_p_values(results = all_results$replogle_large_skew, dataset_name = "replogle_pos_control skew")
plot_pairwise_p_values(results = all_results$replogle_large_resamp, dataset_name = "replogle_pos_control resamp")
# Map(plot_pairwise_p_values, all_results, names(all_results))
```

FR-Perturb has very poor agreement with sceptre. As sceptre agrees with Mixscale, and has generally (much) smaller p-values for the on-targets, I think sceptre is doing better. But it's not clear to me why the disagreement is so strong. 

For replogle_medium, they have some agreement, for both methods although moreso for skew-t. 

But for gasperini, essentially none at all.

```{r}
lapply(all_results, function(results) cor(results$p_value_scep, results$p_value_frpert, use = "complete.obs"))
```



```{r, eval=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(cowplot)
library(scales)



make_scatter <- function(df, xcol, ycol, xname, yname) {
  ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
    geom_point(alpha = 0.4, size = 0.6) +
    scale_x_continuous(trans = revlog_trans(10)) +
    scale_y_continuous(trans = revlog_trans(10)) +
    theme_bw() +
    geom_abline(slope = 1, intercept = 0, col = "grey", linetype="dashed") +
    labs(
      x = paste0(xname, " p-value (−log10)"),
      y = paste0(yname, " p-value (−log10)")
    )
}

plot_pairwise_logfc <- function(results, dataset_name) {
  dffc <- results %>%
    dplyr::select(log2fc_scep, log2fc_frpert) 
  
  mixscale_used <- "log2fc_mix" %in% colnames(results)
  
  if(mixscale_used) {
    dfp$log2fc_mix <- results$log2fc_mix
  }
  
  cmplt <- complete.cases(dffc)
  if(!all(cmplt)) {
    dffc <- dffc[cmplt,]
    cat(sum(!cmplt), "rows removed due to incomplete cases.\n")
  }
  
  
  
  if(mixscale_used) {
  pltlist <- list(
    p1 = make_scatter(dffc, "log2fc_scep",   "log2fc_mix",    "SCEPTRE", "Mixscale"),
    p2 = make_scatter(dffc, "log2fc_scep",   "log2fc_frpert", "SCEPTRE", "FR-Pert"),
    p3 = make_scatter(dffc, "log2fc_mix",    "log2fc_frpert", "Mixscale","FR-Pert")
  )
} else {
  pltlist <- list(p1 = make_scatter(dffc, "log2fc_scep",   "log2fc_frpert", "SCEPTRE", "FR-Pert"))
}
  
  
  
  
  
  
  
  
  
  pltlist <- if(mixscale_used) {
    list(
      p1 = make_scatter(dfp, "p_value_scep",   "p_value_mix",    "SCEPTRE", "Mixscale"),
      p2 = make_scatter(dfp, "p_value_scep",   "p_value_frpert", "SCEPTRE", "FR-Pert"),
      p3 = make_scatter(dfp, "p_value_mix",    "p_value_frpert", "Mixscale","FR-Pert")
    )
  } else {
    list(p1 = make_scatter(dfp, "p_value_scep",   "p_value_frpert", "SCEPTRE", "FR-Pert"))
  }
  
  
  title <- cowplot::ggdraw() +
    cowplot::draw_label(
      paste("Pairwise comparison of on-target p-values for", dataset_name),
      fontface = "bold",
      x = 0.5, hjust = 0.5
    ) +
    theme(plot.margin = margin(0, 0, 0, 0))
  
  grid <- cowplot::plot_grid(plotlist = pltlist, nrow = ifelse(mixscale_used, 2, 1), ncol=ifelse(mixscale_used, 2, 1), align = "hv")
  
  cowplot::plot_grid(
    title,
    grid,
    ncol = 1,
    rel_heights = c(0.12, 1)
  )
}



dffc <- results %>%
  dplyr::select(log2fc_scep, log2fc_frpert)
if(mixscale_used) {
  dffc$log2fc_mix <- results$log2fc_mix
}
  # mutate(across(everything(), ~ pmax(.x, eps)))   # avoid 0s on log scale

make_scatter <- function(df, xcol, ycol, xname, yname) {
  ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
    geom_point(alpha = 0.4, size = 0.6) +
    # scale_x_continuous(trans = revlog_trans(10)) +
    # scale_y_continuous(trans = revlog_trans(10)) +
    theme_bw() +
    geom_abline(slope = 1, intercept = 0, col = "grey", linetype="dashed") +
    labs(
      x = paste0(xname, " log2 fold change"),
      y = paste0(yname, " log2 fold change")
    )
}

if(mixscale_used) {
  pltlist <- list(
    p1 = make_scatter(dffc, "log2fc_scep",   "log2fc_mix",    "SCEPTRE", "Mixscale"),
    p2 = make_scatter(dffc, "log2fc_scep",   "log2fc_frpert", "SCEPTRE", "FR-Pert"),
    p3 = make_scatter(dffc, "log2fc_mix",    "log2fc_frpert", "Mixscale","FR-Pert")
  )
} else {
  pltlist <- list(p1 = make_scatter(dffc, "log2fc_scep",   "log2fc_frpert", "SCEPTRE", "FR-Pert"))
}

title <- cowplot::ggdraw() +
  cowplot::draw_label(
    paste("Pairwise comparison of on-target log2 FC for", dataset_name),
    fontface = "bold",
    x = 0.5, hjust = 0.5
  ) +
  theme(plot.margin = margin(0, 0, 0, 0))


grid <- cowplot::plot_grid(plotlist = pltlist, nrow = ifelse(mixscale_used, 2, 1), ncol=ifelse(mixscale_used, 2, 1), align = "hv")

cowplot::plot_grid(
  title,
  grid,
  ncol = 1,
  rel_heights = c(0.12, 1)
)
```


violin plots of p-values
```{r, eval=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

p_floor <- 1e-250

make_violin <- function(results, title) {
  plot_df <- results %>%
  transmute(
    sceptre    = p_value_scep,
    # mixscale   = p_value_mix,
    `FR-Perturb`  = p_value_frpert
  ) %>%
  pivot_longer(everything(), names_to = "method", values_to = "p") %>%
  mutate(
    p = pmax(p, p_floor),              # avoid -Inf from p=0
    neglog10_p = -log10(p),
    method = factor(method, levels = c("sceptre",  "FR-Perturb")) # "mixscale",
  )

ggplot(plot_df, aes(x = method, y = p)) +
  geom_violin(trim = FALSE) +
  scale_y_continuous(trans = revlog_trans(10)) +
  labs(x = NULL, y = "p-value (-log10)", title = title) +
  theme_bw()

}

make_violin(all_results$gasperini_skew, paste0("p-value distribution by method for gasperini skew"))
make_violin(all_results$gasperini_resamp, paste0("p-value distribution by method for gasperini resamp"))
# TODO combine all on a single violin plot
```

```{r, echo=FALSE}
# sanity check: at least they are decidedly not uniform
# hist(all_results$gasperini_resamp$p_value_frpert, 30, main = "gasperini resamp")



par(mfcol = c(2,1))
hist(-log10(all_results$gasperini_skew$p_value_frpert), 50, main = "gasperini skew")
hist(-log10(all_results$gasperini_skew$p_value_scep), 50, main = "gasperini skew")
par(mfcol = c(1,1))
# 
# tmp <- -log10(all_results$gasperini_skew$p_value_frpert)
# hist(tmp[tmp < 15], 50, main = "truncated gasperini skew", xlab = "p-values (-log10) (truncated to < 15)")
# 
# 
# qqplot(-log10(all_results$gasperini_skew$p_value_frpert), -log10(all_results$gasperini_skew$p_value_scep))
# abline(0,1)
# 
# # sanity check: there is pretty good correlation between the p-values for the two ones
# plot(all_results$gasperini_resamp$p_value_frpert, all_results$gasperini_skew$p_value_frpert)
# plot(-log10(all_results$gasperini_resamp$p_value_frpert), -log10(all_results$gasperini_skew$p_value_frpert))
# abline(0,1)
```



plot: i want to indicate the sample size somehow
```{r, echo=FALSE}
# run_name <- "gasperini_skew"
# dataset_name <- "gasperini_pos_control"
run_name <- "replogle_large_skew"
dataset_name <- "replogle-rd7_pos_control_large"
results <- all_results[[run_name]]
eps <- 1e-250
# eps <- 1e-50
# i want to do pairwise p-value plot but with sample size indicated somehow
pdat <- results |> dplyr::select(p_value_scep, p_value_frpert, target) |>
  mutate(p_value_scep = pmax(p_value_scep, eps), p_value_frpert = pmax(p_value_frpert, eps))



cell_info <- read_csv(file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/input_data", dataset_name, "cell_info.csv"))
head(cell_info)

# so... what represents my sample size here? the total number of cells that have that grna as a target?
sample_sizes <- cell_info |> group_by(grna_target) |> summarize(n = n()) |> dplyr::select(target = grna_target, n)
pdat %<>% left_join(sample_sizes, by = "target")

pdat |> transmute(neglog_p_value_scep = -log10(p_value_scep), neglog_p_value_frpert = -log10(p_value_frpert), n = n) %>%
  filter(complete.cases(.)) |>
  cor()

# this doesn't show much
# pdat |>
#   # mutate(n = cut(n, breaks = 3)) |>
#   # dplyr::filter(p_value_frpert > eps) |>
#   ggplot(aes(x = p_value_scep, y = p_value_frpert, color = n)) +
#   geom_point(alpha = 0.25, size = 0.7) +
#   # facet_grid(~n) +
#   scale_color_viridis_c(option = "C", end = 0.95) +
#   scale_x_continuous(trans = revlog_trans(10)) +
#   scale_y_continuous(trans = revlog_trans(10)) +
#   geom_abline(slope = 1, intercept = 0, col = "grey60", linetype = "dashed") +
#   theme_bw() +
#   labs(
#     x = "SCEPTRE p-value (−log10)",
#     y = "FR-Pert p-value (−log10)",
#     color = "n"
#   )
breaks <- list(
  gasperini_pos_control = c(339, 1050, 1800, 2509),
  `replogle-rd7_pos_control_large` = c(16, 63, 117, 3782)
)[[dataset_name]]

pdat |>
  mutate(cells_per_target = cut(n, breaks = breaks, dig.lab=4, ordered_result = TRUE)) %>%
  split(f = .$cells_per_target) |>
  lapply(function(df) data.frame(cor = cor(-log10(df$p_value_scep), -log10(df$p_value_frpert), use="complete.obs") |> round(2))) %>%
  do.call(what = rbind) %>%
  mutate(cells_per_target = rownames(.)) |>
  `rownames<-`(NULL)

n_max <- Inf
data.frame(
  p_value = -log10(c(pdat$p_value_scep, pdat$p_value_frpert)),
  method = rep(c("sceptre", "FR_perturb"), each = nrow(pdat)),
  n = rep(pdat$n, times = 2)
) |>
  filter(n <= n_max) |>
  ggplot(aes(y=  p_value, x  =n, color = method)) +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~method) +
   geom_smooth(method = "lm", se = FALSE) +
  # scale_y_continuous(trans = revlog_trans(10)) +
labs(x = "total number of cells expressing this target", 
     y = "p-value (-log10)",
     title = paste0("On-target sample size vs p-value for ", dataset_name)
) +
  theme_bw()
pdat |>
  dplyr::select(p_value_scep, )
```



plot ideas
- can i do a single plot with all 3 QQ plots vs uniform? Use sceptre's qq plot layers
- do a violin plot with all datasets and methods in one
- how can i show the correlation between methods, and maybe also with sample size? can i do something like for each n, get cor[p1,p2|n], then smooth that? or, bigger range of n --> smooth on the n scale. 


overall goals for this writeup: 
- i want this to be my bundled up pos control section, showing how we're approaching it, and giving a convincing show that this effort is nearly concluded. 
- for TODOs, i guess more datasets but i don't have any insight into that right now. 

TODO

- computational assessment: make downsampled datasets for this that are more representative, like similar scale of gene expression, etc to be kind of "unbiased"


