---
title: "Association: Positive Control Analysis"
output:
  pdf_document: default
  html_document: default
date: "2026-01-17"
header-includes:
  - \usepackage{booktabs}
  - \usepackage{colortbl}
  - \usepackage{xcolor}
---


```{r, echo=FALSE, include=FALSE}
library(tidyverse)
library(magrittr)
library(purrr)
library(cowplot)
library(scales)
library(sceptre)
```

Methods

1. Mixscale (low MOI only)

Preprocessing:
- Standard Seurat workflow: normalize data, find variable features, scale data, run PCA

Perturbation Signature Calculation (CalcPerturbSig):

- For each cell, find its 20 nearest non-targeting (NT) neighbors in PCA space with the same biological replicate

- Compute perturbation vector: p_i = x_i - Avg(neighbors)

- This controls for biological replicate and systematic effects

Mixscale Scoring (RunMixscale):

- For each target, test all other genes (not the target itself) for differential expression

- If at least 5 genes are marginally DE, proceed; otherwise skip to simpler test

- Keep only those DE genes' measurements (at most 100 by default) in the perturbation vectors

- Compute perturbation scores by projecting onto the line from the perturbed cells' center to NT cells' center

- Compute standardized perturbation scores by z-transforming using the mean and std dev from NT cells

- NT cells centered around 0, perturbed cells "significantly" far away

Differential Expression Test (Run_wmvRegDE):

- For each target-gene pair that got Mixscale scores, run negative binomial regression:

- Response: gene expression values

- Predictors: perturbation scores (0 for NT cells, standardized score for perturbed cells) + log10 library size

- The p-value comes from testing whether the coefficient on the perturbation score is significant

- This is a weighted DE test where the weight reflects how perturbed each cell is


Output:

- Returns p-values for each target-gene pair (currently all pairs, though on-target only would be more efficient)

2. FR-Perturb

- We compare two versions: using their skew-t approximation, or resampling with 2000 samples.

3. SCEPTRE

TODO

Datasets:


```{r, eval=FALSE, include=FALSE}
# not run
dataset_names <- c("replogle-rd7_large", "gasperini_pos_control")
for(dataset in dataset_names) {
  base_path <- file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/input_data", dataset)

  response_mat <- read_rds(file.path(base_path, "sceptre/response_matrix.rds"))
  # cell_info <- read_csv(file.path(base_path, "cell_info.csv"), show_col_types = FALSE)
  grna_target_df <- read_csv(file.path(base_path, "sceptre/grna_target_data_frame.csv"), show_col_types = FALSE)
  
  cat("Dataset name:", dataset, "\n")
  cat("Number of cells:", ncol(response_mat), "; number of genes:", nrow(response_mat),
      "; number of unique guides:",  length(unique(grna_target_df$grna_id)), "\n")
}
```

Dataset name: `replogle-rd7_large`                                                                                                                                                                        
Number of cells: 210110 ; number of genes: 1619 ; number of unique guides: 1813 

Dataset name: `gasperini_pos_control`                                                                                                                                                                          
Number of cells: 185171 ; number of genes: 377 ; number of unique guides: 855 

1. Subsampled Replogle 

- run `sceptre-pipeline` on the full Replogle-rd7. method = mixture, with the default formula used, so we regressed on `response_n_nonzero`, `response_n_umis`, `grna_n_nonzero`, and `grna_n_umis` (all log or log1p as needed). `batch` was not used because it has 56 levels. 

Then, to get the desired subset of the response matrix:

- Filter to low MOI cells: Keep only cells expressing exactly one gRNA

  - Select positive control cells:
  
    - Iterate through response genes in dataset order
    
    - For each gene, collect all cells expressing gRNAs targeting that gene
    
    - Skip genes with fewer than min_num_cells_per_target cells (default: 20, used for "large")
    
    - Continue until reaching num_pc_cells positive control cells (e.g., 200,000 for "large")
    
  - Add non-targeting cells: Include cells expressing non-targeting gRNAs up to num_nt_cells (e.g., 10,000 for "large"), up to all of them
  
  - Extract expression data: Subset the response matrix to selected genes and cells
  
  - Quality filtering: Remove cells with zero expression across all genes, then remove any gRNAs left with zero cells
  
  - Export in multiple formats: Write data for sceptre (binary indicator matrix), mixscale (named vectors), and
  FR-Perturb (h5ad)


The resulting `response_matrix` resembles a block diagonal one, with ambient noise off the diagonal, as the first gene's targeting cells are the first cells in, and so on, until the final columns of NT cells. 

2. Subsampled Gasperini

- Run `sceptre-pipeline` on the full Gasperini data, method = mixture, with the default formula used, so we regressed on `response_n_nonzero`, `response_n_umis`, `grna_n_nonzero`, `grna_n_umis` (all log or log1p as needed), and `batch` (2 levels)

- Identify on-target genes: Find genes that are both targeted by gRNAs and present in the response matrix

  - Collect cells with on-target gRNAs:
  
    - For each on-target gene, get all gRNAs targeting that gene
    
    - Collect all cells expressing any of these gRNAs (no filtering to single gRNA per cell - this is high MOI)
    
  - Add non-targeting cells: Include all cells expressing non-targeting gRNAs
  
  - Create cell-level assignments: Build long-form dataframe mapping each cell to its expressed gRNAs (cells can have
  multiple gRNAs)
  
  - Extract expression data: Subset response matrix to on-target genes and selected cells
  
  - Compute gRNA metrics: Calculate grna_n_nonzero_subset and grna_n_umis_subset from the gRNA UMI counts for the cell
  subset
  
  - Export in multiple formats:
  
    - sceptre: Binary indicator matrix, cell covariates including prep_batch
    
    - FR-Perturb: h5ad file with colon-delimited multi-perturbation column (e.g., "GENE1:GENE2" for cells with multiple
  gRNAs), includes prep_batch covariate
  

```{r, echo = FALSE, include = FALSE}
rm(list=ls())


revlog_trans <- function (base = exp(1)) 
{
    trans <- function(x) {
        -log(x, base)
    }
    inv <- function(x) {
        base^(-x)
    }
    scales::trans_new(name = paste("revlog-", base, sep = ""), 
        transform = trans, inverse = inv, breaks = scales::log_breaks(base = base), 
        domain = c(1e-250, Inf))
}

datasets <- list(
  gasperini_skew = list(run_name = "run_all_skew-t", dataset_name = "gasperini_pos_control"),
  replogle_large_skew = list(run_name = "run_all_skew-t", dataset_name = "replogle-rd7_pos_control_large"),
  
  gasperini_resamp = list(run_name = "run_all_resamp", dataset_name = "gasperini_pos_control"),
  replogle_large_resamp = list(run_name = "run_all_resamp", dataset_name = "replogle-rd7_pos_control_large")
)

base_fp <- file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/outputs")

load_associations <- function(base_fp, dataset_name) {
  scep = read.csv(paste0(base_fp, "/association_on_target_sceptre_", dataset_name, ".csv"))
  if(file.exists(mixscale_fp <- paste0(base_fp, "/association_on_target_mixscale_", dataset_name, ".csv"))) {
    mix = read.csv(mixscale_fp)
  } else {
    mix = NULL
  }
  
  
  # FR-Perturb: p-values and logFC (base e, not base 2, so we convert) 
  frpert_pvals_raw <- read.table(paste0(base_fp, "/frperturb/frperturb_results_", dataset_name, "_pvals.txt"))
  stopifnot("non.targeting" %in% colnames(frpert_pvals_raw), !"non-targeting" %in% rownames(frpert_pvals_raw))
  frpert_lfc_raw <- read.table(paste0(base_fp, "/frperturb/frperturb_results_", dataset_name, "_LFCs.txt"))
  stopifnot(rownames(frpert_pvals_raw) == rownames(frpert_lfc_raw))
  frpert <- data.frame(
    target = rownames(frpert_pvals_raw),
    p_value_frpert = sapply(rownames(frpert_pvals_raw), function(target) frpert_pvals_raw[target,target]),
    # convert from natural log to base 2
    log2fc_frpert = sapply(rownames(frpert_pvals_raw), function(target) frpert_lfc_raw[target,target]) / log(2)
  )

  out = list(scep=scep, mix=mix, frpert = frpert)
  out[!sapply(out, is.null)]
}

collect_results <- function(base_fp, dataset_name) {
  assocs = load_associations(base_fp, dataset_name)
  mixscale_used <- "mix" %in% names(assocs)
  results <- dplyr::select(assocs$scep, log2fc_scep = log_2_fold_change, p_value_scep = p_value, target = grna_target) |>
    full_join(assocs$frpert, by = "target")
  if(mixscale_used) {
    results <- results |> 
      full_join(
        dplyr::select(assocs$mix, target=gene_ID, log2fc_mix = log2FC, p_value_mix = p_weight, mixscale_method = DE_method),
        by = "target"
      ) 
  }
  results
}

# results <- collect_results(base_fp, dataset_name)

all_results <- lapply(datasets, function(run_params) {
  run_name = run_params$run_name
  dataset_name = run_params$dataset_name
  fp <- file.path(base_fp, run_name)
  collect_results(fp, dataset_name)
}) |>
  setNames(names(datasets))

# now i want to match in the # expressed cells to each one
for(result_name in names(all_results)) {
  dataset_name <- datasets[[result_name]]$dataset_name
  data_fp <- file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/input_data", dataset_name)
  cell_info <- read_csv(file.path(data_fp, "cell_info.csv"), show_col_types = FALSE)
  sample_sizes <- cell_info |> group_by(grna_target) |> summarize(n = n()) |>
    dplyr::select(target = grna_target, num_expressed_cells = n)
  all_results[[result_name]] <- left_join(
    all_results[[result_name]],
    sample_sizes,
    by = "target"
    )
}
objs_to_keep <- c("objs_to_keep", "all_results", "datasets", "revlog_trans")
```



Beginning with some tables:

```{r, echo=FALSE}

# -------------------------
# Build a single long table of (dataset, method, target, n, p)
# -------------------------
all_long <- bind_rows(
  # GASPERINI
  all_results$gasperini_skew %>%
    transmute(dataset = "gasperini", target,
              n = num_expressed_cells,
              method = "SCEPTRE",
              p = as.numeric(p_value_scep)),
  all_results$gasperini_skew %>%
    transmute(dataset = "gasperini", target,
              n = num_expressed_cells,
              method = "FR-Perturb (skew)",
              p = as.numeric(p_value_frpert)),
  all_results$gasperini_resamp %>%
    transmute(dataset = "gasperini", target,
              n = num_expressed_cells,
              method = "FR-Perturb (resamp)",
              p = as.numeric(p_value_frpert)),

  # REPLOGLE
  all_results$replogle_large_skew %>%
    transmute(dataset = "replogle", target,
              n = num_expressed_cells,
              method = "SCEPTRE",
              p = as.numeric(p_value_scep)),
  all_results$replogle_large_skew %>%
    transmute(dataset = "replogle", target,
              n = num_expressed_cells,
              method = "Mixscale",
              p = as.numeric(p_value_mix)),
  all_results$replogle_large_skew %>%
    transmute(dataset = "replogle", target,
              n = num_expressed_cells,
              method = "FR-Perturb (skew)",
              p = as.numeric(p_value_frpert)),
  all_results$replogle_large_resamp %>%
    transmute(dataset = "replogle", target,
              n = num_expressed_cells,
              method = "FR-Perturb (resamp)",
              p = as.numeric(p_value_frpert))
) %>%
  mutate(
    p = pmin(pmax(p, 0, na.rm = TRUE), 1)  # clamp into [0,1]
  ) %>%
  filter(is.finite(n), is.finite(p), !is.na(n), !is.na(p))

# =========================================================
# (1) Table over each (dataset, method)
#     - mean p-value
#     - Spearman cor(n, -p)   [so "bigger is better"]
#     - fraction p <= 0.05
# =========================================================
tbl_dataset_method <- all_long %>%
  group_by(dataset, method) %>%
  summarise(
    n_tests = n(),
    mean_p = mean(p, na.rm = TRUE),
    spearman_cor_n_negp = suppressWarnings(cor(n, -p, method = "spearman", use = "complete.obs")),
    frac_p_le_0.05 = mean(p <= 0.05, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(dataset, method)

# tbl_dataset_method


# =========================================================
# (2) Table over each (dataset, method1, method2) with method1<method2
#     - frac(p1 < p2)
#     - Spearman cor(p1, p2)
# =========================================================
make_pair_table <- function(df_long_one_dataset) {
  wide <- df_long_one_dataset %>%
    select(target, method, p) %>%
    distinct() %>%
    pivot_wider(names_from = method, values_from = p)

  methods <- setdiff(names(wide), "target")
  pairs <- combn(methods, 2, simplify = FALSE)

  bind_rows(lapply(pairs, function(pp) {
    m1 <- pp[1]; m2 <- pp[2]
    x <- wide[[m1]]
    y <- wide[[m2]]
    ok <- is.finite(x) & is.finite(y) & !is.na(x) & !is.na(y)

    tibble(
      method1 = m1,
      method2 = m2,
      n_targets = sum(ok),
      frac_p1_lt_p2 = mean(x[ok] < y[ok]),
      spearman_cor_p = suppressWarnings(cor(x[ok], y[ok], method = "spearman", use = "complete.obs"))
    )
  }))
}

tbl_dataset_pairs <- all_long %>%
  group_split(dataset) %>%
  map_dfr(function(df) {
    ds <- unique(df$dataset)
    make_pair_table(df) %>% mutate(dataset = ds, .before = 1)
  }) %>%
  arrange(dataset, method1, method2)

# tbl_dataset_pairs
```

```{r, echo=FALSE, eval=FALSE}

tbl_dataset_method %>%
  transmute(
    Dataset = str_to_title(dataset),
    Method = method,
    `Num. Targets` = n_tests,
    `Mean p-value` = signif(mean_p, 3),
    `Corr. with n` = round(spearman_cor_n_negp, 3),
    `Pcnt. <= 0.05` = paste0(round(frac_p_le_0.05, 3) * 100, "%")
  ) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Summary statistics by dataset and method. The column 'Corr. with n' gives the Spearman correlation of the negative p-value with the number of cells expressing each target. 'Pcnt. $\\leq$ 0.05' gives the percentage of p-values at or below 0.05. Both datasets are downsampled to only have on-targets."
  ) %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "striped"))
```

\begin{table}[!h]
\centering
\caption{Summary statistics by dataset and method. The column 'Corr. with n' gives the Spearman correlation of the negative p-value with the number of cells expressing each target. 'Pcnt. $\leq$ 0.05' gives the percentage of p-values at or below 0.05. Both datasets are downsampled to only have on-targets.}
\centering
\begin{tabular}[t]{llrrrl}
\toprule
Dataset & Method & Num. Targets & Mean p-value & Corr. with n & Pcnt. $\leq$ 0.05\\
\midrule
\cellcolor{gray!10}{Gasperini} & \cellcolor{gray!10}{FR-Perturb (resamp)} & \cellcolor{gray!10}{377} & \cellcolor{gray!10}{0.2510} & \cellcolor{gray!10}{0.031} & \cellcolor{gray!10}{43\%}\\
Gasperini & FR-Perturb (skew) & 377 & 0.2530 & 0.022 & 43.5\%\\
\cellcolor{gray!10}{Gasperini} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{377} & \cellcolor{gray!10}{0.0113} & \cellcolor{gray!10}{0.177} & \cellcolor{gray!10}{98.1\%}\\
\addlinespace
Replogle & FR-Perturb (resamp) & 1619 & 0.2850 & 0.068 & 33.5\%\\
\cellcolor{gray!10}{Replogle} & \cellcolor{gray!10}{FR-Perturb (skew)} & \cellcolor{gray!10}{1619} & \cellcolor{gray!10}{0.2840} & \cellcolor{gray!10}{0.067} & \cellcolor{gray!10}{34.1\%}\\
Replogle & Mixscale & 1619 & 0.1650 & 0.313 & 78.1\%\\
\cellcolor{gray!10}{Replogle} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{1619} & \cellcolor{gray!10}{0.0456} & \cellcolor{gray!10}{0.318} & \cellcolor{gray!10}{87.7\%}\\
\bottomrule
\end{tabular}
\end{table}

```{r, echo=FALSE, eval=FALSE}
tbl_dataset_pairs %>%
  transmute(
    Dataset = str_to_title(dataset),
    `Method 1` = method1,
    `Method 2` = method2,
    `Pcnt. 1 <= 2` = paste0(100 * round(frac_p1_lt_p2, 3), "%"),
    `Corr.` = round(spearman_cor_p, 3)
  ) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Pairwise method comparisons within each dataset. 'Pcnt. 1 $\\leq$ 2' gives the percentage of times the p-value for Method 1 is at or below the p-value for Method 2. 'Corr.' gives the Spearman correlation between the p-values for Methods 1 and 2. Both datasets are downsampled to only have on-targets."
  ) %>%
  kableExtra::kable_styling(latex_options = c("hold_position", "striped"))
```

\begin{table}[!h]
\centering
\caption{Pairwise method comparisons within each dataset. 'Pcnt. 1 $\leq$ 2' gives the percentage of times the p-value for Method 1 is at or below the p-value for Method 2. 'Corr.' gives the Spearman correlation between the p-values for Methods 1 and 2. Both datasets are downsampled to only have on-targets.}
\centering
\begin{tabular}[t]{llllr}
\toprule
Dataset & Method 1 & Method 2 & Pcnt. 1 $\leq$ 2 & Corr.\\
\midrule
\cellcolor{gray!10}{Gasperini} & \cellcolor{gray!10}{FR-Perturb (skew)} & \cellcolor{gray!10}{FR-Perturb (resamp)} & \cellcolor{gray!10}{59.4\%} & \cellcolor{gray!10}{0.990}\\
Gasperini & SCEPTRE & FR-Perturb (resamp) & 98.4\% & 0.233\\
\cellcolor{gray!10}{Gasperini} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{FR-Perturb (skew)} & \cellcolor{gray!10}{98.1\%} & \cellcolor{gray!10}{0.238}\\
\addlinespace
Replogle & FR-Perturb (skew) & FR-Perturb (resamp) & 55.8\% & 0.996\\
\cellcolor{gray!10}{Replogle} & \cellcolor{gray!10}{Mixscale} & \cellcolor{gray!10}{FR-Perturb (resamp)} & \cellcolor{gray!10}{77.9\%} & \cellcolor{gray!10}{0.007}\\

Replogle & Mixscale & FR-Perturb (skew) & 76\% & 0.005\\
\cellcolor{gray!10}{Replogle} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{FR-Perturb (resamp)} & \cellcolor{gray!10}{89\%} & \cellcolor{gray!10}{0.014}\\
Replogle & SCEPTRE & FR-Perturb (skew) & 86.8\% & 0.013\\
\cellcolor{gray!10}{Replogle} & \cellcolor{gray!10}{SCEPTRE} & \cellcolor{gray!10}{Mixscale} & \cellcolor{gray!10}{81.1\%} & \cellcolor{gray!10}{0.926}\\
\bottomrule
\end{tabular}
\end{table}

# Visualization of on-target p-values

We begin with violin plots to show the marginal distribution on the -log10 scale. Then QQ plots give more insight.

```{r, echo = FALSE}
clip_p <- function(p, eps = 1e-250) pmax(as.numeric(p), eps, na.rm = TRUE)

# -------------------------
# Build long table exactly as you described
# -------------------------
long_p <- bind_rows(
  # GASPERINI: sceptre (take from either skew or resamp; they should match for sceptre)
  all_results$gasperini_skew %>%
    transmute(
      screen = "gasperini",
      method = "sceptre",
      neglog10p = -log10(clip_p(p_value_scep))
    ),
  # GASPERINI: FR-Perturb (skew)
  all_results$gasperini_skew %>%
    transmute(
      screen = "gasperini",
      method = "FR-Perturb (skew)",
      neglog10p = -log10(clip_p(p_value_frpert))
    ),
  # GASPERINI: FR-Perturb (resamp)
  all_results$gasperini_resamp %>%
    transmute(
      screen = "gasperini",
      method = "FR-Perturb (resamp)",
      neglog10p = -log10(clip_p(p_value_frpert))
    ),

  # REPLOGLE: sceptre (again, take from either)
  all_results$replogle_large_skew %>%
    transmute(
      screen = "replogle",
      method = "sceptre",
      neglog10p = -log10(clip_p(p_value_scep))
    ),
  # REPLOGLE: mixscale
  all_results$replogle_large_skew %>%
    transmute(
      screen = "replogle",
      method = "mixscale",
      neglog10p = -log10(clip_p(p_value_mix))
    ),
  # REPLOGLE: FR-Perturb (skew)
  all_results$replogle_large_skew %>%
    transmute(
      screen = "replogle",
      method = "FR-Perturb (skew)",
      neglog10p = -log10(clip_p(p_value_frpert))
    ),
  # REPLOGLE: FR-Perturb (resamp)
  all_results$replogle_large_resamp %>%
    transmute(
      screen = "replogle",
      method = "FR-Perturb (resamp)",
      neglog10p = -log10(clip_p(p_value_frpert))
    )
) %>%
  filter(is.finite(neglog10p)) %>%
  mutate(
    method = factor(
      method,
      levels = c("sceptre", "mixscale", "FR-Perturb (skew)", "FR-Perturb (resamp)")
    ),
    screen = factor(screen, levels = c("gasperini", "replogle"))
  )
```

To begin with, we simply show the marginal distributions of the p-values side by side.

```{r, echo=FALSE}
long_p_nice <- long_p %>%
  mutate(
    screen = recode(screen,
                    gasperini = "Gasperini (on-target)",
                    replogle   = "Replogle (on-target)")
  )
ggplot(long_p_nice, aes(x = method, y = neglog10p)) +
  geom_violin(trim = TRUE, scale = "width") +
  facet_wrap(~ screen, scales = "free") +   # <-- free x AND y
  scale_x_discrete(drop = TRUE) +
  labs(x = "Method", y = "p-value (-log10 scale)", title = "Comparison of on-target p-values") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    panel.grid.minor = element_blank()
  )
```


```{r, echo=FALSE}

# Gasperini p-values (original scale)
gasperini_long <- bind_rows(
  all_results$gasperini_skew %>%
    transmute(method = "SCEPTRE", p = p_value_scep),
  all_results$gasperini_skew %>%
    transmute(method = "FR-Perturb (skew)", p = p_value_frpert),
  all_results$gasperini_resamp %>%
    transmute(method = "FR-Perturb (resamp)", p = p_value_frpert)
) %>%
  mutate(
    p = as.numeric(p),
    p = pmin(pmax(p, 1e-300, na.rm = TRUE), 1),
    method = factor(method, levels = c("SCEPTRE", "FR-Perturb (skew)", "FR-Perturb (resamp)"))
  ) %>%
  filter(is.finite(p), !is.na(p))

# Compute expected uniform quantiles within each method
qq_df_gas <- gasperini_long %>%
  group_by(method) %>%
  arrange(p, .by_group = TRUE) %>%
  mutate(
    n = n(),
    expected = (row_number() - 0.5) / n
  ) %>%
  ungroup()

# QQ plot: observed p vs expected uniform p, one facet per method
qq_gasp <- ggplot(qq_df_gas, aes(x = expected, y = p)) +
  geom_point(alpha = 0.5, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  facet_wrap(~ method, ncol = 3) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +  # <- square panels
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Gasperini: QQ plots of p-values vs Uniform(0,1)",
    x = "Expected Uniform(0,1) quantiles",
    y = "Observed p-values"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

# Replogle p-values (original scale)
replogle_long <- bind_rows(
  all_results$replogle_large_skew %>%
    transmute(method = "SCEPTRE", p = p_value_scep),
  all_results$replogle_large_skew %>%
    transmute(method = "Mixscale", p = p_value_mix),
  all_results$replogle_large_skew %>%
    transmute(method = "FR-Perturb (skew)", p = p_value_frpert),
  all_results$replogle_large_resamp %>%
    transmute(method = "FR-Perturb (resamp)", p = p_value_frpert)
) %>%
  mutate(
    p = as.numeric(p),
    p = pmin(pmax(p, 1e-300, na.rm = TRUE), 1),   # clamp to (0,1]
    method = factor(method, levels = c("SCEPTRE", "Mixscale", "FR-Perturb (skew)", "FR-Perturb (resamp)"))
  ) %>%
  filter(is.finite(p), !is.na(p))

# Helper to compute expected uniform quantiles within each method
qq_df <- replogle_long %>%
  group_by(method) %>%
  arrange(p, .by_group = TRUE) %>%
  mutate(
    n = n(),
    expected = (row_number() - 0.5) / n
  ) %>%
  ungroup()

# QQ plot: observed p vs expected uniform p, one facet per method
qq_repl <- ggplot(qq_df, aes(x = expected, y = p)) +
  geom_point(alpha = 0.5, size = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = 2) +
  facet_wrap(~ method, ncol = 2) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +  # <- square panels
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Replogle: QQ plots of p-values vs Uniform(0,1)",
    x = "Expected Uniform(0,1) quantiles",
    y = "Observed p-values"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

print(qq_gasp)
print(qq_repl)
```


```{r, echo=FALSE}

neglog10 <- function(p, eps = 1e-250) -log10(pmax(as.numeric(p), eps, na.rm = TRUE))

# Build long Replogle table: one row per target per method
replogle_long <- bind_rows(
  all_results$replogle_large_skew %>%
    transmute(target,
              n = num_expressed_cells,
              method = "SCEPTRE",
              neglog10p = neglog10(p_value_scep)),
  all_results$replogle_large_skew %>%
    transmute(target,
              n = num_expressed_cells,
              method = "Mixscale",
              neglog10p = neglog10(p_value_mix)),
  all_results$replogle_large_skew %>%
    transmute(target,
              n = num_expressed_cells,
              method = "FR-Perturb (skew)",
              neglog10p = neglog10(p_value_frpert)),
  all_results$replogle_large_resamp %>%
    transmute(target,
              n = num_expressed_cells,
              method = "FR-Perturb (resamp)",
              neglog10p = neglog10(p_value_frpert))
) %>%
  filter(is.finite(n), is.finite(neglog10p), !is.na(n), !is.na(neglog10p)) %>%
  mutate(method = factor(method,
                         levels = c("SCEPTRE", "Mixscale", "FR-Perturb (skew)", "FR-Perturb (resamp)")))

# Single plot with all 4 methods + LOESS per method
ggplot(replogle_long, aes(x = n, y = neglog10p, color = method)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9, formula = y~x) +
  scale_x_continuous(trans = "log10") +   # comment out if you want raw n
  labs(
    title = "Replogle: -log10(p) vs number of on-target cells",
    x = "Number of on-target cells (log10 scale)",
    y = expression(-log[10](p)),
    color = NULL
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())


# Build long Gasperini table: one row per target per method
gasperini_long <- bind_rows(
  all_results$gasperini_skew %>%
    transmute(target,
              n = num_expressed_cells,
              method = "SCEPTRE",
              neglog10p = neglog10(p_value_scep)),
  all_results$gasperini_skew %>%
    transmute(target,
              n = num_expressed_cells,
              method = "FR-Perturb (skew)",
              neglog10p = neglog10(p_value_frpert)),
  all_results$gasperini_resamp %>%
    transmute(target,
              n = num_expressed_cells,
              method = "FR-Perturb (resamp)",
              neglog10p = neglog10(p_value_frpert))
) %>%
  filter(is.finite(n), is.finite(neglog10p), !is.na(n), !is.na(neglog10p)) %>%
  mutate(method = factor(method, levels = c("SCEPTRE", "FR-Perturb (skew)", "FR-Perturb (resamp)")))

ggplot(gasperini_long, aes(x = n, y = neglog10p, color = method)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.9,  formula = y~x) +
  scale_x_continuous(trans = "log10") +   # comment out for raw n
  labs(
    title = "Gasperini: -log10(p) vs number of on-target cells",
    x = "Number of on-target cells (log10 scale)",
    y = expression(-log[10](p)),
    color = NULL
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```



Next, we want to explore the extent to which these p-values agree.

```{r, echo = FALSE}
rm(list = setdiff(ls(), objs_to_keep))
neglog10 <- function(p, eps = 1e-250) -log10(pmax(as.numeric(p), eps, na.rm = TRUE))

# --- wide tables on -log10(p) scale (one row per target) ---
gasperini_wide <- all_results$gasperini_skew %>%
  transmute(
    target,
    `SCEPTRE`             = neglog10(p_value_scep),
    `FR-Perturb (skew)`   = neglog10(p_value_frpert)
  ) %>%
  inner_join(
    all_results$gasperini_resamp %>%
      transmute(target, `FR-Perturb (resamp)` = neglog10(p_value_frpert)),
    by = "target"
  )

replogle_wide <- all_results$replogle_large_skew %>%
  transmute(
    target,
    `SCEPTRE`             = neglog10(p_value_scep),
    `Mixscale`            = neglog10(p_value_mix),
    `FR-Perturb (skew)`   = neglog10(p_value_frpert)
  ) %>%
  inner_join(
    all_results$replogle_large_resamp %>%
      transmute(target, `FR-Perturb (resamp)` = neglog10(p_value_frpert)),
    by = "target"
  )

# Use shared axis limits within each dataset (optional but nice for comparability)
gas_max <- max(as.matrix(select(gasperini_wide, -target)), na.rm = TRUE)
rep_max <- max(as.matrix(select(replogle_wide, -target)),  na.rm = TRUE)

make_scatter <- function(df, xcol, ycol, maxv, cor_method = "spearman") {
  x <- df[[xcol]]
  y <- df[[ycol]]
  r <- suppressWarnings(cor(x, y, use = "complete.obs", method = cor_method))

  ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
    geom_point(alpha = 0.3, size = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    coord_equal() +
    scale_x_continuous(limits = c(0, maxv)) +
    scale_y_continuous(limits = c(0, maxv)) +
    labs(
      title = sprintf("Spearman = %.1f%%", 100 * r),
      x = paste0(xcol, "\n(-log10 p)"),
      y = paste0(ycol, "\n(-log10 p)")
    ) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      axis.title = element_text(size = 9),
      axis.text  = element_text(size = 7),
      plot.title = element_text(size = 10)
    )
}

# library(hexbin)  # needed for geom_hex()
# 
# make_scatter <- function(df, xcol, ycol, maxv, cor_method = "spearman", bins = 40) {
#   x <- df[[xcol]]
#   y <- df[[ycol]]
#   r <- suppressWarnings(cor(x, y, use = "complete.obs", method = cor_method))
# 
#   ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
#     geom_hex(bins = bins) +
#     scale_fill_continuous(trans = "log10", name = "count") +
#     geom_abline(slope = 1, intercept = 0, linetype = 2) +
#     coord_equal() +
#     scale_x_continuous(limits = c(0, maxv)) +
#     scale_y_continuous(limits = c(0, maxv)) +
#     labs(
#       title = sprintf("Spearman = %.1f%%", 100 * r),
#       x = paste0(xcol, "\n(-log10 p)"),
#       y = paste0(ycol, "\n(-log10 p)"),
#       fill = "count"
#     ) +
#     theme_bw() +
#     theme(
#       panel.grid.minor = element_blank(),
#       axis.title = element_text(size = 9),
#       axis.text  = element_text(size = 7),
#       plot.title = element_text(size = 10)
#     )
# }
# -------------------------
# GASPERINI: 3 panels
# -------------------------
gas_pairs <- list(
  c("SCEPTRE", "FR-Perturb (skew)"),
  c("SCEPTRE", "FR-Perturb (resamp)"),
  c("FR-Perturb (skew)", "FR-Perturb (resamp)")
)

gas_plots <- lapply(gas_pairs, \(p) make_scatter(gasperini_wide, p[1], p[2], gas_max))

p_gasperini <- cowplot::plot_grid(plotlist = gas_plots, ncol = 3, align = "hv")
# p_gasperini

# -------------------------
# REPLOGLE: 6 panels
# -------------------------
rep_pairs <- list(
  c("SCEPTRE", "Mixscale"),
  c("SCEPTRE", "FR-Perturb (skew)"),
  c("SCEPTRE", "FR-Perturb (resamp)"),
  c("Mixscale", "FR-Perturb (skew)"),
  c("Mixscale", "FR-Perturb (resamp)"),
  c("FR-Perturb (skew)", "FR-Perturb (resamp)")
)

rep_plots <- lapply(rep_pairs, \(p) make_scatter(replogle_wide, p[1], p[2], rep_max))

p_replogle <- cowplot::plot_grid(plotlist = rep_plots, ncol = 3, align = "hv")
#

p_gasperini_titled <- ggdraw() +
  draw_plot(p_gasperini, x = 0, y = 0, width = 1, height = 0.94) +
  draw_label(
    "Comparison of Gasperini on-target p-values",
    x = 0.5, y = 0.75,
    hjust = 0.5, vjust = 1,
    fontface = "bold", size = 14
  )

p_replogle_titled <- ggdraw() +
  draw_plot(p_replogle, x = 0, y = 0, width = 1, height = 0.94) +
  draw_label(
    "Comparison of Replogle on-target p-values",
    x = 0.5, y = 0.985,
    hjust = 0.5, vjust = 1,
    fontface = "bold", size = 14
  )



print(p_gasperini_titled)
print(p_replogle_titled)


```


```{r, eval = FALSE, include = FALSE, echo = FALSE}

# Wide table of raw p-values for Replogle
replogle_pwide <- all_results$replogle_large_skew %>%
  transmute(
    target,
    `SCEPTRE`            = as.numeric(p_value_scep),
    `Mixscale`           = as.numeric(p_value_mix),
    `FR-Perturb (skew)`  = as.numeric(p_value_frpert)
  ) %>%
  inner_join(
    all_results$replogle_large_resamp %>%
      transmute(target, `FR-Perturb (resamp)` = as.numeric(p_value_frpert)),
    by = "target"
  )

# Convert each method's p-values to rank percentiles in [0,1]:
# 1 = most significant (smallest p), 0 = least significant
replogle_rankwide <- replogle_pwide %>%
  mutate(across(
    -target,
    ~ {
      p <- pmin(pmax(., 1e-300, na.rm = TRUE), 1)  # clamp
      n <- sum(is.finite(p) & !is.na(p))
      r <- rank(p, ties.method = "average", na.last = "keep") # smallest p => rank 1
      1 - (r - 0.5) / n
    }
  ))

# Bland–Altman style on rank percentiles (mean vs difference)
methods <- c("SCEPTRE", "Mixscale", "FR-Perturb (skew)", "FR-Perturb (resamp)")
pairs <- combn(methods, 2, simplify = FALSE)

ba_rank_df <- bind_rows(lapply(pairs, function(pp) {
  m1 <- pp[1]; m2 <- pp[2]
  replogle_rankwide %>%
    transmute(
      pair = paste0(m1, " vs ", m2),
      mean_rank = 0.5 * (.data[[m1]] + .data[[m2]]),
      diff_rank = .data[[m1]] - .data[[m2]]
    )
})) %>%
  filter(is.finite(mean_rank), is.finite(diff_rank))

ggplot(ba_rank_df, aes(x = mean_rank, y = diff_rank)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ pair, ncol = 3) +
  labs(
    title = "Replogle: Bland–Altman on rank percentiles (1 = most significant)",
    x = "Mean rank percentile (two methods)",
    y = "Difference in rank percentile (method1 − method2)"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```


```{r, eval = FALSE, include = FALSE, echo = FALSE}

# Wide table of -log10(p) for Gasperini
gasperini_wide <- all_results$gasperini_skew %>%
  transmute(
    target,
    `SCEPTRE`            = neglog10(p_value_scep),
    `FR-Perturb (skew)`  = neglog10(p_value_frpert)
  ) %>%
  inner_join(
    all_results$gasperini_resamp %>%
      transmute(target, `FR-Perturb (resamp)` = neglog10(p_value_frpert)),
    by = "target"
  )

# Pairwise Bland–Altman data (mean vs difference) for all method pairs
methods <- c("SCEPTRE", "FR-Perturb (skew)", "FR-Perturb (resamp)")
pairs <- combn(methods, 2, simplify = FALSE)

ba_df_gas <- bind_rows(lapply(pairs, function(pp) {
  m1 <- pp[1]; m2 <- pp[2]
  gasperini_wide %>%
    transmute(
      pair = paste0(m1, " vs ", m2),
      mean_neglog10p = 0.5 * (.data[[m1]] + .data[[m2]]),
      diff_neglog10p = .data[[m1]] - .data[[m2]]
    )
})) %>%
  filter(is.finite(mean_neglog10p), is.finite(diff_neglog10p))

ggplot(ba_df_gas, aes(x = mean_neglog10p, y = diff_neglog10p)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ pair, ncol = 3) +
  labs(
    title = "Gasperini: Bland–Altman style agreement of -log10(p) across methods",
    x = "Mean of -log10(p) (two methods)",
    y = "Difference of -log10(p) (method1 − method2)"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())


# Wide table of -log10(p) for Replogle
replogle_wide <- all_results$replogle_large_skew %>%
  transmute(
    target,
    `SCEPTRE`             = neglog10(p_value_scep),
    `Mixscale`            = neglog10(p_value_mix),
    `FR-Perturb (skew)`   = neglog10(p_value_frpert)
  ) %>%
  inner_join(
    all_results$replogle_large_resamp %>%
      transmute(target, `FR-Perturb (resamp)` = neglog10(p_value_frpert)),
    by = "target"
  )

# Pairwise Bland–Altman data (mean vs difference) for all method pairs
methods <- c("SCEPTRE", "Mixscale", "FR-Perturb (skew)", "FR-Perturb (resamp)")
pairs <- combn(methods, 2, simplify = FALSE)

ba_df <- bind_rows(lapply(pairs, function(pp) {
  m1 <- pp[1]; m2 <- pp[2]
  replogle_wide %>%
    transmute(
      pair = paste0(m1, " vs ", m2),
      mean_neglog10p = 0.5 * (.data[[m1]] + .data[[m2]]),
      diff_neglog10p = .data[[m1]] - .data[[m2]]
    )
})) %>%
  filter(is.finite(mean_neglog10p), is.finite(diff_neglog10p))

# Plot: one facet per pair
ggplot(ba_df, aes(x = mean_neglog10p, y = diff_neglog10p)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ pair, ncol = 3) +
  labs(
    title = "Replogle: Bland–Altman style agreement of -log10(p) across methods",
    x = "Mean of -log10(p) (two methods)",
    y = "Difference of -log10(p) (method1 − method2)"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())


```





























```{r, echo=FALSE}

# Convert columns (except target) to "significance percentile":
# 1 = most significant, 0 = least significant
to_sig_percentile <- function(wide_df, ties_method = "average") {
  n <- nrow(wide_df)
  wide_df %>%
    mutate(across(
      -target,
      ~ {
        r <- rank(-., ties.method = ties_method)    # larger -log10p => more significant => smaller rank
        1 - (r - 0.5) / n                           # map rank to (0,1); larger = more significant
      }
    ))
}

gasperini_pct <- to_sig_percentile(gasperini_wide)
replogle_pct  <- to_sig_percentile(replogle_wide)

# shared axis limits (now always ~[0,1], but compute anyway)
gas_max <- max(as.matrix(select(gasperini_pct, -target)), na.rm = TRUE)
rep_max <- max(as.matrix(select(replogle_pct,  -target)), na.rm = TRUE)

make_scatter <- function(df, xcol, ycol, maxv) {
  x <- df[[xcol]]
  y <- df[[ycol]]
  r <- suppressWarnings(cor(x, y, use = "complete.obs", method = "pearson"))

  ggplot(df, aes(x = .data[[xcol]], y = .data[[ycol]])) +
    geom_point(alpha = 0.3, size = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    coord_equal() +
    scale_x_continuous(limits = c(0, maxv)) +
    scale_y_continuous(limits = c(0, maxv)) +
    labs(
      title = sprintf("Pearson = %.1f%%", 100 * r),
      x = paste0(xcol, "\n(significance percentile)"),
      y = paste0(ycol, "\n(significance percentile)")
    ) +
    theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      axis.title = element_text(size = 9),
      axis.text  = element_text(size = 7),
      plot.title = element_text(size = 10)
    )
}

# --- GASPERINI grid (3 panels) ---
gas_pairs <- list(
  c("SCEPTRE", "FR-Perturb (skew)"),
  c("SCEPTRE", "FR-Perturb (resamp)"),
  c("FR-Perturb (skew)", "FR-Perturb (resamp)")
)
gas_plots <- lapply(gas_pairs, \(p) make_scatter(gasperini_pct, p[1], p[2], gas_max))
p_gasperini_rank <- cowplot::plot_grid(plotlist = gas_plots, ncol = 3, align = "hv")
p_gasperini_rank

# --- REPLOGLE grid (6 panels) ---
rep_pairs <- list(
  c("SCEPTRE", "Mixscale"),
  c("SCEPTRE", "FR-Perturb (skew)"),
  c("SCEPTRE", "FR-Perturb (resamp)"),
  c("Mixscale", "FR-Perturb (skew)"),
  c("Mixscale", "FR-Perturb (resamp)"),
  c("FR-Perturb (skew)", "FR-Perturb (resamp)")
)
rep_plots <- lapply(rep_pairs, \(p) make_scatter(replogle_pct, p[1], p[2], rep_max))
p_replogle_rank <- cowplot::plot_grid(plotlist = rep_plots, ncol = 3, align = "hv")
p_replogle_rank
```
















plot: i want to indicate the sample size somehow
```{r, echo=FALSE, eval=FALSE}
# run_name <- "gasperini_skew"
# dataset_name <- "gasperini_pos_control"
run_name <- "replogle_large_skew"
dataset_name <- "replogle-rd7_pos_control_large"
results <- all_results[[run_name]]
eps <- 1e-250
# eps <- 1e-50
# i want to do pairwise p-value plot but with sample size indicated somehow
pdat <- results |> dplyr::select(p_value_scep, p_value_frpert, target) |>
  mutate(p_value_scep = pmax(p_value_scep, eps), p_value_frpert = pmax(p_value_frpert, eps))



cell_info <- read_csv(file.path(.get_config_path("LOCAL_BENCHMARKING_DIR"), "association/pos-control/input_data", dataset_name, "cell_info.csv"))
head(cell_info)

# so... what represents my sample size here? the total number of cells that have that grna as a target?
sample_sizes <- cell_info |> group_by(grna_target) |> summarize(n = n()) |> dplyr::select(target = grna_target, n)
pdat %<>% left_join(sample_sizes, by = "target")

pdat |> transmute(neglog_p_value_scep = -log10(p_value_scep), neglog_p_value_frpert = -log10(p_value_frpert), n = n) %>%
  filter(complete.cases(.)) |>
  cor()

# this doesn't show much
# pdat |>
#   # mutate(n = cut(n, breaks = 3)) |>
#   # dplyr::filter(p_value_frpert > eps) |>
#   ggplot(aes(x = p_value_scep, y = p_value_frpert, color = n)) +
#   geom_point(alpha = 0.25, size = 0.7) +
#   # facet_grid(~n) +
#   scale_color_viridis_c(option = "C", end = 0.95) +
#   scale_x_continuous(trans = revlog_trans(10)) +
#   scale_y_continuous(trans = revlog_trans(10)) +
#   geom_abline(slope = 1, intercept = 0, col = "grey60", linetype = "dashed") +
#   theme_bw() +
#   labs(
#     x = "SCEPTRE p-value (-log10)",
#     y = "FR-Pert p-value (-log10)",
#     color = "n"
#   )
breaks <- list(
  gasperini_pos_control = c(339, 1050, 1800, 2509),
  `replogle-rd7_pos_control_large` = c(16, 63, 117, 3782)
)[[dataset_name]]

pdat |>
  mutate(cells_per_target = cut(n, breaks = breaks, dig.lab=4, ordered_result = TRUE)) %>%
  split(f = .$cells_per_target) |>
  lapply(function(df) data.frame(cor = cor(-log10(df$p_value_scep), -log10(df$p_value_frpert), use="complete.obs") |> round(2))) %>%
  do.call(what = rbind) %>%
  mutate(cells_per_target = rownames(.)) |>
  `rownames<-`(NULL)

n_max <- Inf
data.frame(
  p_value = -log10(c(pdat$p_value_scep, pdat$p_value_frpert)),
  method = rep(c("sceptre", "FR_perturb"), each = nrow(pdat)),
  n = rep(pdat$n, times = 2)
) |>
  filter(n <= n_max) |>
  ggplot(aes(y=  p_value, x  =n, color = method)) +
  geom_point(alpha = 0.4, size = 0.8) +
  facet_wrap(~method) +
   geom_smooth(method = "lm", se = FALSE) +
  # scale_y_continuous(trans = revlog_trans(10)) +
labs(x = "total number of cells expressing this target", 
     y = "p-value (-log10)",
     title = paste0("On-target sample size vs p-value for ", dataset_name)
) +
  theme_bw()
pdat |>
  dplyr::select(p_value_scep, )
```

TODO

- More methods?

- More datasets?

- Connect to guide assignment? Try with different assignments? Is it circular to use sceptre's for sceptre?

- Use covariates like total sequencing depth for guides or genes in the p-value assessment
