---
title: "assignment-analysis"
output:
  pdf_document: default
date: "2025-10-23"
---

# 1. loading the data

```{r, echo=FALSE, include=FALSE}
source("~/.Rprofile")
library(tidyverse)

# takes sceptre assignment matrix and gets a list
# with one element per gRNA, and each element contains
# the indices of the cells assigned to that gRNA
# TODO rewrite this to be more efficient and directly work with sparse matrix form
sparse_matrix_to_list <- function(mat) {
  result <- lapply(1:nrow(mat), function(i) {
    which(mat[i, ])
  })
  names(result) <- rownames(mat)
  return(result)
}

# for crispat, we have our output as a csv like
# 'data.frame':	7143379 obs. of  2 variables:
#  $ cell_id: chr  "CELL_394" "CELL_759" "CELL_814" "CELL_911" ...
#  $ grna_id: chr  "AAGCCCGGCCAGCCGACCGG" "AAGCCCGGCCAGCCGACCGG" "AAGCCCGGCCAGCCGACCGG" "AAGCCCGGCCAGCCGACCGG" ...
# so we split on grna_id, and turn that into the integer cell index by processing
# "CELL_j" to j
csv_to_list_of_ints <- function(df) {
  # Group by gRNA and collect cell indices
  split(df$cell_id, df$grna_id) |>
    lapply(function(cell_names) gsub("CELL_", "", cell_names) |> as.integer())
}

base_dir <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/outputs/gasperini_and_replogle-rd7/"
)
# base_dir <- "~/data/projects/sceptre3/benchmarking/guide_assignment/outputs/gasperini_and_replogle-rd7/gasperini_and_replogle-rd7"
# might have more later
dataset_names <- c("gasperini", "replogle-rd7")

## doing sceptre ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# NOTE: since sceptre has the full assignment matrices, i use this to get the grna ids.


sceptre_assignments_raw <- lapply(dataset_names, function(fname) {
  fpath <- file.path(base_dir, paste0("assignment_matrix_sceptre_", fname, ".rds"))
  readRDS(fpath)
}) |>
  setNames(dataset_names)

sceptre_assignments <- lapply(sceptre_assignments_raw, sparse_matrix_to_list)

grna_names <- lapply(sceptre_assignments_raw, rownames)

## doing crispat ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# outline:
# 1. load the assignment csvs and process them to have fast grna_id lookup
# 2. make an empty list of lists that has every grna_id in it, using the sceptre assignment matrices
# 3. fill that with the results of step (1) so that we know grna_id order and names
#     line up with different methods


# 1. loading the assignment CSVs
crispat_assignments_raw <- lapply(dataset_names, function(fname) {
  file.path(base_dir, paste0("assignments_crispat_", fname, ".csv")) |>
    read.csv() |>
    csv_to_list_of_ints()
}) |>
  setNames(dataset_names)

# 2. making an empty data struc to hold them all with no missing names
crispat_assignments <- lapply(dataset_names, function(name) {
  lapply(grna_names[[name]], function(i) integer(0)) |>
    setNames(grna_names[[name]])
}) |>
  setNames(dataset_names)

# 3. filling the empty data structure
for(name in dataset_names) {
  for(grna_name in names(crispat_assignments_raw[[name]])) {
    crispat_assignments[[name]][[grna_name]] <- crispat_assignments_raw[[name]][[grna_name]]
  }
}

```

loading grna matrices

```{r}
grna_base_dir <- file.path(
  .get_config_path("LOCAL_BENCHMARKING_DIR"),
  "guide_assignment/input_data"
)

grna_gasp <- Matrix::readMM(file.path(grna_base_dir, "gasperini/cleanser/grna_matrix.mtx")) 
grna_repl <- Matrix::readMM(file.path(grna_base_dir, "replogle-rd7/cleanser/grna_matrix.mtx")) 

grna_mats <- list(gasperini = grna_gasp, `replogle-rd7` = grna_repl)

num_cells <- list(gasperini = ncol(grna_gasp), `replogle-rd7` = ncol(grna_repl))

```


# 2. computing similarities
```{r, echo=FALSE, include=FALSE}
jaccard <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  if (union == 0) return(NA_real_)
  return(intersection / union)
}

sims <- lapply(dataset_names, function(name) {
  sapply(grna_names[[name]], function(grna) jaccard(sceptre_assignments[[name]][[grna]], crispat_assignments[[name]][[grna]]))
}) |>
  setNames(dataset_names)


# checking for NAs
na_names_gasp <- which(is.na(sims$gasperini)) |> names()
sceptre_assignments[["gasperini"]][na_names_gasp]
crispat_assignments[["gasperini"]][na_names_gasp]

# ok cool, we only get NAs where BOTH have no cells

na_names_repl <- which(is.na(sims$`replogle-rd7`)) |> names()
# there aren't any! This means every gRNA got at least one cell assigned to it by some method
```


# 3. analysis of similarities

## 3.1 plot 1: distribution of similarities
```{r, echo=FALSE}
library(tidyverse)
theme_set(theme_bw(base_size = 12))

# Tidy similarities: one row per gRNA per dataset
df_sims <- bind_rows(lapply(names(sims), function(nm) {
  tibble(dataset = nm, similarity = sims[[nm]])
})) %>%
  filter(!is.na(similarity))

# Plot: two panels (one per dataset)
p <- ggplot(df_sims, aes(x = similarity)) +
  geom_histogram(bins = 50) +
  facet_wrap(dataset ~ ., ncol = 1, scales = "free") +
  labs(
    title = "Jaccard similarites for each gRNA",
    x = "Jaccard similarity",
    y = "Count"
  ) +
  coord_cartesian(xlim = c(0, 1))

print(p)

```


## 3.2: similarity vs sparsity



```{r}
# get the sparsity of each gene
sparsity_gasp <- tabulate(grna_gasp@i + 1L, nbins = nrow(grna_gasp))
sparsity_repl <- tabulate(grna_repl@i + 1L, nbins = nrow(grna_repl))

plotdata <- data.frame(
  grna_id = unlist(grna_names),
  data = rep(dataset_names, sapply(sims, length)),
  sparsity = c(sparsity_gasp, sparsity_repl),
  jaccard = unlist(sims)
)

ggplot(plotdata, aes(x = log(sparsity+1), y = jaccard)) +
  geom_point() +
  facet_grid(~data, scales="free") +
  theme_bw() +
  labs(x = "Num. cells with nonzero UMI (log scale)", title = "Jaccard vs Sparsity")

```




## 3.3 more nuance for what a value of jaccard means

there is a bigger range of jaccards for replogle, so the same seq won't make sense
```{r, echo=FALSE, include=FALSE}

# sims: named numeric vector in [0, 1]
# sim_vals: numeric vector of target similarities
# m: how many closest names to return per target
preimage_names <- function(similarities, sim_vals, m) {
  stopifnot(is.numeric(similarities), !is.null(names(sims)))
  # drop NAs if any
  similarities <- similarities[!is.na(similarities)]
  out <- setNames(vector("list", length(sim_vals)), as.character(sim_vals))
  for (k in seq_along(sim_vals)) {
    t <- sim_vals[k]
    ord <- order(abs(similarities - t))
    out[[k]] <- names(similarities)[ord][seq_len(min(m, length(ord)))]
  }
  out
}
preimage_names(sims$`replogle-rd7`, c(.8, .9, .95), 5)

# ---- helper: build 2x2 "membership table" counts for one gRNA ----
quad_counts <- function(a_idx, b_idx, N = NULL) {
  a <- if (length(a_idx)) unique(a_idx) else integer(0)
  b <- if (length(b_idx)) unique(b_idx) else integer(0)
  inter <- length(intersect(a, b))
  a_only <- length(setdiff(a, b))
  b_only <- length(setdiff(b, a))
  union_sz <- inter + a_only + b_only
  if (is.null(N)) N <- max(c(a, b, 0L))  # fallback when total #cells unknown
  neither <- max(0L, N - union_sz)

  tibble(
    crispat = c(0, 1, 0, 1),                   # x-axis: CRISPAT membership (0/1)
    sceptre = c(0, 0, 1, 1),                   # y-axis: SCEPTRE membership (0/1)
    region  = c("Neither", "CRISPAT only", "SCEPTRE only", "Intersection"),
    count   = c(neither, a_only, b_only, inter)
  )
}

# helper for %||%
`%||%` <- function(a, b) if (!is.null(a)) a else b
```




```{r, echo=FALSE}

m <- 5
dataset <- "replogle-rd7"
N_cells <- num_cells[[dataset]]  
# sim_vals <- quantile(sims[[dataset]], c(.1, .5, .9), na.rm=TRUE) |> round(3)
sim_vals <- c(.8, .9, .95)

sel <- preimage_names(sims[[dataset]], sim_vals, m)  

# ---- assemble tiles for all panels ----
tiles <- map_dfr(seq_along(sim_vals), function(r) {
  tval <- sim_vals[r]
  grnas <- sel[[as.character(tval)]]

  map2_dfr(grnas, seq_along(grnas), function(gid, k) {
    a_idx <- crispat_assignments[[dataset]][[gid]]
    b_idx <- sceptre_assignments[[dataset]][[gid]]
    qc <- quad_counts(a_idx %||% integer(0), b_idx %||% integer(0), N = N_cells)
    qc %>%
      mutate(
        target = factor(tval, levels = sim_vals),  # row facet
        k = factor(k, levels = 1:m),               # column facet (1..5)
        grna = gid,
        sim  = sims[[dataset]][[gid]]
      )
  })
})



# Labels to print inside each panel
panel_labels <- tiles %>%
  distinct(target, k, grna, sim) %>%
  mutate(label = paste0(grna, "\n(sim=", sprintf("%.3f", sim), ")"),
         x_lab = 1.5,  # between 1 (no) and 2 (yes) after coercion to numeric from factor below
         y_lab = 2.2)

# ---- plot ----
ggplot(tiles, aes(x = factor(crispat), y = factor(sceptre), fill = count)) +
  geom_tile() +
  geom_text(aes(label = format(count, big.mark=",")), fontface = "bold") +
  facet_grid(rows = vars(target), cols = vars(k)) +
  # put the gRNA name + sim at the top of each panel
  # geom_text(data = panel_labels, aes(x = x_lab, y = y_lab, label = label),
  #           inherit.aes = FALSE, hjust = 0.5, vjust = 0, size = 3.1) +
  scale_x_discrete(name = "CRISPAT membership", labels = c("0", "1")) +
  scale_y_discrete(name = "SCEPTRE membership", labels = c("0", "1"),
                   expand = expansion(mult = c(0.1, 0.35))) +  # add headroom for label
  guides(fill = "none") +
  labs(title = paste("Overlap of assigned cells per gRNA for", dataset)) +
  coord_cartesian(clip = "off") # allow labels above tiles

```



```{r, echo=FALSE}
m <- 8
dataset <- "gasperini"
N_cells <- num_cells[[dataset]]  
# sim_vals <- quantile(sims[[dataset]], c(.1, .5, .9), na.rm=TRUE) |> round(3)
sim_vals <- c(.925, .95, .99)

sel <- preimage_names(sims[[dataset]], sim_vals, m)  

# ---- assemble tiles for all panels ----
tiles <- map_dfr(seq_along(sim_vals), function(r) {
  tval <- sim_vals[r]
  grnas <- sel[[as.character(tval)]]

  map2_dfr(grnas, seq_along(grnas), function(gid, k) {
    a_idx <- crispat_assignments[[dataset]][[gid]]
    b_idx <- sceptre_assignments[[dataset]][[gid]]
    qc <- quad_counts(a_idx %||% integer(0), b_idx %||% integer(0), N = N_cells)
    qc %>%
      mutate(
        target = factor(tval, levels = sim_vals),  # row facet
        k = factor(k, levels = 1:m),               # column facet (1..5)
        grna = gid,
        sim  = sims[[dataset]][[gid]]
      )
  })
})



# Labels to print inside each panel
panel_labels <- tiles %>%
  distinct(target, k, grna, sim) %>%
  mutate(label = paste0(grna, "\n(sim=", sprintf("%.3f", sim), ")"),
         x_lab = 1.5,  # between 1 (no) and 2 (yes) after coercion to numeric from factor below
         y_lab = 2.2)

# ---- plot ----
ggplot(tiles, aes(x = factor(crispat), y = factor(sceptre), fill = count)) +
  geom_tile() +
  geom_text(aes(label = format(count, big.mark=",")), fontface = "bold") +
  facet_grid(rows = vars(target), cols = vars(k)) +
  # put the gRNA name + sim at the top of each panel
  # geom_text(data = panel_labels, aes(x = x_lab, y = y_lab, label = label),
  #           inherit.aes = FALSE, hjust = 0.5, vjust = 0, size = 3.1) +
  scale_x_discrete(name = "CRISPAT membership", labels = c("0", "1")) +
  scale_y_discrete(name = "SCEPTRE membership", labels = c("0", "1"),
                   expand = expansion(mult = c(0.1, 0.35))) +  # add headroom for label
  guides(fill = "none") +
  labs(title = paste("Overlap of assigned cells per gRNA for", dataset)) +
  coord_cartesian(clip = "off") # allow labels above tiles

```




# 4. UMI count analysis

## 4.1: number of cells assigned to each gRNA

```{r, echo=FALSE}
# Build a tidy data frame from your two method lists
df <- bind_rows(
  tibble(method = "SCEPTRE", dataset = "gasperini",
         size = sapply(sceptre_assignments[["gasperini"]], length)),
  tibble(method = "SCEPTRE", dataset = "replogle-rd7",
         size = sapply(sceptre_assignments[["replogle-rd7"]], length)),
  tibble(method = "CRISPAT", dataset = "gasperini",
         size = sapply(crispat_assignments[["gasperini"]], length)),
  tibble(method = "CRISPAT", dataset = "replogle-rd7",
         size = sapply(crispat_assignments[["replogle-rd7"]], length))
) %>% filter(!is.na(size))

# Fix facet order to ensure a 2x2 layout
df <- df %>%
  mutate(
    method  = factor(method, levels = c("SCEPTRE", "CRISPAT")),
    dataset = factor(dataset, levels = c("gasperini", "replogle-rd7"))
  )

# Plot: 2 x 2 histograms
p <- ggplot(df, aes(x = size)) +
  geom_histogram(bins = 30, boundary = 0, closed = "left") +
  facet_grid(rows = vars(method), cols = vars(dataset), drop = FALSE, scales = "free") +
  labs(
    title = "Cells assigned to each gRNA",
    x = "Number of cells assigned to a gRNA",
    y = "Count"
  ) +
  theme_bw(base_size = 12)

print(p)

```

## 4.2: MOI: number of gRNAs assigned to each cell 

```{r, echo=FALSE}
library(tidyverse)
theme_set(theme_bw(base_size = 12))

# Invert gRNA -> cells into cell -> #gRNAs for one dataset
counts_per_cell <- function(grna_to_cells, n_cells = NULL) {
  if (is.null(grna_to_cells) || length(grna_to_cells) == 0) {
    return(tibble(cell = integer(), n_grnas = integer()))
  }
  all_cells <- unlist(grna_to_cells, use.names = FALSE)
  if (length(all_cells) == 0) {
    return(tibble(cell = integer(), n_grnas = integer()))
  }
  nbins <- if (is.null(n_cells)) max(all_cells) else n_cells
  counts <- tabulate(all_cells, nbins = nbins)
  out <- tibble(cell = seq_len(nbins), n_grnas = counts)
  if (is.null(n_cells)) out <- dplyr::filter(out, n_grnas > 0)
  out
}

# Build tidy data
df_cells <- bind_rows(
  counts_per_cell(sceptre_assignments[["gasperini"]]) %>% mutate(method = "SCEPTRE", dataset = "gasperini"),
  counts_per_cell(sceptre_assignments[["replogle-rd7"]]) %>% mutate(method = "SCEPTRE", dataset = "replogle-rd7"),
  counts_per_cell(crispat_assignments[["gasperini"]]) %>% mutate(method = "CRISPAT", dataset = "gasperini"),
  counts_per_cell(crispat_assignments[["replogle-rd7"]]) %>% mutate(method = "CRISPAT", dataset = "replogle-rd7")
) %>%
  mutate(  
    method  = factor(method,  levels = c("SCEPTRE", "CRISPAT")),
    dataset = factor(dataset, levels = c("gasperini", "replogle-rd7"))
  )

# Panel-wise means
means_df <- df_cells %>%
  group_by(dataset, method) %>%
  summarise(mean_n = mean(n_grnas, na.rm = TRUE), .groups = "drop")

p <- ggplot(df_cells, aes(x = n_grnas)) +
  geom_histogram(bins = 50) +
  facet_grid(
    rows = vars(dataset),
    cols = vars(method),
    scales = "free_y"
  ) +
  # vertical line at the mean (per panel)
  geom_vline(data = means_df, aes(xintercept = mean_n), linetype = "dashed", color="blue") +
  # vertical line at 1 (same for all panels)
  geom_vline(xintercept = 1, linetype = "dotted", color="red") +
  labs(
    title = "gRNAs assigned per cell",
    x = "# gRNAs assigned to a cell",
    y = "Number of cells"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_bw()

print(p)

```


## 4.3 bimodality
# ```{r}
# # library(Matrix)
# # library(tidyverse)
# # theme_set(theme_bw(base_size = 12))
# 
# # assignments is a named list of method -> dataset -> gRNA -> integer cell indices
# # e.g., list(SCEPTRE = sceptre_assignments, CRISPAT = crispat_assignments, ...)
# 
# plot_grna_count_hists <- function(dataset, method, grna_id, assignments, grna_matrix, bins = 50) {
#   # cell groups
#   assigned <- assignments[[method]][[dataset]][[grna_id]]
#   if (is.null(assigned)) assigned <- integer(0)
#   n_cells <- ncol(grna_matrix)
#   not_assigned <- setdiff(seq_len(n_cells), assigned)  # complement; union == 1:n_cells
# 
#   # extract the row once (stays sparse) and pull just needed columns
#   row_sel <- grna_matrix[which(grna_names[[dataset]] == grna_id), , drop = FALSE]
#   c_assigned <- as.numeric(row_sel[, assigned, drop = FALSE])
#   c_not      <- as.numeric(row_sel[, not_assigned, drop = FALSE])
# 
#   df <- tibble(
#     group = rep(c("Assigned", "Not assigned"), c(length(c_assigned), length(c_not))),
#     count = c(c_assigned, c_not)
#   )
# 
#   ggplot(df, aes(x = count)) +
#     geom_histogram(bins = bins, boundary = 0, closed = "left") +
#     facet_wrap(~ group, nrow = 1, scales = "free_y") +
#     labs(
#       title = sprintf("UMI counts for %s — %s (%s)", substr(grna_id, 1, 10), method, dataset),
#       x = sprintf("UMIs for %s", grna_id),
#       y = "Number of cells"
#     )
# }
# 
# dataset <- "replogle-rd7"
# grna_matrix <- grna_mats[[dataset]]
# num_grnas <- 16
# set.seed(123)
# grnas <- sample(grna_names[[dataset]], size = num_grnas, replace = FALSE)
# 
# 
# 
# 
# 
# # Example:
# plot_grna_count_hists("replogle-rd7", "SCEPTRE",
#   grnas[1],
#   assignments = list(SCEPTRE = sceptre_assignments, CRISPAT = crispat_assignments),
#   grna_matrix = grna_matrix, bins = 50)
# 
# ```
# 

```{r}


plot_grna_count_hists_2x2 <- function(dataset, grna_id,
                                      sceptre_assignments, crispat_assignments,
                                      grna_matrix, bins = 50) {
  
  `%||%` <- function(a,b) if (!is.null(a)) a else b
  n_cells <- ncol(grna_matrix)
  row_sel <- grna_matrix[which(grna_names[[dataset]] == grna_id), , drop = FALSE]
  
  extract_counts <- function(cols) {
    if (length(cols)) as.numeric(row_sel[, cols, drop = FALSE]) else numeric(0)
  }
  
  # SCEPTRE
  s_assigned <- sceptre_assignments[[dataset]][[grna_id]] %||% integer(0)
  s_not <- setdiff(seq_len(n_cells), s_assigned)
  s_df <- bind_rows(
    tibble(method = "SCEPTRE", group = "Assigned",     count = extract_counts(s_assigned)),
    tibble(method = "SCEPTRE", group = "Not assigned", count = extract_counts(s_not))
  )

  # CRISPAT
  c_assigned <- crispat_assignments[[dataset]][[grna_id]] %||% integer(0)
  c_not <- setdiff(seq_len(n_cells), c_assigned)
  c_df <- bind_rows(
    tibble(method = "CRISPAT", group = "Assigned",     count = extract_counts(c_assigned)),
    tibble(method = "CRISPAT", group = "Not assigned", count = extract_counts(c_not))
  )

  df <- bind_rows(s_df, c_df) %>%
    mutate(
      method = factor(method, levels = c("SCEPTRE", "CRISPAT")),
      group  = factor(group,  levels = c("Assigned", "Not assigned"))
    )

  ggplot(df, aes(x = count)) +
    geom_histogram(bins = bins, boundary = 0, closed = "left") +
    facet_wrap(vars(method, group), nrow = 2, scales = "free_y") +  # <- per-panel free y
    labs(
      title = sprintf("gRNA counts for %s — %s", substr(as.character(grna_id), 1, 10), dataset),
      x = sprintf("UMIs for %s", grna_id),
      y = "Number of cells"
    ) +
    scale_x_continuous(trans = "log1p")
}

set.seed(1)
dataset <- "replogle-rd7"
grna_matrix <- grna_mats[[dataset]]
num_grnas <- 16
grnas <- sample(grna_names[[dataset]], size = num_grnas, replace = FALSE)


# Example call:
for(i in 1:10) {
  plt <- plot_grna_count_hists_2x2(dataset,
  grnas[i],
  sceptre_assignments, crispat_assignments, grna_matrix, bins = 50)
  print(plt)
}


```

```{r}
set.seed(1)
dataset <- "gasperini"
grna_matrix <- grna_mats[[dataset]]
num_grnas <- 16
grnas <- sample(grna_names[[dataset]], size = num_grnas, replace = FALSE)


# Example call:
for(i in 1:5) {
  plt <- plot_grna_count_hists_2x2(dataset,
  grnas[i],
  sceptre_assignments, crispat_assignments, grna_matrix, bins = 50)
  print(plt)
}

```


are unassigned all zero? or just looks that way?
nope not all zero.
```{r}
dataset <- "replogle-rd7"
N <- num_cells[[dataset]]
all_idx <- seq_len(N)
grna_matrix <- grna_mats[[dataset]]
set.seed(1)
num_grnas_to_check <- 50
grna_locs <- sample(seq_along(grna_names[[dataset]]), size=num_grnas_to_check, replace=FALSE) |> sort()

results <- data.frame(scep = numeric(num_grnas_to_check), crisp = 0, grna="", jaccard = 0)
for(i in 1:num_grnas_to_check) {
  grna <- grna_names[[dataset]][grna_locs[i]]
  scep_unas <- setdiff(all_idx, sceptre_assignments[[dataset]][[grna]])
  crisp_unas <- setdiff(all_idx, crispat_assignments[[dataset]][[grna]])
  results$scep[i] <- sum(grna_matrix[which(grna_names[[dataset]] == grna), scep_unas] > 0)
  results$crisp[i] <- sum(grna_matrix[which(grna_names[[dataset]] == grna), crisp_unas] > 0)
  results$grna[i] <- grna
  results$jaccard[i] <- sims[[dataset]][grna]
}

plot(results$scep, results$crisp, cex = results$jaccard + 0.5)
abline(0,1)

hist(results$scep)
hist(results$crisp)
hist(results$jaccard)
cor(results$scep, results$crisp)
```


# compare low moi to maximum

```{r}
library(testthat)

# col_which_max_sparse <- function(X, tie = c("random","first"),
#                                  empty = c("random","NA","first"),
#                                  seed = NULL) {
#   tie   <- match.arg(tie)
#   empty <- match.arg(empty)
#   if (!is.null(seed)) set.seed(seed)
# 
#   Xc <- as(X, "dgCMatrix")
#   p  <- Xc@p; i <- Xc@i; x <- Xc@x
#   nC <- ncol(Xc); nR <- nrow(Xc)
#   out <- integer(nC)
# 
#   # helpers for choosing among zero rows (complement of nz)
#   first_zero <- function(nz) {
#     # nz must be sorted, 1-based
#     if (length(nz) == 0L) return(1L)
#     cur <- 1L
#     for (z in nz) {
#       if (cur < z) return(cur)
#       if (cur == z) cur <- cur + 1L
#     }
#     if (cur <= nR) cur else nR  # fallback; cur should always <= nR
#   }
#   random_zero <- function(nz) {
#     k <- nR - length(nz)              # number of zero rows
#     r <- sample.int(k, 1L)
#     prev <- 0L
#     for (z in c(nz, nR + 1L)) {       # include sentinel
#       start <- prev + 1L
#       end   <- z - 1L
#       if (end >= start) {
#         gap <- end - start + 1L
#         if (r <= gap) return(start + r - 1L)
#         r <- r - gap
#       }
#       prev <- z
#     }
#     1L
#   }
# 
#   for (j in seq_len(nC)) {
#     a <- p[j] + 1L; b <- p[j + 1L]
#     if (a > b) {
#       out[j] <- switch(empty,
#         random = sample.int(nR, 1L),
#         `NA`   = NA_integer_,
#         first  = 1L
#       )
#       next
#     }
# 
#     idx <- a:b
#     v   <- x[idx]
#     nzr <- i[idx] + 1L                 # sorted row indices with nonzeros
# 
#     # best among stored nonzeros
#     m <- max(v)
#     if (m > 0) {
#       if (tie == "first") {
#         out[j] <- nzr[ which.max(v) ]
#       } else {
#         w <- which(v == m)
#         out[j] <- nzr[ if (length(w) == 1L) w else sample(w, 1L) ]
#       }
#     } else {
#       # zeros beat all-negative (or hypothetically zero) columns
#       out[j] <- if (tie == "first") first_zero(nzr) else random_zero(nzr)
#     }
#   }
#   out
# }

# X <- sparseMatrix(
#   i = c(2,4,  3,  2,5,  1,2,3,  5),
#   j = c(2,2,  3,  4,4,  5,5,5,  6),
#   x = c(1,1,  5, -1,-1, 2,3,1,  7),
#   dims = c(5,6),
#   giveCsparse = FALSE # dgTMatrix (triplet), like your input
# )
# 
# test_that("tie='first', empty='first' matches base which.max on dense", {
#   expected <- apply(as.matrix(X), 2, which.max)
#   got <- col_which_max_sparse(X, tie = "first", empty = "first")
#   expect_identical(got, expected)
# })
# 
# test_that("tie='random' picks among ties and is reproducible with seed", {
#   got1 <- col_which_max_sparse(X, tie = "random", empty = "first", seed = 123)
#   got2 <- col_which_max_sparse(X, tie = "random", empty = "first", seed = 123)
#   expect_identical(got1, got2)  # deterministic with seed
# 
#   # Column 2 has a positive tie at rows 2 and 4.
#   expect_true(got1[2] %in% c(2,4))
# 
#   # Column 4 has only negatives; zeros win → choose among zero rows {1,3,4}.
#   expect_true(got1[4] %in% c(1,3,4))
# })
# 
# test_that("empty column handling works: NA vs random", {
#   # Column 1 is empty
#   na_res <- col_which_max_sparse(X, tie = "first", empty = "NA")
#   expect_true(is.na(na_res[1]))
# 
#   rnd_res <- col_which_max_sparse(X, tie = "first", empty = "random", seed = 1)
#   expect_true(rnd_res[1] >= 1 && rnd_res[1] <= nrow(X))
#   # Non-empty columns should still match 'first'
#   first_res <- col_which_max_sparse(X, tie = "first", empty = "first")
#   expect_identical(rnd_res[-1], first_res[-1])
# })

```

```{r}
library(Matrix)

# Column-wise which.max with "first" tie-breaking, respecting implicit zeros.
col_which_max_sparse_first <- function(X) {
  Xc <- as(X, "dgCMatrix")           # compress once (dedupes triplets)
  p  <- Xc@p                         # col pointers (length ncol+1)
  i  <- Xc@i                         # 0-based row indices
  x  <- Xc@x                         # values
  nR <- nrow(Xc); nC <- ncol(Xc)
  out <- integer(nC)

  for (j in seq_len(nC)) {
    a <- p[j] + 1L; b <- p[j + 1L]   # nnz slice for column j
    if (a > b) {                     # empty column -> first row (like which.max(rep(0,nR)))
      out[j] <- 1L
      next
    }

    idx <- a:b
    nzr <- i[idx] + 1L               # sorted row indices with nonzeros
    v   <- x[idx]

    k  <- which.max(v)               # first index of the max among nonzeros
    mv <- v[k]

    if (mv >= 0 || length(nzr) == nR) {
      # If best nonzero >= 0 OR there are no implicit zeros, return its row
      out[j] <- nzr[k]
    } else {
      # All stored values < 0 and there exists at least one implicit zero:
      # choose the FIRST zero row (smallest row not in nzr), without densifying
      if (nzr[1] != 1L) {
        out[j] <- 1L
      } else {
        found <- FALSE
        for (t in 2:length(nzr)) {
          if (nzr[t] > nzr[t - 1] + 1L) { out[j] <- nzr[t - 1] + 1L; found <- TRUE; break }
        }
        if (!found) out[j] <- nzr[length(nzr)] + 1L  # next row after the last nonzero
      }
    }
  }
  out
}


X <- sparseMatrix(
  i = c(2,4,  3,  2,5,  1,2,3,  5),
  j = c(2,2,  3,  4,4,  5,5,5,  6),
  x = c(1,1,  5, -1,-1, 2,3,1,  7),
  dims = c(5,6),
  giveCsparse = FALSE # dgTMatrix input (like yours)
)

expected <- apply(as.matrix(X), 2, which.max)
got <- col_which_max_sparse_first(X)
stopifnot(identical(got, expected))  # should pass
```


```{r}
max_repl <- col_which_max_sparse_first(grna_repl)
# now what do i get from sceptre and crispat?
dataset <- "replogle-rd7"
# flat <- data.frame(
#   grna_id = rep(names(crispat_assignments[[dataset]]), sapply(crispat_assignments[[dataset]], length)),
#   cells = unlist(crispat_assignments[[dataset]])
# )
flat <- data.frame(
  grna_id = rep(names(sceptre_assignments[[dataset]]), sapply(sceptre_assignments[[dataset]], length)),
  cells = unlist(sceptre_assignments[[dataset]])
)
dim(flat)

cell2grna <- split(flat$grna_id, f=flat$cells)
cell2grna[[1]]
cells_to_use <- as.numeric(names(cell2grna))


actual_max <- sapply(cell2grna, function(i) i[1])  # taking first one
max_repl_named <- grna_names[[dataset]][max_repl][cells_to_use]

mean(actual_max == max_repl_named)
```

pretty interesting! 